### 2017-11-27 latex/egsr17-cgf/introduction.tex
added latex project for text

\section{Introduction}

\subsection*{Contributions}
\begin{itemize}
	\item System for generation of stencils for coupled PDE's directly from mathematical expressions.
	\item Staggered-grid solver for time-independent PN-equations.
	\item Novel PN-guided Monte-Carlo estimator for participating media.
\end{itemize}

### 2017-11-27 latex/egsr17-cgf/main-body.tex
added latex project for text

% ---------------------------------------------------------------------
% EG author guidelines plus sample file for EG publication using LaTeX2e input
% D.Fellner, v2.02, Jan 25, 2017

%\title[EG \LaTeX\ Author Guidelines]%
%{\LaTeX\ Author Guidelines for EUROGRAPHICS Proceedings Manuscripts}

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
%\author[D. Fellner \& S. Behnke]
%{\parbox{\textwidth}{\centering D.\,W. Fellner\thanks{Chairman Eurographics Publications Board}$^{1,2}$
%        and S. Behnke$^{2}$ 
%        S. Spencer$^2$\thanks{Chairman Siggraph Publications Board}
%        }
%        \\
%% For Computer Graphics Forum: Please use the abbreviation of your first name.
%{\parbox{\textwidth}{\centering $^1$TU Darmstadt \& Fraunhofer IGD, Germany\\
%         $^2$Graz University of Technology, Institute of Computer Graphics and Knowledge Visualization, Austria
%%        $^2$ Another Department to illustrate the use in papers from authors
%%             with different affiliations
%       }
%}
%}

% ------------------------------------------------------------------------

% if the Editors-in-Chief have given you the data, you may uncomment
% the following five lines and insert it here
%
% \volume{36}   % the volume in which the issue will be published;
% \issue{1}     % the issue number of the publication
% \pStartPage{1}      % set starting page

%-------------------------------------------------------------------------
\begin{document}

% uncomment for using teaser
% \teaser{
%  \includegraphics[width=\linewidth]{eg_new}
%  \centering
%   \caption{New EG Logo}
% \label{fig:teaser}
%}

%\maketitle

%-------------------------------------------------------------------------

%\begin{abstract}
%TODO

%\begin{classification} % according to http://www.acm.org/class/1998/
%\CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Raytracing}
%\end{classification}

%\end{abstract}

%-------------------------------------------------------------------------
\input{introduction.tex}

%-------------------------------------------------------------------------
\section{Previous work}

\subsection*{Deterministic Methods}

\begin{itemize}
  \item Lattice Boltzmann Methods \cite{Geist04}
  \item Discrete Ordinates
  \item Diffusion \cite{Stam95} \cite{Jakob10} \cite{Koerner14}
  \item $PN$-Method \cite{Kajiya84} 
  \item Heuristics \cite{Kaplanyan10} \cite{Elek14}
\end{itemize}

\subsection*{Path-guiding}

\begin{itemize}
  \item Russian roulette and splitting \cite{Vorba16}
  \item Learning using SD-Trees \cite{Pegoraro08} \cite{Mueller17} or Gaussian mixture models \cite{Vorba14}
  \item Zero-Variance sampling \cite{Krivanek14}
  \item Reinforcement learning \cite{Dahm17}
\end{itemize}

%-------------------------------------------------------------------------
\section{Radiative Transfer and $PN$-Equations}

In order to compute a rendered image, the radiance field $L$ is evaluated. Its evolution throught space is given by the radiative transfer equation (RTE):

-give RTE
%\begin{}
%\end{}

The radiance field $L$ is typically solved with probabilistic Monte-Carlo methods, such as path-tracing. With these methods, computational demand is directly linked to the average length of paths and the presence of light transport paths which are hard to find with random sampling. The path length increases for dense, highly scattering media and hard-to-sample paths exist with highly anisotropic phase functions or complex visbility in the scene.

Deterministic methods are derived by discretizing the radiance field $L$ and the RTE in spatial- and spherical domain. The result is a system of linear equations of the form $A\vec{x}=b$ where $\vec{x}$ represents the radiance field in the discretized problem. The matrix $A$ and right hand side $b$ are found by assembling a global system from all discretization elements. Deterministic methods suffer from discretization artefacts and only give an accurate solution in the discretization limit. However, they are fast to compute and noise free.

The $PN$-equations are found by projecting the radiance field and the RTE into a truncated spherical harmonics (SH). A regular grid is used for spatial discretization. The SH-expansion of the radiance field $L$ is:

\begin{equation}
L\left(x, \omega \right) = \sum_{l=0}^{N}
\sum_{m=-l}^{l}
L\left(x\right)^{lm}
Y\left(\omega\right)^{l,m}
\end{equation}

where $N$ is the truncation order, $L^{lm}$ is a set SH coefficients and $Y^{l,m}$ is the SH-basis function.

\section{System Overview}
\section{Stencil Generation}
\section{Staggered Grid Solver}
\section{Path-guiding}
\section{Results}
\section{Conclusion and future work}

\begin{itemize}
  \item Extension to surface rendering by dealing vacuum regions and extending to anisotropic RTE.
\end{itemize}

%-------------------------------------------------------------------------

\bibliographystyle{eg-alpha}
%\bibliographystyle{eg-alpha-doi}

\bibliography{bibliography}

%-------------------------------------------------------------------------
%\newpage

\end{document}

### 2017-11-29 cpp/pnsolver/include/MultigridSolver.h
reworked multigrid code and cda stuff

@@ -1,0 +1,222 @@
+#pragma once
+#include <iostream>
+#include <util/timer.h>
+#include <Eigen/Sparse>
+
+
+
+#include <PNSystem.h>
+
+
+
+
+
+
+struct MultigridSolver
+{
+	// we have rowmajor matrix so that we can iterate over row elements
+	typedef Eigen::SparseMatrix<double, Eigen::RowMajor> RealMatrix;
+	typedef Eigen::VectorXd RealVector;
+
+	struct Level
+	{
+		Level( int index = 0 ):
+		    coarse(0),
+		    index(index),
+		    numPreSmoothingIterations(1),
+		    numPostSmoothingIterations(1)
+		{
+		}
+
+		double computeRMSE()
+		{
+			//std::cout << A.rows() << " " << A.cols() << std::endl;
+			//std::cout << x.rows() << " " << x.cols() << std::endl;
+			//std::cout << b.rows() << " " << b.cols() << std::endl;
+			return (b-A*x).norm();
+		}
+
+		int index;
+		int numPreSmoothingIterations;
+		int numPostSmoothingIterations;
+		RealMatrix A;
+		RealVector b;
+		RealVector x;
+		RealVector r;
+		RealMatrix upsample; // converts x to the next coarser grid
+		RealMatrix downsample; // converts from the next coarser grid to this grid
+		Level* coarse; // link to the next coarser level
+
+
+		// debugging
+		PNSystem* sys_level;
+	};
+
+	MultigridSolver()
+	{
+	}
+
+	MultigridSolver( int numLevels )
+	{
+		for( int i=0;i<numLevels;++i )
+			addLevel();
+	}
+
+	~MultigridSolver();
+
+	Level* addLevel();
+	Level* getLevel(int index);
+	int getNumLevels()const;
+
+	// runs a number of Gauss-Seidel iterations on the given problem
+	void run_gs_iterations( RealMatrix& A, RealVector& b, RealVector& x, int numIterations )
+	{
+		for( int k=0;k<numIterations;++k )
+		{
+			// iterate all rows
+			for( int i=0;i<A.rows();++i )
+			{
+				double aii = 0.0;
+				double sum = b(i);
+				// iterate all non-zero column elements (this is why we need RowMajor storage order for A)
+				for( RealMatrix::InnerIterator it(A, i);it;++it )
+				{
[…]

### 2017-11-29 cpp/pnsolver/include/PNVolume.h

@@ -22,7 +22,9 @@ struct PNVolume
 	//void setAbsorptionScattering( Field3d::Ptr absorption, Field3d::Ptr scattering );
 
 	void setEmission(int l, int m, Field3d::Ptr field);
+	void setEmission(int sh_index, Field3d::Ptr field);
 	void setPhase( int l, int m, Field3d::Ptr field);
+	void setPhase( int sh_index, Field3d::Ptr field);
 
 
 	V3d evalExtinction( const P3d& pWS, bool debug = false )const;
@@ -32,15 +34,9 @@ struct PNVolume
 	V3d evalEmission(int l, int m , const P3d& pWS)const;
 	V3d evalPhase(int l, int m , const P3d& pWS)const;
 
-	/*
-	P3d localToWorld(const P3d& pLS)const;
-	P3d worldToLocal(const P3d& pWS)const;
-
-	void setLocalToWorld( const Transformd& localToWorld );
-	void setBound( Box3d boundWS );
-	const Box3d& getBound()const;
-	*/
 	Domain& getDomain();
+
+	Ptr downsample(); // this creates the next coarser mipmap level of this problem
 private:
 	Domain m_domain;
 	// rgb dependent albedo and extinction values (evaluated in local space)

### 2017-11-29 cpp/pnsolver/include/common/Domain.h

@@ -101,25 +101,6 @@ struct Domain
 	{
 		return m_bound.getExtents();
 	}
-
-	Domain downsample()const
-	{
-		V3i res_fine = getResolution();
-
-		bool is2D = res_fine[2] == 1;
-
-		// for multigrid, we require the resolution to be even,
-		// so that we can do restriction and interpolation straigh forwardly
-		if( (res_fine[0]%2!=0)||(res_fine[1]%2!=0)||(!is2D && (res_fine[2]%2!=0)))
-			throw std::runtime_error("Domain::downsample currently requires even resolution");
-
-		V3i res_coarse( res_fine[0]/2, res_fine[1]/2, is2D ? 1:res_fine[2]/2 );
-
-		return Domain(  getBound().getExtents(),
-						res_coarse,
-						getBound().min );
-	}
-
 private:
 	V3i m_resolution;
 	V3d m_voxelsize;

### 2017-11-29 cpp/pnsolver/include/fields/Field.h

@@ -13,6 +13,10 @@ struct Field
 	// this is required for delta tracking where we need to know the maximum
 	//virtual std::pair<T, T> getValueRange()const=0;
 	virtual T getMaxValue()const=0;
+
+	// returns a coarser mimap version of this field
+	// this is used by the multigrid solver
+	virtual Ptr downsample()const=0;
 };
 
 typedef Field<double> Fieldd;
@@ -43,6 +47,12 @@ struct ConstantField : public Field<T>
 		return m_value;
 	}
 
+	// returns a coarser mimap version of this field
+	// this is used by the multigrid solver
+	virtual Field::Ptr downsample()const
+	{
+		return std::make_shared<ConstantField>(m_value);
+	}
 private:
 	T m_value;
 };

### 2017-11-29 cpp/pnsolver/include/fields/VoxelGridField.h

@@ -96,6 +96,56 @@ struct VoxelGridField : public Field<T>
 		m_voxelgrid.save(filename);
 	}
 
+	// returns a coarser mimap version of this field
+	// this is used by the multigrid solver
+	virtual Field::Ptr downsample()const
+	{
+		V3i res_fine = m_voxelgrid.getResolution();
+		bool is2D = res_fine[2] == 1;
+
+		// for downsampling, we require the resolution to be even,
+		// so that we can do it straight forwardly
+		if( (res_fine[0]%2!=0)||(res_fine[1]%2!=0)||(!is2D && (res_fine[2]%2!=0)))
+			throw std::runtime_error("VoxelGridField::downsample currently requires even resolution");
+
+		V3i res_coarse( res_fine[0]/2, res_fine[1]/2, is2D ? 1:res_fine[2]/2 );
+
+		Ptr result = std::make_shared<VoxelGridField<T>>( res_coarse, (T*)0 );
+
+		// create coarse from fine voxels
+		if(is2D)
+		{
+			for(int i_coarse=0;i_coarse<res_coarse[0];++i_coarse)
+				for(int j_coarse=0;j_coarse<res_coarse[1];++j_coarse)
+				{
+					auto v = m_voxelgrid.sample(i_coarse*2+0, j_coarse*2+0, 0);
+					v+= m_voxelgrid.sample(i_coarse*2+1, j_coarse*2+0, 0);
+					v+= m_voxelgrid.sample(i_coarse*2+1, j_coarse*2+1, 0);
+					v+= m_voxelgrid.sample(i_coarse*2+0, j_coarse*2+1, 0);
+					result->getVoxelGrid().lvalue( i_coarse, j_coarse, 0 ) = v/4.0;
+				}
+		}else
+		{
+			for(int i_coarse=0;i_coarse<res_coarse[0];++i_coarse)
+				for(int j_coarse=0;j_coarse<res_coarse[1];++j_coarse)
+					for(int k_coarse=0;k_coarse<res_coarse[2];++k_coarse)
+					{
+						auto v = m_voxelgrid.sample(i_coarse*2+0, j_coarse*2+0, k_coarse*2+0);
+						v+= m_voxelgrid.sample(i_coarse*2+1, j_coarse*2+0, k_coarse*2+0);
[…]

### 2017-11-29 cpp/pnsolver/src/MultigridSolver.cpp

@@ -1,0 +1,518 @@
+#include <MultigridSolver.h>
+
+
+
+
+
+
+
+MultigridSolver::~MultigridSolver()
+{
+	for( auto& lvl:m_levels )
+		delete lvl;
+}
+
+MultigridSolver::Level* MultigridSolver::addLevel()
+{
+	int index = m_levels.size();
+	Level* lvl = new Level(index);
+
+	if(!m_levels.empty())
+		m_levels.back()->coarse = lvl;
+
+	m_levels.push_back(lvl);
+	return lvl;
+}
+
+MultigridSolver::Level* MultigridSolver::getLevel(int index)
+{
+	return m_levels[index];
+}
+
+int MultigridSolver::getNumLevels()const
+{
+	return m_levels.size();
+}
+
+
+void buildUpAndDownsamplingMatrices2( PNSystem& sys_fine, MultigridSolver::RealMatrix& downsampleMatrix, MultigridSolver::RealMatrix& upsampleMatrix )
+{
+	PNSystem::VoxelManager& vm_fine = sys_fine.getVoxelManager();
+	PNSystem::VoxelManager vm_coarse = vm_fine.downsample();
+
+
+	PNSystem::MatrixBuilderd downsampleMatrixBuilder;
+	PNSystem::MatrixBuilderd upsampleMatrixBuilder;
+
+	// 2d stamps
+	std::vector<std::vector<std::pair<V3i, double>>> stamps(8);
+
+	if( vm_fine.is2D() )
+	{
+		stamps[0].push_back(std::make_pair(V3i(-1, -1, 0), 0.5*0.5*1.0));
+		stamps[0].push_back(std::make_pair(V3i(-1, 0, 0), 0.5*1.0*1.0));
+		stamps[0].push_back(std::make_pair(V3i(-1, 1, 0), 0.5*0.5*1.0));
+		stamps[0].push_back(std::make_pair(V3i(0, -1, 0), 1.0*0.5*1.0));
+		stamps[0].push_back(std::make_pair(V3i(0, 0, 0), 1.0*1.0*1.0));
+		stamps[0].push_back(std::make_pair(V3i(0, 1, 0), 1.0*0.5*1.0));
[…]

### 2017-11-29 cpp/pnsolver/src/PNVolume.cpp

@@ -48,18 +48,26 @@ void PNVolume::setExtinctionAlbedo( Field3d::Ptr extinction, Field3d::Ptr albedo
 
 void PNVolume::setEmission( int l, int m, Field3d::Ptr field)
 {
-	int index = sph::index(l, m);
-	if( index>=m_field_q.size() )
-		m_field_q.resize(index+1);
-	m_field_q[index] = field;
+	setEmission(sph::index(l, m), field);
+}
+
+void PNVolume::setEmission(int sh_index, Field3d::Ptr field)
+{
+	if( sh_index>=m_field_q.size() )
+		m_field_q.resize(sh_index+1);
+	m_field_q[sh_index] = field;
 }
 
 void PNVolume::setPhase(int l, int m, Field3d::Ptr field)
 {
-	int index = sph::index(l, m);
-	if( index>=m_field_p.size() )
-		m_field_p.resize(index+1);
-	m_field_p[sph::index(l, m)] = field;
+	setPhase(sph::index(l, m), field);
+}
+
+void PNVolume::setPhase( int sh_index, Field3d::Ptr field)
+{
+	if( sh_index>=m_field_p.size() )
+		m_field_p.resize(sh_index+1);
+	m_field_p[sh_index] = field;
 }
 
 
@@ -107,4 +115,43 @@ V3d PNVolume::evalPhase( int l, int m, const P3d& pWS )const
 	if( index < m_field_p.size() && m_field_p[index] )
 		return m_field_p[index]->eval(m_domain.worldToLocal(pWS));
 	return V3d(0.0);
+}
+
+// this creates the next coarser mipmap level of this problem
+// used for multigrid solver
+PNVolume::Ptr PNVolume::downsample()
+{
+	// downsample domain and check if we can actually do it
+	V3i res_fine = m_domain.getResolution();
+
+	bool is2D = res_fine[2] == 1;
+
+	// for multigrid, we require the resolution to be even,
+	// so that we can do restriction and interpolation straigh forwardly
+	if( (res_fine[0]%2!=0)||(res_fine[1]%2!=0)||(!is2D && (res_fine[2]%2!=0)))
+		return PNVolume::Ptr();
+
+	V3i res_coarse( res_fine[0]/2, res_fine[1]/2, is2D ? 1:res_fine[2]/2 );
+
+	// we stop if any dimension reduces to less than 2.
+	if( (res_coarse[0]<2)||(res_coarse[1]<2)||(!is2D && (res_coarse[2]<2)))
[…]

### 2017-11-29 cpp/pnsolver/src/main.cpp

@@ -22,10 +22,12 @@
 #include <PNVolume.h>
 
 #include <solver.h>
+#include <MultigridSolver.h>
 
 
 namespace py = pybind11;
 
+void save_solution( const std::string& filename, PNSystem& sys, const Eigen::VectorXd& x );
 
 
 void setup_solver( Solver& mg, PNSystem& sys, int numLevels = 1 )
@@ -176,7 +178,7 @@ std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_multigrid2(P
 	return std::make_tuple(std::get<0>(result), std::get<1>(result), std::get<2>(result));
 }
 
-
+/*
 std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_gs(PNSystem& sys)
 {
 	Solver solver;
@@ -184,6 +186,7 @@ std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_gs(PNSystem&
 	auto result = solver.solve_gs();
 	return std::make_tuple(std::get<0>(result), std::get<1>(result), std::get<2>(result));
 }
+*/
 
 std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_blockgs(PNSystem& sys)
 {
@@ -193,9 +196,13 @@ std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_blockgs(PNSy
 	return std::make_tuple(std::get<0>(result), std::get<1>(result), std::get<2>(result));
 }
 
-std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_cg(PNSystem& sys, double tol )
+// this will solve the normal form of Ax=b by explicitly computing A^TA and A^Tb and using CG to solve it
+std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_ls_cg(PNSystem& sys, double tol )
 {
 	sys.build();
+
+	std::vector<double> list_rmse;
+	std::vector<double> list_time;
 
 	Timer timer;
 
@@ -207,6 +214,61 @@ std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_cg(PNSystem&
 	std::cout << "computing A^T*A...\n";std::flush(std::cout);
[…]

### 2017-11-29 cpp/renderer/include/camera.h

@@ -135,7 +135,7 @@ struct PerspectiveCamera : public Camera
 		Eigen::Matrix4d cameraToClip;
 		cameraToClip <<
 			cot, 0,   0,   0,
-			0, cot,   0,   0,
+			0, -cot,   0,   0,
 			0,   0,   m_z_direction*-(m_nearClip+m_farClip)/(m_nearClip-m_farClip), m_z_direction*-2.0*m_nearClip*m_farClip/(m_nearClip-m_farClip),
 			0,   0,   m_z_direction*1,   0;
 
@@ -147,12 +147,19 @@ struct PerspectiveCamera : public Camera
 					);
 
 		m_rasterToCamera = cameraToRaster.inverse();
+
+
+		V3d test_d(0.0, 1.0, 0.0);
+		P3d test_rasterP = cameraToRaster*test_d;
+		std::cout << "test_rasterP=" << test_rasterP.toString() << std::endl;
 	}
 
 
 
 	virtual void sampleRay( const Point2d& rasterP, Ray3d& ray, bool debug = false )const override
 	{
+		if(debug)
+			std::cout << "Camera::sampleRay: rasterP=" << rasterP.toString() << std::endl;
 		P2d tmp(0.0, 0.0);
 		double m_focusDistance = m_farClip;
 
@@ -160,17 +167,24 @@ struct PerspectiveCamera : public Camera
 		/* Compute the corresponding position on the
 		   near plane (in local camera space) */
 		P3d nearP = m_rasterToCamera * P3d( rasterP.x(), rasterP.y(), 0.0f);
+		if(debug)
+			std::cout << "Camera::sampleRay: nearP=" << nearP.toString() << std::endl;
 
 
 		P3d apertureP(tmp.x(), tmp.y(), 0.0f);
 
-		/* Sampled position on the focal plane */
-		P3d focusP = nearP * (m_z_direction*m_focusDistance / nearP.z());
+		// Sampled position on the focal plane
+		// we basically scale the point
+		P3d focusP = nearP * (m_focusDistance / std::abs(nearP.z()));
+		if(debug)
+			std::cout << "Camera::sampleRay: focusP=" << focusP.toString() << std::endl;
 
 		/* Aperture position */
 		/* Turn these into a normalized ray direction, and
 		   adjust the ray interval accordingly */
 		V3d d = (focusP - apertureP).normalized();
+		if(debug)
[…]

### 2017-11-29 cpp/renderer/include/fields/VoxelGridField.h

@@ -55,6 +55,9 @@ struct VoxelGridField : public Field<T>
 
 	virtual T eval( const P3d& pLS )const override
 	{
+		//std::cout << "VoxelGridField::eval pLS=" << pLS.toString() << std::endl;
+		//std::cout << "VoxelGridField::eval pVS=" << localToVoxel(pLS).toString() << std::endl;
+
 		return m_voxelgrid.evaluate(localToVoxel(pLS));
 	}
 

### 2017-11-29 cpp/renderer/include/integrator.h

@@ -50,7 +50,7 @@ struct Integrator
 
 
 // returns sigma_t at sampled position (is invalid when we exceeded maxt)
-double delta_tracking( const Scene* scene, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t );
+double delta_tracking(const Scene* scene, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t , bool debug = false);
 
 struct Vertex
 {
@@ -118,7 +118,7 @@ struct TraceInfo
 
 	bool intersect()
 	{
-		return scene->intersect(Ray3d( current_vertex.getPosition(), current_direction ), its);
+		return scene->intersect(Ray3d( current_vertex.getPosition(), current_direction ), its, debug);
 	}
 
 
@@ -127,7 +127,7 @@ struct TraceInfo
 	{
 		Ray3d ray( current_vertex.getPosition(), current_direction );
 		V3d sigma_t;
-		double distance = delta_tracking( scene, ray, its.t, 0, rng, sigma_t);
+		double distance = delta_tracking( scene, ray, its.t, 0, rng, sigma_t, debug);
 
 		if( distance < its.t )
 		{
@@ -143,6 +143,9 @@ struct TraceInfo
 			if( debug )
 			{
 				std::cout << "TraceInfo::propagate: got scattering event!\n";
+				std::cout << "ray=" << ray.toString() << std::endl;
+				std::cout << "t=" << its.t << std::endl;
+				std::cout << "pWS=" << pWS.toString() << std::endl;
 			}
 		}else
 		{

### 2017-11-29 cpp/renderer/include/integrators/directpn.h

@@ -1,0 +1,56 @@
+#pragma once
+
+
+
+#include <integrator.h>
+
+#include <scene.h>
+#include <math/frame.h>
+
+#include <volume.h>
+
+
+#include <PNSolution.h>
+
+
+// intersects the volume bound and starts volume path tracing within the volume
+// will stop once the path exits the volume
+struct DirectPN : public Integrator
+{
+	typedef std::shared_ptr<DirectPN> Ptr;
+
+	DirectPN( PNSolution* pns,
+	          bool doSingleScattering = true,
+	          bool doMultipleScattering = true );
+
+	V3d trace( TraceInfo& ti, RNGd& rng )const;
+
+	virtual V3d Li( const Scene* scene, RadianceQuery& rq, RNGd& rng )const override;
+	virtual V3d sample_transmittance( const Scene* scene, const Ray3d& ray, double maxt, RNGd& rng )const override
+	{
+		V3d sigma_t;
+		double distance = delta_tracking( scene, ray, maxt, 0, rng, sigma_t );
+
+		if( distance < maxt )
+			// since we return 0, we dont need to produce the pdf in the denominator
+			return V3d(0.0, 0.0, 0.0);
+
+		//NB: transmittance sampling pdf cancels out with the transmittance term
+
+		return V3d(1.0, 1.0, 1.0);
+	}
+
+	virtual std::string toString()const override
+	{
+		std::ostringstream ss;
+		ss << "DirectPN singleScattering=" << m_doSingleScattering << " multipleScattering=" << m_doMultipleScattering << std::endl;
+		return ss.str();
+	}
+
+private:
+	//PNSolution::Ptr m_pns;
+	PNSolution* m_pns;
+
+	bool m_doSingleScattering;
+	bool m_doMultipleScattering;
+};

### 2017-11-29 cpp/renderer/include/scene.h

@@ -51,7 +51,7 @@ struct Scene
 	const Integrator* integrator;
 
 
-	bool intersect( const Ray3d& ray, Intersection& its )const;
+	bool intersect(const Ray3d& ray, Intersection& its , bool debug=false)const;
 	V3d sample_attenuated_directlight( LightSample& ls, RNGd& rng, bool debug = false )const;
 };
 

### 2017-11-29 cpp/renderer/src/Integrator.cpp

@@ -11,7 +11,7 @@
 
 
 // returns sigma_t at sampled position (is invalid when we exceeded maxt)
-double delta_tracking( const Scene* scene, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t )
+double delta_tracking( const Scene* scene, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t, bool debug )
 {
 	double sigma_t_max = scene->volume->getMaxExtinction()[component];
 
@@ -24,7 +24,7 @@ double delta_tracking( const Scene* scene, const Ray3d& ray, double maxt, int co
 		if(t>= maxt)
 			break;
 
-		sigma_t = scene->volume->evalExtinction(ray(t));
+		sigma_t = scene->volume->evalExtinction(ray(t), debug);
 
 		// russian roulette
 		if(rng.next1D()<sigma_t[component]/sigma_t_max)

### 2017-11-29 cpp/renderer/src/PNSolution.cpp

@@ -103,6 +103,13 @@ double PNSolution::eval( const P3d& pWS, const V3d& direction )const
 double PNSolution::evalCoefficient(const P3d &pWS, int coeff_index)const
 {
 	P3d pVS = worldToVoxel(pWS);
+	/*
+	P3d pLS = worldToLocal(pWS);
+	std::cout << "PNSolution::::evalCoefficient pLS=" << pLS.x() << " " << pLS.y() << " " << pLS.z() << std::endl;
+	std::cout << "PNSolution::::evalCoefficient vsP=" << pVS.x() << " " << pVS.y() << " " << pVS.z() << std::endl;
+	std::cout << "PNSolution::::evalCoefficient m_bound=" << m_bound.min.toString() << " " << m_bound.max.toString() << std::endl;
+	*/
+
 
 	// take sample location within voxel into account
 	pVS -= V3d(0.5, 0.5, 0.5);

### 2017-11-29 cpp/renderer/src/integrators/directpn.cpp

@@ -1,0 +1,183 @@
+#include <integrators/directpn.h>
+
+
+
+
+#include <houio/HouGeoIO.h>
+#include <fields/VoxelGridField.h>
+
+houio::ScalarField::Ptr g_temp;
+
+
+DirectPN::DirectPN( PNSolution* pns,
+					bool doSingleScattering,
+					bool doMultipleScattering):
+					Integrator(),
+					m_pns(pns),
+					m_doSingleScattering(doSingleScattering),
+					m_doMultipleScattering(doMultipleScattering)
+{
+	// debug
+	/*
+	{
+		std::string filename = "C:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_p1_2.pns.bgeo";
+		g_temp = houio::HouGeoIO::importVolume(filename);
+
+		P3d pWS( -0.034216, 0.321882, -0.204602 );
+
+
+		houio::math::V3f pWS2( float(pWS.x()), float(pWS.y()), float(pWS.z()) );
+		float e = g_temp->evaluate( g_temp->worldToVoxel(pWS2) );
+		std::cout << "houio e=" << e << std::endl;
+
+
+		double pns_e = m_pns->evalCoefficient(pWS, 0);
+		std::cout << "pns e=" << pns_e << std::endl;
+	}
+	*/
+
+	/*
+	int coeff_index = 0;
+	int numCoeffs = pns->getNumCoeffs();
+	V3i res = pns->getResolution();
+	std::vector<double> data(res[0]*res[1]*res[2]);
+
+	for( int i=0;i<res[0];++i )
+		for( int j=0;j<res[1];++j )
+			for( int k=0;k<res[2];++k )
+			{
+				int index_dst = i*res[2]*res[1] + j*res[2] + k;
+				int index_src = pns->getIndex(V3i(i, j, k))+coeff_index;
+				data[index_dst] = pns->data()[index_src];
+			}
+
+	VoxelGridField3d t( res, data.data() );
+	*/
+
+}
+
+
+
+V3d DirectPN::trace( TraceInfo& ti, RNGd& rng )const
+{
+	V3d L(0.0f, 0.0f, 0.0f);
+
+	if(ti.debug)
+		std::cout <<"VolumePathTracer::trace\n";
+
+	// we only do a single bounce
+	{
+		if(ti.debug)
+			std::cout <<"VolumePathTracer::trace depth=" << ti.depth << std::endl;
+
[…]

### 2017-11-29 cpp/renderer/src/integrators/simplept.cpp

@@ -32,7 +32,7 @@ V3d SimplePT::Li( const Scene* scene, RadianceQuery& rq, RNGd& rng )const
 		ti.current_vertex = Vertex();
 		P3d pWS = rq.ray(mint+Epsilon);
 		if(rq.volume)
-			ti.current_vertex.setPosition(pWS, rq.volume->evalExtinction(pWS), rq.volume->evalAlbedo(pWS));
+			ti.current_vertex.setPosition(pWS, rq.volume->evalExtinction(pWS, rq.debug), rq.volume->evalAlbedo(pWS));
 		else
 			ti.current_vertex.setPosition(pWS, V3d(0.0, 0.0, 0.0), V3d(0.0, 0.0, 0.0));
 		ti.current_direction = rq.ray.d;

### 2017-11-29 cpp/renderer/src/main.cpp

@@ -17,6 +17,7 @@
 #include <integrators/simplept.h>
 #include <integrators/pnispt.h>
 #include <integrators/jispt.h>
+#include <integrators/directpn.h>
 
 #include <fields/VoxelGridField.h>
 
@@ -230,6 +231,12 @@ void render_thread( MonteCarloTaskInfo& ti, const GlobalRenderInfo* gi )
 			gi->scene->camera->sampleRay( P2d(x+0.5, y+0.5), rayWS, debug );
 			//ti.g.scene->camera->sampleRay( P2d(x+ti.rng.next1D(), y+ti.rng.next1D()), rayWS );
 
+			if(debug)
+			{
+				std::cout << "rayWS=" << rayWS.toString() << std::endl;
+				std::cout << "cam2world=\n" << gi->scene->camera->m_cameraToWorld.getMatrix() << std::endl;
+			}
+
 			// do raycast ---
 			try
 			{
@@ -289,14 +296,25 @@ Image::Ptr render( Volume::Ptr volume, Light::Ptr light, Camera::Ptr camera, Int
 	gi.scene = &scene;
 	gi.image = result.get();
 	gi.crop_window = Box2i( V2i(0,0), camera->getResolution() );
-	//gi.debug_pixel = V2i(1, 0);
+	//gi.debug_pixel = V2i(384, 155);
 
 	//int numSamples = 1;
 
+	/*
+	{
+		P3d pWS(0.229053, 0.408338, 0.433446);
+		volume->evalExtinction(pWS, true);
+	}
+	*/
+
+
+
+	///*
 	Terminator terminator(numSamples);
 	//Terminator terminator(90.0);
 	std::cout << "rendering image..."<< std::endl;std::flush(std::cout);
 	runGenericTasks<MonteCarloTaskInfo, GlobalRenderInfo>( render_thread, &gi, terminator, ThreadPool::getNumSystemCores() );
+	//*/
 
 
 	return result;
@@ -487,12 +505,18 @@ Camera::Ptr create_sphere_camera( const Eigen::Vector3d& pWS, int xres, int yres
 	return std::make_shared<SphereCamera>( pWS, xres, yres );
 }
 
-
-Integrator::Ptr create_simplept_integrator()
+Integrator::Ptr create_directpn_integrator( PNSolution::Ptr pns, bool doSingleScattering, bool doMultipleScattering )
 {
-	int maxDepth = std::numeric_limits<int>::max();
-	bool doSingleScattering = true;
[…]

### 2017-11-29 cpp/renderer/src/scene.cpp

@@ -2,15 +2,24 @@
 #include <volume.h>
 #include <integrator.h>
 
-bool Scene::intersect( const Ray3d& ray, Intersection& its )const
+bool Scene::intersect( const Ray3d& ray, Intersection& its, bool debug )const
 {
 	its.reset();
 
+	if(debug)
+	{
+		std::cout << "Scene::intersect ray=" << ray.toString() << std::endl;
+	}
+
 	double mint, maxt;
 	// TODO: take into account whether vertex is surface or volume (Epsilon)...
-	bool result = volume->intersectBound(ray, mint, maxt);
+	bool result = volume->intersectBound(ray, mint, maxt, debug);
 	if( result )
 	{
+		if(debug)
+		{
+			std::cout << "Scene::intersect: got intersection at t=" << maxt << std::endl;
+		}
 		//TODO: here we assume that mint is behind are exactly at the ray origin
 		// if the ray is outside the box, then mint would be the correct choice
 		its.t = maxt;

### 2017-11-29 cpp/renderer/src/volume.cpp

@@ -71,6 +71,11 @@ const Box3d& Volume::getBound()const
 }
 V3d Volume::evalExtinction( const P3d& pWS, bool debug )const
 {
+	if(debug)
+	{
+		std::cout << "Volume::evalExtinction pWS=" << pWS.toString() << std::endl;
+		std::cout << "Volume::evalExtinction pLS=" << worldToLocal(pWS).toString() << std::endl;
+	}
 	return m_field_extinction->eval(worldToLocal(pWS));
 }
 V3d Volume::evalAlbedo( const P3d& pWS, bool debug )const

### 2017-11-29 python/pnsolver/problems.py

@@ -243,14 +243,14 @@ def checkerboard3d():
 
 	return t
 
-def pointsource3d():
+def pointsource3d( res = 80 ):
 
 	# here we set some general parameters 
 	size = 2.0
 	#res = 20
 	#res = 100
 	#res = 100
-	res = 80
+	#res = 80
 	#res = 71
 	#res = 2
 	#res = 200
@@ -284,7 +284,7 @@ def pointsource3d():
 def nebulae():
 	size = 1.0
 	res = 64
-	domain = pnsolver.Domain( np.array([size, size, size]), np.array([res, res, res]), np.array([0.0, 0.0, 0.0]))
+	domain = pnsolver.Domain( np.array([size, size, size]), np.array([res, res, res]), np.array([-0.5, -0.5, -0.5]))
 	t = pnsolver.PNVolume( domain )
 
 	#sigma_t = 8.0
@@ -293,6 +293,11 @@ def nebulae():
 	#extinction_field = pnsolver.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64.bgeo")
 	extinction_field = pnsolver.load_voxelgridfield3d("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64_extinction.grid")
 	
+	# temp
+	#sigma_t = 8.0
+	#extinction_field = pnsolver.ConstantField3d(np.array([sigma_t, sigma_t, sigma_t]))
+
+
 	albedo = 0.9
 	albedo_field = pnsolver.ConstantField3d(np.array([albedo, albedo, albedo]))
 	t.setExtinctionAlbedo( extinction_field, albedo_field )

### 2017-11-29 python/pnsolver/render.py

@@ -84,8 +84,24 @@ def create_scene_nebulae():
 
 	volume = renderer.create_volume()
 	volume.setBound( np.array([-0.5, -0.5, -0.5+offset]), np.array([0.5, 0.5, 0.5+offset]) )
+
+	test = np.zeros((64, 64, 64, 3))
+	#for i in range(test.shape[0]):
+	#	for j in range(test.shape[1]):
+	#		for k in range(test.shape[2]):
+	#			t = float(j)/float(test.shape[1])
+	#			test[i, j, k, 0] = i
+	#			test[i, j, k, 1] = j
+	#			test[i, j, k, 2] = k
+	#test[3, 3, 3, :] = 1
+	#test[2, 3, 3, :] = 1
+	test[46, 57, 59, :] = 123.4
+
+	#sigma_t_field = renderer.VoxelGridField3d(test)
 	sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64.bgeo")
-	#sigma_t_field.save("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64_extinction.grid")
+	#sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae200.bgeo")
+	#sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/test.bgeo")
+	sigma_t_field.save("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64_extinction.grid")
 
 	#sigma_t_field = renderer.load_bgeo("test_sphere.bgeo")
 	albedo_field = renderer.create_constant_field3d( np.array([albedo, albedo, albedo]) )
@@ -107,7 +123,7 @@ def create_scene_nebulae():
 def validate_pointsource_fluence():
 	# assemble scene to render
 	#camera = load_camera( "results/pointsource/result_pointsource.scn", "cam1" )
-	integrator = renderer.create_simplept_integrator()
+	integrator = renderer.create_simplept_integrator(True, -1)
 	power = 4.0*np.pi
 	sigma_t = 8.0
 	albedo = 0.9
@@ -169,7 +185,7 @@ def validate_pointsource_fluence():
 
 def render_groundtruth_radial_distribution( pWS, filename ):
 
-	integrator = renderer.create_simplept_integrator()
[…]

### 2017-12-04 latex/egsr17-cgf/main-body.tex
continued working on higher order tests with nebulae

@@
   \item Lattice Boltzmann Methods \cite{Geist04}
-  \item Discrete Ordinates
-  \item Diffusion \cite{Stam95} \cite{Jakob10} \cite{Koerner14}
+  \item Discrete Ordinates \cite{Languenou95}
   \item $PN$-Method \cite{Kajiya84} 
+    \begin{itemize}
+        \item Kajia briefly explained the theory, but did not give any details on implementation or how to solve it.
+In fact, as Max~\cite{Max95} points out, it is not clear if he succeeded at all, as all of Kajia's results were produced with a simpler method.
+        \item We decided to use the $P_N$-method, because it has not been studied and applied in the context of computer graphics before.
+    \end{itemize}
+  \item Diffusion \cite{Stam95} \cite{Jakob10}
+  \item Variable Eddington Factors (VEF) \cite{Koerner14}
   \item Heuristics \cite{Kaplanyan10} \cite{Elek14}
@@
 
--give RTE
-%\begin{}
-%\end{}
+\begin{itemize}
+  \item give RTE
+  \item also introduce operator notation (used later when talking about least-squares form of the RTE)
+\end{itemize}
 
@@
 \section{Stencil Generation}
-\section{Staggered Grid Solver}
+\begin{itemize}
+  \item Manually generating the stencil code for higher orders is intractable due to the increasing number of coefficient equations.
+We therefore automatically generate the stencil code directly from a computer algebra representation of the expanded RTE.
+  \item Arbitrary order.
+Staggered and collocated grids.
+Neumann and dirichlet boundary conditions.
+Generalized to arbitrary coupled PDE's.
+Could be extended to other problem domains, such as shallow water simulation.
+\end{itemize}
+
+\section{System Solving}
+
+\begin{itemize}
+  \item Staggered grids are required in order to get rid of oscillating artefacts in the $P_N$-solution.
+  \item CDA produces a non-symmetric diagonal-dominant matrix A, which allows the application of Gauss-Seidel method.
+  \item PN-equations produce a non-symmetric, non-diagonal dominant matrix A, which cannot be solved with standard methods.
+  \item Bringing $A\vec{x}=\vec{b}$ into normal form $A^TA\vec{x}=A^T\vec{b}$, produces a symmetric, positive definit system matrix $A^TA$, which allows application of standard methods, such as Conjugate-Gradient.
+  \item Solving the normal form is equivalent to solving the least-squares form of the RTE equation. (maybe give it in operator notation).
+  \item The system matrix of the normal form has a higher condition number than the standard form.
+\end{itemize}
+
+\subsection*{CDA vs. $P_1$}
+
+\begin{itemize}
+  \item CDA is a degenerated form of $P_1$.
+It is derived by isolating the flux-vector on one side of the vector-equation formed by the $l=1$ SH-band equations.
+This isolation requires division by the extinction coefficient, introducing a $\frac{1}{\sigma_t}$ factor which diverges as $\sigma_t$ approaches zero.
+Thresholding to some minimum is required, in order to be able to solve the system for vacuum regions.
+  \item $P_1$ does not require any thresholding of $\sigma_t$ as it does not contain $\sigma_t$ as a denominator. It therefore can deal with vacuum regions without modifications.
+  \item (needs validation) Further, in the presence of vacuum or near vacuum regions, the condition number for CDA is higher than for $P_1$, because of small extinction values in the denominator of the diffusion coefficient.
+  \item Using the normal form for CDA will further increase the condition number when vacuum regions are present and significantly decreases convergence.
+\end{itemize}
+
 \section{Path-guiding}
@@
   \item Extension to surface rendering by dealing vacuum regions and extending to anisotropic RTE.
+  \item Apply our automated stencil generation to other problem domains where coupled PDE's are present.
 \end{itemize}

### 2017-12-04 cpp/pnsolver/include/PNVolume.h

@@ -37,6 +37,8 @@ struct PNVolume
 	Domain& getDomain();
 
 	Ptr downsample(); // this creates the next coarser mipmap level of this problem
+
+	void setExtinctionMinimumThreshold(double min_threshold);
 private:
 	Domain m_domain;
 	// rgb dependent albedo and extinction values (evaluated in local space)
@@ -44,6 +46,7 @@ private:
 	Field3d::Ptr       m_field_albedo;
 	std::vector<Field3d::Ptr> m_field_q; // field for each SH-coefficient of q
 	std::vector<Field3d::Ptr> m_field_p; // field for each SH-coefficient of p (phase function)
+	double m_extinction_minimum_threshold;
 
 	/*
 	//PhaseFunction::Ptr m_phaseFunction;

### 2017-12-04 cpp/pnsolver/src/PNVolume.cpp

@@ -11,7 +11,8 @@
 
 
 PNVolume::PNVolume(const Domain &domain):
-	m_domain(domain)
+	m_domain(domain),
+	m_extinction_minimum_threshold(0.0)
 	//m_worldToLocal(),
 	//m_localToWorld()//,
 	//m_bboxLS( P3d(0.0f,0.0f,0.0f), P3d(1.0f,1.0f,1.0f) ),
@@ -21,6 +22,11 @@ PNVolume::PNVolume(const Domain &domain):
 	setExtinctionAlbedo( std::make_shared<ConstantField3d>(V3d(1.0, 1.0, 1.0)),
 						 std::make_shared<ConstantField3d>(V3d(0.9, 0.9, 0.9)) );
 }
+void PNVolume::setExtinctionMinimumThreshold(double min_threshold)
+{
+	m_extinction_minimum_threshold = min_threshold;
+}
+
 void PNVolume::setExtinctionAlbedo( Field3d::Ptr extinction, Field3d::Ptr albedo )
 {
 	//std::pair<V3d, V3d> valueRange = extinction->getValueRange();
@@ -78,7 +84,18 @@ Domain& PNVolume::getDomain()
 
 V3d PNVolume::evalExtinction( const P3d& pWS, bool debug )const
 {
-	return m_field_extinction->eval(m_domain.worldToLocal(pWS));
+	//return m_field_extinction->eval(m_domain.worldToLocal(pWS));
+
+	///*
+	// for testing we apply here a minimum threshold
+	V3d ext = m_field_extinction->eval(m_domain.worldToLocal(pWS));
+
+	double threshold = m_extinction_minimum_threshold;
+	ext[0] = std::max(ext[0], threshold);
+	ext[1] = std::max(ext[1], threshold);
+	ext[2] = std::max(ext[2], threshold);
+	return ext;
+	//*/
 }
 V3d PNVolume::evalAbsorption( const P3d& pWS, bool debug )const
 {

### 2017-12-04 cpp/pnsolver/src/main.cpp

@@ -515,6 +515,8 @@ std::tuple<Eigen::VectorXd, Eigen::VectorXd, Eigen::VectorXd> solve_ls_lscg(PNSy
 			x = x + alpha*p;
 			r = r - alpha*Ap;
 			double rsnew = r.squaredNorm();
+			if( iteration % 10 == 0 )
+				std::cout << "rmse=" << std::sqrt(rsnew) << std::endl;
 			if( std::sqrt(rsnew) < tol )
 				break;
 			p = r + (rsnew/rsold)*p;
@@ -799,6 +801,12 @@ Eigen::MatrixXcd createRealToComplexConversionMatrix( int order )
 // -convert coefficients from real to complex, to get the true SH coefficients of the radiance field (see p5 in starmap paper)
 void save_solution( const std::string& filename, PNSystem& sys, const Eigen::VectorXd& x )
 {
+	if(sys.getResolution()[2] == 1)
+	{
+		std::cout << "save_solution: unable to create and save a PNSolution file for 2d problem due to different number of coefficients.\n";
+		return;
+	}
+
 	// the solution is given in real valued coefficients. However, the complex->real conversion matrix used during stencil
 	// generation is that from the starmap paper, which is different than our conversion matrix. Therefore we need to first
 	// convert from real to complex using the inverse starmap transform, and then we can use our conversion matrix to convert
@@ -806,14 +814,12 @@ void save_solution( const std::string& filename, PNSystem& sys, const Eigen::Vec
 	// and then convert from complex to real using sph::complex2RealConversionMatrix.
 
 	// TODO: use our own conversion matrix during stencil generation straight away...
-
 	Eigen::VectorXcd x_complex =buildBlockDiagonalMatrix( createRealToComplexConversionMatrix(sys.getStencil().order), sys.getNumVoxels() )*
 								sys.removeStaggering(x);
 
 	Eigen::VectorXd x_real = (buildBlockDiagonalMatrix( sph::buildComplexToRealConversionMatrix(sys.getStencil().order), sys.getNumVoxels() )*
 							 x_complex).real();
[…]

### 2017-12-04 cpp/pnsolver/src/stencil_cda.cpp

@@ -18,14 +18,14 @@ void stencil_cda(PNSystem::Stencil::Context& ctx)
 
 	// for classical diffusion, we need to clamp extinction to avoid division by zero
 	//double extinction_tol = 1.0e-4;
-	double extinction_tol = 6.0;
+	//double extinction_tol = 6.0;
 
-	double D_xph = 1.0/(3.0*std::max(problem.evalExtinction(voxel_center_WS+V3d(vs[0], 0.0, 0.0))[color_channel], extinction_tol) );
-	double D_xmh = 1.0/(3.0*std::max(problem.evalExtinction(voxel_center_WS-V3d(vs[0], 0.0, 0.0))[color_channel], extinction_tol) );
-	double D_yph = 1.0/(3.0*std::max(problem.evalExtinction(voxel_center_WS+V3d(0.0, vs[1], 0.0))[color_channel], extinction_tol) );
-	double D_ymh = 1.0/(3.0*std::max(problem.evalExtinction(voxel_center_WS-V3d(0.0, vs[1], 0.0))[color_channel], extinction_tol) );
-	double D_zph = 1.0/(3.0*std::max(problem.evalExtinction(voxel_center_WS+V3d(0.0, 0.0, vs[2]))[color_channel], extinction_tol) );
-	double D_zmh = 1.0/(3.0*std::max(problem.evalExtinction(voxel_center_WS-V3d(0.0, 0.0, vs[2]))[color_channel], extinction_tol) );
+	double D_xph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(vs[0], 0.0, 0.0))[color_channel]);
+	double D_xmh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(vs[0], 0.0, 0.0))[color_channel]);
+	double D_yph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(0.0, vs[1], 0.0))[color_channel]);
+	double D_ymh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(0.0, vs[1], 0.0))[color_channel]);
+	double D_zph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(0.0, 0.0, vs[2]))[color_channel]);
+	double D_zmh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(0.0, 0.0, vs[2]))[color_channel]);
 
 	// compute diffusion coefficient
 

### 2017-12-04 cpp/pnsolver/src/stencil_fopn_p1_sg.cpp

@@ -38,6 +38,8 @@ void stencil_fopn_p1_sg(PNSystem::Stencil::Context& ctx)
 	//M_2 ---
 	// is constant
 
+	//double lambda_l = std::sqrt( 4.0*M_PI );
+
 	//M_3 ---
 	Eigen::Matrix<double, 4, 4> M_3_real_staggered[8];
 	for( int i=0;i<8;++i )

### 2017-12-04 cpp/renderer/include/fields/VoxelGridField.h

@@ -91,7 +91,7 @@ struct VoxelGridField : public Field<T>
 
 	P3d voxelToLocal(const P3d& pVS)const
 	{
-		return m_voxelgrid.voxelToLocal(pLS);
+		return m_voxelgrid.voxelToLocal(pVS);
 	}
 
 	void save( const std::string& filename )

### 2017-12-04 cpp/renderer/src/PNSolution.cpp

@@ -97,7 +97,8 @@ double PNSolution::eval( const P3d& pWS, const V3d& direction )const
 	sphericalCoordinates(direction, theta, phi);
 
 	double result = sph::eval(theta, phi, coeffs.data(), m_order);
-	return result/std::sqrt(4.0*M_PI);
+	//return result/std::sqrt(4.0*M_PI);
+	return result;
 }
 
 double PNSolution::evalCoefficient(const P3d &pWS, int coeff_index)const
@@ -156,6 +157,7 @@ double PNSolution::evalCoefficient(const P3d &pWS, int coeff_index)const
 	coeff += m_data[voxel_indices[6] + coeff_index]*(1.0-tx)*ty*tz;
 	coeff += m_data[voxel_indices[7] + coeff_index]*tx*ty*tz;
 
+	//return coeff/std::sqrt(4.0*M_PI);
 	return coeff;
 }
 

### 2017-12-04 cpp/renderer/src/integrators/directpn.cpp

@@ -132,9 +132,10 @@ V3d DirectPN::trace( TraceInfo& ti, RNGd& rng )const
 
 				///*
 				// constant phase function allows us to evaluate the zero coefficient directly
+				// there is a std::sqrt(4pi) factor which comes from integrating the sh expanded field over solid angle
 				P3d pWS = ti.current_vertex.getPosition();
 				V3d phase_times_sigma_s = INV_FOURPI*ti.current_vertex.m_sigma_s;
-				double e = m_pns->evalCoefficient(pWS, 0);
+				double e = m_pns->evalCoefficient(pWS, 0)*std::sqrt(4.0*M_PI);
 				L += V3d(e).cwiseProduct(phase_times_sigma_s).cwiseProduct(ti.throughput_over_pdf);
 				//*/
 			}else

### 2017-12-04 cpp/renderer/src/main.cpp

@@ -50,10 +50,14 @@ struct GlobalRenderInfo
 
 struct GlobalComputeFluenceInfo
 {
-	GlobalComputeFluenceInfo()
+	GlobalComputeFluenceInfo():
+		do_direct_light(true),
+		do_indirect_light(true)
 	{
 	}
 
+	bool do_direct_light;
+	bool do_indirect_light;
 	const Scene* scene;
 
 	std::vector<P3d>* sensor_points;
@@ -101,27 +105,35 @@ void compute_fluence_thread( MonteCarloTaskInfo& ti, const GlobalComputeFluenceI
 		bool debug = false;
 
 		P3d pWS = (*gi->sensor_points)[point_index];
-		V3d direction = sampleSphere(ti.rng);
-		double direction_pdf = sampleSpherePDF();
-
-		Ray3d rayWS( pWS, direction );
 
 		// do raycast ---
 		try
 		{
-			RadianceQuery rq;
-			rq.ray = rayWS;
-			rq.pixel = V2i(-1, -1);
-			rq.debug = debug;
-			if(gi->scene->volume->getBound().contains(pWS))
-				rq.volume = gi->scene->volume;
+			// direct light ---
+			if( gi->do_direct_light )
+			{
+				// lightsample is used to sample direct light
+				LightSample ls;
+				ls.refP = pWS;
+				f += gi->scene->sample_attenuated_directlight( ls, ti.rng, debug );
+			}
 
-			// lightsample is used to sample direct light
-			LightSample ls;
-			ls.refP = pWS;
+			// indirect light ---
+			if( gi->do_indirect_light )
+			{
+				V3d direction = sampleSphere(ti.rng);
+				double direction_pdf = sampleSpherePDF();
 
-			f = gi->scene->sample_attenuated_directlight( ls, ti.rng, debug )+ // direct light
-				gi->scene->integrator->Li(gi->scene, rq, ti.rng)/direction_pdf; // indirect light
+				Ray3d rayWS( pWS, direction );
+
+				RadianceQuery rq;
+				rq.ray = rayWS;
+				rq.pixel = V2i(-1, -1);
+				rq.debug = debug;
+				if(gi->scene->volume->getBound().contains(pWS))
+					rq.volume = gi->scene->volume;
+				f += gi->scene->integrator->Li(gi->scene, rq, ti.rng)/direction_pdf;
+			}
 		}
 		catch (std::exception& e)
 		{
[…]

### 2017-12-04 python/pnsolver/check.py

@@ -9,6 +9,20 @@ from matplotlib.colors import LogNorm
 from mpl_toolkits.axes_grid1 import make_axes_locatable
 
 
+
+pns = pnsolver.load_solution("C:/projects/epfl/epfl17/python/pnsolver/results/checkerboard2d/checkerboard2d_p1.pns")
+test = pns.getCoefficientField(0)
+print(test.shape)
+
+
+img = np.clip( test, 1.0e-8, np.max(test) )
+
+fig = plt.figure(figsize=(15,7));
+ax = fig.add_subplot(111)
+img_view = ax.imshow(img[:, :, 0].T, cmap='jet', norm=LogNorm(vmin=np.min(img), vmax=np.max(img)), origin='lower')
+plt.show()
+
+
 '''
 import stencil
 
@@ -35,12 +49,13 @@ for i in range(pni.num_coeffs()):
 
 
 
-filename = "C:/projects/epfl/epfl17/python/pnsolver/test.pns"
-pns = pnsolver.load_solution(filename)
-numCoeffs = pns.getNumCoeffs()
+#filename = "C:/projects/epfl/epfl17/python/pnsolver/test.pns"
+#pns = pnsolver.load_solution(filename)
+#numCoeffs = pns.getNumCoeffs()
 
-x = pnsolver.getSolutionVector(pns)
-x = x.reshape((x.shape[0], 1))
+#x = pnsolver.getSolutionVector(pns)
+#x = x.reshape((x.shape[0], 1))
+
 '''
 for coeff in range(numCoeffs):
     image = util.extract_coefficient_field( x, pns.getResolution(), numCoeffs, coeff )
@@ -60,7 +75,7 @@ for coeff in range(numCoeffs):
 
 '''
 
-
+'''
 def plot_spherical_function( fun, vmin=-0.5, vmax=0.5 ):
     # plot sh functions
     theta = np.arange(0.0, 1.0, 0.01)*np.pi
@@ -89,3 +104,4 @@ cax = divider.append_axes("right", size="5%", pad=0.05)
 plt.colorbar(img_view, cax=cax)
 
 plt.show()
+'''

### 2017-12-04 python/pnsolver/example.py

@@ -8,6 +8,10 @@ import pnsolver
 import itertools
 import scipy.io
 
+import matplotlib.pyplot as plt
+from matplotlib.colors import LogNorm
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+
 
 
 
@@ -39,6 +43,14 @@ def solve( stencil_name, problem, filename, do_neumannBC = False ):
 	# NB: the truncation order has been baked into the cpp code as it is linked directly with the stencil
 	sys = pnsolver.PNSystem(stencil_name, problem, do_neumannBC)
 
+	#problem.setExtinctionMinimumThreshold(4.0)
+	#problem.setExtinctionMinimumThreshold(0.0)
+	#sys.build()
+	#A = sys.get_A()
+	#scipy.io.savemat(filename+".t0.mat", {"A":A})
+	#return
+
+
 	# set the RTE parameter fields ---------------------------------------------
 	#setProblem(sys, problem)
 	#sys.setNeumannBoundaryConditions(do_neumannBC)
@@ -63,8 +75,10 @@ def solve( stencil_name, problem, filename, do_neumannBC = False ):
 
 
 	#tol = 1.0e-10
-	#tol = 1.0e-10
-	tol = 1.0
+	#tol = 1.0e-5
+	#tol = 1.7
+	tol = 4.0
+	#tol = 1.0
 
 	#numLevels = 7
 	#x, convergence, timestamps = pnsolver.solve_multigrid( sys, numLevels )
@@ -98,10 +112,21 @@ def solve( stencil_name, problem, filename, do_neumannBC = False ):
 	#data["convergence_cg"] = convergence
 	#data["timestamps_cg"] = timestamps
 
+	#solver = "ls_cg"
+	solver = "solve_ls_lscg"
+
+	if solver == "ls_cg":
+		x, convergence, timestamps = pnsolver.solve_ls_cg( sys, tol)
+	elif solver == "mg":
+		x, convergence, timestamps = pnsolver.solve_mg( sys, tol, 100)
+	elif solver == "solve_ls_lscg":
+		x, convergence, timestamps = pnsolver.solve_ls_lscg( sys, tol)
+
+
 	#x, convergence, timestamps = pnsolver.solve_ls_cg( sys, tol )
 	#x, convergence, timestamps = pnsolver.solve_gs( sys, tol, 10000 )
 	#x, convergence, timestamps = pnsolver.solve_mg( sys, tol, 100 )
[…]

### 2017-12-04 python/pnsolver/problems.py

@@ -12,7 +12,7 @@ from scipy.ndimage.filters import gaussian_filter
 
 # This is the problem from the starmap paper. An emitting square at the center, surrounded by
 # some highly absorbing squares, embedded within a scattering medium.
-def checkerboard():
+def checkerboard2d( res = 71 ):
 
 	def sigma_a( pWS ):
 		x = pWS[0]
@@ -55,14 +55,16 @@ def checkerboard():
 			return 0.0
 		return 0.0
 
+	size = 7.0
+	#res = 128
+
+	'''
 	problem = {}
 
 	# here we set some general parameters 
-	size = 7.0
 	#res = 20
 	#res = 512
 	#res = 64
-	res = 128
 	#res = 71
 	#res = 2
 	#res = 200
@@ -86,6 +88,28 @@ def checkerboard():
 	problem["q"] = [pnsolver.VoxelGridField( util.rasterize(lambda pWS: source_shcoeffs(0, 0, pWS), domain, dtype=complex), domain, offset*0.5 )]
 
 	return problem
+	'''
+	domain = pnsolver.Domain( np.array([size, size, 1]), np.array([res, res, 1]), np.array([0.0, 0.0, 0.0]))
+	t = pnsolver.PNVolume( domain )
+
+	absorption_voxels = util.rasterize3d( domain, lambda pWS: sigma_a(pWS) )
+	scattering_voxels = util.rasterize3d( domain, lambda pWS: sigma_s(pWS) )
+	extinction_voxels = absorption_voxels+scattering_voxels
+
+	extinction_field = pnsolver.VoxelGridField3d(extinction_voxels)
+	albedo_field = pnsolver.VoxelGridField3d(scattering_voxels/extinction_voxels)
+
+	t.setExtinctionAlbedo( extinction_field, albedo_field )
+	t.setPhase( 0, 0, pnsolver.ConstantField3d(np.array([1.0, 1.0, 1.0])) )
+	emission_voxels = util.rasterize3d( domain, lambda pWS: source_shcoeffs(0, 0, pWS) )
+	t.setEmission( 0, 0, pnsolver.VoxelGridField3d(emission_voxels) )
+
+	return t
+
+
+
+
+
 
 
 def homogeneous():
@@ -143,7 +167,7 @@ def homogeneous():
 
 # This is the problem from the starmap paper. An emitting square at the center, surrounded by
[…]

### 2017-12-04 python/pnsolver/render.py

@@ -67,14 +67,7 @@ def load_camera( filename, id = "cam1" ):
 	return camera
 
 
-def create_scene_pointlight( power, sigma_t, albedo ):
-	volume = renderer.create_volume()
-	volume.setBound( np.array([0.0, 0.0, 0.0]), np.array([2.0, 2.0, 2.0]) )
-	sigma_t_field = renderer.create_constant_field3d( np.array([sigma_t, sigma_t, sigma_t]) )
-	albedo_field = renderer.create_constant_field3d( np.array([albedo, albedo, albedo]) )
-	volume.setExtinctionAlbedo( sigma_t_field, albedo_field )
-	light = renderer.create_point_light( np.array([1.0, 1.0, 1.0]), np.array([power, power, power]) )
-	return volume, light
+
 
 def create_scene_nebulae():
 	albedo = 0.9
@@ -85,23 +78,11 @@ def create_scene_nebulae():
 	volume = renderer.create_volume()
 	volume.setBound( np.array([-0.5, -0.5, -0.5+offset]), np.array([0.5, 0.5, 0.5+offset]) )
 
-	test = np.zeros((64, 64, 64, 3))
-	#for i in range(test.shape[0]):
-	#	for j in range(test.shape[1]):
-	#		for k in range(test.shape[2]):
-	#			t = float(j)/float(test.shape[1])
-	#			test[i, j, k, 0] = i
-	#			test[i, j, k, 1] = j
-	#			test[i, j, k, 2] = k
-	#test[3, 3, 3, :] = 1
-	#test[2, 3, 3, :] = 1
-	test[46, 57, 59, :] = 123.4
 
-	#sigma_t_field = renderer.VoxelGridField3d(test)
 	sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64.bgeo")
 	#sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae200.bgeo")
 	#sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/test.bgeo")
-	sigma_t_field.save("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64_extinction.grid")
+	#sigma_t_field.save("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64_extinction.grid")
 
[…]

### 2017-12-04 python/pnsolver/stencil.py

@@ -1950,7 +1950,8 @@ if __name__ == "__main__":
 	#order = [0,1,2,3,4]
 
 	rte_forms = ["fopn"]
-	order = [1]
+	#order = [1]
+	order = [3,5]
 	#order = [1,2,3,4,5]
 	staggered = [True]
 

### 2017-12-04 python/pnsolver/test_pointsource.py

@@ -1,0 +1,121 @@
+import numpy as np
+import renderer
+import util
+import json
+
+
+# just for plotting
+import matplotlib.pyplot as plt
+from matplotlib.colors import LogNorm
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+
+
+
+def create_scene_pointlight( power, sigma_t, albedo ):
+	volume = renderer.create_volume()
+	volume.setBound( np.array([0.0, 0.0, 0.0]), np.array([2.0, 2.0, 2.0]) )
+	sigma_t_field = renderer.create_constant_field3d( np.array([sigma_t, sigma_t, sigma_t]) )
+	albedo_field = renderer.create_constant_field3d( np.array([albedo, albedo, albedo]) )
+	volume.setExtinctionAlbedo( sigma_t_field, albedo_field )
+	light = renderer.create_point_light( np.array([1.0, 1.0, 1.0]), np.array([power, power, power]) )
+	return volume, light
+
+
+def validate_pointsource_fluence():
+	numSamples = 10000
+	# assemble scene to render
+	#camera = load_camera( "results/pointsource/result_pointsource.scn", "cam1" )
+	integrator = renderer.create_simplept_integrator(True, -1)
+	power = 4.0*np.pi
+	sigma_t = 8.0
+	albedo = 0.9
+	volume, light = create_scene_pointlight(power, sigma_t, albedo)
+
+	# get pnsolution
+	#pns_p5 = renderer.load_pnsolution( "results/pointsource/pointsource_p5.pns" )
+	#pns_p3 = renderer.load_pnsolution( "results/pointsource/pointsource_p3.pns" )
+	pns_p1 = renderer.load_pnsolution( "results/pointsource/pointsource_p1_2.pns" )
+	pns_p1_ms = renderer.load_pnsolution( "results/pointsource/pointsource_p1_2_ms.pns" )
+	#pns_p1_vacuum = renderer.load_pnsolution( "results/pointsource/pointsource_p1_3.pns" )
+	#pns_p1_ms64 = renderer.load_pnsolution( "results/pointsource/pointsource_p1_2_ms64.pns" )
+
+	# render
+	#renderer.render( volume, light, camera, integrator ).save("test.exr")
+
+	# fluence
+	size = 2.0
+	r_list = np.linspace(1.0e-2, size*0.5, 100)
[…]

### 2017-12-06 cpp/renderer/include/Image.h
worked on path guiding for nebulae dataset

@@ -41,12 +41,14 @@ struct Image// : public Eigen::Array<Eigen::Vector3d, Eigen::Dynamic, Eigen::Dyn
 	P2d rasterToUV(const P2d& pRaster)const;
 
 	// manipulation
+	Ptr toLDR();
 
 private:
 	Base m_data;
 };
 
 Image::Ptr blur_image( Image::Ptr image, double stddev );
+Image::Ptr ldr_image( Image::Ptr image, double exposure );
 
 struct ImageSampler
 {

### 2017-12-06 cpp/renderer/include/PNSolution.h

@@ -55,9 +55,9 @@ struct PNSolution
 		double *phiIntegrals(double a, double b);
 
 		// integrates SH function over given block by using precomputed integration tables
-		double integrate(int depth, int zBlock, int phiBlock, const double *coeffs) const;
+		double integrate(int depth, int zBlock, int phiBlock, const double *coeffs, double constant_scale=0.0) const;
 		// integrates SH function over given block by summing the integrals of all its child blocks
-		double integrateChilds(int depth, int i, int j, const double *coeffs) const;
+		double integrateChilds(int depth, int i, int j, const double *coeffs, double constant_scale=0.0) const;
 
 		int indexofSmallestElement(double array[], int size);
 

### 2017-12-06 cpp/renderer/include/integrators/pnispt.h

@@ -1,4 +1,5 @@
 #pragma once
+#include <map>
 #include <integrators/simplept.h>
 
 #include <scene.h>
@@ -7,6 +8,13 @@
 #include <volume.h>
 
 #include <PNSolution.h>
+
+
+
+
+
+
+
 
 
 
@@ -24,8 +32,8 @@ struct TraceInfoPNIS
 	const PNSolution* pns;
 
 	TraceInfoPNIS():
-	    throughput_over_pdf(1.0, 1.0, 1.0),
-	    debug(false)
+		throughput_over_pdf(1.0, 1.0, 1.0),
+		debug(false)
 	{
 	}
 
@@ -82,11 +90,70 @@ struct TraceInfoPNIS
 		if( current_vertex.getType() == Vertex::EVolume )
 		{
 			V3d wi = current_direction;
-			double direction_pdf;
-			current_direction = -pns->sample( current_vertex.getPosition(), direction_pdf, P2d(rng.next1D(), rng.next1D()) );
-			//double phase_over_pdf = scene->volume->samplePhase(current_vertex.getPosition(), wi, current_direction, pdf, rng);
-			double phase_over_pdf = scene->volume->evalPhase(current_vertex.getPosition(), wi, current_direction)/direction_pdf;
+
+			double phase_over_pdf;
+
+
+			// we use one-sample-estimator to decide between phase function and PNS-sampling
+			double prob = 1.0;
+			if( rng.next1D() < prob )
+			{
+				// sample according to PN-Solution
+				double pdf_shsampling;
+				current_direction = -pns->sample( current_vertex.getPosition(), pdf_shsampling, P2d(rng.next1D(), rng.next1D()) );
+
+				if(pdf_shsampling == 0.0)
+					return false;
+
+				double pdf_phase = INV_FOURPI;
+
+				double pdf = prob*pdf_shsampling + (1.0-prob)*pdf_phase;
+
+				phase_over_pdf = scene->volume->evalPhase(current_vertex.getPosition(), wi, current_direction)/pdf;
+
+				if( debug )
+				{
+					dbg()["phase_sampling"] = 0;
+					dbg()["phase_pdf"] = pdf;
+					dbg()["pdf_shsampling"] = pdf_shsampling;
+				}
+			}else
+			{
+				// sample according to phase function
+				double pdf_phase;
[…]

### 2017-12-06 cpp/renderer/include/util/threadpool.h

@@ -234,7 +234,7 @@ void runGenericTasks( typename GenericTask<TaskInfo, GlobalInfo>::RunMethod run,
 
 	for( term.reset();term.keepRunning();term.advance() )
 	{
-		//std::cout << term.m_sample <<" " << term.m_numSamples <<  std::endl;
+		std::cout << term.m_sample <<" " << term.m_numSamples <<  std::endl;
 		if(numThreads>1)
 			ThreadPool::instance()->enqueueAndWait(tasks);
 		else

### 2017-12-06 cpp/renderer/src/Image.cpp

@@ -282,6 +282,37 @@ Image::Ptr blur_image( Image::Ptr image, double stddev )
 	return image_blurred;
 }
 
+Image::Ptr ldr_image( Image::Ptr image, double exposure )
+{
+	Image::Ptr result = std::make_shared<Image>( image->getResolution() );
+
+	double scale = std::pow(2.0, exposure);
+
+	for( int i=0;i<image->getArray().rows();++i )
+	{
+		for( int j=0;j<image->getArray().cols();++j )
+		{
+			Eigen::Vector3d& input_pix = image->getArray().coeffRef(i, j);
+			Eigen::Vector3d output_pix;
+
+			for (int c=0; c<3; ++c)
+			{
+				float value = input_pix[c]*scale;
+
+				if (value <= 0.0031308f)
+					output_pix[c] = 12.92f * value;
+				else
+					output_pix[c] = (1.0f + 0.055f) * std::pow(value, 1.0f/2.4f) -  0.055f;
+			}
+
+			result->getArray().coeffRef(i, j) = output_pix;
+		}
+	}
+
+	return result;
+}
+
+
 
 // ImageSampler -------------------------------
 
@@ -351,3 +382,5 @@ void ImageSampler::buildPDF( Image::Ptr image )
 	}
 	m_colDpdf.normalize();
 }
+
+

### 2017-12-06 cpp/renderer/src/PNSolution.cpp

@@ -470,7 +470,7 @@ PNSolution::SHSampler::SHSampler(int bands, int depth) : m_bands(bands), m_depth
 }
 
 
-double PNSolution::SHSampler::integrate(int depth, int zBlock, int phiBlock, const double* coeffs) const
+double PNSolution::SHSampler::integrate(int depth, int zBlock, int phiBlock, const double* coeffs, double constant_scale) const
 {
 	double result = 0;
 	for (int l=0; l<m_bands; ++l)
@@ -481,6 +481,24 @@ double PNSolution::SHSampler::integrate(int depth, int zBlock, int phiBlock, con
 			result += basisIntegral * sph::get(l, m, coeffs);
 		}
 	}
+
+	// clamp negative integrals to zero
+	result = std::max(result, 0.0);
+
+	// add constant factor
+	int res = (1 << depth);
+	double zStep = -2 / (double) res;
+	double phiStep = 2 * (double) M_PI / (double) res;
+
+	double phi_integral = phiStep;
+
+	double theta_max = 1.0+zBlock*zStep;
+	double theta_min = 1.0+(zBlock+1)*zStep;
+	double theta_integral = -theta_min + theta_max;
+
+	//double scale = 0.001;
+	result += theta_integral*phi_integral*constant_scale;
+
 	return result;
 }
 
@@ -499,17 +517,17 @@ int PNSolution::SHSampler::indexofSmallestElement(double array[], int size)
 	return index;
 }
 
-double PNSolution::SHSampler::integrateChilds(int depth, int i, int j, const double *coeffs) const
+double PNSolution::SHSampler::integrateChilds(int depth, int i, int j, const double *coeffs, double constant_scale) const
 {
 	if(depth<m_depth)
 	{
-		double q00_1 = std::max(integrate(depth+1, i*2, j*2, coeffs), 0.0);
-		double q10_1 = std::max(integrate(depth+1, i*2, j*2+1, coeffs), 0.0);
-		double q01_1 = std::max(integrate(depth+1, i*2+1, j*2, coeffs), 0.0);
-		double q11_1 = std::max(integrate(depth+1, i*2+1, j*2+1, coeffs), 0.0);
+		double q00_1 = integrate(depth+1, i*2, j*2, coeffs, constant_scale);
[…]

### 2017-12-06 cpp/renderer/src/integrators/pnispt.cpp

@@ -3,18 +3,52 @@
 
 
 
-
-
-
-
 V3d PNISPT::Li( const Scene* scene, RadianceQuery& rq, RNGd& rng )const
 {
-/*
-	return V3d( double(rq.pixel[0])/double(scene->camera->getResolutionX()),
-				double(rq.pixel[1])/double(scene->camera->getResolutionY()),
-				0.0);
-*/
+	//rq.transmittance = V3d(1.0, 1.0, 1.0);
 
+	if( rq.debug )
+	{
+		std::cout << "VolumePathTracer::Li\n";
+		std::cout << "VolumePathTracer::Li pix=" << rq.pixel.toString() << std::endl;
+		std::cout << "VolumePathTracer::Li ray=" << rq.ray.toString() << std::endl;
+	}
+
+
+	double mint, maxt;
+	if( scene->volume->intersectBound(rq.ray, mint, maxt, rq.debug) )
+	{
+		// start tracing
+		//TraceInfo ti;
+		TraceInfoPNIS ti;
+		ti.depth = 0;
+		ti.current_vertex = Vertex();
+		P3d pWS = rq.ray(mint+Epsilon);
+		if(rq.volume)
+			ti.current_vertex.setPosition(pWS, rq.volume->evalExtinction(pWS, rq.debug), rq.volume->evalAlbedo(pWS));
+		else
+			ti.current_vertex.setPosition(pWS, V3d(0.0, 0.0, 0.0), V3d(0.0, 0.0, 0.0));
+		ti.current_direction = rq.ray.d;
+		ti.scene = scene;
+		ti.debug = rq.debug;
+		ti.debug = rq.debug;
+		ti.debug_info = &debug_info;
+		ti.pns = m_pns.get();
+		return trace( ti, rng );
+	}else
+	{
+		// no intersection with the medium boundary
+	}
+
+	return V3d(0.0, 0.0, 0.0);
+}
+
+
+
+
+/*
+V3d PNISPT::Li( const Scene* scene, RadianceQuery& rq, RNGd& rng )const
+{
 
 	//rq.transmittance = V3d(1.0, 1.0, 1.0);
 
@@ -30,7 +64,8 @@ V3d PNISPT::Li( const Scene* scene, RadianceQuery& rq, RNGd& rng )const
 	if( scene->volume->intersectBound(rq.ray, mint, maxt, rq.debug) )
 	{
[…]

### 2017-12-06 cpp/renderer/src/integrators/simplept.cpp

@@ -6,13 +6,6 @@
 
 V3d SimplePT::Li( const Scene* scene, RadianceQuery& rq, RNGd& rng )const
 {
-/*
-	return V3d( double(rq.pixel[0])/double(scene->camera->getResolutionX()),
-				double(rq.pixel[1])/double(scene->camera->getResolutionY()),
-				0.0);
-*/
-
-
 	//rq.transmittance = V3d(1.0, 1.0, 1.0);
 
 	if( rq.debug )

### 2017-12-06 cpp/renderer/src/main.cpp

@@ -46,6 +46,7 @@ struct GlobalRenderInfo
 
 	// used for debugging
 	P2i debug_pixel;
+	std::vector<RNGd> *debug_rngs;
 };
 
 struct GlobalComputeFluenceInfo
@@ -220,6 +221,7 @@ Eigen::VectorXd compute_fluence( Volume::Ptr volume,
 
 void render_thread( MonteCarloTaskInfo& ti, const GlobalRenderInfo* gi )
 {
+	//RNGd& rng = ti.rng;
 	int width = gi->scene->camera->getResolutionX();
 	int height = gi->scene->camera->getResolutionY();
 
@@ -232,6 +234,7 @@ void render_thread( MonteCarloTaskInfo& ti, const GlobalRenderInfo* gi )
 		for (int x = gi->crop_window.min.x(); x < gi->crop_window.max.x(); ++x)
 		{
 			int index = y*width + x;
+			RNGd& rng = (*gi->debug_rngs)[ index ];
 			V3d f(0.0, 0.0, 0.0);
 			//Color3f T(1.0f);
 
@@ -248,7 +251,7 @@ void render_thread( MonteCarloTaskInfo& ti, const GlobalRenderInfo* gi )
 
 			Ray3d rayWS;
 			gi->scene->camera->sampleRay( P2d(x+0.5, y+0.5), rayWS, debug );
-			//ti.g.scene->camera->sampleRay( P2d(x+ti.rng.next1D(), y+ti.rng.next1D()), rayWS );
+			//ti.g.scene->camera->sampleRay( P2d(x+rng.next1D(), y+rng.next1D()), rayWS );
 
 			if(debug)
 			{
@@ -263,8 +266,11 @@ void render_thread( MonteCarloTaskInfo& ti, const GlobalRenderInfo* gi )
 				rq.ray = rayWS;
 				rq.pixel = V2i(x, y);
 				rq.debug = debug;
-				f = gi->scene->integrator->Li(gi->scene, rq, ti.rng);
+				f = gi->scene->integrator->Li(gi->scene, rq, rng);
 				//T = rq.transmittance;
+
+				if( debug )
+					std::cout << "f=" << f << std::endl;
 			}
 			catch (std::exception& e)
 			{
@@ -284,7 +290,7 @@ void render_thread( MonteCarloTaskInfo& ti, const GlobalRenderInfo* gi )
 
 			if(debug)
 			{
[…]

### 2017-12-06 python/pnsolver/check.py

@@ -1,6 +1,7 @@
 import os,sys,inspect
 import numpy as np
-import pnsolver
+#import pnsolver
+import renderer
 import util
 import stencil
 
@@ -10,6 +11,23 @@ from mpl_toolkits.axes_grid1 import make_axes_locatable
 
 
 
+pns = renderer.load_pnsolution( "c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_p5_2_ms.pns" )
+
+pWS = np.array([-0.27704304, 0.36083166, -0.22953043])
+#sample = np.array([0.3960667149031164, 0.18571028805648737])
+#sample = np.array([0.8444218515250481, 0.7579544029403025])
+sample = np.array([0.37858849022321517, 0.49207364263387265])
+
+dir, pdf = pns.sample(pWS, sample)
+print("")
+pdf2 = pns.pdf( pWS, dir )
+
+print(  "pdf={:.02} pdf2={:.02}".format(pdf, pdf2))
+
+
+
+
+'''
 pns = pnsolver.load_solution("C:/projects/epfl/epfl17/python/pnsolver/results/checkerboard2d/checkerboard2d_p1.pns")
 test = pns.getCoefficientField(0)
 print(test.shape)
@@ -21,6 +39,7 @@ fig = plt.figure(figsize=(15,7));
 ax = fig.add_subplot(111)
 img_view = ax.imshow(img[:, :, 0].T, cmap='jet', norm=LogNorm(vmin=np.min(img), vmax=np.max(img)), origin='lower')
 plt.show()
+'''
 
 
 '''

### 2017-12-06 python/pnsolver/render.py

@@ -211,15 +211,15 @@ if __name__ == "__main__":
 	'''
 
 	# render pn-solution directly (instead of using it to boost standard MC) -----------------------------
-	#'''
+	'''
 	volume, light = create_scene_nebulae()
 	camera = load_camera("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae.scn")
 	#camera = renderer.create_perspective_camera(512, 512, 45.0)
 	#translate = translation_matrix( np.array([0.0, 0.0, 1.9]) )
 	#camera.setCameraToWorldTransform(translate)
 
-	#numSamples = 100
-	numSamples = 1
+	numSamples = 100
+	#numSamples = 1
 	
 	#integrator_groundtruth_ss = renderer.create_simplept_integrator(True, 1)
 	#img_ss = renderer.render( volume, light, camera, integrator_groundtruth_ss, numSamples )
@@ -229,8 +229,9 @@ if __name__ == "__main__":
 	#img_ms.save("nebulae_groundtruth_ms.exr")
 
 
-	#pns = renderer.load_pnsolution( "c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_p1_2_ms.pns" )	
-	pns = renderer.load_pnsolution( "c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_p3_2_ms.pns" )	
+	pns = renderer.load_pnsolution( "c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_p1_2_ms.pns" )	
+	#pns = renderer.load_pnsolution( "c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_p3_2_ms.pns" )	
+	#pns = renderer.load_pnsolution( "c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_p5_2_ms.pns" )
 	#pns = renderer.load_pnsolution( "c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_p5.pns" )	
 	#pns = renderer.load_pnsolution( "c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae_cda3.pns" )	
 
@@ -240,19 +241,85 @@ if __name__ == "__main__":
 
[…]

### 2018-01-05 latex/egsr17-cgf/introduction.tex
started working on paper text

@@
 \begin{itemize}
-	\item System for generation of stencils for coupled PDE's directly from mathematical expressions.
-	\item Staggered-grid solver for time-independent PN-equations.
-	\item Novel PN-guided Monte-Carlo estimator for participating media.
+	\item We derive the time-independent real-valued PN-equations.
+	\item We present a novel system for generating finite difference stencils for coupled PDE's, directly from a computer algebra representation of mathematical expressions.
+	\item We introduce a staggered-grid solver for time-independent PN-equations and use it for rendering participating media.
+	\item Finally, we compare higher order PN against Flux-Limited Diffusion.
 \end{itemize}
+
+In the next section we will discuss related work and why it motivates our contribution.
+Section~\ref{sec:pnequations} introduces the real-valued $P_N$-equations, which we solve using our solver described in section~\ref{sec:pnsolver}.
+We briefly recapitulate variable eddington factor methods and flux-limited diffusion in particular in section~\ref{sec:vef}.
+Our solvers for $P_N$ and FLD are run on a set of standard problems in section~\ref{sec:results}.
+We finish with our conclusion in section~\ref{sec:conclusion}.

### 2018-01-05 latex/egsr17-cgf/main-body.tex
started working on paper text

--- 2017-11-27 latex/egsr17-cgf/main-body.tex
@@
 
-%\title[EG \LaTeX\ Author Guidelines]%
-%{\LaTeX\ Author Guidelines for EUROGRAPHICS Proceedings Manuscripts}
+\title[Draft]%
+{Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media}
 
@@
 
-%\maketitle
+\maketitle
 
@@
 
-%\begin{abstract}
-%TODO
+\begin{abstract}
+
+Classical diffusion approximation (CDA) is very competitive for rendering highly scattering participating media, because in these scenarios, Monte Carlo based methods suffer from high computational cost.
+Variable Eddington Factor methods, such as flux-limited diffusion (FLD), have shown to further improve the results from CDA.
+In this paper, we look at CDA as a degenerate case of the $P_1$-method and investigate the open question, if going higher order gives a viable alternative to FLD.
+We derive the time-independent real-valued $P_N$-equations and introduce a framework for generating finite difference stencils directly from the computer algebra representation of the $P_N$-equations for arbitrary order.
+We further present a staggered-grid $P_N$-solver and compare it against FLD for various standard problems.
 
@@
 
-%\end{abstract}
+\end{abstract}
 
@@
 
-\subsection*{Deterministic Methods}
+Light transport in participating media is governed by the radiative transfer equation(RTE), which was first introduced in astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}.
+Today, this equation is typically solved using Monte Carlo based methods.
+However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become unfeasible, due to high computational cost which comes from tracing long particle paths with many scattering events.
 
-\begin{itemize}
-  \item Lattice Boltzmann Methods \cite{Geist04}
-  \item Discrete Ordinates
-  \item Diffusion \cite{Stam95} \cite{Jakob10} \cite{Koerner14}
-  \item $PN$-Method \cite{Kajiya84} 
-  \item Heuristics \cite{Kaplanyan10} \cite{Elek14}
-\end{itemize}
+In contrast to path-tracing, the $P_N$-method is a deterministic method, which gives a solution by solving a system of linear equations.
+It is derived by discretizing the angular variable of the radiative transfer equation into spherical harmonics (SH).
+This gives rise to a set of coupled, complex-valued partial differential equations, called $P_N$-equations.
+The subscript $N$ refers to the truncation order of the SH.
 
-\subsection*{Path-guiding}
+The $P_N$-method has a long history in other fields and was never applied in graphics.
+Kajia~\cite{Kajiya84} explained the theory, but did not give any details on implementation or how to solve it.
+In fact, as Max~\cite{Max95} pointed out, it is not clear if Kajiya succeeded at all at applying the method, as all of the results in the paper were produced with a simpler method.
+This is further strengthened by our observation, that a straight forward finite difference discretization of the $P_N$-equations produces unuseable results, due to oscillation artefacts in the solution.
+In our paper, we derive the real-valued $P_N$-equations and present a staggered-grid solver, which produces an artefact free solution.
 
-\begin{itemize}
-  \item Russian roulette and splitting \cite{Vorba16}
-  \item Learning using SD-Trees \cite{Pegoraro08} \cite{Mueller17} or Gaussian mixture models \cite{Vorba14}
-  \item Zero-Variance sampling \cite{Krivanek14}
-  \item Reinforcement learning \cite{Dahm17}
-\end{itemize}
+Similar to $P_N$, the classical diffusion approximation (CDA) is a deterministic method, which arrives at a solution by solving a system of linear equations.
+It is derived as a degenerate case of the $P_1$-equations (truncation after the first SH order), which collapses to a simple diffusion equation giving the method its name.
+CDA has a long history in other domains, such as astrophysics~\cite{Ishimaru78} and was introduced to graphics by Stam \cite{Stam95}.
+It was further extended to anisotropic radiative transfer by \cite{Jakob10}.
+
+CDA suffers from severe energy loss close to regions with strong density gradients.
+This problem is addressed by the variable eddington factor(VEF)-formalism, among which flux-limited diffusion (FLD) is the most prominent.
+FLD was first introduced in astrophysics by~\cite{Pomraining78} and was introduced to graphics by Koerner et al.~\cite{Koerner14}.
+Unfortunately, both CDA and FLD do not support anisotropic phase functions.
+
+The idea behind VEF is to replace the diffusion coefficient in the diffusion equation of CDA by a non-linear diffusion coefficient.
+This diffusion coefficient is being derived by seeing it as a mean to interpolate between the diffusive and pure transport regime.
+The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
+In the absence of scattering (e.g.
+pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.
+
+It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion.
+This question is also been raised in other domains~\cite{Olson00} and resolving it is the main motivation for this paper.
+
+%\begin{itemize}
+  %\item Alternative deterministic methods
+  %\begin{itemize}
+  %  \item Heuristics \cite{Kaplanyan10} \cite{Elek14}
+  %  \item Lattice Boltzmann Methods \cite{Geist04}
+  %  \item Discrete Ordinates \cite{Languenou95}
+  %\end{itemize}
+  %However, among all the deterministic methods, the diffusion approximation has been the most popular due to its intuition and simplicity.
+
+%\end{itemize}
 
 %-------------------------------------------------------------------------
-\section{Radiative Transfer and $PN$-Equations}
+\section{Real-valued $P_N$-Equations}
+\label{sec:pnequations}
 
-In order to compute a rendered image, the radiance field $L$ is evaluated. Its evolution throught space is given by the radiative transfer equation (RTE):
+The $P_N$-equations are derived from the radiative transfer equation~(RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
+\begin{align}
+%\label{eq:rte}
+\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
+=&
+-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\\
+&
++\sigma_s\left(\vec{x}\right) \int_{\Omega}
+{
+p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega \right)\ud\omega'
+}\\
+&
++Q\left(\vec{x}, \omega\right)
+\end{align}
 
--give RTE
-%\begin{}
-%\end{}
+The LHS is the transport term and we refer to the terms on the RHS as collision, scattering and source term respectively.
 
-The radiance field $L$ is typically solved with probabilistic Monte-Carlo methods, such as path-tracing.
-With these methods, computational demand is directly linked to the average length of paths and the presence of light transport paths which are hard to find with random sampling.
-The path length increases for dense, highly scattering media and hard-to-sample paths exist with highly anisotropic phase functions or complex visbility in the scene.
+The $P_N$-method uses SH to discretize the angular domain into a set of coupled equations.
+We further discretize the spatial variable using a finite difference (FD) grid.
+Each term represents a change of the radiance field $L$.
+Therefore, after discretizing the continuous variables, we can represent the RTE as a sequence of linear transformations, which can be collapsed into a single coefficient matrix $A$ and a RHS vector $\vec{Q}$:
+\begin{align}
+(T+C-S)\vec{u} = A\vec{u} = \vec{Q}
+\end{align}
 
-Deterministic methods are derived by discretizing the radiance field $L$ and the RTE in spatial- and spherical domain.
-The result is a system of linear equations of the form $A\vec{x}=b$ where $\vec{x}$ represents the radiance field in the discretized problem.
-The matrix $A$ and right hand side $b$ are found by assembling a global system from all discretization elements.
-Deterministic methods suffer from discretization artefacts and only give an accurate solution in the discretization limit.
-However, they are fast to compute and noise free.
+$T$, $C$, $S$ are matrices, which result from the discretized transport, collision and scattering terms in the $P_N$-equations respectively.
+All SH coefficients for every voxel of our FD-grid are flattened into the solution vector $\vec{u}$, which is the unknown we want to solve for.
+The RHS vector likewise contains the flattened SH coefficients of the emission field $Q$ of every voxel.
 
-The $PN$-equations are found by projecting the radiance field and the RTE into a truncated spherical harmonics (SH).
-A regular grid is used for spatial discretization.
-The SH-expansion of the radiance field $L$ is:
+In order to derive the real-valued $P_N$-equations from which we can construct $A$ and $\vec{Q}$, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
+\begin{align}
+\SHBR^{l,m}=
+\left\{
+\begin{array}{lr}
+\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
+\SHBC^{l,m}, & \text{for } m = 0\\
+\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}-\left(-1\right)^m\SHBC^{l,m}\right), & \text{for } m > 0
+\end{array}
+\right.
+\end{align}
 
-\begin{equation}
-L\left(x, \omega \right) = \sum_{l=0}^{N}
+First, angular dependent quantities in the RTE, such as the radiance field $L$ (or the emission field $Q$) are projected into SH coefficients:
+\begin{align}
+L^{l,m}\left(\vec{x}\right)
+=
+\int_\Omega{L\left(\vec{x}, \omega\right)\SHBR^{l,m}\ud\omega}
+\end{align}
+
+The reconstruction $\hat{L}$ will have an error due to truncation at order $N$:
+\begin{align}
+L\left(\vec{x}, \omega\right)
+\approx
+\hat{L}\left(\vec{x}, \omega\right) =
+\sum_{l=0}^{N}
+{
 \sum_{m=-l}^{l}
-L\left(x\right)^{lm}
-Y\left(\omega\right)^{l,m}
-\end{equation}
+{
+L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
+}
+}
+=
+\sum_{l,m}
+{
+L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
+}
+\end{align}
 
-where $N$ is the truncation order, $L^{lm}$ is a set SH coefficients and $Y^{l,m}$ is the SH-basis function.
+The angular dependent quantities are replaced with their SH reconstruction in the RTE.
+The $P_N$-equations are found by multiplying each term with the conjugate complex of the real-valued SH basis function $\overline{\SHBR^{l,m}}$, followed by integration over solid angle.
+In order to arrive at a form, which does not depend on angle anymore, a set of transformations and identities are applied.
+Of particular importance are the orthogonality property
+\begin{align}
+\int_{\Omega}{ \SHBC^{l,m}\overline{\SHBC^{l',m'}}\ud\omega} = \delta_{l,l'}\delta_{m,m'}
+\end{align}
 
-\section{System Overview}
-\section{Stencil Generation}
-\section{Staggered Grid Solver}
-\section{Path-guiding}
+and the following recursion relation
+\begin{align}
+\omega_x\overline{\SHBC^{l, m}}=
+&
+-\frac{1}{2}c^{{l-1,m-1}}\overline{Y^{l-1, m-1}}+\frac{1}{2}d^{{l+1,m-1}}\overline{Y^{l+1, m-1}}\\
+&
++\frac{1}{2}e^{{l-1,m+1}}\overline{Y^{l-1, m+1}}-\frac{1}{2}f^{{l+1,m+1}}\overline{Y^{l+1, m+1}}\\
+\omega_y\SHBC^{l, m}=
+&
+\frac{i}{2}c^{{l-1,m-1}}\overline{Y^{l-1, m-1}}-\frac{i}{2}d^{{l+1,m-1}}\overline{Y^{l+1, m-1}}\\
+&
++\frac{i}{2}e^{{l-1,m+1}}\overline{Y^{l-1, m+1}}-\frac{i}{2}f^{{l+1,m+1}}\overline{Y^{l+1, m+1}}\\
+\omega_y\SHBC^{l, m}=
+&
+a^{{l-1,m}}\overline{Y^{l-1, m}}+b^{{l+1,m}}\overline{Y^{l+1, m}}
+\end{align}
+
+with 
+\begin{align}
+&a^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}},
+b^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}\\
+&c^{l,m}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}},
+d^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}\\
+&e^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}},
+f^{l,m}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
+\end{align}
+
+%Finally, the spatially dependent variables are discretized (using finite difference stencils for derivatives).
+The resulting equations are factorized according to the unknowns ($l,m$-coefficient within voxel $i,j,k$).
+The coefficients to these unknowns
+
+Note that the real-valued SH basis functions are different, depending on $m$, which means that we get three different versions of the $P_N$-equations (for $m<0$, $m=0$ and $m>0$).
+Further note, that we need to expand the real-valued SH basis functions in order to be able to apply the given identities above.
+This causes the equations to become very large and difficult to deal with when applying transformations and identities.
+
+It becomes even more of a problem later, when we want to expand the spatially discretized equations for a particular choice of truncation order $N$.
+Manually creating a FD stencil from hundreds of terms, including complex numbers and derivatives, is very challenging for first order and becomes impossible for higher order.
+
+\subsection*{Computer Algebra Representation}
+
+In order to tackle this problem, we created a system for representing algebraic expressions as an expression tree in Python.
+Our framework can represent terms, integrals, derivatives, sums, variables, functions and special symbols, such as the Kronecker delta or imaginary number.
+It supports manipulations, such as the application of distributive law, substitution, rearranging nested integrals and application of the identities given above.
+
+After entering each term of the RTE into our framework, we arrive at the $P_N$-equations by applying a set of manipulations and identities using API commands provided by our framework.
+For debugging purposes, the expression tree can be rendered directly into latex code.
+This allows us to deal with pages of dense equations very easily.
+
+<show some codesnippet for generating expressions and applying manipulations?>
+
+The computer algebra representation of the $P_N$-equations is input to our $P_N$-solver, which we will discuss in the next section.
+
+\section{$P_N$-Solver}
+\label{sec:pnsolver}
+
+In this section we discuss the implementation of our $P_N$-solver.
+The inputs are the $P_N$-equations in form of our computer algebra representation from section~\ref{sec:pnequations} and the scalar fields for $\sigma_t$ and $\sigma_s$.
+The angular dependent emission field $Q\left(\vec{x}, \omega\right)$ is projected into SH, which gives a scalar field for each SH coefficient as another input.
+We assume the phase function is not spatially varying and we therefore only require a single set of coefficients $p^{l,m}$ as input.
+
+Solving for the unknown vector $\vec{u}$, which gives the SH coefficients of the radiance function at every voxel of our FD grid, requires building the coefficient matrix $A$ and RHS $\vec{Q}$.
+We seperate this build process into two steps.
+The first step does not need any RTE parameters and is also independent of spatial grid resolution.
+This is the stencil generation and it can be done once for every truncation order $N$ in advance.
+
+The second step uses the precomputed stencil to compute the coefficients of $A$ and $\vec{Q}$ from the RTE-parameter fields and a given resolution for the FD-grid.
+Then $\vec{u}$ is found by solving $A\vec{u}=\vec{Q}$, using a standard linear system solver.
+
+\subsection{Stencil Generation}
+
+The stencil generation builds directly on top of our computer algebra framework for representing mathematical expressions.
+After specifying a truncation order $N$, the sums in the equations are expanded into individual terms.
+These are then further factorized according to the unknowns.
+
+Finally, the equations are discretized in spatial domain.
+Since all manipulations are done by recursively parsing the expression tree and replacing subtrees as needed, discretization can be handled elegantly by our system.
+The $P_N$-equations are discretized in stencil space (not in world space).
+We start parsing the expression at the origin of the stencil (meaning $\vec{x}=(0,0,0)$).
+
+Whenever we encounter a derivative operator, we parse its contained sub-expression multiple times with shifted voxel offsets according to the finite difference direction.
+The resulting expressions are multiplied with an expression representing the finite difference weight (which will contain the voxelsize as a variable).
+The parser keeps track of the current voxel offset in a stack.
+Nested derivatives are handled naturally and produce second order finite difference stencils as expected.
+All occurances of the continuous positional variable $\vec{x}$ are replaced by the current voxel offset in the stack.
+
+As we will see later in section~\ref{sec:results}, our solver needs to support staggered grids in order to avoid severe oscillation artefacts which arise, when all coefficients are located at the voxel center.
+With staggered grids, coefficients can live on the intersections of a subgrid with double resolution.
+So instead of voxel centers alone, coefficients can live on the center of voxel faces, voxel edges or the intersection between voxels.
+
+Our solver supports placement of coefficients on arbitrary staggered grid locations (including collocation).
+Whenever an unknown is encountered during expression tree parsing, the discretization code looks at the staggerd grid location at which the unknown is to be evaluated and the location at which the unknown lives on the grid.
+Depending on how the two are located to each other, the parser returns an expression, which interpolates the unknown at the requested offset from its surrounding locations.
+It returns the unknown itself, if both happen to coincide at the same location.
+
+After discretization, the final expression tree is used to generate C++ code for the stencil.
+The unknowns in the terms give information about the coefficient index and voxel offset and therefore identify a column offset in the matrix $A$.
+The other factors in each term are rendered into C++ expressions, which are executed during runtime to set the coefficients in matrix $A$ and $\vec{Q}$.
+The stencil is executed per voxel and contains the code for computing the cofficients for all rows in $A$ and $\vec{Q}$ which belong to a single voxel.
+
+\subsection{System Building and Solving}
+
+The matrix $A$ and RHS $Q$ are built and solved using a C++ framework, which is being compiled with the generated stencil code.
+The build process consists of iterating over all voxels and executing the stencil.
+The generated stencil code will evaluate RTE parameter fields and access coefficients in $A$ and $\vec{Q}$.
+
+Since the stencil has been generated by discretizing the $P_N$-equation relative to the stencil origin, we can apply it to every voxel by simply transforming all queries from stencil space to world space and similarily shift all write operations into $A$ and $\vec{Q}$ according to the current voxel for which the stencil is being executed.
+With the offset information from the stencil and the current voxel, absolute row and column locations in $A$ are determined.
+
+The stencil has been generated without any notion of domain boundaries.
+Our solver supports Neumann boundary conditions (BC) and Dirichlet BC.
+They are handled transparently by the C++ framework through reindexing.
+Whenever the stencil writes into $A$ for a SH coefficient outside the domain, the framework will return the row and column in $A$ for the SH coefficient with the same index, which is the closest inside the domain (Neumann BC) or it will simply ignore the write operation (Dirichlet BC).
+
+Additional active coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are required to respect the boundary correctly (see Figure~\ref{fig:staggeredgrid}).
+This requires careful managegment of additional coefficients in boundary voxels, which is done transparently by the C++ framework.
+
+\todo[inline]{The original todo note withouth changed colours.\newline Here's another line.}
+
+\begin{figure}[h]
+\centering
+\missingfigure{\Huge Nice this triangle?}
+\caption{Some caption}
+\label{nicefigure}
+\end{figure}
+
+<picture showing staggered boundary coefficients>
+%\label{fig:staggeredgrid}
+
+%Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.
+The generated linear system becomes very large for high order and dense FD grids.
+The number of rows is determined by the number of voxels times the number of SH coefficients per voxel.
+The matrix $A$ is squared and fortunately very sparse.
+It further is non-symmetric and also not diagonal dominant, which rules out many standard methods for solving linear systems.
+We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead.
+This gives a symmetric and positive definit system matrix $A^TA$ (although with a higher condition number).
+
+%\subsection*{CDA vs. $P_1$}
+%\begin{itemize}
+  %\item CDA is a degenerated form of $P_1$.
+It is derived by isolating the flux-vector on one side of the vector-equation formed by the $l=1$ SH-band equations.
+This isolation requires division by the extinction coefficient, introducing a $\frac{1}{\sigma_t}$ factor which diverges as $\sigma_t$ approaches zero.
+Thresholding to some minimum is required, in order to be able to solve the system for vacuum regions.
+  %\item $P_1$ does not require any thresholding of $\sigma_t$ as it does not contain $\sigma_t$ as a denominator. It therefore can deal with vacuum regions without modifications.
+  %\item (needs validation) Further, in the presence of vacuum or near vacuum regions, the condition number for CDA is higher than for $P_1$, because of small extinction values in the denominator of the diffusion coefficient.
+  %\item Using the normal form for CDA will further increase the condition number when vacuum regions are present and significantly decreases convergence.
+%\end{itemize}
+
+\section{Flux-limited Diffusion}
+\label{sec:vef}
+
+Our FLD solver follows the implementation in~\cite{Koerner14} very closely (though our solver runs on CPU).
+
 \section{Results}
-\section{Conclusion and future work}
+\label{sec:results}
 
-\begin{itemize}
-  \item Extension to surface rendering by dealing vacuum regions and extending to anisotropic RTE.
-\end{itemize}
+\subsection{Pointsource problem}
+\subsection{Nebulae/Cloud}
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+%\begin{itemize}
+  %\item Extension to surface rendering by dealing vacuum regions and extending to anisotropic RTE.
+  %\item Apply our automated stencil generation to other problem domains where coupled PDE's are present.
+%\end{itemize}
 

### 2018-01-23 python/pnsolver/example2.py
finished real valued pn equations for m<0

@@ -344,7 +344,8 @@ if __name__ == "__main__":
 	#order = [5]
 	staggered = [True]
 	#staggered = [False]
-	boundary_conditions = [False]
+	#boundary_conditions = [False]
+	boundary_conditions = [True]
 
 	#rte_forms = []
 	#order = []

### 2018-01-23 python/pnsolver/meh.py

@@ -429,9 +429,10 @@ class Kronecker(Function):
 		return "\\delta_{{{},{}}}".format(self.getArgument(0).toLatex(), self.getArgument(1).toLatex())
 
 class SHBasis(Function):
-	def __init__(self, l, m, omega, conjugate_complex=False):
+	def __init__(self, l, m, omega, conjugate_complex=False, is_real = False):
 		super().__init__( "Y", [l, m, omega])
 		self.conjugate_complex = conjugate_complex
+		self.is_real = is_real
 
 	def get_l(self):
 		return self.getChildren(0)
@@ -441,22 +442,28 @@ class SHBasis(Function):
 		return self.getChildren(2)
 	def is_conjugate_complex(self):
 		return self.conjugate_complex
-
+	def isReal(self):
+		return self.is_real
+	def isComplex(self):
+		return not self.is_real
 	def deep_copy(self):
 		l = self.get_l()
 		m = self.get_m()
 		omega = self.get_direction_argument()
-		return SHBasis( self.get_l().deep_copy(), self.get_m().deep_copy(), self.get_direction_argument().deep_copy(), self.is_conjugate_complex() )
+		return SHBasis( self.get_l().deep_copy(), self.get_m().deep_copy(), self.get_direction_argument().deep_copy(), self.is_conjugate_complex(), self.isReal() )
 
 	def toLatex(self):
 		l = self.get_l()
 		m = self.get_m()
 		omega = self.get_direction_argument()
+		basistype = "C"
+		if self.isReal():
+			basistype = "R"
 		#print([l,m,omega])
 		if self.is_conjugate_complex():
-			return "\overline{Y^{" + l.toLatex() + ", " + m.toLatex() + "}}(" + omega.toLatex() + ")"
+			return "\overline{Y_{\mathbb{" + basistype + "}}^{" + l.toLatex() + ", " + m.toLatex() + "}}(" + omega.toLatex() + ")"
 		else:
-			return "Y^{" + l.toLatex() + ", " + m.toLatex() + "}(" + omega.toLatex() + ")"
+			return "Y_{\mathbb{" + basistype + "}}^{" + l.toLatex() + ", " + m.toLatex() + "}(" + omega.toLatex() + ")"
 
 
 class SHCoefficient(Function):
@@ -483,6 +490,8 @@ class Operator( Expression ):
 		return self.numChildren()
 	def getOperand(self, index):
 		return self.children[index]
[…]

### 2018-01-23 python/pnsolver/rte_terms.py

@@ -626,9 +626,9 @@ class fopn(object):
 		#	meh.print_expr(expr)
 
 		expr = meh.apply_recursive(expr, meh.DistributiveLaw())
-		if debug == True:
-			print("------------------------------")
-			meh.print_expr(expr)
+		#if debug == True:
+		#	print("------------------------------")
+		#	meh.print_expr(expr)
 
 		expr = meh.apply_recursive(expr, meh.SHRecursiveRelation())
 		expr = meh.apply_recursive(expr, meh.SplitIntegrals())
@@ -744,7 +744,58 @@ class CountImaginaryUnits(object):
 
 class fopn_real(object):
 	@staticmethod
-	def transport_term_expand( sh_real_basis, order, debug = False ):
+	def transport_term_expand2( m, order, debug = False, gg = None ):
+
+		x = meh.tensor("\\vec{x}", rank=1, dimension=3)
+		x.setComponent(0, meh.var("x"))
+		x.setComponent(1, meh.var("y"))
+		x.setComponent(2, meh.var("z"))
+		x.collapsed = True
+
+		coeff_c = meh.SHRecursiveRelation.c( meh.sub(meh.var("l'"), meh.num(1)), meh.sub( meh.var("m'"), meh.num(1) ) )
+		coeff_d = meh.SHRecursiveRelation.d( meh.add(meh.var("l'"), meh.num(1)), meh.sub( meh.var("m'"), meh.num(1) ) )
+		coeff_e = meh.SHRecursiveRelation.e( meh.sub(meh.var("l'"), meh.num(1)), meh.add( meh.var("m'"), meh.num(1) ) )
+		coeff_f = meh.SHRecursiveRelation.f( meh.add(meh.var("l'"), meh.num(1)), meh.add( meh.var("m'"), meh.num(1) ) )
+		coeff_a = meh.SHRecursiveRelation.a( meh.sub(meh.var("l'"), meh.num(1)), meh.var("m'"))
+		coeff_b = meh.SHRecursiveRelation.b( meh.add(meh.var("l'"), meh.num(1)), meh.var("m'"))
+
+		L0 = meh.SHCoefficient( 'L', meh.sub(meh.var("l'"), meh.num(1)), meh.add(meh.neg(meh.var("m'")), meh.num(1)), x )
[…]

### 2018-01-23 python/pnsolver/stencil2.py

@@ -88,6 +88,8 @@ class PNInfo3D(object):
 		return self.numCoeffs
 
 	def getIndex( self, l, m ):
+		if l > self.order:
+			return None
 		return l*(l+1)+m 
 
 	def getLMIndex( self, index ):
@@ -136,8 +138,11 @@ class PNInfo2D(object):
 		return self.numCoeffs
 
 	def getIndex( self, l, m ):
+		#print( "getIndex l={} m{}".format(l,m) )
 		index3d = self.pni3d.getIndex(l,m)
-		return self.index3d_to_index2d[index3d]
+		if not index3d is None:
+			return self.index3d_to_index2d[index3d]
+		return None
 
 	def getLMIndex( self, index2d ):
 		index3d = self.index2d_to_index3d[index2d]
@@ -542,6 +547,9 @@ def apply_spatial_discretization_recursive( expr, info, level = 0 ):
 		# function_offset is the staggered grid location at which the function is defined
 		if expr.getSymbol() == info.pni.getUnknownSymbol():
 			coeff_index = getUnknownIndex(info.pni, expr.getArguments())
+			if coeff_index == None:
+				# the coefficient vanishes (this happens in 2d case)
+				return meh.num(0)				
 			function_offset = info.pni.getOffset(coeff_index)
 		else:
 			# all other functions are always defined at cell centers
@@ -577,6 +585,10 @@ def apply_spatial_discretization_recursive( expr, info, level = 0 ):
 			# so we just return the expression as is
 			return u
 		else:
+			#print(eval_offset)
+			#print(coeff_index)
+			#print(function_offset)
+			#raise ValueError("interpolation not expected")
 			# now we interpolate from the neighbous along the defined local axes
 			offset_combinations = itertools.product(*[[-1, 1] for d in range(numAxes)])
 			num_offset_combinations = 2**numAxes
@@ -898,11 +910,20 @@ def generate_stencil_code( pne, pni, stencil_name ):
 		e = pne_discretized[coeff_index]
 		l,m = pni.getLMIndex(coeff_index)
 
+		debug = False
+		#if l==1 and m == -1:
+		#	debug = True
+
+		if debug == True:
+			print("TEST --------------")
+			meh.print_expr(e)
+
+
 		#if coeff_index == 1:
 		#	print("\n---------------------")
 		#	print("l={} m={}".format(l,m))
 		#	meh.print_expr(e)
[…]

### 2018-01-23 latex/egsr17-cgf/main-body.tex
started working on supplemental document with extended derivations

@@
 \title[Draft]%
-{Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media}
+{
+%Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media
+A Deterministic Method for Simulation of Light Transport in Participating Media based on Higher Order Spherical Harmonics
+}
 
@@
 % uncomment for using teaser
-% \teaser{
-%  \includegraphics[width=\linewidth]{eg_new}
-%  \centering
-%   \caption{New EG Logo}
-% \label{fig:teaser}
-%}
+\teaser
+{
+\centering
+\begin{subfigure}{0.2\linewidth}
+\missingfigure{}
+\caption{$CDA/P_1$}
+\end{subfigure}%
+\hspace{0.05\linewidth}
+\begin{subfigure}{0.2\linewidth}
+\missingfigure{}
+\caption{FLD}
+\end{subfigure}%
+\hspace{0.05\linewidth}
+\begin{subfigure}{0.2\linewidth}
+\missingfigure{}
+\caption{$P_5$}
+\end{subfigure}%
+\hspace{0.05\linewidth}
+\begin{subfigure}{0.2\linewidth}
+\missingfigure{}
+\caption{Groundtruth}
+\end{subfigure}%
+%\includegraphics[width=\linewidth]{figures/eg_new}
+\icaption
+{
+Flux-limited diffusion (b) is an ad-hoc extension to the classical diffusion approximation (a).
+It can be considered the state of the art among deterministic methods for simulating light transport in highly scattering participating media, which is a common challenge for brute-force path tracing (d).
+We add the $P_N$-method (c) to the arsenal of available deterministic methods in rendering and investigate its benefits and tradeoffs against flux-limited diffusion.
+}
+\label{fig:teaser}
+}
 
@@
 \begin{abstract}
-
-Classical diffusion approximation (CDA) is very competitive for rendering highly scattering participating media, because in these scenarios, Monte Carlo based methods suffer from high computational cost.
+%Classical diffusion approximation (CDA) is very competitive for rendering highly scattering participating media, because in these scenarios, Monte Carlo based methods suffer from high computational cost.
 Variable Eddington Factor methods, such as flux-limited diffusion (FLD), have shown to further improve the results from CDA.
@@
 We further present a staggered-grid $P_N$-solver and compare it against FLD for various standard problems.
+Rendering highly scattering participating media is a common challenge for brute-force path-tracing.
+Classical diffusion approximation (CDA) is a competitive alternative, as it reduces the problem to solving a simple diffusion equation.
+Variable eddington factor methods, such as flux-limited diffusion (FLD), improve CDA by making the diffusion coefficient non-linear.
+Both methods are based on the spherical harmonics expansion of the radiance field, truncated after the first order.
+In this paper, we investigate the open question, if going higher order provides a viable alternative to the ad-hoc fixes to diffusion by FLD.
+Increasing the order introduces a growing set of increasingly complex coupled partial differential equations, called $P_N$-equations, which become very difficult to work with for higher order.
+We introduce a computer algebra framework for representing and manipulating mathematical equations and use it to derive the time-independent real-valued $P_N$-equations at arbitrary order.
+We further present a staggered-grid $P_N$-solver and generate its stencil code directly from the expression tree of the $P_N$-equations.
+Both, the computer algebra framework, as well as the solver are made avaiable open source.
+Finally, we discuss how our method compares against FLD for various standard problems.
 
@@
 
-Light transport in participating media is governed by the radiative transfer equation(RTE), which was first introduced in astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}.
+Light transport in participating media is governed by the radiative transfer equation~(RTE), which was first introduced in astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}.
 Today, this equation is typically solved using Monte Carlo based methods.
-However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become unfeasible, due to high computational cost which comes from tracing long particle paths with many scattering events.
+However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become unfeasible, due to the high computational cost which comes from tracing long particle paths with many scattering events.
 
@@
 This problem is addressed by the variable eddington factor(VEF)-formalism, among which flux-limited diffusion (FLD) is the most prominent.
-FLD was first introduced in astrophysics by~\cite{Pomraining78} and was introduced to graphics by Koerner et al.~\cite{Koerner14}.
-Unfortunately, both CDA and FLD do not support anisotropic phase functions.
+FLD was first introduced in astrophysics by~\cite{Levermore81} and was introduced to graphics by Koerner et al.~\cite{Koerner14}.
 
 The idea behind VEF is to replace the diffusion coefficient in the diffusion equation of CDA by a non-linear diffusion coefficient.
-This diffusion coefficient is being derived by seeing it as a mean to interpolate between the diffusive and pure transport regime.
+This new diffusion coefficient is being derived by seeing it as a mean to spatially blend between the diffusive and pure transport regime.
 The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
@@
 
-The $P_N$-equations are derived from the radiative transfer equation~(RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
+In this section we present how to optain the real-valued $P_N$-equations.
+For brevity and readability, we only give a brief outline here and refer to the supplemental material for the detailed derivation.
+
+The $P_N$-equations are derived from the radiative transfer equation, which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
 \begin{align}
@@
 =&
--\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\\
+-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\nonumber\\
 &
@@
 p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega \right)\ud\omega'
-}\\
+}\nonumber\\
 &
-+Q\left(\vec{x}, \omega\right)
++Q\left(\vec{x}, \omega\right)\nonumber
 \end{align}
@@
 The LHS is the transport term and we refer to the terms on the RHS as collision, scattering and source term respectively.
+The symbols $\sigma_t$, $\sigma_s$, $p$ and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term respectively.
 
@@
 
+\begin{figure}[h]
+\centering
+\missingfigure{test}
+\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
+\end{figure}
+
 In order to derive the real-valued $P_N$-equations from which we can construct $A$ and $\vec{Q}$, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
@@
 
-First, angular dependent quantities in the RTE, such as the radiance field $L$ (or the emission field $Q$) are projected into SH coefficients:
+First, the angular dependent quantities in the RTE, such as the radiance field $L$ (or the emission field $Q$) are projected into SH coefficients:
 \begin{align}
@@
 
-The reconstruction $\hat{L}$ will have an error due to truncation at order $N$:
+The reconstruction $\hat{L}$ has an error due to truncation at order $N$:
 \begin{align}
@@
 The angular dependent quantities are replaced with their SH reconstruction in the RTE.
-The $P_N$-equations are found by multiplying each term with the conjugate complex of the real-valued SH basis function $\overline{\SHBR^{l,m}}$, followed by integration over solid angle.
+The $P_N$-equations are found by multiplying each term with the conjugate complex of the real-valued SH basis function, followed by integration over solid angle.
 In order to arrive at a form, which does not depend on angle anymore, a set of transformations and identities are applied.
@@
 
-and the following recursion relation
+and recursion relations such as the following:
 \begin{align}
-\omega_x\overline{\SHBC^{l, m}}=
-&
--\frac{1}{2}c^{{l-1,m-1}}\overline{Y^{l-1, m-1}}+\frac{1}{2}d^{{l+1,m-1}}\overline{Y^{l+1, m-1}}\\
-&
-+\frac{1}{2}e^{{l-1,m+1}}\overline{Y^{l-1, m+1}}-\frac{1}{2}f^{{l+1,m+1}}\overline{Y^{l+1, m+1}}\\
-\omega_y\SHBC^{l, m}=
-&
-\frac{i}{2}c^{{l-1,m-1}}\overline{Y^{l-1, m-1}}-\frac{i}{2}d^{{l+1,m-1}}\overline{Y^{l+1, m-1}}\\
-&
-+\frac{i}{2}e^{{l-1,m+1}}\overline{Y^{l-1, m+1}}-\frac{i}{2}f^{{l+1,m+1}}\overline{Y^{l+1, m+1}}\\
-\omega_y\SHBC^{l, m}=
+\omega_z\SHBC^{l, m}=
 &
@@
 &a^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}},
-b^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}\\
-&c^{l,m}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}},
-d^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}\\
-&e^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}},
-f^{l,m}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
+b^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
 \end{align}
@@
 Note that the real-valued SH basis functions are different, depending on $m$, which means that we get three different versions of the $P_N$-equations (for $m<0$, $m=0$ and $m>0$).
-Further note, that we need to expand the real-valued SH basis functions in order to be able to apply the given identities above.
-This causes the equations to become very large and difficult to deal with when applying transformations and identities.
+Further note, that we need to expand the real-valued SH basis functions into their complex-valued definition, in order to be able to apply the given identities above.
+This causes the equations to become large and difficult to deal with when applying transformations and identities.
 
 It becomes even more of a problem later, when we want to expand the spatially discretized equations for a particular choice of truncation order $N$.
-Manually creating a FD stencil from hundreds of terms, including complex numbers and derivatives, is very challenging for first order and becomes impossible for higher order.
+Manually creating a FD stencil from so many terms is challenging for first order and becomes increasingly difficult and error prone for higher order.
 
@@
 
-In order to tackle this problem, we created a system for representing algebraic expressions as an expression tree in Python.
+\begin{figure}[h]
+\centering
+\begin{subfigure}{0.45\columnwidth}
+\missingfigure{test}
+\end{subfigure}%
+\hspace{0.05\columnwidth}
+\begin{subfigure}{0.45\columnwidth}
+\missingfigure{test2}
+\end{subfigure}%
+
+\begin{subfigure}{0.45\columnwidth}
+\missingfigure{test}
+\end{subfigure}%
+\hspace{0.05\columnwidth}
+\begin{subfigure}{0.45\columnwidth}
+\missingfigure{test2}
+\end{subfigure}%
+
+\icaption{The computer algebra representation is constructed using API commands provided by our framework (top left) and represented internally as an expression tree (top right).
+Additional API commands (bottem left) allow valid transformations, which modify the expression tree accordingly (bottom right).}
+\end{figure}
+
+In order to tackle this problem, we created a system for representing algebraic expressions as an expression tree.
 Our framework can represent terms, integrals, derivatives, sums, variables, functions and special symbols, such as the Kronecker delta or imaginary number.
@@
 
-<show some codesnippet for generating expressions and applying manipulations?>
-
 The computer algebra representation of the $P_N$-equations is input to our $P_N$-solver, which we will discuss in the next section.
@@
 \label{sec:pnsolver}
+
+\begin{figure*}[t]
+\centering
+\missingfigure{$P_N$-solver overview: generate stencil code $\rightarrow$ build system $\rightarrow$ solve $\rightarrow$ render}
+\icaption{Overview of our $P_N$-solver.
+After generating the stencil source code from the expression trees representing the $P_N$-equations, the linear system $A\vec{u}=\vec{Q}$ is built using RTE parameter fields and additional user input, such as grid resolution and type of boundary conditions.
+The resulting system is solved for $\vec{u}$, which is then used in our rendering application.}
+\label{fig:pnsolver}
+\end{figure*}
 
@@
 
-As we will see later in section~\ref{sec:results}, our solver needs to support staggered grids in order to avoid severe oscillation artefacts which arise, when all coefficients are located at the voxel center.
+As we see in figure~\ref{fig:artefacts}, our solver needs to support staggered grids in order to avoid severe oscillation artefacts which arise, when all coefficients are located at the voxel center.
 With staggered grids, coefficients can live on the intersections of a subgrid with double resolution.
@@
 It returns the unknown itself, if both happen to coincide at the same location.
+
+\begin{figure}[h]
+\centering
+\begin{subfigure}{0.45\columnwidth}
+\missingfigure{test}
+\end{subfigure}%
+\hspace{0.05\columnwidth}
+\begin{subfigure}{0.45\columnwidth}
+\missingfigure{test2}
+\end{subfigure}%
+\icaption{The solution of our solver for the 2D-checkerboard problem with collocated grids (left), where all coefficients are located at the cell centers.
+It suffers from severe oscillating artefacts, which are removed by offseting the coefficient locations using staggered grids (right).}
+\label{fig:artefacts}
+\end{figure}
 
@@
 Additional active coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are required to respect the boundary correctly (see Figure~\ref{fig:staggeredgrid}).
-This requires careful managegment of additional coefficients in boundary voxels, which is done transparently by the C++ framework.
+This requires careful managegment of additional coefficients at boundary voxels, which is done transparently by the C++ framework.
 
-\todo[inline]{The original todo note withouth changed colours.\newline Here's another line.}
+%\todo[inline]{The original todo note withouth changed colours.\newline Here's another line.}
 
@@
 \centering
-\missingfigure{\Huge Nice this triangle?}
-\caption{Some caption}
-\label{nicefigure}
+\begin{subfigure}{0.45\columnwidth}
+\missingfigure{test}
+\end{subfigure}%
+\hspace{0.05\columnwidth}
+\begin{subfigure}{0.45\columnwidth}
+\missingfigure{test2}
+\end{subfigure}%
+\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
+\label{fig:staggeredgrid}
 \end{figure}
 
-<picture showing staggered boundary coefficients>
-%\label{fig:staggeredgrid}
-
 %Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.
-The generated linear system becomes very large for high order and dense FD grids.
+The generated linear system becomes very large for high order and resolution.
 The number of rows is determined by the number of voxels times the number of SH coefficients per voxel.
@@
 
+\subsection{Rendering}
+
+\begin{itemize}
+	\item explain that we use single scattered light as emission field
+	\item explain our single scattering path tracer using the $P_N$-solution as indirect light
+\end{itemize}
+
+\subsection{2D checkerboard}
 \subsection{Pointsource problem}

### 2018-02-01 latex/egsr17-cgf/main-body.tex
continued writing the paper. finished theory section

@@
 \missingfigure{}
-\caption{$CDA/P_1$}
+\caption{Classical diffusion}
 \end{subfigure}%
@@
 \missingfigure{}
-\caption{FLD}
+\caption{Flux-limited diffusion}
 \end{subfigure}%
@@
 \missingfigure{}
-\caption{$P_5$}
+\caption{ours}
 \end{subfigure}%
@@
 \missingfigure{}
-\caption{Groundtruth}
+\caption{Ground truth}
 \end{subfigure}%
@@
 It can be considered the state of the art among deterministic methods for simulating light transport in highly scattering participating media, which is a common challenge for brute-force path tracing (d).
-We add the $P_N$-method (c) to the arsenal of available deterministic methods in rendering and investigate its benefits and tradeoffs against flux-limited diffusion.
+We add the $P_N$-method (c) to the toolbox of deterministic methods in rendering and investigate its benefits and tradeoffs against flux-limited diffusion.
 }
@@
 Rendering highly scattering participating media is a common challenge for brute-force path-tracing.
-Classical diffusion approximation (CDA) is a competitive alternative, as it reduces the problem to solving a simple diffusion equation.
-Variable eddington factor methods, such as flux-limited diffusion (FLD), improve CDA by making the diffusion coefficient non-linear.
+Classical diffusion approximation is a competitive alternative, as it reduces the problem to solving a simple diffusion equation.
+Extensions, such as flux-limited diffusion (FLD), improve the former by switching to a non-linear diffusion coefficient.
 Both methods are based on the spherical harmonics expansion of the radiance field, truncated after the first order.
-In this paper, we investigate the open question, if going higher order provides a viable alternative to the ad-hoc fixes to diffusion by FLD.
-Increasing the order introduces a growing set of increasingly complex coupled partial differential equations, called $P_N$-equations, which become very difficult to work with for higher order.
-We introduce a computer algebra framework for representing and manipulating mathematical equations and use it to derive the time-independent real-valued $P_N$-equations at arbitrary order.
+In this paper, we investigate the open question of wether higher orders provide a viable alternative to these two approaches.
+Increasing the order introduces a set of increasingly complex coupled partial differential equations, whose growing number and complexity make them very difficult to work with for higher order.
+We introduce a computer algebra framework for representing and manipulating the underlying mathematical equations and use it to derive the time-independent real-valued $P_N$-equations for arbitrary orders.
 We further present a staggered-grid $P_N$-solver and generate its stencil code directly from the expression tree of the $P_N$-equations.
-Both, the computer algebra framework, as well as the solver are made avaiable open source.
-Finally, we discuss how our method compares against FLD for various standard problems.
-
+Finally, we discuss how our method compares against prior work for various standard problems.
+We will release our computer algebra system, solver, and data as open source to ensure full reproducibility of all of our results.
 %\begin{classification} % according to http://www.acm.org/class/1998/
@@
 %\end{classification}
-
 \end{abstract}
@@
 
-Light transport in participating media is governed by the radiative transfer equation~(RTE), which was first introduced in astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}.
+Light transport in participating media is governed by the radiative transfer equation~(RTE), first studied in the context of astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}.
 Today, this equation is typically solved using Monte Carlo based methods.
-However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become unfeasible, due to the high computational cost which comes from tracing long particle paths with many scattering events.
+However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become prohibitive, due to the high computational cost which comes from tracing long particle paths with many scattering events.
 
@@
 In fact, as Max~\cite{Max95} pointed out, it is not clear if Kajiya succeeded at all at applying the method, as all of the results in the paper were produced with a simpler method.
-This is further strengthened by our observation, that a straight forward finite difference discretization of the $P_N$-equations produces unuseable results, due to oscillation artefacts in the solution.
-In our paper, we derive the real-valued $P_N$-equations and present a staggered-grid solver, which produces an artefact free solution.
+This is further strengthened by our observation, that a straight forward finite difference discretization of the $P_N$-equations produces unuseable results, due to oscillation artifacts in the solution.
+In our paper, we derive the real-valued $P_N$-equations along with a staggered-grid solver, that produces artifact free solutions.
 
@@
 CDA suffers from severe energy loss close to regions with strong density gradients.
-This problem is addressed by the variable eddington factor(VEF)-formalism, among which flux-limited diffusion (FLD) is the most prominent.
-FLD was first introduced in astrophysics by~\cite{Levermore81} and was introduced to graphics by Koerner et al.~\cite{Koerner14}.
+The problem is addressed by a modification known as Variable Eddington factor, which nonlinearly adjust the diffusion coefficient to improve the solution near density gradients and low-density regions.
+Flux-limited diffusion, developed in the context of astrophysics by Levermore et al.~\cite{Levermore81} and later introduced to graphics by Koerner et al.~\cite{Koerner14}, is the most prominent example.
 
@@
 
-The LHS is the transport term and we refer to the terms on the RHS as collision, scattering and source term respectively.
-The symbols $\sigma_t$, $\sigma_s$, $p$ and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term respectively.
+The left hand side (LHS) is the transport term and we refer to the terms on the right hand side (RHS) as collision, scattering and source term respectively.
+The symbols $\sigma_t$, $\sigma_s$, $p$ and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term.
 
-The $P_N$-method uses SH to discretize the angular domain into a set of coupled equations.
-We further discretize the spatial variable using a finite difference (FD) grid.
-Each term represents a change of the radiance field $L$.
-Therefore, after discretizing the continuous variables, we can represent the RTE as a sequence of linear transformations, which can be collapsed into a single coefficient matrix $A$ and a RHS vector $\vec{Q}$:
+The RTE is often given in operator notation, where each term is expressed as an operator, acting on the radiance field $L$:
 \begin{align}
-(T+C-S)\vec{u} = A\vec{u} = \vec{Q}
+T\left(L\right) = -C\left(L\right) + S\left(L\right) + Q
 \end{align}
 
-$T$, $C$, $S$ are matrices, which result from the discretized transport, collision and scattering terms in the $P_N$-equations respectively.
-All SH coefficients for every voxel of our FD-grid are flattened into the solution vector $\vec{u}$, which is the unknown we want to solve for.
-The RHS vector likewise contains the flattened SH coefficients of the emission field $Q$ of every voxel.
-
-\begin{figure}[h]
-\centering
-\missingfigure{test}
-\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
-\end{figure}
-
-In order to derive the real-valued $P_N$-equations from which we can construct $A$ and $\vec{Q}$, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
-\begin{align}
+The $P_N$-method expands the angular dependence into spherical harmonics.
+For the real-valued $P_N$-equations, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
+\begin{align*}
 \SHBR^{l,m}=
@@
 \right.
+\end{align*}
+
+We express the projection into the SH basis functions with the projection operator $\mathcal{S}$:
+\begin{align}
+\mathcal{P}^{l, m}(f) = \int_{\Omega}f(\mathbf{x}, \omega) \SHBR^{l,m}(\omega)\,\mathrm{d}\omega = f^{l,m}
+\nonumber
 \end{align}
 
-First, the angular dependent quantities in the RTE, such as the radiance field $L$ (or the emission field $Q$) are projected into SH coefficients:
-\begin{align}
-L^{l,m}\left(\vec{x}\right)
-=
-\int_\Omega{L\left(\vec{x}, \omega\right)\SHBR^{l,m}\ud\omega}
-\end{align}
-
-The reconstruction $\hat{L}$ has an error due to truncation at order $N$:
-\begin{align}
-L\left(\vec{x}, \omega\right)
-\approx
-\hat{L}\left(\vec{x}, \omega\right) =
+The $P_N$-equations are derived by first replacing the radiance field $L$ by its SH reconstruction $\widehat{L}$, which introduces an error due to truncation at order $N$:
+\begin{align*}
+\widehat{L}\left(\vec{x}, \omega\right) =
 \sum_{l=0}^{N}
@@
 }
+\approx
+L\left(\vec{x}, \omega\right)
+\end{align*}
+
+After similarly replacing all angular dependent RTE parameters, we project each term into SH, using the projection operator $\mathcal{S}$.
+This produces a single equation for each $l,m$-pair.
+The $P_N$-equations therefore can be written as the following set:
+\begin{align}
+\mathcal{P}^{l,m}\mathcal{T}\left(\widehat{L}\right)
 =
-\sum_{l,m}
-{
-L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
-}
+-\mathcal{P}^{l,m}\mathcal{C}\left(\widehat{L}\right) 
++\mathcal{P}^{l,m}\mathcal{S}\left(\widehat{L}\right)
++\mathcal{P}^{l,m}\left(Q\right)
+\nonumber
 \end{align}
 
-The angular dependent quantities are replaced with their SH reconstruction in the RTE.
-The $P_N$-equations are found by multiplying each term with the conjugate complex of the real-valued SH basis function, followed by integration over solid angle.
-In order to arrive at a form, which does not depend on angle anymore, a set of transformations and identities are applied.
-Of particular importance are the orthogonality property
+Expanding the operators and writing the real-valued $P_N$-equations in compact form, equires a string of expansions, manipulations and applications of identities.
+We here give the final result and refer to the supplemental material for the detailed derivation.
+Since the real-valued SH basis functions differ, depending on $m<0$, $m=0$ or $m>0$, we get different projections $\mathcal{S}^{l,m}$ as well.
+
+Therefore, the real-valued $P_N$-equations are for $m<0$:
+\begin{align*}
+&-\frac{1}{2}c^{\scaleto{l-1,m-1}{4pt}}
+\partial_y
+L^{\scaleto{l-1,-m+1}{4pt}}
+%\\
++\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
+\partial_y
+L^{\scaleto{l+1,-m+1}{4pt}}
+%\\
+-\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m+1}{4pt}}
+\partial_y
+L^{\scaleto{l-1,-m-1}{4pt}}
+\\&
++\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{\scaleto{l+1,m+1}{4pt}}
+\partial_y
+L^{\scaleto{l+1,-m-1}{4pt}}
+%\\
++\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}c^{\scaleto{l-1,m-1}{4pt}}
+\partial_x
+L^{\scaleto{l-1,m-1}{4pt}}
+\\&
+-\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}e^{{l-1,m+1}}
+\partial_x
+L^{\scaleto{l-1,m+1}{4pt}}
+%\\
++\frac{1}{2}f^{\scaleto{l+1,m+1}{4pt}}
+\partial_x
+L^{\scaleto{l+1,m+1}{4pt}}
+%\\
+-\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
+\partial_x
+L^{\scaleto{l+1,m-1}{4pt}}
+\\&
++a^{\scaleto{l-1,m}{4pt}}
+\partial_z
+L^{\scaleto{l-1,m}{4pt}}
+%\\
++b^{\scaleto{l+1,m}{4pt}}
+\partial_z
+L^{\scaleto{l+1,m}{4pt}}
+%\\
++\sigma_t L^{\scaleto{l,m}{4pt}}
+%\\
+-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
+%\\
+= Q^{\scaleto{l,m}{4pt}}
+\end{align*}
+
+for $m=0$:
+\begin{align*}
+&
+\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_x L^{\scaleto{l-1,1}{4pt}}
+-\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_x L^{\scaleto{l+1,1}{4pt}}
+%\\&
+\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_y L^{\scaleto{l-1,-1}{4pt}}
+\\&
+-\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_y L^{\scaleto{l+1,-1}{4pt}}
+%\\&
+a^{\scaleto{l-1,0}{4pt}}\partial_z L^{\scaleto{l-1,0}{4pt}}
++b^{\scaleto{l+1,0}{4pt}}\partial_z L^{\scaleto{l+1,0}{4pt}}
+\\&
++\sigma_t L^{\scaleto{l,m}{4pt}}
+-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
+= Q^{\scaleto{l,m}{4pt}}
+\end{align*}
+
+and for $m>0$:
 \begin{align}
-\int_{\Omega}{ \SHBC^{l,m}\overline{\SHBC^{l',m'}}\ud\omega} = \delta_{l,l'}\delta_{m,m'}
-\end{align}
-
-and recursion relations such as the following:
-\begin{align}
-\omega_z\SHBC^{l, m}=
 &
-a^{{l-1,m}}\overline{Y^{l-1, m}}+b^{{l+1,m}}\overline{Y^{l+1, m}}
+\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_x L^{\scaleto{l-1,m+1}{4pt}}
+%\\&
+-\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_x L^{\scaleto{l+1,m+1}{4pt}}
+%\\&
+-\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m-1}{4pt}}\partial_x L^{\scaleto{l-1,m-1}{4pt}}
+\nonumber
+\\&
++\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{l+1,-m+1}\partial_x L^{\scaleto{l+1,m-1}{4pt}}
+%\\&
++\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_y L^{\scaleto{l-1,-m-1}{4pt}}
+\nonumber
+\\&
+-\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_y L^{\scaleto{l+1,-m-1}{4pt}}
+%\nonumber
+%\\&
++\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}e^{\scaleto{l-1,-m+1}{4pt}}\partial_y L^{\scaleto{l-1,-m+1}{4pt}}
+\nonumber
+\\&
+-\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}f^{\scaleto{l+1,-m+1}{4pt}}\partial_y L^{\scaleto{l+1,-m+1}{4pt}}
+%\\&
++a^{\scaleto{l-1,-m}{4pt}}\partial_z L^{\scaleto{l-1,m}{4pt}}
+%\nonumber
+%\\&
++b^{\scaleto{l+1,-m}{4pt}}\partial_z L^{\scaleto{l+1,m}{4pt}}
+\nonumber
+\\&
++\sigma_t L^{\scaleto{l,m}{4pt}}
+-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
+= Q^{\scaleto{l,m}{4pt}}
+\nonumber
 \end{align}
@@
 with
+\begin{align*}
+\label{eq:real_sh_basis}
+\beta^{x}=
+\left\{
+\begin{array}{lr}
+\frac{2}{\sqrt{2}}, & \text{for } \vert x\vert = 1\\
+1, & \text{for } \vert x\vert \neq 1
+\end{array}
+\right.
+,\quad
+\delta_{x\neq y}=
+\left\{
+\begin{array}{lr}
+1, & \text{for } x \neq y \\
+0, & \text{for } x = y
+\end{array}
+\right.
+\end{align*}
+
+and
+\begin{align*}
+&
+a^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}} \qquad
+b^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
+\\&
+c^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \qquad
+d^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
+\\&
+e^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \qquad
+f^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
+\end{align*}
+\begin{align*}
+\lambda_l=\sqrt{\frac{4\pi}{2l+1}}
+\end{align*}
+
+The spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxelgrid.
+The radiance field $L$, after full discretization, is represented as a set of coefficients per voxel.
+Flattening these into a single vector gives the solution vector $\vec{u}$.
+The RHS vector $\vec{Q}$ is produced similarly.
+The differential operators are discretized using a finite difference stencil.
+This allows the projected operators to be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$:
 \begin{align}
-&a^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}},
-b^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
+(T+C-S)\vec{u} = A\vec{u} = \vec{Q}
 \end{align}
 
-%Finally, the spatially dependent variables are discretized (using finite difference stencils for derivatives).
-The resulting equations are factorized according to the unknowns ($l,m$-coefficient within voxel $i,j,k$).
-The coefficients to these unknowns
+$T$, $C$, $S$ are matrices, which result from the discretized transport, collision and scattering term in the $P_N$-equations respectively.
 
-Note that the real-valued SH basis functions are different, depending on $m$, which means that we get three different versions of the $P_N$-equations (for $m<0$, $m=0$ and $m>0$).
-Further note, that we need to expand the real-valued SH basis functions into their complex-valued definition, in order to be able to apply the given identities above.
-This causes the equations to become large and difficult to deal with when applying transformations and identities.
+\begin{figure}[h]
+\centering
+\missingfigure{test}
+\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
+\end{figure}
 
-It becomes even more of a problem later, when we want to expand the spatially discretized equations for a particular choice of truncation order $N$.
-Manually creating a FD stencil from so many terms is challenging for first order and becomes increasingly difficult and error prone for higher order.
+In the next section, we will present our solver for solving the real-valued $P_N$-equations.
 
-\subsection*{Computer Algebra Representation}
+\section{$P_N$-Solver}
+\label{sec:pnsolver}
+
+\subsection{Computer Algebra Representation}
+
+Expanding and discretizing the $P_N$-equations is arduous and error prone if done by hand and gets increasingly complex for higher order.
+We tackle this problem by expressing the RTE in a computer algebra representation and automatically apply the discretization in angular and spatial domain using manipulation rules on this representation.
 
@@
 
+\subsection{Angular Discretization}
+
 After entering each term of the RTE into our framework, we arrive at the $P_N$-equations by applying a set of manipulations and identities using API commands provided by our framework.
@@
 The computer algebra representation of the $P_N$-equations is input to our $P_N$-solver, which we will discuss in the next section.
-
-\section{$P_N$-Solver}
-\label{sec:pnsolver}
 
@@
 
-\subsection{Stencil Generation}
+\subsection{Spatial Discretization}
 
@@
 
-\subsection{System Building and Solving}
+\subsection{Solving}
 

### 2018-02-07 latex/egsr17-cgf/conclusion.tex
continued working on draft

\section{Conclusion}
\label{sec:conclusion}

In this paper, we introduced the $P_N$-method to the toolbox of deterministic methods for rendering participating media in computer graphics. We derived and presented the real-valued $P_N$-equations, along with a staggered grid solver for solving them. Unique to our system is, that it uses a computer algebra representation of the equation to solve as input. Discretization in angular and spatial domain is done without any user interference, using manipulation passes on the representation. The stencil code, which is used to propagate the system of linear equations, is automatically generated from the expression tree.

This way, we can easily deal with complex coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand. Our approach helps especially for higher order, when the number of equations increase and become more complex.

We further show, how to use the results in a rendering system and run our solver for various standard problems for validation.

\begin{itemize}
	\item Mention performance problems and improving of solver performance as future work.
	\item mention vaccum convergence problem and the challenge to couple deterministic methods with surface rendering problems.
	\item Apply our automated stencil generation to other problem domains where coupled PDE's are present.
\end{itemize}

### 2018-02-07 latex/egsr17-cgf/introduction.tex
continued working on draft

--- 2017-11-27 latex/egsr17-cgf/introduction.tex
@@
 
-\subsection*{Contributions}
+Simulating light transport in paticipating media is still a challenging problem for image synthesis in computer graphics.
+Monte Carlo based techniques have established as the standard approach, because they can produce unbiased results and are conceptionally easy to implement.
+They become computationally demanding however, for participating media, which exhibits strong scattering or anisotropy.
+
+Deterministic methods on the other hand are unpopular.
+This is because they suffer from discretization artifacts, produce biased results, cannot be coupled easily with surface rendering problems and are trickier to implement.
+However, their appeal lies in the fact that they have better performance for certain problems and produce a global solution across the whole domain.
+
+The work on path guiding techniques from recent years has shown, how discretized approximate representations of the correct solution can be used to accelerate Monte Carlo integration techniques, such as path tracing.
+Instead of generating these approximate representations using correct methods, deterministic methods may offer a viable alternative and pave the way for hybrid methods, which combine the performance benefits of deterministic methods with correct and exact Monte Carlo techniques.
+Further, deterministic methods lend themselfs to applications, where a fast approximate solution is often preferred over correct, but slowly converging results.
+Under this light, it is important to study deterministic methods and have a solid understanding about their characteristics and performance traits for typical rendering problems.
+
+The $P_N$-method is a deterministic method, which is popular in other fields like medical imaging and nuclear sciences, but has not been explored for applications in graphics yet.
+The purpose and main contribution of our paper is, to gain a solid understanding of its foundations and present a method for application in rendering.
+In particular, we present the following theoretical and practical contributions:
 \begin{itemize}
-	\item System for generation of stencils for coupled PDE's directly from mathematical expressions.
-	\item Staggered-grid solver for time-independent PN-equations.
-	\item Novel PN-guided Monte-Carlo estimator for participating media.
+	\item We derive and present the time-independent real-valued $P_N$-equations.
+	\item We introduce a staggered-grid solver, for which we generate stencil code automatically from a computer algebra representation of the $P_N$-equations.
+This allows us to deal with the increasingly complex equations, which the $P_N$-method produces for higher order.
+It further allows our solver to be used for any (potentially coupled) partial differential equation, which results in a system of linear equations after discretization.
+	\item Finally, we compare the $P_N$-method for higher orders against flux-limited diffusion, and answer the long standing question about which approach should be preferred.
 \end{itemize}
+
+In the next section, we will discuss related work and why it motivates our contribution.
+Section~\ref{sec:pnequations} introduces the real-valued $P_N$-equations, which we solve using our solver described in section~\ref{sec:pnsolver}.
+We explain in section~\ref{sec:rendering}, how the solution is being applied in a rendering application.
+Our $P_N$-solver is run and compared against flux-limited diffusion for a set of standard problems in section~\ref{sec:results}.
+We finish with our conclusion in section~\ref{sec:conclusion}.

### 2018-02-07 latex/egsr17-cgf/main-body.tex
continued working on draft

--- 2018-01-23 latex/egsr17-cgf/main-body.tex
@@
 \begin{document}
-
 % uncomment for using teaser
@@
 \missingfigure{}
-\caption{$CDA/P_1$}
+\vspace{-0.17in}
+\caption{Classical diffusion}
 \end{subfigure}%
@@
 \missingfigure{}
-\caption{FLD}
+\vspace{-0.17in}
+\caption{Flux-limited diffusion}
 \end{subfigure}%
@@
 \missingfigure{}
-\caption{$P_5$}
+\vspace{-0.17in}
+\caption{ours}
+\label{fig:nebulae_ours}
 \end{subfigure}%
@@
 \missingfigure{}
-\caption{Groundtruth}
+\vspace{-0.17in}
+\caption{Ground truth}
 \end{subfigure}%
 %\includegraphics[width=\linewidth]{figures/eg_new}
+\vspace{-0.1in}
 \icaption
@@
 We further present a staggered-grid $P_N$-solver and compare it against FLD for various standard problems.
-Rendering highly scattering participating media is a common challenge for brute-force path-tracing.
-Classical diffusion approximation (CDA) is a competitive alternative, as it reduces the problem to solving a simple diffusion equation.
-Variable eddington factor methods, such as flux-limited diffusion (FLD), improve CDA by making the diffusion coefficient non-linear.
+Rendering highly scattering participating media is a common challenge for brute-force path tracing.
+Classical diffusion approximation is a competitive alternative, as it reduces the problem to solving a simple diffusion equation.
+Extensions, such as flux-limited diffusion, improve the former by switching to a non-linear diffusion coefficient.
 Both methods are based on the spherical harmonics expansion of the radiance field, truncated after the first order.
-In this paper, we investigate the open question, if going higher order provides a viable alternative to the ad-hoc fixes to diffusion by FLD.
-Increasing the order introduces a growing set of increasingly complex coupled partial differential equations, called $P_N$-equations, which become very difficult to work with for higher order.
-We introduce a computer algebra framework for representing and manipulating mathematical equations and use it to derive the time-independent real-valued $P_N$-equations at arbitrary order.
+In this paper, we investigate the open question of wether higher orders provide a viable alternative to these two approaches.
+Increasing the order introduces a set of increasingly complex coupled partial differential equations, whose growing number and complexity make them very difficult to work with for higher order.
+We introduce a computer algebra framework for representing and manipulating the underlying mathematical equations and use it to derive the time-independent real-valued $P_N$-equations for arbitrary orders.
 We further present a staggered-grid $P_N$-solver and generate its stencil code directly from the expression tree of the $P_N$-equations.
-Both, the computer algebra framework, as well as the solver are made avaiable open source.
-Finally, we discuss how our method compares against FLD for various standard problems.
-
+Finally, we discuss how our method compares against prior work for various standard problems.
+We will release our computer algebra system, solver, and data as open source to ensure full reproducibility of all of our results.
 %\begin{classification} % according to http://www.acm.org/class/1998/
@@
 %-------------------------------------------------------------------------
-\section{Previous work}
 
-Light transport in participating media is governed by the radiative transfer equation~(RTE), which was first introduced in astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}.
-Today, this equation is typically solved using Monte Carlo based methods.
-However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become unfeasible, due to the high computational cost which comes from tracing long particle paths with many scattering events.
-
-In contrast to path-tracing, the $P_N$-method is a deterministic method, which gives a solution by solving a system of linear equations.
-It is derived by discretizing the angular variable of the radiative transfer equation into spherical harmonics (SH).
-This gives rise to a set of coupled, complex-valued partial differential equations, called $P_N$-equations.
-The subscript $N$ refers to the truncation order of the SH.
-
-The $P_N$-method has a long history in other fields and was never applied in graphics.
-Kajia~\cite{Kajiya84} explained the theory, but did not give any details on implementation or how to solve it.
-In fact, as Max~\cite{Max95} pointed out, it is not clear if Kajiya succeeded at all at applying the method, as all of the results in the paper were produced with a simpler method.
-This is further strengthened by our observation, that a straight forward finite difference discretization of the $P_N$-equations produces unuseable results, due to oscillation artefacts in the solution.
-In our paper, we derive the real-valued $P_N$-equations and present a staggered-grid solver, which produces an artefact free solution.
-
-Similar to $P_N$, the classical diffusion approximation (CDA) is a deterministic method, which arrives at a solution by solving a system of linear equations.
-It is derived as a degenerate case of the $P_1$-equations (truncation after the first SH order), which collapses to a simple diffusion equation giving the method its name.
-CDA has a long history in other domains, such as astrophysics~\cite{Ishimaru78} and was introduced to graphics by Stam \cite{Stam95}.
-It was further extended to anisotropic radiative transfer by \cite{Jakob10}.
-
-CDA suffers from severe energy loss close to regions with strong density gradients.
-This problem is addressed by the variable eddington factor(VEF)-formalism, among which flux-limited diffusion (FLD) is the most prominent.
-FLD was first introduced in astrophysics by~\cite{Levermore81} and was introduced to graphics by Koerner et al.~\cite{Koerner14}.
-
-The idea behind VEF is to replace the diffusion coefficient in the diffusion equation of CDA by a non-linear diffusion coefficient.
-This new diffusion coefficient is being derived by seeing it as a mean to spatially blend between the diffusive and pure transport regime.
-The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
-In the absence of scattering (e.g.
-pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.
-
-It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion.
-This question is also been raised in other domains~\cite{Olson00} and resolving it is the main motivation for this paper.
-
-%\begin{itemize}
-  %\item Alternative deterministic methods
-  %\begin{itemize}
-  %  \item Heuristics \cite{Kaplanyan10} \cite{Elek14}
-  %  \item Lattice Boltzmann Methods \cite{Geist04}
-  %  \item Discrete Ordinates \cite{Languenou95}
-  %\end{itemize}
-  %However, among all the deterministic methods, the diffusion approximation has been the most popular due to its intuition and simplicity.
-
-%\end{itemize}
+\input{previous_work.tex}
 
 %-------------------------------------------------------------------------
-\section{Real-valued $P_N$-Equations}
-\label{sec:pnequations}
+\input{pn_equations.tex}
 
-In this section we present how to optain the real-valued $P_N$-equations.
-For brevity and readability, we only give a brief outline here and refer to the supplemental material for the detailed derivation.
+%-------------------------------------------------------------------------
+\input{solver.tex}
 
-The $P_N$-equations are derived from the radiative transfer equation, which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
-\begin{align}
-%\label{eq:rte}
-\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
-=&
--\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\nonumber\\
-&
-+\sigma_s\left(\vec{x}\right) \int_{\Omega}
-{
-p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega \right)\ud\omega'
-}\nonumber\\
-&
-+Q\left(\vec{x}, \omega\right)\nonumber
-\end{align}
+%-------------------------------------------------------------------------
+\input{rendering.tex}
 
-The LHS is the transport term and we refer to the terms on the RHS as collision, scattering and source term respectively.
-The symbols $\sigma_t$, $\sigma_s$, $p$ and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term respectively.
+%-------------------------------------------------------------------------
+\input{results.tex}
 
-The $P_N$-method uses SH to discretize the angular domain into a set of coupled equations.
-We further discretize the spatial variable using a finite difference (FD) grid.
-Each term represents a change of the radiance field $L$.
-Therefore, after discretizing the continuous variables, we can represent the RTE as a sequence of linear transformations, which can be collapsed into a single coefficient matrix $A$ and a RHS vector $\vec{Q}$:
-\begin{align}
-(T+C-S)\vec{u} = A\vec{u} = \vec{Q}
-\end{align}
-
-$T$, $C$, $S$ are matrices, which result from the discretized transport, collision and scattering terms in the $P_N$-equations respectively.
-All SH coefficients for every voxel of our FD-grid are flattened into the solution vector $\vec{u}$, which is the unknown we want to solve for.
-The RHS vector likewise contains the flattened SH coefficients of the emission field $Q$ of every voxel.
-
-\begin{figure}[h]
-\centering
-\missingfigure{test}
-\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
-\end{figure}
-
-In order to derive the real-valued $P_N$-equations from which we can construct $A$ and $\vec{Q}$, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
-\begin{align}
-\SHBR^{l,m}=
-\left\{
-\begin{array}{lr}
-\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
-\SHBC^{l,m}, & \text{for } m = 0\\
-\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}-\left(-1\right)^m\SHBC^{l,m}\right), & \text{for } m > 0
-\end{array}
-\right.
-\end{align}
-
-First, the angular dependent quantities in the RTE, such as the radiance field $L$ (or the emission field $Q$) are projected into SH coefficients:
-\begin{align}
-L^{l,m}\left(\vec{x}\right)
-=
-\int_\Omega{L\left(\vec{x}, \omega\right)\SHBR^{l,m}\ud\omega}
-\end{align}
-
-The reconstruction $\hat{L}$ has an error due to truncation at order $N$:
-\begin{align}
-L\left(\vec{x}, \omega\right)
-\approx
-\hat{L}\left(\vec{x}, \omega\right) =
-\sum_{l=0}^{N}
-{
-\sum_{m=-l}^{l}
-{
-L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
-}
-}
-=
-\sum_{l,m}
-{
-L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
-}
-\end{align}
-
-The angular dependent quantities are replaced with their SH reconstruction in the RTE.
-The $P_N$-equations are found by multiplying each term with the conjugate complex of the real-valued SH basis function, followed by integration over solid angle.
-In order to arrive at a form, which does not depend on angle anymore, a set of transformations and identities are applied.
-Of particular importance are the orthogonality property
-\begin{align}
-\int_{\Omega}{ \SHBC^{l,m}\overline{\SHBC^{l',m'}}\ud\omega} = \delta_{l,l'}\delta_{m,m'}
-\end{align}
-
-and recursion relations such as the following:
-\begin{align}
-\omega_z\SHBC^{l, m}=
-&
-a^{{l-1,m}}\overline{Y^{l-1, m}}+b^{{l+1,m}}\overline{Y^{l+1, m}}
-\end{align}
-
-with 
-\begin{align}
-&a^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}},
-b^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
-\end{align}
-
-%Finally, the spatially dependent variables are discretized (using finite difference stencils for derivatives).
-The resulting equations are factorized according to the unknowns ($l,m$-coefficient within voxel $i,j,k$).
-The coefficients to these unknowns
-
-Note that the real-valued SH basis functions are different, depending on $m$, which means that we get three different versions of the $P_N$-equations (for $m<0$, $m=0$ and $m>0$).
-Further note, that we need to expand the real-valued SH basis functions into their complex-valued definition, in order to be able to apply the given identities above.
-This causes the equations to become large and difficult to deal with when applying transformations and identities.
-
-It becomes even more of a problem later, when we want to expand the spatially discretized equations for a particular choice of truncation order $N$.
-Manually creating a FD stencil from so many terms is challenging for first order and becomes increasingly difficult and error prone for higher order.
-
-\subsection*{Computer Algebra Representation}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-
-\icaption{The computer algebra representation is constructed using API commands provided by our framework (top left) and represented internally as an expression tree (top right).
-Additional API commands (bottem left) allow valid transformations, which modify the expression tree accordingly (bottom right).}
-\end{figure}
-
-In order to tackle this problem, we created a system for representing algebraic expressions as an expression tree.
-Our framework can represent terms, integrals, derivatives, sums, variables, functions and special symbols, such as the Kronecker delta or imaginary number.
-It supports manipulations, such as the application of distributive law, substitution, rearranging nested integrals and application of the identities given above.
-
-After entering each term of the RTE into our framework, we arrive at the $P_N$-equations by applying a set of manipulations and identities using API commands provided by our framework.
-For debugging purposes, the expression tree can be rendered directly into latex code.
-This allows us to deal with pages of dense equations very easily.
-
-The computer algebra representation of the $P_N$-equations is input to our $P_N$-solver, which we will discuss in the next section.
-
-\section{$P_N$-Solver}
-\label{sec:pnsolver}
-
-\begin{figure*}[t]
-\centering
-\missingfigure{$P_N$-solver overview: generate stencil code $\rightarrow$ build system $\rightarrow$ solve $\rightarrow$ render}
-\icaption{Overview of our $P_N$-solver.
-After generating the stencil source code from the expression trees representing the $P_N$-equations, the linear system $A\vec{u}=\vec{Q}$ is built using RTE parameter fields and additional user input, such as grid resolution and type of boundary conditions.
-The resulting system is solved for $\vec{u}$, which is then used in our rendering application.}
-\label{fig:pnsolver}
-\end{figure*}
-
-In this section we discuss the implementation of our $P_N$-solver.
-The inputs are the $P_N$-equations in form of our computer algebra representation from section~\ref{sec:pnequations} and the scalar fields for $\sigma_t$ and $\sigma_s$.
-The angular dependent emission field $Q\left(\vec{x}, \omega\right)$ is projected into SH, which gives a scalar field for each SH coefficient as another input.
-We assume the phase function is not spatially varying and we therefore only require a single set of coefficients $p^{l,m}$ as input.
-
-Solving for the unknown vector $\vec{u}$, which gives the SH coefficients of the radiance function at every voxel of our FD grid, requires building the coefficient matrix $A$ and RHS $\vec{Q}$.
-We seperate this build process into two steps.
-The first step does not need any RTE parameters and is also independent of spatial grid resolution.
-This is the stencil generation and it can be done once for every truncation order $N$ in advance.
-
-The second step uses the precomputed stencil to compute the coefficients of $A$ and $\vec{Q}$ from the RTE-parameter fields and a given resolution for the FD-grid.
-Then $\vec{u}$ is found by solving $A\vec{u}=\vec{Q}$, using a standard linear system solver.
-
-\subsection{Stencil Generation}
-
-The stencil generation builds directly on top of our computer algebra framework for representing mathematical expressions.
-After specifying a truncation order $N$, the sums in the equations are expanded into individual terms.
-These are then further factorized according to the unknowns.
-
-Finally, the equations are discretized in spatial domain.
-Since all manipulations are done by recursively parsing the expression tree and replacing subtrees as needed, discretization can be handled elegantly by our system.
-The $P_N$-equations are discretized in stencil space (not in world space).
-We start parsing the expression at the origin of the stencil (meaning $\vec{x}=(0,0,0)$).
-
-Whenever we encounter a derivative operator, we parse its contained sub-expression multiple times with shifted voxel offsets according to the finite difference direction.
-The resulting expressions are multiplied with an expression representing the finite difference weight (which will contain the voxelsize as a variable).
-The parser keeps track of the current voxel offset in a stack.
-Nested derivatives are handled naturally and produce second order finite difference stencils as expected.
-All occurances of the continuous positional variable $\vec{x}$ are replaced by the current voxel offset in the stack.
-
-As we see in figure~\ref{fig:artefacts}, our solver needs to support staggered grids in order to avoid severe oscillation artefacts which arise, when all coefficients are located at the voxel center.
-With staggered grids, coefficients can live on the intersections of a subgrid with double resolution.
-So instead of voxel centers alone, coefficients can live on the center of voxel faces, voxel edges or the intersection between voxels.
-
-Our solver supports placement of coefficients on arbitrary staggered grid locations (including collocation).
-Whenever an unknown is encountered during expression tree parsing, the discretization code looks at the staggerd grid location at which the unknown is to be evaluated and the location at which the unknown lives on the grid.
-Depending on how the two are located to each other, the parser returns an expression, which interpolates the unknown at the requested offset from its surrounding locations.
-It returns the unknown itself, if both happen to coincide at the same location.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-\icaption{The solution of our solver for the 2D-checkerboard problem with collocated grids (left), where all coefficients are located at the cell centers.
-It suffers from severe oscillating artefacts, which are removed by offseting the coefficient locations using staggered grids (right).}
-\label{fig:artefacts}
-\end{figure}
-
-After discretization, the final expression tree is used to generate C++ code for the stencil.
-The unknowns in the terms give information about the coefficient index and voxel offset and therefore identify a column offset in the matrix $A$.
-The other factors in each term are rendered into C++ expressions, which are executed during runtime to set the coefficients in matrix $A$ and $\vec{Q}$.
-The stencil is executed per voxel and contains the code for computing the cofficients for all rows in $A$ and $\vec{Q}$ which belong to a single voxel.
-
-\subsection{System Building and Solving}
-
-The matrix $A$ and RHS $Q$ are built and solved using a C++ framework, which is being compiled with the generated stencil code.
-The build process consists of iterating over all voxels and executing the stencil.
-The generated stencil code will evaluate RTE parameter fields and access coefficients in $A$ and $\vec{Q}$.
-
-Since the stencil has been generated by discretizing the $P_N$-equation relative to the stencil origin, we can apply it to every voxel by simply transforming all queries from stencil space to world space and similarily shift all write operations into $A$ and $\vec{Q}$ according to the current voxel for which the stencil is being executed.
-With the offset information from the stencil and the current voxel, absolute row and column locations in $A$ are determined.
-
-The stencil has been generated without any notion of domain boundaries.
-Our solver supports Neumann boundary conditions (BC) and Dirichlet BC.
-They are handled transparently by the C++ framework through reindexing.
-Whenever the stencil writes into $A$ for a SH coefficient outside the domain, the framework will return the row and column in $A$ for the SH coefficient with the same index, which is the closest inside the domain (Neumann BC) or it will simply ignore the write operation (Dirichlet BC).
-
-Additional active coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are required to respect the boundary correctly (see Figure~\ref{fig:staggeredgrid}).
-This requires careful managegment of additional coefficients at boundary voxels, which is done transparently by the C++ framework.
-
-%\todo[inline]{The original todo note withouth changed colours.\newline Here's another line.}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
-\label{fig:staggeredgrid}
-\end{figure}
-
-%Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.
-The generated linear system becomes very large for high order and resolution.
-The number of rows is determined by the number of voxels times the number of SH coefficients per voxel.
-The matrix $A$ is squared and fortunately very sparse.
-It further is non-symmetric and also not diagonal dominant, which rules out many standard methods for solving linear systems.
-We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead.
-This gives a symmetric and positive definit system matrix $A^TA$ (although with a higher condition number).
-
-%\subsection*{CDA vs. $P_1$}
-%\begin{itemize}
-  %\item CDA is a degenerated form of $P_1$.
-It is derived by isolating the flux-vector on one side of the vector-equation formed by the $l=1$ SH-band equations.
-This isolation requires division by the extinction coefficient, introducing a $\frac{1}{\sigma_t}$ factor which diverges as $\sigma_t$ approaches zero.
-Thresholding to some minimum is required, in order to be able to solve the system for vacuum regions.
-  %\item $P_1$ does not require any thresholding of $\sigma_t$ as it does not contain $\sigma_t$ as a denominator. It therefore can deal with vacuum regions without modifications.
-  %\item (needs validation) Further, in the presence of vacuum or near vacuum regions, the condition number for CDA is higher than for $P_1$, because of small extinction values in the denominator of the diffusion coefficient.
-  %\item Using the normal form for CDA will further increase the condition number when vacuum regions are present and significantly decreases convergence.
-%\end{itemize}
-
-\section{Flux-limited Diffusion}
-\label{sec:vef}
-
-Our FLD solver follows the implementation in~\cite{Koerner14} very closely (though our solver runs on CPU).
-
-\section{Results}
-\label{sec:results}
-
-\subsection{Rendering}
-
-\begin{itemize}
-	\item explain that we use single scattered light as emission field
-	\item explain our single scattering path tracer using the $P_N$-solution as indirect light
-\end{itemize}
-
-\subsection{2D checkerboard}
-\subsection{Pointsource problem}
-\subsection{Nebulae/Cloud}
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-%\begin{itemize}
-  %\item Extension to surface rendering by dealing vacuum regions and extending to anisotropic RTE.
-  %\item Apply our automated stencil generation to other problem domains where coupled PDE's are present.
-%\end{itemize}
+%-------------------------------------------------------------------------
+\input{conclusion.tex}
 

### 2018-02-07 latex/egsr17-cgf/pn_equations.tex
continued working on draft

\section{Real-valued $P_N$-Equations}
\label{sec:pnequations}

In this section we present how to optain the real-valued $P_N$-equations. For brevity and readability, we only give a brief outline here and refer to the supplemental material for the detailed derivation.

The $P_N$-equations are derived from the radiative transfer equation, which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
\begin{align}
%\label{eq:rte}
\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
=&
-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\nonumber\\
&
+\sigma_s\left(\vec{x}\right) \int_{\Omega}
{
p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega \right)\ud\omega'
}\nonumber\\
&
+Q\left(\vec{x}, \omega\right)\nonumber
\end{align}

The left hand side (LHS) is the transport term and we refer to the terms on the right hand side (RHS) as collision, scattering and source term respectively. The symbols $\sigma_t$, $\sigma_s$, $p$ and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term.

The RTE is often given in operator notation, where transport, collision and scattering are expressed as operators $\mathcal{T}$, $\mathcal{C}$ and $\mathcal{S}$, which are applied to the radiance field $L$:
\begin{align}
\mathcal{T}\left(L\right) = -\mathcal{C}\left(L\right) + \mathcal{S}\left(L\right) + Q
\end{align}

The $P_N$-method expands the angular dependence into spherical harmonics. Since the radiance field $L$ is real only, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
\begin{align*}
\SHBR^{l,m}=
\left\{
\begin{array}{lr}
\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
\SHBC^{l,m}, & \text{for } m = 0\\
\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}-\left(-1\right)^m\SHBC^{l,m}\right), & \text{for } m > 0
\end{array}
\right.
\end{align*}

We express the projection into the SH basis functions with the projection operator $\mathcal{P}$:
\begin{align}
\mathcal{P}^{l, m}(f) = \int_{\Omega}f(\mathbf{x}, \omega) \SHBR^{l,m}(\omega)\,\mathrm{d}\omega = f^{l,m}
\nonumber
\end{align}

The $P_N$-equations are derived by first replacing the radiance field $L$ by its SH reconstruction $\widehat{L}$, which introduces an error due to truncation at order $N$:
\begin{align*}
\widehat{L}\left(\vec{x}, \omega\right) =
\sum_{l=0}^{N}
{
\sum_{m=-l}^{l}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
}
}
\approx
L\left(\vec{x}, \omega\right)
\end{align*}

After similarly replacing all angular dependent RTE parameters, we project each term into SH, using the projection operator $\mathcal{P}$. This produces a single equation for each $l,m$-pair. The $P_N$-equations therefore can be written as:
\begin{align}
\mathcal{P}^{l,m}\mathcal{T}\left(\widehat{L}\right)
=
-\mathcal{P}^{l,m}\mathcal{C}\left(\widehat{L}\right) 
+\mathcal{P}^{l,m}\mathcal{S}\left(\widehat{L}\right)
+\mathcal{P}^{l,m}\left(Q\right)
\nonumber
\end{align}

Expanding the operators and writing the real-valued $P_N$-equations in compact form, requires a string of expansions, manipulations and applications of identities. We here give the final result and refer to the supplemental material for the detailed derivation. Since the real-valued SH basis have different definitions, depending on $m<0$, $m=0$ or $m>0$, we get different projections $\mathcal{S}^{l,m}$ depending no the sign of $m$.

The real-valued $P_N$-equations are for $m<0$:
\begin{align}
&-\frac{1}{2}c^{\scaleto{l-1,m-1}{4pt}}
\partial_y
L^{\scaleto{l-1,-m+1}{4pt}}
%\\
+\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
\partial_y
L^{\scaleto{l+1,-m+1}{4pt}}
%\\
-\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m+1}{4pt}}
\partial_y
L^{\scaleto{l-1,-m-1}{4pt}}
\nonumber
\\&
+\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{\scaleto{l+1,m+1}{4pt}}
\partial_y
L^{\scaleto{l+1,-m-1}{4pt}}
%\\
+\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}c^{\scaleto{l-1,m-1}{4pt}}
\partial_x
L^{\scaleto{l-1,m-1}{4pt}}
\nonumber
\\&
-\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}e^{{l-1,m+1}}
\partial_x
L^{\scaleto{l-1,m+1}{4pt}}
%\\
+\frac{1}{2}f^{\scaleto{l+1,m+1}{4pt}}
\partial_x
L^{\scaleto{l+1,m+1}{4pt}}
%\\
-\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
\partial_x
L^{\scaleto{l+1,m-1}{4pt}}
\nonumber
\\&
+a^{\scaleto{l-1,m}{4pt}}
\partial_z
L^{\scaleto{l-1,m}{4pt}}
%\\
+b^{\scaleto{l+1,m}{4pt}}
\partial_z
L^{\scaleto{l+1,m}{4pt}}
%\\
+\sigma_t L^{\scaleto{l,m}{4pt}}
%\\
-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
%\\
= Q^{\scaleto{l,m}{4pt}}
\label{eq:rpn_m_<_z}
\end{align}

for $m=0$:
\begin{align}
&
\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_x L^{\scaleto{l-1,1}{4pt}}
-\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_x L^{\scaleto{l+1,1}{4pt}}
%\\&
\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_y L^{\scaleto{l-1,-1}{4pt}}
\nonumber
\\&
-\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_y L^{\scaleto{l+1,-1}{4pt}}
%\\&
a^{\scaleto{l-1,0}{4pt}}\partial_z L^{\scaleto{l-1,0}{4pt}}
+b^{\scaleto{l+1,0}{4pt}}\partial_z L^{\scaleto{l+1,0}{4pt}}
\nonumber
\\&
+\sigma_t L^{\scaleto{l,m}{4pt}}
-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
= Q^{\scaleto{l,m}{4pt}}
\label{eq:rpn_m_=_z}
\end{align}

and for $m>0$:
\begin{align}
&
\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_x L^{\scaleto{l-1,m+1}{4pt}}
%\\&
-\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_x L^{\scaleto{l+1,m+1}{4pt}}
%\\&
-\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m-1}{4pt}}\partial_x L^{\scaleto{l-1,m-1}{4pt}}
\nonumber
\\&
+\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{l+1,-m+1}\partial_x L^{\scaleto{l+1,m-1}{4pt}}
%\\&
+\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_y L^{\scaleto{l-1,-m-1}{4pt}}
\nonumber
\\&
-\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_y L^{\scaleto{l+1,-m-1}{4pt}}
%\nonumber
%\\&
+\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}e^{\scaleto{l-1,-m+1}{4pt}}\partial_y L^{\scaleto{l-1,-m+1}{4pt}}
\nonumber
\\&
-\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}f^{\scaleto{l+1,-m+1}{4pt}}\partial_y L^{\scaleto{l+1,-m+1}{4pt}}
%\\&
+a^{\scaleto{l-1,-m}{4pt}}\partial_z L^{\scaleto{l-1,m}{4pt}}
%\nonumber
%\\&
+b^{\scaleto{l+1,-m}{4pt}}\partial_z L^{\scaleto{l+1,m}{4pt}}
\nonumber
\\&
+\sigma_t L^{\scaleto{l,m}{4pt}}
-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
= Q^{\scaleto{l,m}{4pt}}
%\nonumber
\label{eq:rpn_m_>_z}
\end{align}

with
\begin{align*}
\label{eq:real_sh_basis}
\beta^{x}=
\left\{
\begin{array}{lr}
\frac{2}{\sqrt{2}}, & \text{for } \vert x\vert = 1\\
1, & \text{for } \vert x\vert \neq 1
\end{array}
\right.
,\quad
\delta_{x\neq y}=
\left\{
\begin{array}{lr}
1, & \text{for } x \neq y \\
0, & \text{for } x = y
\end{array}
\right.
\end{align*}

and
\begin{align*}
&
a^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}} \qquad
b^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
\\&
c^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \qquad
d^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
\\&
e^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \qquad
f^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
\end{align*}
\begin{align*}
\lambda_l=\sqrt{\frac{4\pi}{2l+1}}
\end{align*}

\begin{figure}[h]
\centering
\missingfigure{test}
\vspace{-0.2in}
\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
\end{figure}

\begin{figure*}[t]
\centering
\missingfigure{$P_N$-solver overview: generate stencil code $\rightarrow$ build system $\rightarrow$ solve $\rightarrow$ render}
\vspace{-0.2in}
\icaption{Overview of our $P_N$-solver. After generating the stencil source code from the expression trees representing the $P_N$-equations, the linear system $A\vec{u}=\vec{Q}$ is built using RTE parameter fields and additional user input, such as grid resolution and type of boundary conditions. The resulting system is solved for $\vec{u}$, which is then used in our rendering application.}
\label{fig:pnsolver}
\end{figure*}

The spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxelgrid. The radiance field $L$, after full discretization, is represented as a set of coefficients per voxel. Flattening these over all voxels into a single vector gives the solution vector $\vec{u}$. The RHS vector $\vec{Q}$ is produced similarly. The differential operators are discretized using central differences. This allows the projected operators to be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$:
\begin{align}
(T+C-S)\vec{u} = A\vec{u} = \vec{Q}
\end{align}

$T$, $C$, $S$ are matrices, which result from the discretized transport, collision and scattering operators in the $P_N$-equations respectively.

In the next section, we will present our solver for solving the real-valued $P_N$-equations.

### 2018-02-07 latex/egsr17-cgf/previous_work.tex
continued working on draft

\section{Previous work}

Light transport in participating media is governed by the radiative transfer equation~(RTE), first studied in the context of astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}. Today, this equation is typically solved using Monte Carlo based methods. However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become prohibitive, due to the high computational cost, which comes from tracing long particle paths with many scattering events.

In contrast to path-tracing, the $P_N$-method is a deterministic method, which gives a solution by solving a system of linear equations. It is derived by discretizing the angular variable of the radiative transfer equation into spherical harmonics (SH). This gives rise to a set of coupled, complex-valued partial differential equations, called $P_N$-equations. The subscript $N$ refers to the truncation order of the SH.

The $P_N$-method has a long history in other fields and was never applied in graphics. Kajia~\cite{Kajiya84} explained the theory, but did not give any details on implementation or how to solve it. In fact, as Max~\cite{Max95} pointed out, it is not clear if Kajiya succeeded at all at applying the method, as all of the results in the paper were produced with a simpler method. This is further strengthened by our observation, that a straight forward finite difference discretization of the $P_N$-equations produces unuseable results, due to oscillation artifacts in the solution. In our paper, we derive the real-valued $P_N$-equations along with a staggered-grid solver, that produces artifact free solutions.

Similar to $P_N$, the classical diffusion approximation (CDA) is a deterministic method, which arrives at a solution by solving a system of linear equations. It is derived as a degenerate case of the $P_1$-equations (truncation after the first SH order), which collapses to a simple diffusion equation giving the method its name. CDA has a long history in other domains, such as astrophysics~\cite{Ishimaru78} and was introduced to graphics by Stam \cite{Stam95}. It was further extended to anisotropic radiative transfer by \cite{Jakob10}.

CDA suffers from severe energy loss close to regions with strong density gradients. The problem is addressed by a modification known as Variable Eddington factor (VEF), which nonlinearly adjusts the diffusion coefficient to improve the solution near density gradients and low-density regions. Flux-limited diffusion, developed in the context of astrophysics by Levermore et al.~\cite{Levermore81} and later introduced to graphics by Koerner et al.~\cite{Koerner14}, is the most prominent example.

The idea behind VEF is to replace the diffusion coefficient in the diffusion equation of CDA by a non-linear diffusion coefficient. This new diffusion coefficient is being derived by seeing it as a mean to spatially blend between the diffusive and pure transport regime. The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines. In the absence of scattering (e.g. pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.

It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion. This question is also been raised in other domains~\cite{Olson00} and resolving it is the main motivation for this paper.

%\begin{itemize}
  %\item Alternative deterministic methods
  %\begin{itemize}
  %  \item Heuristics \cite{Kaplanyan10} \cite{Elek14}
  %  \item Lattice Boltzmann Methods \cite{Geist04}
  %  \item Discrete Ordinates \cite{Languenou95}
  %\end{itemize}
  %However, among all the deterministic methods, the diffusion approximation has been the most popular due to its intuition and simplicity.

%\end{itemize}

### 2018-02-07 latex/egsr17-cgf/rendering.tex
continued working on draft

\section{Rendering}
\label{sec:rendering}

In order to evaluate our results, we chose to use our render the solution directly. We use an approach similar to Koerner et al.~\cite{Koerner14}, where we seperate the radiance field into single scattered light $L_{ss}$ and multiple scattered light $L_{ms}$:
\begin{align}
L\left(\vec{x},\omega\right) = L_{ss}\left(\vec{x},\omega\right) + L_{ms}\left(\vec{x},\omega\right)
\end{align}

The single scattered light is folded into the emission term $Q$:
\begin{align}
Q(\vec{x}, \omega) = L_{ss}(\vec{x}, \omega) = \sigma_s\left(\vec{x}\right)\int_\Omega{ p\left(\omega'\rightarrow\omega\right) L_{u}\left(\vec{x}, \omega'\right)\ud\omega' }
\end{align}

This means that, our solver will solve for the multiple scattered light $L_{ms}$. The quantity $L_u$ is the uncollided light, which was emitted from the lightsource and attenuated by the volume without undergoing any scattering event. It is easily computed for delta light sources, such as directional lights, in homogeneous volumes. Raymarching can be used in heterogeneous volumes. For more complex light sources, the emission field can be precomputed with a number of Monte Carlo iterations.

Running the solver gives solution vector $\vec{u}$. The staggering is removed by interpolating all coefficients to voxel centers. Also the additional coefficients at boundary voxels are removed. This operation is represented as a matrix and results in a simple three-dimensional voxelgrid with SH coefficients for order $N$ at the center of each voxel.

For rendering, we use a simple forward path tracing approach, where we start tracing from the camera. At the first scattering event, we use next event estimation to account for $L_{ss}$. Then we sample a new direction according to the phase function. Instead of continuing tracing into the new direction, we simply evaluate the inscattering integral using $\widehat{L}_{ms}$. The SH coefficients at $\vec{x}$ are found by trilinear interpolation from the voxelgrid of SH coefficients.

### 2018-02-07 latex/egsr17-cgf/results.tex
continued working on draft

\section{Results}
\label{sec:results}

In this section, we present results for a set of standard problems in order to validate and evaluate our method. We also compare against classical diffusion (CDA) and flux-limited diffusion (FLD).

Classical diffusion is based on the diffusion equation, which is derived by collapsing the $P_1$-equations:
\begin{align} 
\nabla\left(\frac{1}{3\sigma_t}\nabla L^{0,0}\right)  = -Q^{0,0}
\label{eq:cda}
\end{align}

Since our solver can work with any PDE, which results in a linear system of equations, we only needed to put equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which would generate the correct stencil code automatically.

Since FLD is based on a non-linear diffusion equation, we were not able to use our system in the same way. Our implementation closely follows the implementation in~\cite{Koerner14} (though ours runs on CPU) and we refer to their paper for more details.

\subsection{2D checkerboard}

First we run our solver on the 2D checkerboard, a very common test case in other fields. This allows us to validate our solver against the solver from Seibold et al.~\cite{Seibold14}.

Their solver solves for the time-dependent and complex-valued $P_N$-equations in the 2D case, which is derived by assuming that all SH coefficients, RTE parameters and boundary conditions are z-independent. This causes all SH coefficients and moment equations for which $l+m$ is odd to vanish. Due to the time-dependency, their approach is to do explicit incremental steps in time. We run their solver for many timesteps, in order to get a result which is close to steady state.

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\vspace{-0.2in}
\icaption{Comparison of the result for the checkerboard test using Starmap~\cite{Seibold14} (left) against ours (right). Our result is in excellent agreement with the reference.}
\label{fig:vs_starmap}
\end{figure}

As can be seen in figure~\ref{fig:vs_starmap}, the results from our solver are in excellent agreement with the results from Seibold et al.~\cite{Seibold14} and verify the correctness of our implementation.

\subsection{Pointsource problem}

We also run our solver for the point source problem, a single point light in a homogeneous medium. This not only helps to validate our implementation for the 3D case against an analytical ground truth, but also allows to get a clear assessment of the methods accuracy.

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\vspace{-0.2in}
\icaption{Lineplot through the 3D solution of our solver for the point source problem for various order $N$ (left). Solution for $P_5$ compared against classical diffusion and flux-limited diffusion (right).}
\label{fig:pointsource}
\end{figure}

For our test case we choose a FD resolution of $80\times80\times80$, an extinction coefficient $\sigma_t=8.0$ and albedo $\alpha=0.9$. We run the solver for different truncation values $N$ and in figure~\ref{fig:pointsource}, we see that the solution becomes increasingly accurate for higher truncation order. The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even. We further see that the $P_1$ solution exactly matches the results from CDA, which makes sense, as the latter is only a collapsed version of the former.

\begin{itemize}
	\item performance/computation time
\end{itemize}

\subsection{Procedural Clouds}

Finally, we run our solver on different procedural cloud datasets to get an idea of its performance in more practical applications.

Figure~\ref{fig:teaser}~\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function. We see that at order $N=5$, our method can capture indirect illumination similarily well as FLD and is significantly better than CDA as expected.

\begin{itemize}
	\item mention performance.
	\item run for various values of $N$
	\item convergence behaviour in vacuum. Clearify
	\item run for anisotropic problem
\end{itemize}

### 2018-02-07 latex/egsr17-cgf/solver.tex
continued working on draft

\section{$P_N$-Solver}
\label{sec:pnsolver}

Expanding, discretizing and working with the $P_N$-equations gets increasingly complex for higher order and is arduous and error prone if done by hand. We tackle this problem by expressing the RTE in a computer algebra representation and apply the discretization in angular and spatial domain using manipulation passes on the representation.

The discretized $P_N$-equations are then rendered into source code and compiled. We refer to this as the stencil code. It allows fast construction of the linear system $A\vec{u}=\vec{Q}$ for arbitrary resolution and RTE parameters and has no dependency on the computer algebra framework anymore.

Finally, the generated linear system is solved using standard methods. The solution $\vec{u}$ represents the discretized radiance field $\widehat{L}$, which is evaluated using using trilinear interpolation of the SH coefficients. An overview of the solver is given in figure~\ref{fig:pnsolver}. Each step will be described in more detail in the following subsections.

\subsection{Computer Algebra Representation}

Our computer algebra representation is based on mathematical expression trees, hierarchies composed out of expression instances. Those can be basic primitives, such as numbers, variables or special symbols, such as the Kronecker delta or imaginary number. Other expression types, such as integrals, derivatives, sums, products and functions build the hierarchy by referencing child expressions.

Our framework further provides manipulators, which can be executed on expressions. We implemented all manipulations, which were required for the derivation of the $P_N$-equations. This includes operations, such as application of the distributive law, substitution, constant folding, reordering of nested integrals, application of identities up to more complex operations, such as factorization of unknowns and discretization of spatial variables. The derivation steps of the real-valued $P_N$-equations in the supplemental material were all carried out using our framework.

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%

\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\vspace{-0.2in}
\icaption{The computer algebra representation is constructed using API commands provided by our framework (top left) and represented internally as an expression tree (top right). Additional API commands (bottem left) allow valid transformations, which modify the expression tree accordingly (bottom right).}
\end{figure}

Finally, frontends allow rendering the expression tree into different forms. We implemented frontends for rendering expression trees to \LaTeX and C++ source code. The equations in the supplemental material were almost all rendered by the former. The latter was used for generating the stencil code used by our solver.

Note that our framework is different from a computer algebra system (CAS), in that our framework only deals with representation and manipulation of expressions, and is not concerned with finding answers to mathematical problems.

\subsection{Discretization}

The result of the discretization in angular domain are the real-valued $P_N$-equations. We arrived at the consice form in equations~\ref{eq:rpn_m_<_z}-~\ref{eq:rpn_m_>_z} using our framework. However, once the compact form was found, we used it directly as input to our solver.

The discretization code iterates over all $l,m$-pairs up to order $N$ and chooses either equation~\ref{eq:rpn_m_<_z}, equation~\ref{eq:rpn_m_=_z} or equation~\ref{eq:rpn_m_>_z}, according to the sign of $m$. All occurances of $l$ and $m$ are replaced with their actual values.

The spatial discretization is done by parsing the expression tree of each equation from the root. All occurances of the continuous positional variable $\vec{x}$ are replaced by discrete voxel coordinates $i,j,k$, whose values are retrieved from the top of a stack. The stack is kept by the parser and is initialized with the voxel at $(0,0,0)$. We refer to the space of the discretized $P_N$-equations as stencil space.

Whenever the parser encounters a differential operator, the current voxel position is shifted and pushed to the stack. Then the child expression is parsed and multiplied with a weight value, followed by popping the stack. This is done for each element of a central difference stencil, which provides the offset and weight values (referencing the voxelsize variable). The dimension of the stencil is given by the derivation variable of the operator. Nested derivatives are handled naturally and produce higher order finite difference stencils as expected.

Another aspect to consider during discretization is the location of the variables and unknowns. As we will see later, our solver needs to support staggered grids, where the SH coefficients are not all located at the voxel center (collocated). Some also can be located at the faces, edges and intersections of voxels (staggered).

Our solver therefore supports placement of coefficients at arbitrary staggered grid locations. Whenever a coefficient is encountered during expression tree parsing for discretization, the parser checks how its location is related to the current location on the top of the stack. Depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at the requested offset from its surrounding locations, or it returns the coefficient itself, if it happens to coincide. This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center. The position stack of the parser is initialized with the staggered grid location defined by the coefficient, which is associated with the $l,m$-pair of the equation currently being discretized.

\subsection{Stencil Code Generation and System Building}

After applying the discretization step to the expression tree of the $P_N$-equations, it is used to generate the stencil code. In numerical analysis, a stencil is an arrangement of voxels and weights, that relate values at different locations to each other and form the basis for propagating rows in the system matrix $A$ and RHS vector $\vec{Q}$ with values. The name comes from the fact, that the geometric structure and weights of the configuration do not change, when applied to different voxels. The same is true for the $P_N$-equations and we use this fact to generate a single function, which propagates rows in $A$ and $\vec{Q}$ for a given voxel.

The $P_N$-equations express, how each coefficient of a voxel depends on other coefficients of the same or adjacent voxels. The unknowns in the terms give information about the coefficient index and voxel offset, and therefore identify a column offset in the matrix $A$. The factors to these coefficients may contain evaluations of RTE parameters, such as $\sigma_t$. Therefore, these factors can not be determined during stencil generation, but are rendered into code expressions, which are executed as part of the stencil function during runtime. Because the stencil code has been generated in stencil space relative to the voxel at $(0,0,0)$, we can run the same stencil code for every voxel, by simply applying an offset accordingly.

The $P_N$-equations have been discretized, and the stencil has been generated, without any notion of domain boundaries and boundary conditions (BC). Our solver supports Neumann BC and Dirichlet BC. They are handled transparently by the code which runs the stencil. Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or return the row and column in $A$ for the same coefficient of the closest voxel inside the domain (Neumann BC).

In order to respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels (see Figure~\ref{fig:staggeredgrid}). This requires careful managegment and bookkeeping of coefficient indices, which is done transparently by the code running the stencil.

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\vspace{-0.2in}
\icaption{The solution of our solver for the 2D-checkerboard problem with collocated grids (left), where all coefficients are located at the cell centers. It suffers from severe oscillating artefacts, which are removed by offseting the coefficient locations using staggered grids (right).}
\label{fig:artefacts}
\end{figure}

\subsection{Solving}

After running the stencil code for every voxel, the system matrix $A$
and RHS vector $\vec{Q}$ are propagated with values and can be used with standard methods for solving systems of linear equations. The number of rows is determined by the number of voxels times the number of coefficients per voxel and can therefore become very large for high resolution and high truncation order. The matrix $A$ is squared and fortunately very sparse, due to finite differences and the structure of the $P_N$-equations. Unfortunately, it also is non-symmetric (due to the transport term) and not diagonal dominant, which rules out many standard methods for solving linear systems. We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number.

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\vspace{-0.2in}
\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
\label{fig:staggeredgrid}
\end{figure}

%Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.

%\subsection*{CDA vs. $P_1$}
%\begin{itemize}
  %\item CDA is a degenerated form of $P_1$. It is derived by isolating the flux-vector on one side of the vector-equation formed by the $l=1$ SH-band equations. This isolation requires division by the extinction coefficient, introducing a $\frac{1}{\sigma_t}$ factor which diverges as $\sigma_t$ approaches zero. Thresholding to some minimum is required, in order to be able to solve the system for vacuum regions.
  %\item $P_1$ does not require any thresholding of $\sigma_t$ as it does not contain $\sigma_t$ as a denominator. It therefore can deal with vacuum regions without modifications.
  %\item (needs validation) Further, in the presence of vacuum or near vacuum regions, the condition number for CDA is higher than for $P_1$, because of small extinction values in the denominator of the diffusion coefficient.
  %\item Using the normal form for CDA will further increase the condition number when vacuum regions are present and significantly decreases convergence.
%\end{itemize}

### 2018-02-21 latex/egsr17-cgf/introduction.tex
started working on anisotropy test case

@@
 Simulating light transport in paticipating media is still a challenging problem for image synthesis in computer graphics.
-Monte Carlo based techniques have established as the standard approach, because they can produce unbiased results and are conceptionally easy to implement.
+Monte Carlo based techniques are established as the standard approach, because they can produce unbiased results and are conceptionally easy to implement.
 They become computationally demanding however, for participating media, which exhibits strong scattering or anisotropy.

### 2018-02-21 latex/egsr17-cgf/main-body.tex
started working on anisotropy test case

@@
 
+\author
+{
+Compiled on \today
+%David Koerner$^{1,2,3}$
+%\qquad Jan Nov\'ak$^{1}$
+%\qquad Peter Kutz$^{2}$
+%\qquad Ralf Habel$^{2}$
+%\qquad Wojciech Jarosz$^{1,4}$
+\\
+% For Computer Graphics Forum: Please use the abbreviation of your first name.
+%$^1$Disney Research
+%\qquad $^2$Walt Disney Animation Studios
+%\qquad $^3$University of Stuttgart
+%\qquad $^4$Dartmouth College
+}
+
 % ------------------------------------------------------------------------
@@
 \centering
-\begin{subfigure}{0.2\linewidth}
-\missingfigure{}
+\begin{subfigure}{0.24\linewidth}
+%\missingfigure{}
+\includegraphics[width=\linewidth]{images/nebulae_ms_cda.png}
 \vspace{-0.17in}
 \caption{Classical diffusion}
+\label{fig:teaser_cda}
 \end{subfigure}%
-\hspace{0.05\linewidth}
-\begin{subfigure}{0.2\linewidth}
-\missingfigure{}
+\hspace{0.002\linewidth}
+\begin{subfigure}{0.24\linewidth}
+\includegraphics[width=\linewidth]{images/nebulae_ms_p5.png}
+\vspace{-0.17in}
+\caption{$P_5$ (ours)}
+\label{fig:nebulae_ours}
+\label{fig:teaser_pn}
+\end{subfigure}%
+\hspace{0.002\linewidth}
+\begin{subfigure}{0.24\linewidth}
+%\missingfigure{}
+\includegraphics[width=\linewidth]{images/nebulae_ms_fld.png}
 \vspace{-0.17in}
 \caption{Flux-limited diffusion}
+\label{fig:teaser_fld}
 \end{subfigure}%
-\hspace{0.05\linewidth}
-\begin{subfigure}{0.2\linewidth}
-\missingfigure{}
-\vspace{-0.17in}
-\caption{ours}
-\label{fig:nebulae_ours}
-\end{subfigure}%
-\hspace{0.05\linewidth}
-\begin{subfigure}{0.2\linewidth}
-\missingfigure{}
+\hspace{0.002\linewidth}
+\begin{subfigure}{0.24\linewidth}
+%\missingfigure{}
+\includegraphics[width=\linewidth]{images/nebulae_ms_groundtruth.png}
 \vspace{-0.17in}
 \caption{Ground truth}
+\label{fig:teaser_gt}
 \end{subfigure}%
@@
 {
-Flux-limited diffusion (b) is an ad-hoc extension to the classical diffusion approximation (a).
-It can be considered the state of the art among deterministic methods for simulating light transport in highly scattering participating media, which is a common challenge for brute-force path tracing (d).
-We add the $P_N$-method (c) to the toolbox of deterministic methods in rendering and investigate its benefits and tradeoffs against flux-limited diffusion.
+Flux-limited diffusion (\subref{fig:teaser_fld}) is an ad-hoc extension to the classical diffusion approximation (\subref{fig:teaser_cda}).
+It can be considered the state of the art among deterministic methods for simulating light transport in highly scattering participating media, which is a common challenge for brute-force path tracing (\subref{fig:teaser_gt}).
+We add the $P_N$-method (\subref{fig:teaser_pn}) to the toolbox of deterministic methods in rendering and investigate its benefits and tradeoffs against flux-limited diffusion.
 }

### 2018-02-21 latex/egsr17-cgf/previous_work.tex
started working on anisotropy test case

@@
 Kajia~\cite{Kajiya84} explained the theory, but did not give any details on implementation or how to solve it.
-In fact, as Max~\cite{Max95} pointed out, it is not clear if Kajiya succeeded at all at applying the method, as all of the results in the paper were produced with a simpler method.
+In fact, as Max~\cite{Max95} pointed out, it is not clear if Kajiya succeeded at all at applying the method, as all of the results in his paper were produced with a simpler method.
 This is further strengthened by our observation, that a straight forward finite difference discretization of the $P_N$-equations produces unuseable results, due to oscillation artifacts in the solution.

### 2018-02-21 latex/egsr17-cgf/solver.tex
started working on anisotropy test case

@@
 Finally, the generated linear system is solved using standard methods.
-The solution $\vec{u}$ represents the discretized radiance field $\widehat{L}$, which is evaluated using using trilinear interpolation of the SH coefficients.
+The solution $\vec{u}$ represents the discretized radiance field $\widehat{L}$, which is evaluated using trilinear interpolation of the SH coefficients.
 An overview of the solver is given in figure~\ref{fig:pnsolver}.
@@
 
-Our computer algebra representation is based on mathematical expression trees, hierarchies composed out of expression instances.
+Our computer algebra representation is based on mathematical expression trees, which are hierarchies of expression instances.
 Those can be basic primitives, such as numbers, variables or special symbols, such as the Kronecker delta or imaginary number.
@@
 They are handled transparently by the code which runs the stencil.
-Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or return the row and column in $A$ for the same coefficient of the closest voxel inside the domain (Neumann BC).
+Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or write into the row and column in $A$ of the same coefficient at the closest voxel inside the domain (Neumann BC).
 
@@
 \begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
+%\missingfigure{test}
+\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_collocated.png}
 \end{subfigure}%
@@
 \begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
+%\missingfigure{test2}
+\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_staggered.png}
 \end{subfigure}%
-\vspace{-0.2in}
+\vspace{-0.1in}
 \icaption{The solution of our solver for the 2D-checkerboard problem with collocated grids (left), where all coefficients are located at the cell centers.
-It suffers from severe oscillating artefacts, which are removed by offseting the coefficient locations using staggered grids (right).}
-\label{fig:artefacts}
+It suffers from severe oscillating artifacts, which are removed by offsetting the coefficient locations using staggered grids (right).}
+\label{fig:artifacts}
 \end{figure}

### 2018-02-21 cpp/pnsolver/include/math/sph.h

@@ -38,7 +38,8 @@ namespace sph
 	void staticInit();
 	void staticShutdown();
 	double legendreP(int l, int m, double x);
-	void convolve( const double* coeffs, const double* coeffs_filter, int order, double* coeffs_result );
+	//void convolve( const double* coeffs, const double* coeffs_filter, int order, double* coeffs_result );
+	void convolve( double* coeffs, const double* coeffs_filter, int order );
 	void project(std::function<double(double, double)> fun, int order, int res, double* coeffs_result );
 	double csp( int m );
 	double lambda( int l );

### 2018-02-21 cpp/pnsolver/src/main.cpp

@@ -982,7 +982,6 @@ struct HGPhase
 
 	HGPhase( double g ):m_g(g)
 	{
-
 	}
 
 	double eval( const V3d& wi, const V3d& wo )const
@@ -1815,6 +1814,7 @@ void test2()
 
 PYBIND11_MODULE(pnsolver, m)
 {
+	sph::staticInit();
 	/*
 	m.def( "solve_multigrid", &solve_multigrid);
 	m.def( "solve_multigrid2", &solve_multigrid2);
@@ -1833,6 +1833,8 @@ PYBIND11_MODULE(pnsolver, m)
 
 	m.def( "test", &test);
 	m.def( "test2", &test2);
+
+	m.def( "sph_basis", &sph::basis);
 
 
 	/*

### 2018-02-21 cpp/pnsolver/src/math/sph.cpp

@@ -141,6 +141,7 @@ namespace sph
 		return p_ll;
 	}
 
+	/*
 	void convolve(const double *coeffs, const double *coeffs_filter, int order, double *coeffs_result)
 	{
 		for (int l=0; l<=order; ++l)
@@ -148,6 +149,17 @@ namespace sph
 			double alpha = std::sqrt(4 * (double) M_PI / (2*l + 1));
 			for (int m=-l; m<=l; ++m)
 				get(l, m, coeffs_result) *= alpha * get(l, 0, coeffs_filter);
+		}
+	}
+	*/
+
+	void convolve(double *coeffs, const double *coeffs_filter, int order)
+	{
+		for (int l=0; l<=order; ++l)
+		{
+			double alpha = std::sqrt(4 * (double) M_PI / (2*l + 1));
+			for (int m=-l; m<=l; ++m)
+				get(l, m, coeffs) *= alpha * get(l, 0, coeffs_filter);
 		}
 	}
 

### 2018-02-21 cpp/pnsolver/src/stencil_cda.cpp

@@ -20,12 +20,12 @@ void stencil_cda(PNSystem::Stencil::Context& ctx)
 	//double extinction_tol = 1.0e-4;
 	//double extinction_tol = 6.0;
 
-	double D_xph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(vs[0], 0.0, 0.0))[color_channel]);
-	double D_xmh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(vs[0], 0.0, 0.0))[color_channel]);
-	double D_yph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(0.0, vs[1], 0.0))[color_channel]);
-	double D_ymh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(0.0, vs[1], 0.0))[color_channel]);
-	double D_zph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(0.0, 0.0, vs[2]))[color_channel]);
-	double D_zmh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(0.0, 0.0, vs[2]))[color_channel]);
+	double D_xph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(vs[0]*0.5, 0.0, 0.0))[color_channel]);
+	double D_xmh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(vs[0]*0.5, 0.0, 0.0))[color_channel]);
+	double D_yph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(0.0, vs[1]*0.5, 0.0))[color_channel]);
+	double D_ymh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(0.0, vs[1]*0.5, 0.0))[color_channel]);
+	double D_zph = 1.0/(3.0*problem.evalExtinction(voxel_center_WS+V3d(0.0, 0.0, vs[2]*0.5))[color_channel]);
+	double D_zmh = 1.0/(3.0*problem.evalExtinction(voxel_center_WS-V3d(0.0, 0.0, vs[2]*0.5))[color_channel]);
 
 	// compute diffusion coefficient
 

### 2018-02-21 cpp/pnsolver/src/stencil_fopn_p2_sg.cpp

@@ -1,0 +1,214 @@
+// This file was generated by stencil.py
+
+#include <PNSystem.h>
+
+void stencil_fopn_p2_sg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(1*domain.getVoxelSize()[0]), 1.0/(1*domain.getVoxelSize()[1]), 1.0/(1*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	Eigen::Matrix<std::complex<double>, 9, 9> S;
+	S.coeffRef(0, 0) = std::complex<double>(1.0, 0.0);
+	S.coeffRef(1, 1) = std::complex<double>(0.7071067811865475, 0.0);
+	S.coeffRef(1, 3) = std::complex<double>(-0.7071067811865475, 0.0);
+	S.coeffRef(2, 1) = std::complex<double>(-0.0, -0.7071067811865475);
+	S.coeffRef(2, 3) = std::complex<double>(-0.0, -0.7071067811865475);
+	S.coeffRef(3, 2) = std::complex<double>(1.0, 0.0);
+	S.coeffRef(4, 4) = std::complex<double>(0.7071067811865475, 0.0);
+	S.coeffRef(4, 8) = std::complex<double>(0.7071067811865475, 0.0);
+	S.coeffRef(5, 4) = std::complex<double>(-0.0, -0.7071067811865475);
+	S.coeffRef(5, 8) = std::complex<double>(0.0, 0.7071067811865475);
+	S.coeffRef(6, 5) = std::complex<double>(0.7071067811865475, 0.0);
+	S.coeffRef(6, 7) = std::complex<double>(-0.7071067811865475, 0.0);
+	S.coeffRef(7, 5) = std::complex<double>(-0.0, -0.7071067811865475);
+	S.coeffRef(7, 7) = std::complex<double>(-0.0, -0.7071067811865475);
+	S.coeffRef(8, 6) = std::complex<double>(1.0, 0.0);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_fopn_p3_sg.cpp

@@ -1,382 +1,0 @@
-/*
-// This file was generated by stencil.py
-
-#include <PNSystem.h>
-
-void stencil_fopn_p3_sg(PNSystem::Stencil::Context& ctx)
-{
-	V3i vi = ctx.getVoxelCoord();
-	V3d vd = vi.cast<double>();
-	const Domain& domain = ctx.getDomain();
-	const PNVolume& problem = ctx.getProblem();
-	V3d h_inv( 1.0/(1*domain.getVoxelSize()[0]), 1.0/(1*domain.getVoxelSize()[1]), 1.0/(1*domain.getVoxelSize()[2]) );
-	int color_channel = 0;
-
-	Eigen::Matrix<std::complex<double>, 16, 16> S;
-	S.coeffRef(0, 0) = std::complex<double>(1.0, 0.0);
-	S.coeffRef(1, 1) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(1, 3) = std::complex<double>(-0.7071067811865475, 0.0);
-	S.coeffRef(2, 1) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(2, 3) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(3, 2) = std::complex<double>(1.0, 0.0);
-	S.coeffRef(4, 4) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(4, 8) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(5, 4) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(5, 8) = std::complex<double>(0.0, 0.7071067811865475);
-	S.coeffRef(6, 5) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(6, 7) = std::complex<double>(-0.7071067811865475, 0.0);
-	S.coeffRef(7, 5) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(7, 7) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(8, 6) = std::complex<double>(1.0, 0.0);
-	S.coeffRef(9, 9) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(9, 15) = std::complex<double>(-0.7071067811865475, 0.0);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_fopn_p5_sg.cpp

@@ -1,906 +1,0 @@
-/*
-// This file was generated by stencil.py
-
-#include <PNSystem.h>
-
-void stencil_fopn_p5_sg(PNSystem::Stencil::Context& ctx)
-{
-	V3i vi = ctx.getVoxelCoord();
-	V3d vd = vi.cast<double>();
-	const Domain& domain = ctx.getDomain();
-	const PNVolume& problem = ctx.getProblem();
-	V3d h_inv( 1.0/(1*domain.getVoxelSize()[0]), 1.0/(1*domain.getVoxelSize()[1]), 1.0/(1*domain.getVoxelSize()[2]) );
-	int color_channel = 0;
-
-	Eigen::Matrix<std::complex<double>, 36, 36> S;
-	S.coeffRef(0, 0) = std::complex<double>(1.0, 0.0);
-	S.coeffRef(1, 1) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(1, 3) = std::complex<double>(-0.7071067811865475, 0.0);
-	S.coeffRef(2, 1) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(2, 3) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(3, 2) = std::complex<double>(1.0, 0.0);
-	S.coeffRef(4, 4) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(4, 8) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(5, 4) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(5, 8) = std::complex<double>(0.0, 0.7071067811865475);
-	S.coeffRef(6, 5) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(6, 7) = std::complex<double>(-0.7071067811865475, 0.0);
-	S.coeffRef(7, 5) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(7, 7) = std::complex<double>(-0.0, -0.7071067811865475);
-	S.coeffRef(8, 6) = std::complex<double>(1.0, 0.0);
-	S.coeffRef(9, 9) = std::complex<double>(0.7071067811865475, 0.0);
-	S.coeffRef(9, 15) = std::complex<double>(-0.7071067811865475, 0.0);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p1_2d_cg.cpp

@@ -1,0 +1,69 @@
+// This file was generated by stencil2.py on Sunday, 11. February 2018 04:15PM
+
+#include <PNSystem.h>
+
+void stencil_p1_2d_cg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(2*domain.getVoxelSize()[0]), 1.0/(2*domain.getVoxelSize()[1]), 1.0/(2*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(-1, 0, 0), 2 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 2 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, -1, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
+	ctx.coeff_A( 1, vi+V3i(0, -1, 0), 0 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 1, vi+V3i(0, 1, 0), 0 ) += -(0.57735026919*h_inv[1]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p1_2d_sg.cpp

@@ -1,0 +1,79 @@
+// This file was generated by stencil2.py on Sunday, 11. February 2018 04:15PM
+
+#include <PNSystem.h>
+
+void stencil_p1_2d_sg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(1*domain.getVoxelSize()[0]), 1.0/(1*domain.getVoxelSize()[1]), 1.0/(1*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 2 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 2 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
+	ctx.coeff_A( 1, vi+V3i(0, -1, 0), 0 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 1, vi+V3i(0, 0, 0), 0 ) += -(0.57735026919*h_inv[1]);
+	{
+		double c = 0.0;
+		c+=(0.5*ctx.evalExtinction(0, -1, 0)[color_channel]);
+		c+=(0.5*ctx.evalExtinction(0, 0, 0)[color_channel]);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p1_3d_cg.cpp

@@ -1,0 +1,86 @@
+// This file was generated by stencil2.py on Friday, 16. February 2018 08:43PM
+
+#include <PNSystem.h>
+
+void stencil_p1_3d_cg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(2*domain.getVoxelSize()[0]), 1.0/(2*domain.getVoxelSize()[1]), 1.0/(2*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(-1, 0, 0), 3 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 3 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, -1, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, -1), 2 ) += -(0.57735026919*h_inv[2]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 1), 2 ) += (0.57735026919*h_inv[2]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
+	ctx.coeff_A( 1, vi+V3i(0, -1, 0), 0 ) += (0.57735026919*h_inv[1]);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p1_3d_sg.cpp

@@ -1,0 +1,101 @@
+// This file was generated by stencil2.py on Friday, 16. February 2018 08:43PM
+
+#include <PNSystem.h>
+
+void stencil_p1_3d_sg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(1*domain.getVoxelSize()[0]), 1.0/(1*domain.getVoxelSize()[1]), 1.0/(1*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 3 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 3 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 2 ) += -(0.57735026919*h_inv[2]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 1), 2 ) += (0.57735026919*h_inv[2]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
+	ctx.coeff_A( 1, vi+V3i(0, -1, 0), 0 ) += (0.57735026919*h_inv[1]);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p2_3d_cg.cpp

@@ -1,0 +1,195 @@
+// This file was generated by stencil2.py on Friday, 16. February 2018 08:43PM
+
+#include <PNSystem.h>
+
+void stencil_p2_3d_cg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(2*domain.getVoxelSize()[0]), 1.0/(2*domain.getVoxelSize()[1]), 1.0/(2*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(-1, 0, 0), 3 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 3 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, -1, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, -1), 2 ) += -(0.57735026919*h_inv[2]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 1), 2 ) += (0.57735026919*h_inv[2]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p2_3d_sg.cpp

@@ -1,0 +1,273 @@
+// This file was generated by stencil2.py on Friday, 16. February 2018 08:43PM
+
+#include <PNSystem.h>
+
+void stencil_p2_3d_sg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(1*domain.getVoxelSize()[0]), 1.0/(1*domain.getVoxelSize()[1]), 1.0/(1*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 3 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 3 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 2 ) += -(0.57735026919*h_inv[2]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 1), 2 ) += (0.57735026919*h_inv[2]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
+	ctx.coeff_A( 1, vi+V3i(0, -1, 0), 8 ) += -(0.4472135955*h_inv[1]);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p3_3d_cg.cpp

@@ -1,0 +1,370 @@
+// This file was generated by stencil2.py on Friday, 16. February 2018 08:43PM
+
+#include <PNSystem.h>
+
+void stencil_p3_3d_cg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(2*domain.getVoxelSize()[0]), 1.0/(2*domain.getVoxelSize()[1]), 1.0/(2*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(-1, 0, 0), 3 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 3 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, -1, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, -1), 2 ) += -(0.57735026919*h_inv[2]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 1), 2 ) += (0.57735026919*h_inv[2]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p3_3d_sg.cpp

@@ -1,0 +1,555 @@
+// This file was generated by stencil2.py on Friday, 16. February 2018 08:43PM
+
+#include <PNSystem.h>
+
+void stencil_p3_3d_sg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(1*domain.getVoxelSize()[0]), 1.0/(1*domain.getVoxelSize()[1]), 1.0/(1*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 3 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 3 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 0), 2 ) += -(0.57735026919*h_inv[2]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 1), 2 ) += (0.57735026919*h_inv[2]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
+	ctx.coeff_A( 1, vi+V3i(0, -1, 0), 8 ) += -(0.4472135955*h_inv[1]);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p4_3d_cg.cpp

@@ -1,0 +1,611 @@
+// This file was generated by stencil2.py on Friday, 16. February 2018 08:43PM
+
+#include <PNSystem.h>
+
+void stencil_p4_3d_cg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(2*domain.getVoxelSize()[0]), 1.0/(2*domain.getVoxelSize()[1]), 1.0/(2*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(-1, 0, 0), 3 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 3 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, -1, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, -1), 2 ) += -(0.57735026919*h_inv[2]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 1), 2 ) += (0.57735026919*h_inv[2]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p4_3d_sg.cpp

@@ -2,7 +2,7 @@
 
 #include <PNSystem.h>
 
-void stencil_p5_3d_sg(PNSystem::Stencil::Context& ctx)
+void stencil_p4_3d_sg(PNSystem::Stencil::Context& ctx)
 {
 	V3i vi = ctx.getVoxelCoord();
 	V3d vd = vi.cast<double>();
@@ -591,20 +591,10 @@ void stencil_p5_3d_sg(PNSystem::Stencil::Context& ctx)
 		ctx.coeff_b( 15 ) += c;
 	}
 	// row=16 l=4 m=-4 --------------------------
-	ctx.coeff_A( 16, vi+V3i(0, -1, 0), 35 ) += -(0.476731294623*h_inv[1]);
-	ctx.coeff_A( 16, vi+V3i(0, 0, 0), 35 ) += (0.476731294623*h_inv[1]);
 	ctx.coeff_A( 16, vi+V3i(0, -1, 0), 15 ) += (0.471404520791*h_inv[1]);
 	ctx.coeff_A( 16, vi+V3i(0, 0, 0), 15 ) += -(0.471404520791*h_inv[1]);
-	ctx.coeff_A( 16, vi+V3i(0, -1, 0), 33 ) += -(0.0710669054519*h_inv[1]);
-	ctx.coeff_A( 16, vi+V3i(0, 0, 0), 33 ) += (0.0710669054519*h_inv[1]);
 	ctx.coeff_A( 16, vi+V3i(-1, 0, 0), 9 ) += (0.471404520791*h_inv[0]);
 	ctx.coeff_A( 16, vi+V3i(0, 0, 0), 9 ) += -(0.471404520791*h_inv[0]);
-	ctx.coeff_A( 16, vi+V3i(-1, 0, 0), 27 ) += -(0.0710669054519*h_inv[0]);
-	ctx.coeff_A( 16, vi+V3i(0, 0, 0), 27 ) += (0.0710669054519*h_inv[0]);
-	ctx.coeff_A( 16, vi+V3i(-1, 0, 0), 25 ) += (0.476731294623*h_inv[0]);
-	ctx.coeff_A( 16, vi+V3i(0, 0, 0), 25 ) += -(0.476731294623*h_inv[0]);
-	ctx.coeff_A( 16, vi+V3i(0, 0, 0), 26 ) += -(0.301511344578*h_inv[2]);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p5_2d_cg.cpp

@@ -1,0 +1,495 @@
+// This file was generated by stencil2.py on Sunday, 11. February 2018 04:10PM
+
+#include <PNSystem.h>
+
+void stencil_p5_2d_cg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(2*domain.getVoxelSize()[0]), 1.0/(2*domain.getVoxelSize()[1]), 1.0/(2*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(-1, 0, 0), 2 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 2 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, -1, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
+	ctx.coeff_A( 1, vi+V3i(0, -1, 0), 5 ) += -(0.4472135955*h_inv[1]);
+	ctx.coeff_A( 1, vi+V3i(0, 1, 0), 5 ) += (0.4472135955*h_inv[1]);
[…]

### 2018-02-21 cpp/pnsolver/src/stencil_p5_3d_cg.cpp

@@ -1,0 +1,918 @@
+// This file was generated by stencil2.py on Friday, 16. February 2018 08:43PM
+
+#include <PNSystem.h>
+
+void stencil_p5_3d_cg(PNSystem::Stencil::Context& ctx)
+{
+	V3i vi = ctx.getVoxelCoord();
+	V3d vd = vi.cast<double>();
+	const Domain& domain = ctx.getDomain();
+	const PNVolume& problem = ctx.getProblem();
+	V3d h_inv( 1.0/(2*domain.getVoxelSize()[0]), 1.0/(2*domain.getVoxelSize()[1]), 1.0/(2*domain.getVoxelSize()[2]) );
+	int color_channel = 0;
+
+	// row=0 l=0 m=0 --------------------------
+	ctx.coeff_A( 0, vi+V3i(-1, 0, 0), 3 ) += (0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(1, 0, 0), 3 ) += -(0.57735026919*h_inv[0]);
+	ctx.coeff_A( 0, vi+V3i(0, -1, 0), 1 ) += (0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 1, 0), 1 ) += -(0.57735026919*h_inv[1]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, -1), 2 ) += -(0.57735026919*h_inv[2]);
+	ctx.coeff_A( 0, vi+V3i(0, 0, 1), 2 ) += (0.57735026919*h_inv[2]);
+	{
+		double c = 0.0;
+		c+=ctx.evalExtinction(0, 0, 0)[color_channel];
+		c+=-(3.54490770181*ctx.evalScattering(0, 0, 0)[color_channel]*ctx.evalPhase(0, 0, 0, 0, 0)[color_channel]);
+		ctx.coeff_A( 0, vi+V3i(0, 0, 0), 0 ) += c;
+	}
+	{
+		double c = 0.0;
+		c+=ctx.evalEmission(0, 0, 0, 0, 0)[color_channel];
+		ctx.coeff_b( 0 ) += c;
+	}
+	// row=1 l=1 m=-1 --------------------------
[…]

### 2018-02-21 cpp/renderer/include/fields/VoxelGridField.h

@@ -96,6 +96,7 @@ struct VoxelGridField : public Field<T>
 
 	void save( const std::string& filename )
 	{
+		std::cout << "writing voxelgrid to " << filename << std::endl;
 		m_voxelgrid.save(filename);
 	}
 

### 2018-02-21 cpp/renderer/include/integrator.h

@@ -51,6 +51,7 @@ struct Integrator
 
 // returns sigma_t at sampled position (is invalid when we exceeded maxt)
 double delta_tracking(const Scene* scene, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t , bool debug = false);
+double delta_tracking(const Volume* volume, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t , bool debug = false);
 
 struct Vertex
 {
@@ -169,8 +170,11 @@ struct TraceInfo
 	{
 		if( current_vertex.getType() == Vertex::EVolume )
 		{
-			V3d wi = current_direction;
+			// technically, in a standard forward pt, -current_direciton is the
+			// direction of outgoing light (towards the camera), which would be called wo
+			V3d wi = -current_direction;
 			double pdf;
+			// NB: we assume wi and wo can be flipped, which means we assume phase function only depends on angle
 			double phase_over_pdf = scene->volume->samplePhase(current_vertex.getPosition(), wi, current_direction, pdf, rng);
 			throughput_over_pdf = phase_over_pdf*throughput_over_pdf.cwiseProduct(current_vertex.m_sigma_s);
 			return true;

### 2018-02-21 cpp/renderer/include/math/common.h

@@ -197,15 +197,21 @@ void sincos(T theta, T *_sin, T *_cos)
 }
 
 template<typename T>
-inline double safe_acos( const T& in )
+inline T safe_acos( const T& in )
 {
 	return std::acos( std::max( T(-1.0), std::min(T(1.0), in)) );
 }
 
 template<typename T>
-inline double safe_asin( const T& in )
+inline T safe_asin( const T& in )
 {
 	return std::asin( std::max( T(-1.0), std::min(T(1.0), in)) );
+}
+
+template<typename T>
+inline T safe_sqrt(T value)
+{
+	return std::sqrt(std::max(T(0.0), value));
 }
 
 template<typename T>

### 2018-02-21 cpp/renderer/include/math/sph.h

@@ -38,11 +38,14 @@ namespace sph
 	void staticInit();
 	void staticShutdown();
 	double legendreP(int l, int m, double x);
+	//void convolve( const double* coeffs, const double* coeffs_filter, int order, double* coeffs_result );
 	void convolve( double* coeffs, const double* coeffs_filter, int order );
 	void project(std::function<double(double, double)> fun, int order, int res, double* coeffs_result );
 	double csp( int m );
+	double lambda( int l );
 	Eigen::MatrixXcd buildComplexToRealConversionMatrix( int order );
-	double basis_real( int l, int m, double theta, double phi );
+	double basis( int l, int m, double theta, double phi );
+	std::complex<double> complex_basis( int l, int m, double theta, double phi );
 	double factorial(int x);
 	double computeNormalization(int l, int m);
 	extern double *m_normalization;

### 2018-02-21 cpp/renderer/include/math/sph_.h

@@ -1,58 +1,0 @@
-#pragma once
-
-#include <complex>
-#include <functional>
-
-
-
-namespace sph
-{
-/*
-    typedef std::complex<double> complex;
-	inline int numSHCoeffs( int order )
-	{
-		return (order+1)*(order+1);
-	}
-
-	inline int shIndex( int l, int m )
-	{
-		return l * (l + 1) + m;
-	}
-
-	double P(int l, int m, double x);
-
-    complex sph_basis( int l, int m, double theta, double phi );
-*/
-
-
-
-
-
-
-
-	int numCoeffs(int order);
-	int index( int l, int m );
-	double eval(double theta, double phi, const double *coeffs, int order);
-	double& get( int l, int m, double* coeffs );
-	const double& get( int l, int m, const double* coeffs );
-	void staticInit();
-	void staticShutdown();
-	double legendreP(int l, int m, double x);
-	void convolve( double* coeffs, const double* coeffs_filter, int order );
-	void project(std::function<double(double, double)> fun, int order, int res, double* coeffs_result );
-	double csp( int m );
-	Eigen::MatrixXcd buildComplexToRealConversionMatrix( int order );
-	double basis_real( int l, int m, double theta, double phi );
-	double factorial(int x);
-	double computeNormalization(int l, int m);
-	extern double *m_normalization;
-	extern int m_normTableSize;
-	inline static double normalization(int l, int m)
-	{
-		if (l < m_normTableSize)
-			return m_normalization[l*(l+1)/2 + m];
-		else
-			return computeNormalization(l, m);
-	}
-}
-

### 2018-02-21 cpp/renderer/include/volume.h

@@ -45,17 +45,13 @@ struct IsotropicPhase : public PhaseFunction
 	}
 };
 
+/*
 struct HGPhase : public PhaseFunction
 {
 	typedef std::shared_ptr<HGPhase> Ptr;
 
 	virtual double sample( const V3d& wi, V3d& wo, double& pdf, RNGd& rng ) const override
 	{
-		/*
-		wo = sampleSphere<double>(rng);
-		pdf = INV_FOURPI;
-		return 1.0; // eval/pdf = 1.0
-		*/
 		return 0.0;
 	}
 
@@ -63,6 +59,67 @@ struct HGPhase : public PhaseFunction
 	{
 		double temp = 1.0 + m_g*m_g + 2.0 * m_g * dot(wi, wo);
 		return INV_FOURPI * (1 - m_g*m_g) / (temp * std::sqrt(temp));
+	}
+
+
+private:
+	double m_g;
+};
+*/
+
+struct HGPhase : public PhaseFunction
+{
+	typedef std::shared_ptr<HGPhase> Ptr;
+
+	HGPhase( double g ):m_g(g)
+	{
+
+	}
+
+	virtual double sample( const V3d& wi, V3d& wo, double& pdf, RNGd& rng ) const override
+	{
+		P2d sample(rng.next1D(), rng.next1D());
+
+		double cosTheta;
+		if (std::abs(m_g) < Epsilon)
+		{
+			cosTheta = 1 - 2*sample.x();
+		} else
+		{
+			double sqrTerm = (1 - m_g * m_g) / (1 - m_g + 2 * m_g * sample.x());
+			cosTheta = (1 + m_g * m_g - sqrTerm * sqrTerm) / (2 * m_g);
+		}
+
+		double sinTheta = safe_sqrt(1.0f-cosTheta*cosTheta),sinPhi, cosPhi;
+
+		sincos(2.0*M_PI*sample.y(), &sinPhi, &cosPhi);
+
+		wo = Framed(-wi).toWorld(V3d(
+			sinTheta * cosPhi,
+			sinTheta * sinPhi,
+			cosTheta
+		));
+
+		pdf = eval(wi, wo);
+
+		return 1.0f;
+	}
+
+	double eval( const V3d& wi, const V3d& wo )const
+	{
+		double temp = 1.0 + m_g*m_g + 2.0 * m_g * dot(wi, wo);
+		return INV_FOURPI * (1 - m_g*m_g) / (temp * std::sqrt(temp));
+	}
+
[…]

### 2018-02-21 cpp/renderer/src/Integrator.cpp

@@ -8,12 +8,10 @@
 
 
 
-
-
 // returns sigma_t at sampled position (is invalid when we exceeded maxt)
-double delta_tracking( const Scene* scene, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t, bool debug )
+double delta_tracking( const Volume* volume, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t, bool debug )
 {
-	double sigma_t_max = scene->volume->getMaxExtinction()[component];
+	double sigma_t_max = volume->getMaxExtinction()[component];
 
 	double t = 0.0;
 	while(true)
@@ -24,7 +22,7 @@ double delta_tracking( const Scene* scene, const Ray3d& ray, double maxt, int co
 		if(t>= maxt)
 			break;
 
-		sigma_t = scene->volume->evalExtinction(ray(t), debug);
+		sigma_t = volume->evalExtinction(ray(t), debug);
 
 		// russian roulette
 		if(rng.next1D()<sigma_t[component]/sigma_t_max)
@@ -32,6 +30,12 @@ double delta_tracking( const Scene* scene, const Ray3d& ray, double maxt, int co
 	}
 
 	return t;
+}
+
+// returns sigma_t at sampled position (is invalid when we exceeded maxt)
+double delta_tracking( const Scene* scene, const Ray3d& ray, double maxt, int component, RNGd& rng, V3d& sigma_t, bool debug )
+{
+	return delta_tracking( scene->volume, ray, maxt, component, rng, sigma_t, debug );
 }
 
 

### 2018-02-21 cpp/renderer/src/main.cpp

@@ -234,7 +234,8 @@ void render_thread( MonteCarloTaskInfo& ti, const GlobalRenderInfo* gi )
 		for (int x = gi->crop_window.min.x(); x < gi->crop_window.max.x(); ++x)
 		{
 			int index = y*width + x;
-			RNGd& rng = (*gi->debug_rngs)[ index ];
+			//RNGd& rng = (*gi->debug_rngs)[ index ];
+			RNGd& rng = ti.rng;
 			V3d f(0.0, 0.0, 0.0);
 			//Color3f T(1.0f);
 
@@ -304,6 +305,7 @@ void render_thread( MonteCarloTaskInfo& ti, const GlobalRenderInfo* gi )
 Image::Ptr render( Volume::Ptr volume, Light::Ptr light, Camera::Ptr camera, Integrator::Ptr integrator, int numSamples )
 {
 	std::cout << "rendering:\n";
+
 	std::cout << volume->toString();
 	std::cout << light->toString();
 	std::cout << camera->toString();
@@ -324,7 +326,7 @@ Image::Ptr render( Volume::Ptr volume, Light::Ptr light, Camera::Ptr camera, Int
 	//gi.debug_pixel = V2i(378, 110);
 	//gi.debug_pixel = V2i(256, 256);
 
-
+	/*
 	// for debugging, intitialize rngs
 	std::vector<RNGd> debug_rngs;
 	int index = 0;
@@ -332,7 +334,7 @@ Image::Ptr render( Volume::Ptr volume, Light::Ptr light, Camera::Ptr camera, Int
 		for( int j=0;j<camera->getResolution()[1];++j, ++index )
 			debug_rngs.push_back( RNGd(123+index) );
 	gi.debug_rngs = &debug_rngs;
-
+	*/
 
 	//int numSamples = 1;
 
@@ -507,6 +509,114 @@ VoxelGridField3d::Ptr compute_unscattered_fluence(  Volume::Ptr volume,
 
 	return result;
 }
+struct GlobalBakeUnscatteredDirectionalLightInfo
+{
+	Volume* volume;
+	VoxelGrid3d* grid; // result where all threads write to
+	V3d lightDirection;
+	double lightRadiance;
+	double maxDistance;
+};
+
+
+void bake_unscattered_directional_light_thread( MonteCarloTaskInfo& taskinfo, const GlobalBakeUnscatteredDirectionalLightInfo* gi )
+{
+	RNGd& rng = taskinfo.rng;
[…]

### 2018-02-21 cpp/renderer/src/math/sph.cpp

@@ -12,148 +12,23 @@
 namespace sph
 {
 
-	/*
-	// Number of precomputed factorials and double-factorials that can be
-	// returned in constant time.
-	const int kCacheSize = 16;
-
-
-	// Compute the factorial for an integer @x. It is assumed x is at least 0.
-	// This implementation precomputes the results for low values of x, in which
-	// case this is a constant time lookup.
-	//
-	// The vast majority of SH evaluations will hit these precomputed values.
-	double factorial(int x)
+	std::complex<double> Y_gg( int l, int m, double theta, double phi )
 	{
-		const double factorial_cache[kCacheSize] = {1, 1, 2, 6, 24, 120, 720, 5040,
-												  40320, 362880, 3628800, 39916800,
-												  479001600, 6227020800,
-												  87178291200, 1307674368000};
-
-		if (x < kCacheSize)
-		{
-			return factorial_cache[x];
-		}else
-		{
-			double s = 1.0;
-			for (int n = 2; n <= x; n++)
-			{
-				s *= n;
-			}
-			return s;
-		}
+		return  csp(m)*
+				normalization(l, std::abs(m))*
+				legendreP(l, m, std::cos(theta))*std::complex<double>(std::cos(m*phi), std::sin(m*phi));
 	}
 
-	// Compute the double factorial for an integer @x. This assumes x is at least
-	// 0.  This implementation precomputes the results for low values of x, in
-	// which case this is a constant time lookup.
-	//
-	// The vast majority of SH evaluations will hit these precomputed values.
-	// See http://mathworld.wolfram.com/DoubleFactorial.html
-	double doubleFactorial(int x)
+	std::complex<double> Y_cc( int l, int m, double theta, double phi )
 	{
-	  const double dbl_factorial_cache[kCacheSize] = {1, 1, 2, 3, 8, 15, 48, 105,
-													  384, 945, 3840, 10395, 46080,
-													  135135, 645120, 2027025};
-
-		if (x < kCacheSize)
-		{
-			return dbl_factorial_cache[x];
-		}else
-		{
-			double s = 1.0;
[…]

### 2018-02-21 cpp/renderer/src/math/sph_.cpp

@@ -1,442 +1,0 @@
-#include <math/common.h>
-#include <math/sph.h>
-
-
-
-
-
-
-
-
-
-namespace sph
-{
-
-	/*
-	// Number of precomputed factorials and double-factorials that can be
-	// returned in constant time.
-	const int kCacheSize = 16;
-
-
-	// Compute the factorial for an integer @x. It is assumed x is at least 0.
-	// This implementation precomputes the results for low values of x, in which
-	// case this is a constant time lookup.
-	//
-	// The vast majority of SH evaluations will hit these precomputed values.
-	double factorial(int x)
-	{
-		const double factorial_cache[kCacheSize] = {1, 1, 2, 6, 24, 120, 720, 5040,
-												  40320, 362880, 3628800, 39916800,
-												  479001600, 6227020800,
-												  87178291200, 1307674368000};
-
-		if (x < kCacheSize)
-		{
-			return factorial_cache[x];
-		}else
-		{
-			double s = 1.0;
-			for (int n = 2; n <= x; n++)
-			{
-				s *= n;
-			}
-			return s;
-		}
-	}
-
-	// Compute the double factorial for an integer @x. This assumes x is at least
-	// 0.  This implementation precomputes the results for low values of x, in
-	// which case this is a constant time lookup.
-	//
-	// The vast majority of SH evaluations will hit these precomputed values.
-	// See http://mathworld.wolfram.com/DoubleFactorial.html
-	double doubleFactorial(int x)
-	{
-	  const double dbl_factorial_cache[kCacheSize] = {1, 1, 2, 3, 8, 15, 48, 105,
-													  384, 945, 3840, 10395, 46080,
-													  135135, 645120, 2027025};
-
-		if (x < kCacheSize)
-		{
-			return dbl_factorial_cache[x];
-		}else
-		{
-			double s = 1.0;
-			double n = x;
-			while (n > 1.0)
-			{
-				s *= n;
-				n -= 2.0;
-			}
-			return s;
-		}
-	}
-	// Evaluate the associated Legendre polynomial of degree @l and order @m at
[…]

### 2018-02-21 cpp/renderer/src/volume.cpp

@@ -39,6 +39,11 @@ void Volume::setExtinctionAlbedo( Field3d::Ptr extinction, Field3d::Ptr albedo )
 	m_field_extinction = extinction;
 	m_field_albedo = albedo;
 }
+
+void Volume::setPhaseFunction(PhaseFunction::Ptr phase)
+{
+	m_phaseFunction = phase;
+}
 void Volume::setLocalToWorld( const Transformd& localToWorld )
 {
 	m_localToWorld = localToWorld;

### 2018-02-21 python/pnsolver/example2.py

@@ -48,10 +48,11 @@ def solve( stencil_name, problem, filename, do_neumannBC = False ):
 
 	#sys.build()
 	#A = sys.get_A()
+	#b = sys.get_b()
 	##print(A.shape)
-	#mat_filename = filename+".t0_2.mat"
+	#mat_filename = filename+".mat"
 	#print(mat_filename)
-	#scipy.io.savemat(mat_filename, {"A2":A})
+	#scipy.io.savemat(mat_filename, {"A":A, "b":b, "xref":x})
 	#return
 
 
@@ -78,11 +79,12 @@ def solve( stencil_name, problem, filename, do_neumannBC = False ):
 	#x = x.reshape((x.shape[0], 1))
 
 
-	tol = 1.0e-10
+	#tol = 1.0e-10
 	#tol = 1.0e-5
 	#tol = 1.7
-	#tol = 4.0
+	#tol = 4.0 # nebulae p5
 	#tol = 1.0
+	tol = 0.16 # cloud p5
 
 	#numLevels = 7
 	#x, convergence, timestamps = pnsolver.solve_multigrid( sys, numLevels )
@@ -116,23 +118,32 @@ def solve( stencil_name, problem, filename, do_neumannBC = False ):
 	#data["convergence_cg"] = convergence
 	#data["timestamps_cg"] = timestamps
 
-	# instead of solving...just store matrices...
+	## instead of solving...just store matrices...
 	#sys.build()
 	#data = {}
 	#data["A"] = sys.get_A()
 	#data["b"] = sys.get_b()
-	#scipy.io.savemat("{}.mat".format(filename), data)
+	#mat_filename = "{}.mat".format(filename)
+	#scipy.io.savemat(mat_filename, data)
+	#print("save {}".format(mat_filename))
 	#return
 
 	# solve ----------------
+	solver = None
+	#solver = "gs"
+	#solver = "cg"
 	#solver = "ls_cg"
-	solver = "solve_ls_lscg"
+	solver = "ls_lscg"
 
-	if solver == "ls_cg":
+	if solver == "gs":
+		x, convergence, timestamps = pnsolver.solve_gs( sys, tol, 1000)
+	elif solver == "cg":
+		x, convergence, timestamps = pnsolver.solve_cg( sys, tol)
+	elif solver == "ls_cg":
 		x, convergence, timestamps = pnsolver.solve_ls_cg( sys, tol)
 	elif solver == "mg":
 		x, convergence, timestamps = pnsolver.solve_mg( sys, tol, 100)
[…]

### 2018-02-21 python/pnsolver/meh.py

@@ -1308,7 +1308,7 @@ class SHRecursiveRelation(object):
 			if base < 0.0:
 				return np.nan
 			return np.sqrt(base)
-		b = fun( "b", add(l, num(1)), m )
+		b = fun( "b", l, m )
 		b.setAllSuperScripts()
 		b.body2 = b_body
 		return b
@@ -1318,6 +1318,7 @@ class SHRecursiveRelation(object):
 			base = (l+m+1)*(l+m+2)/((2*l+3)*(2*l+1))
 			if base < 0.0:
 				return np.nan
+			#print("running c_lm {} {}={}".format(l, m, np.sqrt(base)))
 			return np.sqrt(base)
 		c = fun( "c", l, m )
 		c.setAllSuperScripts()

### 2018-02-21 python/pnsolver/plot_sh.py

@@ -1,0 +1,43 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import renderer
+
+#renderer.sh_basis( 0, 0, 0.5, 0.0 )
+#renderer.test()
+#exit(1)
+
+def sh_delta( order, theta, phi ):
+	delta_theta = 0.5*np.pi
+	delta_phi = 0.0
+	sum = 0.0
+	for l in range(0,order+1):
+		for m in range(-l,l+1):
+			g = renderer.sh_basis_conj( l,m, delta_theta, delta_phi )
+			sum += renderer.sh_basis( l,m, theta, phi )*g
+	return sum
+
+
+
+renderer.sh_init()
+
+print( renderer.sh_basis( 0, 0, 0.0, 0.0 ) )
+print( np.sqrt(1.0/(4.0*np.pi)) )
+exit(1)
+
+order = 1
+
+theta = np.arange(0, np.pi, 0.01)
+
+ax = plt.subplot(111, projection='polar')
+
+r = np.zeros(theta.shape)
+for i in range(theta.shape[0]):
+	r[i] = sh_delta( order, theta[i], 0.0 )
+ax.plot(theta, r)
+#ax.set_rmax(2)
+#ax.set_rticks([0.5, 1, 1.5, 2])  # less radial ticks
+#ax.set_rlabel_position(-22.5)  # get radial labels away from plotted line
+#ax.grid(True)
+
+#ax.set_title("A line plot on a polar axis", va='bottom')
+plt.show()

### 2018-02-21 python/pnsolver/plots_pointssource.py

@@ -1,0 +1,135 @@
+import os,sys,inspect
+import random
+
+import renderer
+import util
+
+import numpy as np
+
+import matplotlib.pyplot as plt
+from matplotlib.colors import LogNorm
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+
+sigma_t = 8.0
+albedo = 0.9
+size = 2.0
+pointsource_center = np.array([1.0, 1.0, 1.0])
+
+def grosjean(r, sigma_t, albedo):
+    sigma_a = (1.0-albedo)*sigma_t
+    sigma_s = albedo*sigma_t
+    ss = np.exp(-sigma_t*r)/(4.0*np.pi*r*r)
+    tmp = sigma_a+sigma_s
+    D = (2.0*sigma_a+sigma_s)/(3.0*tmp*tmp)
+    ms = 3.0*sigma_s*sigma_t*np.exp(-np.sqrt(sigma_a/D)*r)/(4.0*np.pi*(2.0*sigma_a+sigma_s)*r)
+    return ss+ms
+    #return ms
+
+def grosjean_ss(r, sigma_t, albedo):
+    sigma_a = (1.0-albedo)*sigma_t
+    sigma_s = albedo*sigma_t
+    ss = np.exp(-sigma_t*r)/(4.0*np.pi*r*r)
+    tmp = sigma_a+sigma_s
+    D = (2.0*sigma_a+sigma_s)/(3.0*tmp*tmp)
+    ms = 3.0*sigma_s*sigma_t*np.exp(-np.sqrt(sigma_a/D)*r)/(4.0*np.pi*(2.0*sigma_a+sigma_s)*r)
+    #return ss+ms
+    return ss
+
+def cda_org( r, sigma_t, albedo ):
+    sigma_a = (1.0-albedo)*sigma_t
+    sigma_tr = np.sqrt(3.0*sigma_a*sigma_t)
+    return 3.0*sigma_t*np.exp(-sigma_tr*r)/(4.0*np.pi*r)
+
+
+pns_cda = renderer.load_pnsolution( "C:/projects/epfl/epfl17/python/pnsolver/results/pointsource/pointsource_cda5.pns" )
+pns_cda_old = renderer.load_pnsolution( "C:/projects/epfl/epfl17/python/pnsolver/results/pointsource/pointsource_cda_old.pns" )
+pns_cda_new = renderer.load_pnsolution( "C:/projects/epfl/epfl17/python/pnsolver/results/pointsource/pointsource_cda_new2.pns" )
[…]

### 2018-02-21 python/pnsolver/problems.py

@@ -457,10 +457,10 @@ def pointsource3d( res = 80 ):
 	# 			  = p00*np.sqrt(1.0/(4.0*np.pi))
 	# 			  = np.sqrt(1.0/(4.0*np.pi))*np.sqrt(1.0/(4.0*np.pi))
 	phase_00 = np.sqrt(1.0/(4.0*np.pi))
-	test = phase_00*lambda_l
+	test = phase_00*lambda_l #  == 1
 	t.setPhase( 0, 0, pnsolver.ConstantField3d(np.array([test, test, test])) )
 
-	'''
+	#'''
 	## the emission field is a 3d voxelgrid where the center voxel is set to an emission value
 	shape = (domain.getResolution()[0], domain.getResolution()[1], domain.getResolution()[2], 3)
 	q_voxels = np.zeros(shape) # real valued 
@@ -468,15 +468,15 @@ def pointsource3d( res = 80 ):
 	pointsource_center_WS = np.array([size*0.5, size*0.5, size*0.5])
 	pointsource_center_VS = domain.worldToVoxel(pointsource_center_WS)
 	center_voxel = np.array([int(pointsource_center_VS[0]), int(pointsource_center_VS[1]), int(pointsource_center_VS[2])])
-	print("test={}".format(vs[0]*vs[1]*vs[2]))
-	print("test={}".format(1.0/(vs[0]*vs[1]*vs[2])))
+	#print("test={}".format(vs[0]*vs[1]*vs[2]))
+	#print("test={}".format(1.0/(vs[0]*vs[1]*vs[2])))
 	q_voxels[center_voxel[0], center_voxel[1], center_voxel[2], :] = 1.0/(vs[0]*vs[1]*vs[2])
 	t.setEmission( 0, 0, pnsolver.VoxelGridField3d(q_voxels) )
-	'''
+	#'''
 
 	# the emission field is a 3d voxelgrid where each voxel is initialized with the single scattered light
 	# this way, the solution will only be for the multiple scattered light
-	#'''
+	'''
 	shape = (domain.getResolution()[0], domain.getResolution()[1], domain.getResolution()[2], 3)
 	q_voxels = np.zeros(shape) # real valued 
 	vs = domain.getVoxelSize()
@@ -507,11 +507,148 @@ def pointsource3d( res = 80 ):
 				#q_voxels[i, j, k, :] = sigma_s*grosjean_ss(r, sigma_t, albedo)#*np.sqrt(4.0*np.pi)
[…]

### 2018-02-21 python/pnsolver/render.py

@@ -10,10 +10,10 @@ import matplotlib.pyplot as plt
 from matplotlib.colors import LogNorm
 from mpl_toolkits.axes_grid1 import make_axes_locatable
 
+# for saving png
+import scipy
 
 
-
-# openexr
 
 def translation_matrix(direction):
 	# Return matrix to translate by direction vector.
@@ -72,11 +72,9 @@ def load_camera( filename, id = "cam1" ):
 def create_scene_nebulae():
 	albedo = 0.9
 
-	#offset = -0.5
-	offset = 0.0
 
 	volume = renderer.create_volume()
-	volume.setBound( np.array([-0.5, -0.5, -0.5+offset]), np.array([0.5, 0.5, 0.5+offset]) )
+	volume.setBound( np.array([-0.5, -0.5, -0.5]), np.array([0.5, 0.5, 0.5]) )
 
 
 	sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64.bgeo")
@@ -101,7 +99,41 @@ def create_scene_nebulae():
 	return volume, light
 
 
+def create_scene_cloud():
+	albedo = 0.9
 
+
+	#volume = renderer.create_volume()
+	#volume.setBound( np.array([-0.5, -0.5, -0.5]), np.array([0.5, 0.5, 0.5]) )
+
+	sigma_t_field, bound_min, bound_max = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/cloud/cloud.bgeo")
+	#sigma_t_field, bound_min, bound_max = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/cloud/cloud64_extinction.bgeo")
+	print(bound_min)
+	print(bound_max)
+
+	volume = renderer.create_volume()
+	volume.setBound( bound_min, bound_max )
+	#volume.setPhaseFunction( renderer.HGPhase(0.8) )
+
+	#sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae200.bgeo")
+	#sigma_t_field = renderer.load_bgeo("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/test.bgeo")
+	#sigma_t_field.save("c:/projects/epfl/epfl17/python/pnsolver/results/nebulae/nebulae64_extinction.grid")
+
+	#sigma_t_field = renderer.load_bgeo("test_sphere.bgeo")
[…]

### 2018-02-21 python/pnsolver/rte_terms.py

@@ -743,54 +743,7 @@ class CountImaginaryUnits(object):
 		return expr
 
 class fopn_real(object):
-	@staticmethod
-	def transport_term_expand2( m, order, debug = False, gg = None ):
 
-		x = meh.tensor("\\vec{x}", rank=1, dimension=3)
-		x.setComponent(0, meh.var("x"))
-		x.setComponent(1, meh.var("y"))
-		x.setComponent(2, meh.var("z"))
-		x.collapsed = True
-
-		coeff_c = meh.SHRecursiveRelation.c( meh.sub(meh.var("l'"), meh.num(1)), meh.sub( meh.var("m'"), meh.num(1) ) )
-		coeff_d = meh.SHRecursiveRelation.d( meh.add(meh.var("l'"), meh.num(1)), meh.sub( meh.var("m'"), meh.num(1) ) )
-		coeff_e = meh.SHRecursiveRelation.e( meh.sub(meh.var("l'"), meh.num(1)), meh.add( meh.var("m'"), meh.num(1) ) )
-		coeff_f = meh.SHRecursiveRelation.f( meh.add(meh.var("l'"), meh.num(1)), meh.add( meh.var("m'"), meh.num(1) ) )
-		coeff_a = meh.SHRecursiveRelation.a( meh.sub(meh.var("l'"), meh.num(1)), meh.var("m'"))
-		coeff_b = meh.SHRecursiveRelation.b( meh.add(meh.var("l'"), meh.num(1)), meh.var("m'"))
-
-		L0 = meh.SHCoefficient( 'L', meh.sub(meh.var("l'"), meh.num(1)), meh.add(meh.neg(meh.var("m'")), meh.num(1)), x )
-		L1 = meh.SHCoefficient( 'L', meh.add(meh.var("l'"), meh.num(1)), meh.add(meh.neg(meh.var("m'")), meh.num(1)), x )
-		L2 = meh.SHCoefficient( 'L', meh.sub(meh.var("l'"), meh.num(1)), meh.sub(meh.neg(meh.var("m'")), meh.num(1)), x )
-		L3 = meh.SHCoefficient( 'L', meh.add(meh.var("l'"), meh.num(1)), meh.sub(meh.neg(meh.var("m'")), meh.num(1)), x )
-
-		L4 = meh.SHCoefficient( 'L', meh.sub(meh.var("l'"), meh.num(1)), meh.sub(meh.var("m'"), meh.num(1)), x )
[…]

### 2018-02-21 python/pnsolver/stencil.py

@@ -1958,9 +1958,9 @@ if __name__ == "__main__":
 
 	terms_fopn = []
 	terms_fopn.append(rte_terms.fopn.transport_term())
-	#terms_fopn.append(rte_terms.fopn.collision_term())
-	#terms_fopn.append(rte_terms.fopn.scattering_term())
-	#terms_fopn.append(rte_terms.fopn.source_term())
+	terms_fopn.append(rte_terms.fopn.collision_term())
+	terms_fopn.append(rte_terms.fopn.scattering_term())
+	terms_fopn.append(rte_terms.fopn.source_term())
 	
 	
 	terms_fopn = rte_terms.splitAddition( terms_fopn )
@@ -1982,10 +1982,12 @@ if __name__ == "__main__":
 	#order = [0,1,2,3,4]
 
 	rte_forms = ["fopn"]
-	order = [1]
+	#order = [1]
+	order = [2]
 	#order = [3,5]
 	#order = [1,2,3,4,5]
-	staggered = [False]
+	#staggered = [False]
+	staggered = [True]
 
 	test = itertools.product(rte_forms, order, staggered)
 	for c in test:
@@ -1999,8 +2001,8 @@ if __name__ == "__main__":
 		stencil_name = "stencil_{}_p{}_{}".format(rte_form_name, order, staggered_id[is_staggered] )
 		filename = "{}/{}.cpp".format(path, stencil_name)
 
-		pni = PNInfo2D(order, is_staggered)
-		#pni = PNInfo3D(order, is_staggered)
+		#pni = PNInfo2D(order, is_staggered)
+		pni = PNInfo3D(order, is_staggered)
 
 
 		generate_stencil_code( stencil_name, filename, rte_form_terms, pni )

### 2018-02-21 python/pnsolver/stencil2.py

@@ -1044,12 +1044,14 @@ def generate_stencil_code( pne, pni, stencil_name ):
 			coeff_expr = coeffs.getExpr()
 			if coeff_expr.__class__ == meh.Addition:
 				cpp += "\t{\n"
-				cpp += "\t\tstd::complex<double> c(0.0, 0.0);\n"
+				#cpp += "\t\tstd::complex<double> c(0.0, 0.0);\n"
+				cpp += "\t\tdouble c = 0.0;\n"
 				ops = coeff_expr.getOperands()
 				for op in ops:
 					code = generate_cpp_recursive(op);
 					cpp += "\t\tc+={};\n".format(code)
-				cpp += "\t\tctx.coeff_A( {}, vi+V3i({}, {}, {}), {} ) += c.real();\n".format(coeff_index, i, j, k, coeff_index_j)
+				#cpp += "\t\tctx.coeff_A( {}, vi+V3i({}, {}, {}), {} ) += c.real();\n".format(coeff_index, i, j, k, coeff_index_j)
+				cpp += "\t\tctx.coeff_A( {}, vi+V3i({}, {}, {}), {} ) += c;\n".format(coeff_index, i, j, k, coeff_index_j)
 				cpp += "\t}\n"
 			else:
 				code = generate_cpp_recursive(coeff_expr);
@@ -1060,7 +1062,8 @@ def generate_stencil_code( pne, pni, stencil_name ):
 		# RHS
 		if not rhs_coeffs is None:
 			cpp += "\t{\n"
-			cpp += "\t\tstd::complex<double> c(0.0, 0.0);\n"
+			#cpp += "\t\tstd::complex<double> c(0.0, 0.0);\n"
+			cpp += "\t\tdouble c = 0.0;\n"
 			if rhs_coeffs.getExpr().__class__ == meh.Addition:
 				ops = rhs_coeffs.getExpr().getOperands()
 				for op in ops:
@@ -1069,7 +1072,8 @@ def generate_stencil_code( pne, pni, stencil_name ):
 			else:
 				code = generate_cpp_recursive(rhs_coeffs.getExpr());
 				cpp += "\t\tc+={};\n".format(code)
-			cpp += "\t\tctx.coeff_b( {} ) += c.real();\n".format(coeff_index)
+			#cpp += "\t\tctx.coeff_b( {} ) += c.real();\n".format(coeff_index)
[…]

### 2018-02-21 cpp/testrenderer/src/volumes/fieldvolume.cpp
updating notebooks

@@ -41,13 +41,13 @@ namespace volumes
 
 		virtual double evalPhase( const P3d& pWS, const V3d& wi, const V3d& wo )const override
 		{
-			P3d pLS = worldToLocal*pLS;
+			P3d pLS = worldToLocal*pWS;
 			return phase_field->eval(pWS, wi, wo);
 		}
 
 		virtual double samplePhase( const P3d& pWS, const V3d& wi, V3d& wo, double& pdf, RNGd& rng )const override
 		{
-			P3d pLS = worldToLocal*pLS;
+			P3d pLS = worldToLocal*pWS;
 			return phase_field->sample(pLS, wi, wo, pdf, rng);
 		}
 

### 2018-04-05 latex/egsr17-cgf/conclusion.tex
updated paper

@@
 We derived and presented the real-valued $P_N$-equations, along with a staggered grid solver for solving them.
+We showed, how to use the results in a rendering system and ran our solver for various standard problems for validation.
+
+We originally set out to understand how non-linear diffusion compares against $P_N$ for increasing order.
+Although the lack of higher moments makes the FLD solution very flat, its energy conserving nature and comparably small resource footprint make it a much better approach when compared to $P_N$, which becomes more and more intractable for higher moment count.
+This is especially true, when compromises in accuracy of the approximation are acceptable.
+
+The literature in other fields often states, that the $P_N$ method---when solved in normal form like we do---is able to deal with vacuum regions.
+This is misleading.
+The $P_N$-method in normal form indeed does not break down completely in the presence of vacuum like diffusion based methods do (due to $\sigma_t^{-1}$ in the diffusion coefficient).
+However, in the presence of vacuum, the condition number of the system matrix becomes infinite and does not converge either.
+Therefore $P_N$ based methods also require minimum thresholding of the extinction coefficient and offer no benefit for vacuum regions.
+
+Much more work needs to be done in order to make our method competitive to alternative solutions.
+We believe this can be made possible by employing a multigrid scheme for solving the linear system of equations.
+We implemented a multigrid solver, but did not find the expected performance improvements.
+This is possibly due to the coupling between coefficients within a voxel, which does not work well together with the smoothing step.
+We want to study this further in the future.
+
 Unique to our system is, that it uses a computer algebra representation of the equation to solve as input.
-Discretization in angular and spatial domain is done without any user interference, using manipulation passes on the representation.
+Discretization in angular and spatial domain is done using manipulation passes on the representation.
 The stencil code, which is used to propagate the system of linear equations, is automatically generated from the expression tree.
+This way, we can easily deal with complex coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand.
 
-This way, we can easily deal with complex coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand.
-Our approach helps especially for higher order, when the number of equations increase and become more complex.
+When researching the application of $P_N$ in other fields, we came across a rich variety of variations on the $P_N$-method, such as simplified $P_N$ ($SP_N$)~\cite{Ryan10}, filtered $P_N$ ($FP_N$)~\cite{Radice13}, diffusion-correction $P_N$ ($DP_N$)~\cite{Schaefer11} and least-squares $P_N$ ($LSP_N$)~\cite{Hansen14}.
+These variations have been introduced to address certain problems of the standard $P_N$-method, such as ringing artefacts, dealing with vacuum regions and general convergence.
+Our solver can be applied to any (potentially coupled) PDE and therefore can generate stencil code for all these variations by simply expressing the respective PDE's in our computer algebra representation and providing this as an input to our solver.
+This would allow an exhaustive comparison of all these methods and we consider this as future work.
 
-We further show, how to use the results in a rendering system and run our solver for various standard problems for validation.
-
-\begin{itemize}
-	\item Mention performance problems and improving of solver performance as future work.
-	\item mention vaccum convergence problem and the challenge to couple deterministic methods with surface rendering problems.
-	\item Apply our automated stencil generation to other problem domains where coupled PDE's are present.
-\end{itemize}
+Finally, since the approach of our solver is very generic, we also would like to explore its application to other simulation problems in computer graphics.

### 2018-04-05 latex/egsr17-cgf/introduction.tex
updated paper

@@
 
-Deterministic methods on the other hand are unpopular.
-This is because they suffer from discretization artifacts, produce biased results, cannot be coupled easily with surface rendering problems and are trickier to implement.
+Deterministic methods on the other hand are unpopular, because they suffer from discretization artifacts, produce biased results, cannot be coupled easily with surface rendering problems and are trickier to implement.
 However, their appeal lies in the fact that they have better performance for certain problems and produce a global solution across the whole domain.
@@
 The $P_N$-method is a deterministic method, which is popular in other fields like medical imaging and nuclear sciences, but has not been explored for applications in graphics yet.
-The purpose and main contribution of our paper is, to gain a solid understanding of its foundations and present a method for application in rendering.
+The purpose and main contribution of our paper is, to gain a solid understanding of its foundations and present a method for using it on problems in rendering.
 In particular, we present the following theoretical and practical contributions:
 \begin{itemize}
-	\item We derive and present the time-independent real-valued $P_N$-equations.
+	\item We derive and present the time-independent real-valued $P_N$-equations. To our knowledge, we are the first to derive and write down their compact form.
 	\item We introduce a staggered-grid solver, for which we generate stencil code automatically from a computer algebra representation of the $P_N$-equations.
@@
 It further allows our solver to be used for any (potentially coupled) partial differential equation, which results in a system of linear equations after discretization.
-	\item Finally, we compare the $P_N$-method for higher orders against flux-limited diffusion, and answer the long standing question about which approach should be preferred.
+	\item Finally, we compare the $P_N$-method for higher orders against flux-limited diffusion and ground truth Monte Carlo integration.
 \end{itemize}
@@
 In the next section, we will discuss related work and why it motivates our contribution.
-Section~\ref{sec:pnequations} introduces the real-valued $P_N$-equations, which we solve using our solver described in section~\ref{sec:pnsolver}.
-We explain in section~\ref{sec:rendering}, how the solution is being applied in a rendering application.
+In section~\ref{sec:discretized_rte} we revisit the deterministic approach to light transport simulation in participating media and outline the discretization using spherical harmonics.
+In section~\ref{sec:car} we introduce our computer algebra representation, which we required to derive the real-valued $P_N$-equations.
+This representation is also a key component of our solver, which we present in section~\ref{sec:pnsolver}.
+We explain, in section~\ref{sec:rendering}, how the solution is being applied in a rendering application.
 Our $P_N$-solver is run and compared against flux-limited diffusion for a set of standard problems in section~\ref{sec:results}.

### 2018-04-05 latex/egsr17-cgf/main-body.tex
updated paper

@@
 %Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media
-A Deterministic Method for Simulation of Light Transport in Participating Media based on Higher Order Spherical Harmonics
+$P_N$-Method for Multiple Scattering in Participating Media
 }
@@
 In this paper, we investigate the open question of wether higher orders provide a viable alternative to these two approaches.
-Increasing the order introduces a set of increasingly complex coupled partial differential equations, whose growing number and complexity make them very difficult to work with for higher order.
+Increasing the order introduces a set of increasingly complex coupled partial differential equations, whose growing number and complexity make them very difficult to work with.
 We introduce a computer algebra framework for representing and manipulating the underlying mathematical equations and use it to derive the time-independent real-valued $P_N$-equations for arbitrary orders.
@@
 %-------------------------------------------------------------------------
-\input{pn_equations.tex}
+\input{section_discretized_rte.tex}
+\input{section_car.tex}
+\input{section_real_valued_pn_equations.tex}
 

### 2018-04-05 latex/egsr17-cgf/previous_work.tex
updated paper

@@
 
+\begin{figure}[h]
+\centering
+\begin{subfigure}{0.45\columnwidth}
+%\missingfigure{test}
+\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_collocated.png}
+\end{subfigure}%
+\hspace{0.05\columnwidth}
+\begin{subfigure}{0.45\columnwidth}
+%\missingfigure{test2}
+\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_staggered.png}
+\end{subfigure}%
+\vspace{-0.1in}
+\icaption{Solving the 2D-checkerboard problem using naive collocated grids produces oscillating artefacts (left). Our solver uses staggered grids and produces artefact free results (right).}
+\label{fig:artifacts}
+\end{figure}
+
 Similar to $P_N$, the classical diffusion approximation (CDA) is a deterministic method, which arrives at a solution by solving a system of linear equations.
@@
 
-The idea behind VEF is to replace the diffusion coefficient in the diffusion equation of CDA by a non-linear diffusion coefficient.
-This new diffusion coefficient is being derived by seeing it as a mean to spatially blend between the diffusive and pure transport regime.
+The idea behind VEF is to derive a better diffusion coefficient by seeing it as a mean to spatially blend between the diffusive and pure transport regime.
 The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
@@
 It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion.
-This question is also been raised in other domains~\cite{Olson00} and resolving it is the main motivation for this paper.
+It is also been raised in other domains~\cite{Olson00} and resolving it is one motivation for this paper.
 

### 2018-04-05 latex/egsr17-cgf/section_car.tex
updated paper

--- 2018-02-21 latex/egsr17-cgf/solver.tex
@@
-\section{$P_N$-Solver}
-\label{sec:pnsolver}
-
-Expanding, discretizing and working with the $P_N$-equations gets increasingly complex for higher order and is arduous and error prone if done by hand.
-We tackle this problem by expressing the RTE in a computer algebra representation and apply the discretization in angular and spatial domain using manipulation passes on the representation.
-
-The discretized $P_N$-equations are then rendered into source code and compiled.
-We refer to this as the stencil code.
-It allows fast construction of the linear system $A\vec{u}=\vec{Q}$ for arbitrary resolution and RTE parameters and has no dependency on the computer algebra framework anymore.
-
-Finally, the generated linear system is solved using standard methods.
-The solution $\vec{u}$ represents the discretized radiance field $\widehat{L}$, which is evaluated using trilinear interpolation of the SH coefficients.
-An overview of the solver is given in figure~\ref{fig:pnsolver}.
-Each step will be described in more detail in the following subsections.
-
-\subsection{Computer Algebra Representation}
+\section{Computer Algebra Representation}
+\label{sec:car}
 
@@
 
-\begin{figure}[h]
-\centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-\vspace{-0.2in}
-\icaption{The computer algebra representation is constructed using API commands provided by our framework (top left) and represented internally as an expression tree (top right).
-Additional API commands (bottem left) allow valid transformations, which modify the expression tree accordingly (bottom right).}
-\end{figure}
-
 Finally, frontends allow rendering the expression tree into different forms.
-We implemented frontends for rendering expression trees to \LaTeX and C++ source code.
+We implemented frontends for rendering expression trees to \LaTeX~and C++ source code.
 The equations in the supplemental material were almost all rendered by the former.
@@
 Note that our framework is different from a computer algebra system (CAS), in that our framework only deals with representation and manipulation of expressions, and is not concerned with finding answers to mathematical problems.
-
-\subsection{Discretization}
-
-The result of the discretization in angular domain are the real-valued $P_N$-equations.
-We arrived at the consice form in equations~\ref{eq:rpn_m_<_z}-~\ref{eq:rpn_m_>_z} using our framework.
-However, once the compact form was found, we used it directly as input to our solver.
-
-The discretization code iterates over all $l,m$-pairs up to order $N$ and chooses either equation~\ref{eq:rpn_m_<_z}, equation~\ref{eq:rpn_m_=_z} or equation~\ref{eq:rpn_m_>_z}, according to the sign of $m$.
-All occurances of $l$ and $m$ are replaced with their actual values.
-
-The spatial discretization is done by parsing the expression tree of each equation from the root.
-All occurances of the continuous positional variable $\vec{x}$ are replaced by discrete voxel coordinates $i,j,k$, whose values are retrieved from the top of a stack.
-The stack is kept by the parser and is initialized with the voxel at $(0,0,0)$.
-We refer to the space of the discretized $P_N$-equations as stencil space.
-
-Whenever the parser encounters a differential operator, the current voxel position is shifted and pushed to the stack.
-Then the child expression is parsed and multiplied with a weight value, followed by popping the stack.
-This is done for each element of a central difference stencil, which provides the offset and weight values (referencing the voxelsize variable).
-The dimension of the stencil is given by the derivation variable of the operator.
-Nested derivatives are handled naturally and produce higher order finite difference stencils as expected.
-
-Another aspect to consider during discretization is the location of the variables and unknowns.
-As we will see later, our solver needs to support staggered grids, where the SH coefficients are not all located at the voxel center (collocated).
-Some also can be located at the faces, edges and intersections of voxels (staggered).
-
-Our solver therefore supports placement of coefficients at arbitrary staggered grid locations.
-Whenever a coefficient is encountered during expression tree parsing for discretization, the parser checks how its location is related to the current location on the top of the stack.
-Depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at the requested offset from its surrounding locations, or it returns the coefficient itself, if it happens to coincide.
-This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center.
-The position stack of the parser is initialized with the staggered grid location defined by the coefficient, which is associated with the $l,m$-pair of the equation currently being discretized.
-
-\subsection{Stencil Code Generation and System Building}
-
-After applying the discretization step to the expression tree of the $P_N$-equations, it is used to generate the stencil code.
-In numerical analysis, a stencil is an arrangement of voxels and weights, that relate values at different locations to each other and form the basis for propagating rows in the system matrix $A$ and RHS vector $\vec{Q}$ with values.
-The name comes from the fact, that the geometric structure and weights of the configuration do not change, when applied to different voxels.
-The same is true for the $P_N$-equations and we use this fact to generate a single function, which propagates rows in $A$ and $\vec{Q}$ for a given voxel.
-
-The $P_N$-equations express, how each coefficient of a voxel depends on other coefficients of the same or adjacent voxels.
-The unknowns in the terms give information about the coefficient index and voxel offset, and therefore identify a column offset in the matrix $A$.
-The factors to these coefficients may contain evaluations of RTE parameters, such as $\sigma_t$.
-Therefore, these factors can not be determined during stencil generation, but are rendered into code expressions, which are executed as part of the stencil function during runtime.
-Because the stencil code has been generated in stencil space relative to the voxel at $(0,0,0)$, we can run the same stencil code for every voxel, by simply applying an offset accordingly.
-
-The $P_N$-equations have been discretized, and the stencil has been generated, without any notion of domain boundaries and boundary conditions (BC).
-Our solver supports Neumann BC and Dirichlet BC.
-They are handled transparently by the code which runs the stencil.
-Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or write into the row and column in $A$ of the same coefficient at the closest voxel inside the domain (Neumann BC).
-
-In order to respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels (see Figure~\ref{fig:staggeredgrid}).
-This requires careful managegment and bookkeeping of coefficient indices, which is done transparently by the code running the stencil.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}{0.45\columnwidth}
-%\missingfigure{test}
-\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_collocated.png}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-%\missingfigure{test2}
-\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_staggered.png}
-\end{subfigure}%
-\vspace{-0.1in}
-\icaption{The solution of our solver for the 2D-checkerboard problem with collocated grids (left), where all coefficients are located at the cell centers.
-It suffers from severe oscillating artifacts, which are removed by offsetting the coefficient locations using staggered grids (right).}
-\label{fig:artifacts}
-\end{figure}
-
-\subsection{Solving}
-
-After running the stencil code for every voxel, the system matrix $A$
-and RHS vector $\vec{Q}$ are propagated with values and can be used with standard methods for solving systems of linear equations.
-The number of rows is determined by the number of voxels times the number of coefficients per voxel and can therefore become very large for high resolution and high truncation order.
-The matrix $A$ is squared and fortunately very sparse, due to finite differences and the structure of the $P_N$-equations.
-Unfortunately, it also is non-symmetric (due to the transport term) and not diagonal dominant, which rules out many standard methods for solving linear systems.
-We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead.
-This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-\vspace{-0.2in}
-\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
-\label{fig:staggeredgrid}
-\end{figure}
-
-%Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.
-
-%\subsection*{CDA vs. $P_1$}
-%\begin{itemize}
-  %\item CDA is a degenerated form of $P_1$.
-It is derived by isolating the flux-vector on one side of the vector-equation formed by the $l=1$ SH-band equations.
-This isolation requires division by the extinction coefficient, introducing a $\frac{1}{\sigma_t}$ factor which diverges as $\sigma_t$ approaches zero.
-Thresholding to some minimum is required, in order to be able to solve the system for vacuum regions.
-  %\item $P_1$ does not require any thresholding of $\sigma_t$ as it does not contain $\sigma_t$ as a denominator. It therefore can deal with vacuum regions without modifications.
-  %\item (needs validation) Further, in the presence of vacuum or near vacuum regions, the condition number for CDA is higher than for $P_1$, because of small extinction values in the denominator of the diffusion coefficient.
-  %\item Using the normal form for CDA will further increase the condition number when vacuum regions are present and significantly decreases convergence.
-%\end{itemize}

### 2018-04-05 latex/egsr17-cgf/section_discretized_rte.tex
updated paper

--- 2018-02-07 latex/egsr17-cgf/pn_equations.tex
@@
-\section{Real-valued $P_N$-Equations}
-\label{sec:pnequations}
+\section{Discretized Radiative Transfer Equation}
+\label{sec:discretized_rte}
 
-In this section we present how to optain the real-valued $P_N$-equations.
-For brevity and readability, we only give a brief outline here and refer to the supplemental material for the detailed derivation.
-
-The $P_N$-equations are derived from the radiative transfer equation, which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
+Our method is derived from the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
 \begin{align}
@@
 +Q\left(\vec{x}, \omega\right)\nonumber
+\label{eq:rte}
 \end{align}
@@
 
-The $P_N$-method expands the angular dependence into spherical harmonics.
+Deterministic methods, like ours, are derived by discretizing the angular and spatial domain.
+This gives rise to a linear system of equations, which can be solved using standard methods.
+For the $P_N$-method, the angular variable is discretized first, using a truncated spherical harmonics expansion.
+This results in the $P_N$-equations, a system of coupled PDE's, which still have a continuous spatial variable.
+
+The number of equations grows with the truncation order $N$.
+This is why the discretization is hard and error prone if done by hand.
+We therefore use a computer algebra representation to automate this process.
+After giving an outline of the general discretization in this section, we will present our computer algebra representation in the next section.
+The $P_N$-equations, as result of our automated discretization, are given in section~\ref{sec:real_valued_pn_eq}.
+
 Since the radiance field $L$ is real only, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
-\begin{align*}
+\begin{align}
 \SHBR^{l,m}=
@@
 \right.
-\end{align*}
+\label{eq:sh_real_basis}
+\end{align}
 
@@
 
-The $P_N$-equations are derived by first replacing the radiance field $L$ by its SH reconstruction $\widehat{L}$, which introduces an error due to truncation at order $N$:
-\begin{align*}
+The $P_N$-equations are derived by first expressing all angular dependent parameters in spherical harmonics.
+The radiance field $L$ in equation~\ref{eq:rte} is therefore replaced by its SH reconstruction $\widehat{L}$, introducing an error due to truncation at order $N$:
+\begin{align}
 \widehat{L}\left(\vec{x}, \omega\right) =
@@
 L\left(\vec{x}, \omega\right)
-\end{align*}
+\nonumber
+\end{align}
 
-After similarly replacing all angular dependent RTE parameters, we project each term into SH, using the projection operator $\mathcal{P}$.
+After substitution, all angular parameters are expressed in terms of spherical harmonics, but still depend on the continuous angular variable $\omega$.
+As a next step, we project each term of the RTE into SH, using the projection operator $\mathcal{P}$.
 This produces a single equation for each $l,m$-pair.
@@
 +\mathcal{P}^{l,m}\left(Q\right)
-\nonumber
+\label{eq:pn_operator_notation}
 \end{align}
 
-Expanding the operators and writing the real-valued $P_N$-equations in compact form, requires a string of expansions, manipulations and applications of identities.
-We here give the final result and refer to the supplemental material for the detailed derivation.
-Since the real-valued SH basis have different definitions, depending on $m<0$, $m=0$ or $m>0$, we get different projections $\mathcal{S}^{l,m}$ depending no the sign of $m$.
+Once the $P_N$-equation have been found, the spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxelgrid (using central differences for differential operators).
 
-The real-valued $P_N$-equations are for $m<0$:
-\begin{align}
-&-\frac{1}{2}c^{\scaleto{l-1,m-1}{4pt}}
-\partial_y
-L^{\scaleto{l-1,-m+1}{4pt}}
-%\\
-+\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
-\partial_y
-L^{\scaleto{l+1,-m+1}{4pt}}
-%\\
--\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m+1}{4pt}}
-\partial_y
-L^{\scaleto{l-1,-m-1}{4pt}}
-\nonumber
-\\&
-+\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{\scaleto{l+1,m+1}{4pt}}
-\partial_y
-L^{\scaleto{l+1,-m-1}{4pt}}
-%\\
-+\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}c^{\scaleto{l-1,m-1}{4pt}}
-\partial_x
-L^{\scaleto{l-1,m-1}{4pt}}
-\nonumber
-\\&
--\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}e^{{l-1,m+1}}
-\partial_x
-L^{\scaleto{l-1,m+1}{4pt}}
-%\\
-+\frac{1}{2}f^{\scaleto{l+1,m+1}{4pt}}
-\partial_x
-L^{\scaleto{l+1,m+1}{4pt}}
-%\\
--\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
-\partial_x
-L^{\scaleto{l+1,m-1}{4pt}}
-\nonumber
-\\&
-+a^{\scaleto{l-1,m}{4pt}}
-\partial_z
-L^{\scaleto{l-1,m}{4pt}}
-%\\
-+b^{\scaleto{l+1,m}{4pt}}
-\partial_z
-L^{\scaleto{l+1,m}{4pt}}
-%\\
-+\sigma_t L^{\scaleto{l,m}{4pt}}
-%\\
--\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
-%\\
-= Q^{\scaleto{l,m}{4pt}}
-\label{eq:rpn_m_<_z}
-\end{align}
-
-for $m=0$:
-\begin{align}
-&
-\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_x L^{\scaleto{l-1,1}{4pt}}
--\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_x L^{\scaleto{l+1,1}{4pt}}
-%\\&
-\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_y L^{\scaleto{l-1,-1}{4pt}}
-\nonumber
-\\&
--\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_y L^{\scaleto{l+1,-1}{4pt}}
-%\\&
-a^{\scaleto{l-1,0}{4pt}}\partial_z L^{\scaleto{l-1,0}{4pt}}
-+b^{\scaleto{l+1,0}{4pt}}\partial_z L^{\scaleto{l+1,0}{4pt}}
-\nonumber
-\\&
-+\sigma_t L^{\scaleto{l,m}{4pt}}
--\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
-= Q^{\scaleto{l,m}{4pt}}
-\label{eq:rpn_m_=_z}
-\end{align}
-
-and for $m>0$:
-\begin{align}
-&
-\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_x L^{\scaleto{l-1,m+1}{4pt}}
-%\\&
--\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_x L^{\scaleto{l+1,m+1}{4pt}}
-%\\&
--\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m-1}{4pt}}\partial_x L^{\scaleto{l-1,m-1}{4pt}}
-\nonumber
-\\&
-+\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{l+1,-m+1}\partial_x L^{\scaleto{l+1,m-1}{4pt}}
-%\\&
-+\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_y L^{\scaleto{l-1,-m-1}{4pt}}
-\nonumber
-\\&
--\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_y L^{\scaleto{l+1,-m-1}{4pt}}
-%\nonumber
-%\\&
-+\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}e^{\scaleto{l-1,-m+1}{4pt}}\partial_y L^{\scaleto{l-1,-m+1}{4pt}}
-\nonumber
-\\&
--\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}f^{\scaleto{l+1,-m+1}{4pt}}\partial_y L^{\scaleto{l+1,-m+1}{4pt}}
-%\\&
-+a^{\scaleto{l-1,-m}{4pt}}\partial_z L^{\scaleto{l-1,m}{4pt}}
-%\nonumber
-%\\&
-+b^{\scaleto{l+1,-m}{4pt}}\partial_z L^{\scaleto{l+1,m}{4pt}}
-\nonumber
-\\&
-+\sigma_t L^{\scaleto{l,m}{4pt}}
--\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
-= Q^{\scaleto{l,m}{4pt}}
-%\nonumber
-\label{eq:rpn_m_>_z}
-\end{align}
-
-with
-\begin{align*}
-\label{eq:real_sh_basis}
-\beta^{x}=
-\left\{
-\begin{array}{lr}
-\frac{2}{\sqrt{2}}, & \text{for } \vert x\vert = 1\\
-1, & \text{for } \vert x\vert \neq 1
-\end{array}
-\right.
-,\quad
-\delta_{x\neq y}=
-\left\{
-\begin{array}{lr}
-1, & \text{for } x \neq y \\
-0, & \text{for } x = y
-\end{array}
-\right.
-\end{align*}
-
-and
-\begin{align*}
-&
-a^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}} \qquad
-b^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
-\\&
-c^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \qquad
-d^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
-\\&
-e^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \qquad
-f^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
-\end{align*}
-\begin{align*}
-\lambda_l=\sqrt{\frac{4\pi}{2l+1}}
-\end{align*}
-
-\begin{figure}[h]
-\centering
-\missingfigure{test}
-\vspace{-0.2in}
-\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
-\end{figure}
-
-\begin{figure*}[t]
-\centering
-\missingfigure{$P_N$-solver overview: generate stencil code $\rightarrow$ build system $\rightarrow$ solve $\rightarrow$ render}
-\vspace{-0.2in}
-\icaption{Overview of our $P_N$-solver.
-After generating the stencil source code from the expression trees representing the $P_N$-equations, the linear system $A\vec{u}=\vec{Q}$ is built using RTE parameter fields and additional user input, such as grid resolution and type of boundary conditions.
-The resulting system is solved for $\vec{u}$, which is then used in our rendering application.}
-\label{fig:pnsolver}
-\end{figure*}
-
-The spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxelgrid.
-The radiance field $L$, after full discretization, is represented as a set of coefficients per voxel.
-Flattening these over all voxels into a single vector gives the solution vector $\vec{u}$.
+The radiance field $L$, after full discretization, is represented as a set of SH coefficients per voxel.
+Flattening these over all voxels into a single vector, gives the solution vector $\vec{u}$.
 The RHS vector $\vec{Q}$ is produced similarly.
-The differential operators are discretized using central differences.
-This allows the projected operators to be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$:
+The projected operators can be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$ (see figure~\ref{fig:matrix_layout}):
 \begin{align}
@@
 
-$T$, $C$, $S$ are matrices, which result from the discretized transport, collision and scattering operators in the $P_N$-equations respectively.
+$T$, $C$, $S$ are matrices, which result from the discretized transport, collision and scattering operators respectively.
 
-In the next section, we will present our solver for solving the real-valued $P_N$-equations.
+\begin{figure}[h]
+\centering
+%\missingfigure{test}
+\includegraphics[width=\columnwidth]{figures/fig_matrix_layout.pdf}
+\vspace{-0.2in}
+\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
+\label{fig:matrix_layout}
+\end{figure}
+
+So far we have given the $P_N$-equations only in high-level operator notation (equation~\ref{eq:pn_operator_notation}).
+Carrying out the full derivation creates large, unwieldy equations and requires a string of expansions, manipulations and applications of identities.
+These are hard to do and error prone if done by hand.
+We therefore developed a computer algebra representation, which allowed us to derive the the $P_N$-equation in a semi-automatic fashion.
+In the next section, we will present our computer algebra representation and give the final result of the angular discretization in section~\ref{sec:real_valued_pn_eq}.

### 2018-04-05 latex/egsr17-cgf/section_results.tex
updated paper

--- 2018-02-07 latex/egsr17-cgf/results.tex
@@
 \centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
+\begin{subfigure}{0.49\columnwidth}
+%\missingfigure{test}
+\includegraphics[width=\columnwidth]{images/checkerboard2d_p5_neumann_staggered_starmap.png}
 \end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
+\hspace{0.01\columnwidth}
+\begin{subfigure}{0.49\columnwidth}
+%\missingfigure{test2}
+\includegraphics[width=\columnwidth]{images/checkerboard2d_p5_neumann_staggered.png}
 \end{subfigure}%
-\vspace{-0.2in}
-\icaption{Comparison of the result for the checkerboard test using Starmap~\cite{Seibold14} (left) against ours (right). Our result is in excellent agreement with the reference.}
+\vspace{-0.1in}
+\icaption{Comparison of the result for the checkerboard test using Starmap~\cite{Seibold14} (left) against ours (right).
+Our result is in excellent agreement with the reference.
+We attribute the difference at the boundary to the lack of convergence of the Starmap solve.}
 \label{fig:vs_starmap}
@@
 We also run our solver for the point source problem, a single point light in a homogeneous medium.
-This not only helps to validate our implementation for the 3D case against an analytical ground truth, but also allows to get a clear assessment of the methods accuracy.
+This not only helps to validate our implementation for the 3D case, but also allows to get a clear assessment of the methods accuracy.
+We use the Grosjean approximation, which was introduced by D'Eon et al.\cite{dEon11}, as a very accurate approximation to the ground truth solution.
 
@@
 \begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
+\includegraphics[width=\columnwidth]{figures/pointsource_pn.pdf}
+%\missingfigure{test}
+\caption{$P_N$ vs. ground truth}
+\label{fig:pointsource_pn}
 \end{subfigure}%
@@
 \begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
+%\missingfigure{test2}
+\includegraphics[width=\columnwidth]{figures/pointsource_p5.pdf}
+\caption{$P_5$ vs. CDA and FLD}
+\label{fig:pointsource_p5}
 \end{subfigure}%
-\vspace{-0.2in}
+\vspace{-0.1in}
 \icaption{Lineplot through the 3D solution of our solver for the point source problem for various order $N$ (left).
-Solution for $P_5$ compared against classical diffusion and flux-limited diffusion (right).}
+Solution for $P_5$ compared against classical diffusion, flux-limited diffusion and analytical solution (right).}
 \label{fig:pointsource}
@@
 
-For our test case we choose a FD resolution of $80\times80\times80$, an extinction coefficient $\sigma_t=8.0$ and albedo $\alpha=0.9$.
-We run the solver for different truncation values $N$ and in figure~\ref{fig:pointsource}, we see that the solution becomes increasingly accurate for higher truncation order.
+For our test case, we choose a FD resolution of $80\times80\times80$, an extinction coefficient $\sigma_t=8.0$ and albedo $\alpha=0.9$.
+We run the solver for different truncation values $N$.
+In figure~\ref{fig:pointsource}~\subref{fig:pointsource_pn}, we see that the solution becomes increasingly accurate for higher truncation order.
 The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even.
-We further see that the $P_1$ solution exactly matches the results from CDA, which makes sense, as the latter is only a collapsed version of the former.
-
-\begin{itemize}
-	\item performance/computation time
-\end{itemize}
+We further see, that the $P_1$ solution exactly matches the results from CDA, which confirmes, that the latter is only a collapsed version of the former.
 

### 2018-04-05 latex/egsr17-cgf/solver.tex
updated paper

--- 2018-02-07 latex/egsr17-cgf/solver.tex
@@
 
-Expanding, discretizing and working with the $P_N$-equations gets increasingly complex for higher order and is arduous and error prone if done by hand.
-We tackle this problem by expressing the RTE in a computer algebra representation and apply the discretization in angular and spatial domain using manipulation passes on the representation.
+The truncation order $N$ is the key input parameter to the solver.
+With higher values, manually coding the solver from the equations would be ardous, error prone and very time consuming.
+We therefore decided to make use of the equations computer algebra representation and designed our solver around it.
 
-The discretized $P_N$-equations are then rendered into source code and compiled.
-We refer to this as the stencil code.
-It allows fast construction of the linear system $A\vec{u}=\vec{Q}$ for arbitrary resolution and RTE parameters and has no dependency on the computer algebra framework anymore.
+The solver consists of two components.
+The first is a precomputation (section~\ref{sec:solver_precomputation}), which is done once for every single value of $N$.
+This steps runs a partial evaluation on the mathematical expression tree and applies the spatial discretization in a reference space we call stencil space.
+The precomputation step automatically generates source code from the resulting expression tree.
 
-Finally, the generated linear system is solved using standard methods.
-The solution $\vec{u}$ represents the discretized radiance field $\widehat{L}$, which is evaluated using using trilinear interpolation of the SH coefficients.
+The generated stencil code is compiled with the runtime component (section~\ref{sec:solver_runtime}) of our solver.
+This component gets the actual problem as input, including gid resolution and RTE parameter fields.
+It builds the linear system and solves it using standard methods.
 An overview of the solver is given in figure~\ref{fig:pnsolver}.
-Each step will be described in more detail in the following subsections.
 
-\subsection{Computer Algebra Representation}
+\begin{figure*}[t!]
+\centering
+%\missingfigure{$P_N$-solver overview: generate stencil code $\rightarrow$ build system $\rightarrow$ solve $\rightarrow$ render}
+\includegraphics[width=\textwidth]{figures/fig_pipeline.pdf}
+\vspace{-0.2in}
+\icaption{Overview of our $P_N$-solver.
+After generating the stencil source code from the expression trees representing the $P_N$-equations, the linear system $A\vec{u}=\vec{Q}$ is built using RTE parameter fields and additional user input, such as grid resolution and type of boundary conditions.
+The resulting system is solved for $\vec{u}$, which is then used in our rendering application.}
+\label{fig:pnsolver}
+\end{figure*}
 
-Our computer algebra representation is based on mathematical expression trees, hierarchies composed out of expression instances.
-Those can be basic primitives, such as numbers, variables or special symbols, such as the Kronecker delta or imaginary number.
-Other expression types, such as integrals, derivatives, sums, products and functions build the hierarchy by referencing child expressions.
+\subsection{Precomputation}
+\label{sec:solver_precomputation}
 
-Our framework further provides manipulators, which can be executed on expressions.
-We implemented all manipulations, which were required for the derivation of the $P_N$-equations.
-This includes operations, such as application of the distributive law, substitution, constant folding, reordering of nested integrals, application of identities up to more complex operations, such as factorization of unknowns and discretization of spatial variables.
-The derivation steps of the real-valued $P_N$-equations in the supplemental material were all carried out using our framework.
+The result of the precomputation is a stencil, which can be used during runtime to build the linear system for a given problem.
+The stencil is a pattern of indices into the solution vector, along with some weights.
+It expresses how the sum of the weighted solution vector components relate to the RHS for a given unknown in the system and therefore contains most information required to propagate a row in the system matrix $A$ and RHS-vector $\vec{Q}$.
+The index pattern does not change for different rows, giving the stencil its name.
 
-\begin{figure}[h]
-\centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
+A stencil generally is crafted by discretizing a PDE at a hypothetical center voxel $ijk$ (assuming to mean the voxel center most of the time).
+Finite differences create weighted references to other voxels (e.g.
+$i+1,jk$).
+After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and $ijk$-offsets.
+$ijk$ will only be known during runtime, when the stencil is executed for a particular unknown (row).
+Then the offsets can be used to find the component index into the solution-vector and weights can be evaluated for concrete world space position.
+We refer to the space with the hypthetical voxel $ijk$ at the center as stencil space.
 
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-\vspace{-0.2in}
-\icaption{The computer algebra representation is constructed using API commands provided by our framework (top left) and represented internally as an expression tree (top right).
-Additional API commands (bottem left) allow valid transformations, which modify the expression tree accordingly (bottom right).}
-\end{figure}
-
-Finally, frontends allow rendering the expression tree into different forms.
-We implemented frontends for rendering expression trees to \LaTeX and C++ source code.
-The equations in the supplemental material were almost all rendered by the former.
-The latter was used for generating the stencil code used by our solver.
-
-Note that our framework is different from a computer algebra system (CAS), in that our framework only deals with representation and manipulation of expressions, and is not concerned with finding answers to mathematical problems.
-
-\subsection{Discretization}
-
-The result of the discretization in angular domain are the real-valued $P_N$-equations.
-We arrived at the consice form in equations~\ref{eq:rpn_m_<_z}-~\ref{eq:rpn_m_>_z} using our framework.
-However, once the compact form was found, we used it directly as input to our solver.
-
-The discretization code iterates over all $l,m$-pairs up to order $N$ and chooses either equation~\ref{eq:rpn_m_<_z}, equation~\ref{eq:rpn_m_=_z} or equation~\ref{eq:rpn_m_>_z}, according to the sign of $m$.
-All occurances of $l$ and $m$ are replaced with their actual values.
+Typically, the stencil is coded by hand, which would be challenging with our $P_N$-equations and higher values of $N$.
+In our system, the stencil code is created automatically from the computer algebra representation.
+Spatial discretization and fatorization into canonical form are done as manipulation passes on the mathematical expression tree.
+The precomputation code iterates over all terms and analyzes the unkown to get the $ijk$-offset.
+The factor-expression is extracted from the tree and used to generate source code for evaluating the factor-expression during runtime (including calls for evaluating RTE-parameter fields).
 
@@
 All occurances of the continuous positional variable $\vec{x}$ are replaced by discrete voxel coordinates $i,j,k$, whose values are retrieved from the top of a stack.
-The stack is kept by the parser and is initialized with the voxel at $(0,0,0)$.
-We refer to the space of the discretized $P_N$-equations as stencil space.
+The stack is kept by the parser and is initialized with the voxel at $ijk=(0,0,0)$.
 
@@
 As we will see later, our solver needs to support staggered grids, where the SH coefficients are not all located at the voxel center (collocated).
-Some also can be located at the faces, edges and intersections of voxels (staggered).
+Coefficients can also be located at face-centers, edges and intersections of voxels (staggered).
 
-Our solver therefore supports placement of coefficients at arbitrary staggered grid locations.
+Our solver supports placement of coefficients at arbitrary staggered grid locations.
 Whenever a coefficient is encountered during expression tree parsing for discretization, the parser checks how its location is related to the current location on the top of the stack.
@@
 
-\subsection{Stencil Code Generation and System Building}
+%\subsection{Stencil Code Generation}
 
-After applying the discretization step to the expression tree of the $P_N$-equations, it is used to generate the stencil code.
+% After applying the discretization step to the expression tree of the $P_N$-equations, it is used to generate the stencil code.
 In numerical analysis, a stencil is an arrangement of voxels and weights, that relate values at different locations to each other and form the basis for propagating rows in the system matrix $A$ and RHS vector $\vec{Q}$ with values.
@@
 
-The $P_N$-equations express, how each coefficient of a voxel depends on other coefficients of the same or adjacent voxels.
+%The $P_N$-equations express, how each coefficient of a voxel depends on other coefficients of the same or adjacent voxels.
 The unknowns in the terms give information about the coefficient index and voxel offset, and therefore identify a column offset in the matrix $A$.
@@
 
-The $P_N$-equations have been discretized, and the stencil has been generated, without any notion of domain boundaries and boundary conditions (BC).
-Our solver supports Neumann BC and Dirichlet BC.
-They are handled transparently by the code which runs the stencil.
-Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or return the row and column in $A$ for the same coefficient of the closest voxel inside the domain (Neumann BC).
+%\subsection{System Building and Solving}
+\subsection{Runtime}
+\label{sec:solver_runtime}
 
-In order to respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels (see Figure~\ref{fig:staggeredgrid}).
-This requires careful managegment and bookkeeping of coefficient indices, which is done transparently by the code running the stencil.
+The generated stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver.
+The runtime executes the stencil for every voxel to propagate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.
 
-\begin{figure}[h]
-\centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-\vspace{-0.2in}
-\icaption{The solution of our solver for the 2D-checkerboard problem with collocated grids (left), where all coefficients are located at the cell centers.
-It suffers from severe oscillating artefacts, which are removed by offseting the coefficient locations using staggered grids (right).}
-\label{fig:artefacts}
-\end{figure}
-
-\subsection{Solving}
-
-After running the stencil code for every voxel, the system matrix $A$
-and RHS vector $\vec{Q}$ are propagated with values and can be used with standard methods for solving systems of linear equations.
-The number of rows is determined by the number of voxels times the number of coefficients per voxel and can therefore become very large for high resolution and high truncation order.
+The number of rows is determined by the number of voxels times the number of coefficients per voxel (see figure~\ref{fig:matrix_layout}) and can therefore become very large for high resolution and high truncation order.
 The matrix $A$ is squared and fortunately very sparse, due to finite differences and the structure of the $P_N$-equations.
@@
 
+The $P_N$-equations have been discretized, and the stencil has been generated, without any notion of domain boundaries and boundary conditions (BC).
+Our solver supports Neumann BC and Dirichlet BC.
+They are handled transparently by the code which runs the stencil.
+Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or write into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC).
+
+In order to respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels (see Figure~\ref{fig:staggeredgrid}).
+This requires careful managegment and bookkeeping of coefficient indices, which is done transparently by the runtime code.
+
 \begin{figure}[h]
 \centering
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test}
-\end{subfigure}%
-\hspace{0.05\columnwidth}
-\begin{subfigure}{0.45\columnwidth}
-\missingfigure{test2}
-\end{subfigure}%
-\vspace{-0.2in}
-\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
+\includegraphics[width=0.5\columnwidth]{figures/fig_staggered_grids.pdf}
+%\vspace{-0.2in}
+\icaption{Staggered grids place coefficients at different locations within the finite difference grid. Additional unknowns are required in boundary cells to produce a correct result.}
 \label{fig:staggeredgrid}
 \end{figure}
+
+%\begin{figure}[h]
+%\centering
+%\begin{subfigure}{0.45\columnwidth}
+%%\missingfigure{test}
+%\includegraphics[width=\columnwidth]{figures/fig_staggered_grids.pdf}
+%\end{subfigure}%
+%\hspace{0.05\columnwidth}
+%\begin{subfigure}{0.45\columnwidth}
+%\missingfigure{test2}
+%\end{subfigure}%
+%\vspace{-0.2in}
+%\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
+%\label{fig:staggeredgrid}
+%\end{figure}
 

### 2018-04-10 latex/egsr17-cgf/introduction.tex
final version

@@
 
-Simulating light transport in paticipating media is still a challenging problem for image synthesis in computer graphics.
-Monte Carlo based techniques are established as the standard approach, because they can produce unbiased results and are conceptionally easy to implement.
-They become computationally demanding however, for participating media, which exhibits strong scattering or anisotropy.
+Simulating light transport in paticipating media remains a challenging problem for image synthesis in computer graphics.
+Due to their ability to produce unbiased results and conceptional simplicity, Monte Carlo based techniques have become the standard approach.
+The main downside of these methods are their computational demands when rendering participating media, which exhibits strong scattering or anisotropy.
 
-Deterministic methods on the other hand are unpopular, because they suffer from discretization artifacts, produce biased results, cannot be coupled easily with surface rendering problems and are trickier to implement.
-However, their appeal lies in the fact that they have better performance for certain problems and produce a global solution across the whole domain.
+Deterministic methods have enjoyed less popularity, because they suffer from discretization artifacts, produce biased results, cannot be coupled easily with surface rendering problems and are trickier to implement.
+However, their appeal lies in the fact that they produce a global solution across the whole domain and have better performance for certain problems.
 
-The work on path guiding techniques from recent years has shown, how discretized approximate representations of the correct solution can be used to accelerate Monte Carlo integration techniques, such as path tracing.
-Instead of generating these approximate representations using correct methods, deterministic methods may offer a viable alternative and pave the way for hybrid methods, which combine the performance benefits of deterministic methods with correct and exact Monte Carlo techniques.
-Further, deterministic methods lend themselfs to applications, where a fast approximate solution is often preferred over correct, but slowly converging results.
-Under this light, it is important to study deterministic methods and have a solid understanding about their characteristics and performance traits for typical rendering problems.
+The work on path guiding techniques from recent years has shown, how approximate representations of the steady-state transport in a scene can be used to accelerate Monte Carlo integration techniques, such as path tracing.
+Instead of generating these approximate representations using Monte Carlo methods, deterministic methods may offer a viable alternative.
+Hybrid methods could combine benefits of deterministic methods while generating unbiased results.
 
-The $P_N$-method is a deterministic method, which is popular in other fields like medical imaging and nuclear sciences, but has not been explored for applications in graphics yet.
-The purpose and main contribution of our paper is, to gain a solid understanding of its foundations and present a method for using it on problems in rendering.
-In particular, we present the following theoretical and practical contributions:
+Deterministic methods also lend themselves to applications where fast approximate solutions are preferable over correct, but slowly converging results.
+For this reason, we find it important to study deterministic methods and have a solid understanding about their characteristics and performance traits for typical rendering problems.
+
+The $P_N$-method is a deterministic method, which is popular in fields like medical imaging and nuclear sciences, but has not found use in computer graphics thus far.
+The purpose and main contribution of our paper is to gain a solid understanding of its foundations and present a method for using it in the context of rendering.
+In particular, we present these theoretical and practical contributions:
 \begin{itemize}
@@
 
-In the next section, we will discuss related work and why it motivates our contribution.
-In section~\ref{sec:discretized_rte} we revisit the deterministic approach to light transport simulation in participating media and outline the discretization using spherical harmonics.
-In section~\ref{sec:car} we introduce our computer algebra representation, which we required to derive the real-valued $P_N$-equations.
-This representation is also a key component of our solver, which we present in section~\ref{sec:pnsolver}.
-We explain, in section~\ref{sec:rendering}, how the solution is being applied in a rendering application.
-Our $P_N$-solver is run and compared against flux-limited diffusion for a set of standard problems in section~\ref{sec:results}.
-We finish with our conclusion in section~\ref{sec:conclusion}.
+In the next section, we will discuss related work and its relation to our contribution.
+In Section~\ref{sec:discretized_rte} we revisit the deterministic approach to light transport simulation in participating media and outline the discretization using spherical harmonics.
+In Section~\ref{sec:car} we introduce our computer algebra representation, which we required to derive the real-valued $P_N$-equations.
+This representation is also a key component of our solver, which we present in Section~\ref{sec:pnsolver}.
+Section~\ref{sec:rendering} discusses application of the solution to the context of rendering.
+We compare our $P_N$-solver against flux-limited diffusion for a set of standard problems in Section~\ref{sec:results}.
+Finally, Section~\ref{sec:conclusion} concludes with a summary and review of future work.

### 2018-04-10 latex/egsr17-cgf/main-body.tex
final version

@@
 {
-Compiled on \today
+%Compiled on \today
+paper1027
 %David Koerner$^{1,2,3}$
@@
 \vspace{-0.17in}
-\caption{Ground truth}
+\caption{Brute force path tracing}
 \label{fig:teaser_gt}
@@
 {
-Flux-limited diffusion (\subref{fig:teaser_fld}) is an ad-hoc extension to the classical diffusion approximation (\subref{fig:teaser_cda}).
-It can be considered the state of the art among deterministic methods for simulating light transport in highly scattering participating media, which is a common challenge for brute-force path tracing (\subref{fig:teaser_gt}).
+Flux-limited diffusion (\subref{fig:teaser_fld}) is an extension to the classical diffusion approximation (\subref{fig:teaser_cda}).
+It improves accuracy, but is based on ad-hoc assumptions about volumetric transport.
 We add the $P_N$-method (\subref{fig:teaser_pn}) to the toolbox of deterministic methods in rendering and investigate its benefits and tradeoffs against flux-limited diffusion.
@@
 We further present a staggered-grid $P_N$-solver and compare it against FLD for various standard problems.
-Rendering highly scattering participating media is a common challenge for brute-force path tracing.
-Classical diffusion approximation is a competitive alternative, as it reduces the problem to solving a simple diffusion equation.
-Extensions, such as flux-limited diffusion, improve the former by switching to a non-linear diffusion coefficient.
-Both methods are based on the spherical harmonics expansion of the radiance field, truncated after the first order.
-In this paper, we investigate the open question of wether higher orders provide a viable alternative to these two approaches.
+Rendering highly scattering participating media using brute force path tracing is a challenge.
+The diffusion approximation reduces the problem to solving a simple linear partial differential equation.
+Flux-limited diffusion introduces nonlinearities to alleviate the approximation error but introduces several ad-hoc assumptions.
+Both methods are based on the spherical harmonics expansion of the radiance field, that is truncated after the first order.
+In this paper, we investigate the open question of whether higher orders provide a viable alternative to these two approaches.
 Increasing the order introduces a set of increasingly complex coupled partial differential equations, whose growing number and complexity make them very difficult to work with.
@@
 Finally, we discuss how our method compares against prior work for various standard problems.
-We will release our computer algebra system, solver, and data as open source to ensure full reproducibility of all of our results.
+We will release our computer algebra system, solver, and data as open source to ensure reproducibility of all of our results.
 %\begin{classification} % according to http://www.acm.org/class/1998/
@@
 %-------------------------------------------------------------------------
-\input{results.tex}
+\input{section_results.tex}
 
 %-------------------------------------------------------------------------
-\input{conclusion.tex}
+\input{section_conclusion.tex}
 

### 2018-04-10 latex/egsr17-cgf/previous_work.tex
final version

@@
 
-Light transport in participating media is governed by the radiative transfer equation~(RTE), first studied in the context of astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}.
-Today, this equation is typically solved using Monte Carlo based methods.
-However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become prohibitive, due to the high computational cost, which comes from tracing long particle paths with many scattering events.
+Light transport in participating media is governed by the radiative transfer equation~(RTE), first studied in the context of astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later introduced to computer graphics by Kajiya~\cite{Kajiya86}.
+Today, this equation is typically solved using Monte Carlo methods.
+In strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become prohibitively expensive.
+Consider a high albedo media like milk, where tracing paths with many scattering events is necessary.
 
-In contrast to path-tracing, the $P_N$-method is a deterministic method, which gives a solution by solving a system of linear equations.
+In contrast to path-tracing, the $P_N$-method gives a solution by solving a system of linear equations.
 It is derived by discretizing the angular variable of the radiative transfer equation into spherical harmonics (SH).
 This gives rise to a set of coupled, complex-valued partial differential equations, called $P_N$-equations.
-The subscript $N$ refers to the truncation order of the SH.
+The subscript $N$ refers to the truncation order of the spherical harmonics.
 
@@
 \vspace{-0.1in}
-\icaption{Solving the 2D-checkerboard problem using naive collocated grids produces oscillating artefacts (left). Our solver uses staggered grids and produces artefact free results (right).}
+\icaption{Solving the 2D checkerboard problem using naive collocated grids produces oscillating artifacts (left). Our solver uses staggered grids and produces artifact free results (right).}
 \label{fig:artifacts}
@@
 Similar to $P_N$, the classical diffusion approximation (CDA) is a deterministic method, which arrives at a solution by solving a system of linear equations.
-It is derived as a degenerate case of the $P_1$-equations (truncation after the first SH order), which collapses to a simple diffusion equation giving the method its name.
+It corresponds to the $P_N$-equations when $N=1$ (truncation after the first SH order), which can be collapsed to a simple diffusion equation, giving the method its name.
 CDA has a long history in other domains, such as astrophysics~\cite{Ishimaru78} and was introduced to graphics by Stam \cite{Stam95}.
-It was further extended to anisotropic radiative transfer by \cite{Jakob10}.
 
 CDA suffers from severe energy loss close to regions with strong density gradients.
-The problem is addressed by a modification known as Variable Eddington factor (VEF), which nonlinearly adjusts the diffusion coefficient to improve the solution near density gradients and low-density regions.
+The problem can be addressed by a modification known as Variable Eddington factor (VEF), which nonlinearly adjusts the diffusion coefficient to improve the solution near density gradients and low-density regions.
 Flux-limited diffusion, developed in the context of astrophysics by Levermore et al.~\cite{Levermore81} and later introduced to graphics by Koerner et al.~\cite{Koerner14}, is the most prominent example.
 
-The idea behind VEF is to derive a better diffusion coefficient by seeing it as a mean to spatially blend between the diffusive and pure transport regime.
-The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
+VEF is based on the observation, that the behavior of volumetric transport is closely linked to the density of the material: In the diffusive regime, thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
 In the absence of scattering (e.g.
 pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.
+The idea behind VEF is to derive a better diffusion coefficient by seeing it as a mean to spatially blend between the diffusive and pure transport regime.
 
 It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion.
-It is also been raised in other domains~\cite{Olson00} and resolving it is one motivation for this paper.
+This question is also been raised in other domains~\cite{Olson00} and resolving it is one of our motivations.
+
+\vspace{1in}
 

### 2018-04-10 latex/egsr17-cgf/rendering.tex
started working on anisotropy test case
final version

@@
 
-In order to evaluate our results, we chose to use our render the solution directly.
 We use an approach similar to Koerner et al.~\cite{Koerner14}, where we seperate the radiance field into single scattered light $L_{ss}$ and multiple scattered light $L_{ms}$:
@@
 This means that, our solver will solve for the multiple scattered light $L_{ms}$.
-The quantity $L_u$ is the uncollided light, which was emitted from the lightsource and attenuated by the volume without undergoing any scattering event.
-It is easily computed for delta light sources, such as directional lights, in homogeneous volumes.
-Raymarching can be used in heterogeneous volumes.
-For more complex light sources, the emission field can be precomputed with a number of Monte Carlo iterations.
+The quantity $L_u$ is the uncollided light, which was emitted from the light source and attenuated by the volume without undergoing any scattering event.
+We compute it using a few light samples per voxel, which quickly converge to a useful result for Dirac Delta light sources.
 
 Running the solver gives solution vector $\vec{u}$.
-The staggering is removed by interpolating all coefficients to voxel centers.
-Also the additional coefficients at boundary voxels are removed.
-This operation is represented as a matrix and results in a simple three-dimensional voxelgrid with SH coefficients for order $N$ at the center of each voxel.
+We then un-stagger the solution by interpolating all coefficients to voxel centers.
+The additional coefficients at boundary voxels are no longer needed.
+This operation is represented as a matrix that produces a three-dimensional voxel grid with SH coefficients for order $N$ at the center of each voxel.
 
@@
 Then we sample a new direction according to the phase function.
-Instead of continuing tracing into the new direction, we simply evaluate the inscattering integral using $\widehat{L}_{ms}$.
-The SH coefficients at $\vec{x}$ are found by trilinear interpolation from the voxelgrid of SH coefficients.
+Instead of continuing tracing into the new direction, we evaluate the inscattering integral using $\widehat{L}_{ms}$.
+The SH coefficients at $\vec{x}$ are found by trilinear interpolation from the voxel grid of SH coefficients.

### 2018-04-10 latex/egsr17-cgf/section_car.tex
final version

@@
 Our computer algebra representation is based on mathematical expression trees, which are hierarchies of expression instances.
-Those can be basic primitives, such as numbers, variables or special symbols, such as the Kronecker delta or imaginary number.
-Other expression types, such as integrals, derivatives, sums, products and functions build the hierarchy by referencing child expressions.
+An expressions can be either a basic primitive, such as a number, variable or special symbol, such as the Kronecker delta or imaginary number.
+Other expression types, such as integrals, derivatives, sums, products and functions, build the hierarchy by referencing child expressions.
 
 Our framework further provides manipulators, which can be executed on expressions.
-We implemented all manipulations, which were required for the derivation of the $P_N$-equations.
-This includes operations, such as application of the distributive law, substitution, constant folding, reordering of nested integrals, application of identities up to more complex operations, such as factorization of unknowns and discretization of spatial variables.
+We implemented all manipulations required to derive the $P_N$-equations.
+This includes application of the distributive law, substitution, constant folding, reordering of nested integrals, application of identities up to more complex operations, such as factorization of unknowns and discretization of spatial variables.
 The derivation steps of the real-valued $P_N$-equations in the supplemental material were all carried out using our framework.
+
+\begin{figure}[h]
+%\vspace{1in}
+\centering
+\includegraphics[width=0.9\columnwidth]{figures/fig_car.pdf}
+\vspace{-0.16in}
+\icaption{Our computer algebra framework allows to represent equations as mathematical expressions trees.
+It further provides a set of functions for manipulating the tree according to valid mathematical operations, such as the binomial expansion above.
+Frontends allow generation of source code from the expression tree.}
+\label{fig:staggeredgrid}
+\end{figure}
+
+%Our computer algebra framework allows to represent equations as mathematical expressions trees.
+It further provides a set of functions, which manipulate the tree according to valid mathematical operations, such as the binomial expansion above.
+Rendering frontends allow generation of source or code from the expression tree.
 
@@
 
-Note that our framework is different from a computer algebra system (CAS), in that our framework only deals with representation and manipulation of expressions, and is not concerned with finding answers to mathematical problems.
+Note that our framework is different from a computer algebra system (CAS), such as Mathematica.
+While these systems probably use similar representations internally, they are aimed at finding answers to mathematical problems and have not been designed to allow application of custom manipulation passes on the expression tree.
+Our framework has been written solely for that purpose.
+It only deals with representation and manipulation of mathematical expressions and therefore is very lightweight and easy to use.

### 2018-04-10 latex/egsr17-cgf/section_conclusion.tex
final version

--- 2018-04-05 latex/egsr17-cgf/conclusion.tex
@@
 We originally set out to understand how non-linear diffusion compares against $P_N$ for increasing order.
-Although the lack of higher moments makes the FLD solution very flat, its energy conserving nature and comparably small resource footprint make it a much better approach when compared to $P_N$, which becomes more and more intractable for higher moment count.
-This is especially true, when compromises in accuracy of the approximation are acceptable.
+Although the lack of higher moments makes the FLD solution very flat, its energy conserving nature and comparably small resource footprint make it a much better approach when compared to $P_N$, which becomes increasingly costly for higher values of $N$.
 
 The literature in other fields often states, that the $P_N$ method---when solved in normal form like we do---is able to deal with vacuum regions.
-This is misleading.
+We found this misleading.
 The $P_N$-method in normal form indeed does not break down completely in the presence of vacuum like diffusion based methods do (due to $\sigma_t^{-1}$ in the diffusion coefficient).
@@
 Discretization in angular and spatial domain is done using manipulation passes on the representation.
-The stencil code, which is used to propagate the system of linear equations, is automatically generated from the expression tree.
-This way, we can easily deal with complex coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand.
+The stencil code, which is used to propagate the system of linear equations, is generated from the expression tree.
+This way, we can easily deal with coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand.
 
 When researching the application of $P_N$ in other fields, we came across a rich variety of variations on the $P_N$-method, such as simplified $P_N$ ($SP_N$)~\cite{Ryan10}, filtered $P_N$ ($FP_N$)~\cite{Radice13}, diffusion-correction $P_N$ ($DP_N$)~\cite{Schaefer11} and least-squares $P_N$ ($LSP_N$)~\cite{Hansen14}.
-These variations have been introduced to address certain problems of the standard $P_N$-method, such as ringing artefacts, dealing with vacuum regions and general convergence.
+These variations have been introduced to address certain problems of the standard $P_N$-method, such as ringing artifacts, dealing with vacuum regions and general convergence.
 Our solver can be applied to any (potentially coupled) PDE and therefore can generate stencil code for all these variations by simply expressing the respective PDE's in our computer algebra representation and providing this as an input to our solver.

### 2018-04-10 latex/egsr17-cgf/section_discretized_rte.tex
final version

@@
 
-Our method is derived from the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
+Our method derives from the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
 \begin{align}
@@
 
-The left hand side (LHS) is the transport term and we refer to the terms on the right hand side (RHS) as collision, scattering and source term respectively.
-The symbols $\sigma_t$, $\sigma_s$, $p$ and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term.
+The left hand side (LHS) is the transport term, and we refer to the terms on the right hand side (RHS) as collision, scattering, and source term, respectively.
+The symbols $\sigma_t$, $\sigma_s$, $p$, and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term.
 
-The RTE is often given in operator notation, where transport, collision and scattering are expressed as operators $\mathcal{T}$, $\mathcal{C}$ and $\mathcal{S}$, which are applied to the radiance field $L$:
+The RTE is often given in operator notation, where transport, collision, and scattering are expressed as operators $\mathcal{T}$, $\mathcal{C}$ and $\mathcal{S}$, which are applied to the radiance field $L$:
 \begin{align}
@@
 This gives rise to a linear system of equations, which can be solved using standard methods.
-For the $P_N$-method, the angular variable is discretized first, using a truncated spherical harmonics expansion.
-This results in the $P_N$-equations, a system of coupled PDE's, which still have a continuous spatial variable.
+For the $P_N$-method, the angular variable is first discretized, using a truncated spherical harmonics expansion.
+This results in the $P_N$-equations, a system of coupled PDE's that still depend on a continuous spatial variable.
 
@@
 After giving an outline of the general discretization in this section, we will present our computer algebra representation in the next section.
-The $P_N$-equations, as result of our automated discretization, are given in section~\ref{sec:real_valued_pn_eq}.
+The $P_N$-equations that result from our automated discretization are given in Section~\ref{sec:real_valued_pn_eq}.
 
-Since the radiance field $L$ is real only, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
+Since the radiance field $L$ is real, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
 \begin{align}
@@
 
-We express the projection into the SH basis functions with the projection operator $\mathcal{P}$:
+We express the projection into the spherical harmonics basis functions with a projection operator $\mathcal{P}$:
 \begin{align}
@@
 
-The $P_N$-equations are derived by first expressing all angular dependent parameters in spherical harmonics.
-The radiance field $L$ in equation~\ref{eq:rte} is therefore replaced by its SH reconstruction $\widehat{L}$, introducing an error due to truncation at order $N$:
+The $P_N$-equations are derived by first expressing all direction-dependent parameters in spherical harmonics.
+The radiance field $L$ in Equation~\ref{eq:rte} is therefore replaced by its SH reconstruction $\widehat{L}$, introducing an error due to truncation at order $N$:
 \begin{align}
@@
 
-After substitution, all angular parameters are expressed in terms of spherical harmonics, but still depend on the continuous angular variable $\omega$.
-As a next step, we project each term of the RTE into SH, using the projection operator $\mathcal{P}$.
+After substitution, all angular parameters are expressed in terms of spherical harmonics, but they still depend on the continuous angular variable $\omega$.
+As a next step, we project each term of the RTE into spherical harmonics, using the projection operator $\mathcal{P}$.
 This produces a single equation for each $l,m$-pair.
@@
 
-Once the $P_N$-equation have been found, the spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxelgrid (using central differences for differential operators).
+Once the $P_N$-equation have been found, the spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxel grid (using central differences for differential operators).
 
-The radiance field $L$, after full discretization, is represented as a set of SH coefficients per voxel.
+Following this discretization, the radiance field $L$, is represented as a set of SH coefficients per voxel.
 Flattening these over all voxels into a single vector, gives the solution vector $\vec{u}$.
@@
 
-So far we have given the $P_N$-equations only in high-level operator notation (equation~\ref{eq:pn_operator_notation}).
+So far, we have only given the $P_N$-equations in high-level operator notation (Equation~\ref{eq:pn_operator_notation}).
 Carrying out the full derivation creates large, unwieldy equations and requires a string of expansions, manipulations and applications of identities.
-These are hard to do and error prone if done by hand.
+These are challenging and error-prone if done by hand.
 We therefore developed a computer algebra representation, which allowed us to derive the the $P_N$-equation in a semi-automatic fashion.
-In the next section, we will present our computer algebra representation and give the final result of the angular discretization in section~\ref{sec:real_valued_pn_eq}.

### 2018-04-10 latex/egsr17-cgf/section_results.tex
final version

@@
 
-Classical diffusion is based on the diffusion equation, which is derived by collapsing the $P_1$-equations:
+Our computer algebra framework and the precomputation has been implemented in Python.
+The runtime component of our solver has been implemented in C++.
+We use a naive implementation of a CG solver, which has been modified such that we do not need to expliticly compute the normal form of the linear system to solve.
+We use the sparse matrix representation and sparse matrix vector product from the Eigen linear algebra library.
+
+The solver for classical diffusion is based on the diffusion equation, which is derived by collapsing the $P_1$-equations:
 \begin{align} 
@@
 
-Since our solver can work with any PDE, which results in a linear system of equations, we only needed to put equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which would generate the correct stencil code automatically.
+Since our solver can work with any PDE, which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which would generate the correct stencil code automatically.
 
@@
 
-First we run our solver on the 2D checkerboard, a very common test case in other fields. This allows us to validate our solver against the solver from Seibold et al.~\cite{Seibold14}.
+First we ran our solver on the 2D checkerboard, a very common test case in other fields.
+The problem has dimensions $7\times 7$ and is discretized with resolution $71\times 71$.
+Unit size blocks are filled with purely absorbing medium with $\sigma_a=10$ in a checkerboard pattern.
+All other blocks are filled with a purely scattering medium with $\sigma_s=1$.
 
-Their solver solves for the time-dependent and complex-valued $P_N$-equations in the 2D case, which is derived by assuming that all SH coefficients, RTE parameters and boundary conditions are z-independent.
+Solving the standard checkerboard problem, allows us to validate our solver against the solver from Seibold et al.~\cite{Seibold14}, which solves for the time-dependent and complex-valued $P_N$-equations in the 2D case only.
+The 2D case is derived by assuming that all SH coefficients, RTE parameters and boundary conditions are z-independent.
 This causes all SH coefficients and moment equations for which $l+m$ is odd to vanish.
@@
 %\missingfigure{test}
-\includegraphics[width=\columnwidth]{images/checkerboard2d_p5_neumann_staggered_starmap.png}
+\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered_starmap.png}
 \end{subfigure}%
@@
 %\missingfigure{test2}
-\includegraphics[width=\columnwidth]{images/checkerboard2d_p5_neumann_staggered.png}
+\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered.png}
 \end{subfigure}%
-\vspace{-0.1in}
-\icaption{Comparison of the result for the checkerboard test using Starmap~\cite{Seibold14} (left) against ours (right).
-Our result is in excellent agreement with the reference.
-We attribute the difference at the boundary to the lack of convergence of the Starmap solve.}
+\vspace{-0.2in}
+\icaption{Comparison of the result for the checkerboard test using Starmaps time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$.
+Our results are in good agreement.}
 \label{fig:vs_starmap}
@@
 
-As can be seen in figure~\ref{fig:vs_starmap}, the results from our solver are in excellent agreement with the results from Seibold et al.~\cite{Seibold14} and verify the correctness of our implementation.
+As can be seen in figure~\ref{fig:vs_starmap}, the results from our solver are in good agreement with the results from Seibold et al.~\cite{Seibold14} and verify the correctness of our implementation.
+Converging to a residual of $10e^{-10}$ takes $0.27s$ for $P_1$ and $25s$ for $P_5$.
 
@@
 We also run our solver for the point source problem, a single point light in a homogeneous medium.
-This not only helps to validate our implementation for the 3D case, but also allows to get a clear assessment of the methods accuracy.
-We use the Grosjean approximation, which was introduced by D'Eon et al.\cite{dEon11}, as a very accurate approximation to the ground truth solution.
+This not only helps to validate our implementation for the 3D case, but also provides information on the accuracy of these methods.
+We use the Grosjean approximation, which was introduced by D'Eon et al.\cite{dEon11} as a very accurate approximation to the ground truth solution.
 
@@
 We run the solver for different truncation values $N$.
-In figure~\ref{fig:pointsource}~\subref{fig:pointsource_pn}, we see that the solution becomes increasingly accurate for higher truncation order.
+In Figure~\ref{fig:pointsource}~\subref{fig:pointsource_pn}, we see that the solution becomes increasingly accurate for higher truncation order.
 The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even.
 We further see, that the $P_1$ solution exactly matches the results from CDA, which confirmes, that the latter is only a collapsed version of the former.
+The time to solve is significant with $10m$ for $P_1$ and $45m$ with $P_5$.
+With these performance characteristics, the $P_N$-method is clearly not competitive in comparison with much diffusion and FLD solver, which are much faster.
 
-\subsection{Procedural Clouds}
+\subsection{Nebulae}
 
-Finally, we run our solver on different procedural cloud datasets to get an idea of its performance in more practical applications.
-
+Finally, we run our solver on a procedural cloud dataset to get an idea of its performance in more practical applications.
 Figure~\ref{fig:teaser}~\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function.
 We see that at order $N=5$, our method can capture indirect illumination similarily well as FLD and is significantly better than CDA as expected.
+The indirectly illuminated region at the bottom appears to be closer to the path-traced result as opposed to the solution form FLD which is very flat in that region.
+However, in many other areas, $P_5$ seems to still suffer a lot from the problem of energy loss near transitions from dense regions to vacuum.
+It appears that going higher order a few steps, mitigates this problem only slowly at an immense cost of compute and storage.
 
-\begin{itemize}
-	\item mention performance.
-	\item run for various values of $N$
-	\item convergence behaviour in vacuum. Clearify
-	\item run for anisotropic problem
-\end{itemize}
+The main characteristic of the nebulae dataset is the presence of vacuum.
+We found that having vacuum regions in the dataset will cause the condition number to become infinite and the solver practically doesnt converge.
+We therefore introduced a minimum threshold for the extinction coefficient $\sigma_t$.
+Every voxel with an extinction coefficient smaller than the threshold would be set to the threshold value.
+In Figure~\ref{fig:results_convergence} we show the effect of the minimum threshold on the convergence behavior.
+As the threshold increases, convergence improves.
+
+%\begin{table}[h]
+%\begin{center}
+%\begin{tabular}{ c | c c c }
+% & \textbf{2D Checkerboard} & \textbf{Point Source} & \textbf{Nebulae} \\ 
+%\hline
+%$P_1$ & 0.27s & 10m (64\times 64\times 64) & cell6 \\  
+%$P_2$ & 2.4s & cell8 & cell9 \\
+%$P_3$ & 5.8s & cell8 & cell9 \\
+%$P_4$ & 12.7s & cell8 & cell9 \\
+%$P_5$ & 25s & ? & ?
+%\end{tabular}
+%\end{center}
+%\icaption{Performance measurements for different datasets and varying truncation order $N$}
+%\label{table2}
+%\end{table}
+
+\begin{figure}[h]
+\centering
+%\missingfigure{test}
+\includegraphics[width=0.9\columnwidth]{figures/fig_nebulae_p1_convergence.pdf}
+%\vspace{-0.2in}
+\vspace{-0.1in}
+\icaption{Convergence behavior of our solver with $N=1$ for the nebulae dataset and for varying minimum thresholds of the extinction coefficient $\sigma_t$.
+Threshold values and an estimate for the condition number of $A$ (Matlabs condest function) are shown next to the plots.
+The convergence deteriorates as the threshold decreases.
+Once it reaches zero, the presence of pure vaccuum makes the condition number infinite.}
+\label{fig:results_convergence}
+\end{figure}
+
+%\begin{itemize}
+%	\item mention performance.
+%	\item run for various values of $N$
+%	\item convergence behaviour in vacuum. Clearify
+%	\item run for anisotropic problem
+%\end{itemize}

### 2018-04-10 latex/egsr17-cgf/solver.tex
final version

@@
-\section{$P_N$-Solver}
-\label{sec:pnsolver}
-
-The truncation order $N$ is the key input parameter to the solver.
-With higher values, manually coding the solver from the equations would be ardous, error prone and very time consuming.
-We therefore decided to make use of the equations computer algebra representation and designed our solver around it.
-
-The solver consists of two components.
-The first is a precomputation (section~\ref{sec:solver_precomputation}), which is done once for every single value of $N$.
-This steps runs a partial evaluation on the mathematical expression tree and applies the spatial discretization in a reference space we call stencil space.
-The precomputation step automatically generates source code from the resulting expression tree.
-
-The generated stencil code is compiled with the runtime component (section~\ref{sec:solver_runtime}) of our solver.
-This component gets the actual problem as input, including gid resolution and RTE parameter fields.
-It builds the linear system and solves it using standard methods.
-An overview of the solver is given in figure~\ref{fig:pnsolver}.
-
 \begin{figure*}[t!]
@@
 
+\section{$P_N$-Solver}
+\label{sec:pnsolver}
+
+The truncation order $N$ is the key input parameter to the solver.
+With higher values, manual implementation of the solver from the equations would be ardous, error-prone and time-consuming.
+We therefore decided to make use of the computer algebra representation and designed our solver around it.
+
+The solver consists of two components.
+The first is a precomputation (Section~\ref{sec:solver_precomputation}), which is excuted once for every single value of $N$.
+This steps runs a partial evaluation on the mathematical expression tree and applies the spatial discretization in a reference space we call stencil space.
+The precomputation step automatically generates source code from the resulting expression tree.
+
+The generated stencil code is compiled with the runtime component (Section~\ref{sec:solver_runtime}) of our solver.
+This component receives the actual problem as input, including grid resolution and RTE parameter fields.
+It then builds the linear system and solves it using standard methods.
+An overview of the solver is given in Figure~\ref{fig:pnsolver}.
+
 \subsection{Precomputation}
@@
 The result of the precomputation is a stencil, which can be used during runtime to build the linear system for a given problem.
-The stencil is a pattern of indices into the solution vector, along with some weights.
-It expresses how the sum of the weighted solution vector components relate to the RHS for a given unknown in the system and therefore contains most information required to propagate a row in the system matrix $A$ and RHS-vector $\vec{Q}$.
-The index pattern does not change for different rows, giving the stencil its name.
+The stencil is a pattern of indices into the solution vector, along with values.
+It expresses how the sum of the weighted solution vector, components relate to the RHS for a given unknown in the system and therefore contains most information required to fill the system matrix $A$ and RHS-vector $\vec{Q}$ row by row.
+Note that while the coefficients may change, the sparsity pattern of the stencil is identical for different rows.
 
-A stencil generally is crafted by discretizing a PDE at a hypothetical center voxel $ijk$ (assuming to mean the voxel center most of the time).
+\begin{figure}[b]
+\centering
+%\missingfigure{test}
+\includegraphics[width=0.7\columnwidth]{figures/fig_stencil_pipeline.pdf}
+%\vspace{-0.2in}
+\icaption{Creating the stencil code requires several steps, usually done by hand. We express the given problem in a computer algebra representation and use it to fully automate the process.}
+\label{fig:stencile_pipeline}
+\end{figure}
+
+Stencil generation entails discretizing a PDE at a hypothetical center voxel $(i, j, k)$ (assuming to mean the voxel center most of the time).
 Finite differences create weighted references to other voxels (e.g.
-$i+1,jk$).
-After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and $ijk$-offsets.
-$ijk$ will only be known during runtime, when the stencil is executed for a particular unknown (row).
-Then the offsets can be used to find the component index into the solution-vector and weights can be evaluated for concrete world space position.
-We refer to the space with the hypthetical voxel $ijk$ at the center as stencil space.
+$i+1, j, k$).
+After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and offsets (Figure~\ref{fig:stencile_pipeline}).
+$(i, j, k)$ will only be known during runtime, when the stencil is executed for a particular unknown (row).
+Then the offsets can be used to find the component index into the solution-vector, and weights can be evaluated for concrete world space position.
+We refer to the space with the hypthetical voxel $(i, j, k)$ at the center as stencil space.
 
-Typically, the stencil is coded by hand, which would be challenging with our $P_N$-equations and higher values of $N$.
-In our system, the stencil code is created automatically from the computer algebra representation.
-Spatial discretization and fatorization into canonical form are done as manipulation passes on the mathematical expression tree.
-The precomputation code iterates over all terms and analyzes the unkown to get the $ijk$-offset.
-The factor-expression is extracted from the tree and used to generate source code for evaluating the factor-expression during runtime (including calls for evaluating RTE-parameter fields).
+\begin{wrapfigure}{r}{0.6\columnwidth}
+%\begin{center}
+\hspace{-.2in}
+\includegraphics[width=0.6\columnwidth]{figures/fig_discretization.pdf}
+%\end{center}
+\end{wrapfigure}The spatial discretization is done by parsing the expression tree from the root.
+The discrete substitute for the continuous position variable $\vec{x}$ is initialized with $ijk$.
+Differential operator nodes are replaced by a subtree, which expresses the finite difference approximation (including voxelsize factor $h$).
+The subtree of the differential operator node is duplicated for different offsets to the discrete variable $(i, j, k)$.
+Since this offset only applies to the substree, a stack is maintained by the parser to manage scope.
+Whenever the parser encounters the continuous variable $\vec{x}$ in the expression tree, its node in the tree is replaced by the discrete substitute, currently on top of the stack.
+Nested differential operators yield higher order finite difference stencils as expected.
 
-The spatial discretization is done by parsing the expression tree of each equation from the root.
-All occurances of the continuous positional variable $\vec{x}$ are replaced by discrete voxel coordinates $i,j,k$, whose values are retrieved from the top of a stack.
-The stack is kept by the parser and is initialized with the voxel at $ijk=(0,0,0)$.
+\begin{wrapfigure}{r}{0.6\columnwidth}
+%\begin{center}
+\hspace{-.2in}
+\includegraphics[width=0.6\columnwidth]{figures/fig_codegen.pdf}
+%\end{center}
+\end{wrapfigure}
+Factorization into canonical form is done as a manipulation pass on the mathematical expression tree.
+The result allows to implement the code generation in a straight forward fashion.
+For each term, the $ijk$-offset is retrieved from the unknown.
+The factor-expression, which is multipled with the unknown, is extracted from the tree and used to generate source code for evaluating the factor-expression during runtime (including calls for evaluating RTE-parameter fields).
 
-Whenever the parser encounters a differential operator, the current voxel position is shifted and pushed to the stack.
-Then the child expression is parsed and multiplied with a weight value, followed by popping the stack.
-This is done for each element of a central difference stencil, which provides the offset and weight values (referencing the voxelsize variable).
-The dimension of the stencil is given by the derivation variable of the operator.
-Nested derivatives are handled naturally and produce higher order finite difference stencils as expected.
+\begin{figure}[h]
 
-Another aspect to consider during discretization is the location of the variables and unknowns.
-As we will see later, our solver needs to support staggered grids, where the SH coefficients are not all located at the voxel center (collocated).
-Coefficients can also be located at face-centers, edges and intersections of voxels (staggered).
+%\vspace{1in}
+\centering
+\includegraphics[width=0.9\columnwidth]{figures/fig_staggered_grids.pdf}
+%\vspace{-0.2in}
+\icaption{Staggered grids place coefficients at different locations within the finite difference grid.}
+\label{fig:staggeredgrid}
+\end{figure}
 
-Our solver supports placement of coefficients at arbitrary staggered grid locations.
-Whenever a coefficient is encountered during expression tree parsing for discretization, the parser checks how its location is related to the current location on the top of the stack.
-Depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at the requested offset from its surrounding locations, or it returns the coefficient itself, if it happens to coincide.
+\begin{wrapfigure}{r}{0.6\columnwidth}
+%\begin{center}
+\hspace{-.1in}
+\includegraphics[width=0.6\columnwidth]{figures/fig_staggered_interpolation.pdf}
+\vspace{-.1in}
+%\end{center}
+\end{wrapfigure}Our solver supports placement of coefficients at arbitrary staggered grid locations (see Figure~\ref{fig:staggeredgrid}).
+This means, that during the discretization step, the discrete location $(i, j, k)$ (at which the unknown is meant to be evaluated) might not coincide with the location of the unknown.
+To solve this, depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at $(i, j, k)$ from their defined locations.
+It those happen to coincide, the coefficient itsself is returned.
 This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center.
-The position stack of the parser is initialized with the staggered grid location defined by the coefficient, which is associated with the $l,m$-pair of the equation currently being discretized.
 
@@
 
-The generated stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver.
+The stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver.
 The runtime executes the stencil for every voxel to propagate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.
 
-The number of rows is determined by the number of voxels times the number of coefficients per voxel (see figure~\ref{fig:matrix_layout}) and can therefore become very large for high resolution and high truncation order.
-The matrix $A$ is squared and fortunately very sparse, due to finite differences and the structure of the $P_N$-equations.
-Unfortunately, it also is non-symmetric (due to the transport term) and not diagonal dominant, which rules out many standard methods for solving linear systems.
+The number of rows is determined by the number of voxels times the number of coefficients per voxel (see Figure~\ref{fig:matrix_layout}) and can therefore become very large for high resolution and high truncation order.
+The matrix $A$ is square and extremely sparse, due to the local structure of the finite differences discretization.
+Unfortunately, it is non-symmetric due to the transport term and not diagonal dominant, which rules out many standard methods for solving linear systems.
 We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead.
 This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number.
+Investigation of other solution schemes (e.g.
+multigrid) would be an interesting avenue for future work.
 
-The $P_N$-equations have been discretized, and the stencil has been generated, without any notion of domain boundaries and boundary conditions (BC).
-Our solver supports Neumann BC and Dirichlet BC.
-They are handled transparently by the code which runs the stencil.
-Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or write into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC).
+Our solver supports Neumann and Dirichlet boundary conditions.
+They are handled transparently by the code which generates the stencil.
+Whenever the stencil accumulates coefficients into a boundary location, the framework either ignores the write operation (Dirichlet BC) or accumulates into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC).
+This is done changing the index of the written component under the hood.
 
-In order to respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels (see Figure~\ref{fig:staggeredgrid}).
+%To respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels.
 This requires careful managegment and bookkeeping of coefficient indices, which is done transparently by the runtime code.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\columnwidth]{figures/fig_staggered_grids.pdf}
-%\vspace{-0.2in}
-\icaption{Staggered grids place coefficients at different locations within the finite difference grid. Additional unknowns are required in boundary cells to produce a correct result.}
-\label{fig:staggeredgrid}
-\end{figure}
 

### 2018-06-05 latex/egsr17-cgf/main-body.tex
minor revision

@@
 
-\title[Draft]%
+\title%
 {
@@
 
-\author
+\author[D. Koerner \& J. Portsmouth \& W. Jakob]
 {
 %Compiled on \today
-paper1027
-%David Koerner$^{1,2,3}$
-%\qquad Jan Nov\'ak$^{1}$
-%\qquad Peter Kutz$^{2}$
+%paper1027
+David Koerner$^{1}$
+\qquad Jamie Portsmouth$^{2}$
+\qquad Wenzel Jakob$^{3}$
 %\qquad Ralf Habel$^{2}$
@@
 % For Computer Graphics Forum: Please use the abbreviation of your first name.
-%$^1$Disney Research
-%\qquad $^2$Walt Disney Animation Studios
-%\qquad $^3$University of Stuttgart
+$^1$University of Stuttgart
+\qquad $^2$Solid Angle
+\qquad $^3$\'{E}cole polytechnique f\'{e}d\'{e}rale de Lausanne (EPFL)
 %\qquad $^4$Dartmouth College
@@
 Increasing the order introduces a set of increasingly complex coupled partial differential equations, whose growing number and complexity make them very difficult to work with.
-We introduce a computer algebra framework for representing and manipulating the underlying mathematical equations and use it to derive the time-independent real-valued $P_N$-equations for arbitrary orders.
+We \rev{use} a computer algebra framework for representing and manipulating the underlying mathematical equations and use it to derive the time-independent real-valued $P_N$-equations for arbitrary orders.
 We further present a staggered-grid $P_N$-solver and generate its stencil code directly from the expression tree of the $P_N$-equations.
@@
 
+% --------------------------------------------------------------------------------------------------
+\vspace{-0.1in}
+\section*{Acknowledgements}
+\label{sec:acknowledgements}
+\vspace{-0.05in}
+We thank the anonymous reviewers for their valuable and encouraging comments and feedback.
+We also thank Martin Frank and Benjamin Seibold for the very valuable discussions and answers to our questions.
+We thank Bernd Eberhardt for feedback and support.
+This project has been partially funded by the MSC-BW project.
+
 %-------------------------------------------------------------------------

### 2018-06-05 latex/egsr17-cgf/section_car.tex
minor revision

--- 2018-04-05 latex/egsr17-cgf/section_car.tex
@@
 
-Our computer algebra representation is based on mathematical expression trees, which are hierarchies of expression instances.
-Those can be basic primitives, such as numbers, variables or special symbols, such as the Kronecker delta or imaginary number.
-Other expression types, such as integrals, derivatives, sums, products and functions build the hierarchy by referencing child expressions.
+\rev{We use a computer algebra representation for working with the $P_N$-equations.
+It represents the equations using a tree of mathematical expressions, which represent numbers, symbols and other expression types, such as integrals, derivatives, sums, products and functions.
+Further, manipulators can be executed on these expression trees to perform substitution, constant folding, reordering of nested integrals, application of identities and more complex operations.
+Finally, frontends allow rendering the expression tree into different forms, such as \LaTeX~and C++ source code.
+While we ended up implementing our own lightweight framework, off-the-shelf packages, such as SymPy (www.sympy.org), exist and would be equally suitable for our use case.}
 
-Our framework further provides manipulators, which can be executed on expressions.
-We implemented all manipulations, which were required for the derivation of the $P_N$-equations.
-This includes operations, such as application of the distributive law, substitution, constant folding, reordering of nested integrals, application of identities up to more complex operations, such as factorization of unknowns and discretization of spatial variables.
-The derivation steps of the real-valued $P_N$-equations in the supplemental material were all carried out using our framework.
+\begin{figure}[h]
+%\vspace{1in}
+\centering
+\includegraphics[width=0.9\columnwidth]{figures/fig_car.pdf}
+\vspace{-0.16in}
+\icaption{\rev{A} computer algebra framework allows to represent equations as mathematical expressions trees.
+It further provides a set of functions for manipulating the tree according to valid mathematical operations, such as the binomial expansion above.
+Frontends allow generation of source code from the expression tree.}
+\label{fig:staggeredgrid}
+\end{figure}
 
-Finally, frontends allow rendering the expression tree into different forms.
-We implemented frontends for rendering expression trees to \LaTeX~and C++ source code.
-The equations in the supplemental material were almost all rendered by the former.
-The latter was used for generating the stencil code used by our solver.
-
-Note that our framework is different from a computer algebra system (CAS), in that our framework only deals with representation and manipulation of expressions, and is not concerned with finding answers to mathematical problems.
+\rev{Using the computer algebra representation, we perform the derivation steps required to arrive at the real-valued $P_N$-equations (the derivation steps shown in the supplemental material were almost all rendered from the expression tree).
+More importantly, we use the representation to perform automated discretization and generate the stencil code used by our solver.
+This is detailed in section~\ref{sec:solver_precomputation}}.

### 2018-06-05 latex/egsr17-cgf/solver.tex
pre_finalization
minor revision

@@
 \end{figure*}
+
+\vspace{-0.75in}
 
@@
 
+\vspace{0.6in}
+
 Stencil generation entails discretizing a PDE at a hypothetical center voxel $(i, j, k)$ (assuming to mean the voxel center most of the time).
@@
 We refer to the space with the hypthetical voxel $(i, j, k)$ at the center as stencil space.
+
+\vspace{0.2in}
 
@@
 To solve this, depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at $(i, j, k)$ from their defined locations.
-It those happen to coincide, the coefficient itsself is returned.
+If those happen to coincide, the coefficient itsself is returned.
 This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center.
@@
 The matrix $A$ is square and extremely sparse, due to the local structure of the finite differences discretization.
-Unfortunately, it is non-symmetric due to the transport term and not diagonal dominant, which rules out many standard methods for solving linear systems.
-We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead.
+Unfortunately, it is non-symmetric due to the transport term and not diagonal dominant, which rules out many \rev{iterative} methods for solving linear systems.
+\rev{Iterative methods are useful, as they allow balancing accuracy against performance by tweaking the convergence threshold.} We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead.
 This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number.
@@
 multigrid) would be an interesting avenue for future work.
+\rev{However, more importantly, in the presence of vacuum regions, the matrix $A$ becomes singular and the system cannot be solved at all.
+This requires the introduction of a minimum threshold for the extinction coefficient $\sigma_t$.}
 

### 2018-06-05 latex/egsr17-cgf/response/review_response.tex
added review resonse letter

\documentclass[11pt,letterpaper,headsepline,DIV15]{scrartcl}

\def\bf{\normalfont\bfseries}

\usepackage{enumitem}
\usepackage{bm} % for bold greek symbols
\usepackage{color}

\usepackage{alltt}
%\renewcommand{\ttdefault}{txtt}

%checkmark
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\color{green}\ding{51}\color{black}}%
\newcommand{\xmark}{\color{red}\ding{55}\color{black}}%

% strike out text
\usepackage{soul,xcolor}

\definecolor{red}{RGB}{120,0,0}
\definecolor{gray}{RGB}{150,150,150}
\definecolor{todoColor}{RGB}{120,30,60}

\def\todo#1{{\textbf{\em \textcolor{todoColor}{\{TODO: #1\}}}}}
\def\todogm#1{{\textbf{\em \textcolor{todoColor}{\{TODO-G: #1\}}}}}
\def\todofs#1{{\textbf{\em \textcolor{todoColor}{\{TODO-F: #1\}}}}}

\definecolor{answerColor}{RGB}{40,60,140}
%\newcounter{aCounter}
%\setcounter{aCounter}{0}
%%\newcommand{\answer}[1]{\textcolor{answerColor}{\textbf{#1}}}
%%\newcommand{\answer}[0]{\noindent\stepcounter{aCounter}\textcolor{answerColor}{Reply	 \arabic{aCounter} }}
%\newcommand{\answer}[1]{\noindent\stepcounter{aCounter}\textcolor{answerColor}{\bf Reply \arabic{aCounter}\\}{\textcolor{answerColor}{#1}}}
%%\newcommand{\answer}[2]{\noindent\stepcounter{aCounter}\textcolor{answerColor}{\label{#1}\bf Reply \arabic{aCounter}\\}{\textcolor{answerColor}{#2}}}

% -------- comment --------------
\usepackage{verbatim}
\newcounter{cmntcounter}
\renewcommand{\thecmntcounter}{\arabic{cmntcounter}}

\newcommand{\nextcomment}[1]{\refstepcounter{cmntcounter}\label{#1}\textcolor{red}{\textbf{comment-\thecmntcounter}}}

% -------- reply --------------
\newcounter{rplcounter}
%\renewcommand{\thefpcounter}{\thechapter.\arabic{fpcounter}}
%\renewcommand{\thefpcounter}{\thesection.\arabic{fpcounter}}
\renewcommand{\therplcounter}{\arabic{rplcounter}}

\newenvironment{reply}[2]{%
\refstepcounter{rplcounter}%
\label{#1}%
\noindent\textbf{\textcolor{answerColor}{Reply~\therplcounter}\\}%
\color{answerColor}
}%
{}%

\newcommand{\tocomment}[1]{\bf#1}

\usepackage[T1]{fontenc}
\newcommand{\changefont}[3]{
\fontfamily{#1} \fontseries{#2} \fontshape{#3} \selectfont}
\changefont{phv}{m}{n}
\renewcommand{\familydefault}{\sfdefault}

%\title{{\small Cover Letter for VisWeek 2013 -- Paper Id 130, Minor Revision}\\

% {\Large Coronal Holes Extraction Based on Isocline Surfaces}}

\title{\normalsize Cover Letter for EGSR2018-- Submission 1027 (EI\&I), Revised Version}

\vspace{1cm}
%\author{M. \"Uffinger, F. Sadlo, T. Ertl}
%\date{2013-11-22}
\date{}

\begin{document}

\maketitle
\noindent Dear Program Chairs, Dear Reviewers,
\\

\noindent thank you for reviewing our submission and accepting it for publication at EGSR2018 EI\&I track with a minor revision.
\\

\noindent Common among the reviews is the request for more rendering evaluations, which unfortunately we have not available yet. While we took care in presenting results, which validated the theory (pointsource problem) and demonstrated applicability in rendering (Nebulae), we agree that more results are needed. The lack thereof is one of the reasons, why we submitted to the EI\&I track. We are planning to have more renderings of different datasets for the next stage of this project.
\\

\noindent In the following, we cite the original comments with non-proportional black font and use bold font for the parts we address in the subsequent replies, which are in blue color. Attached to this response, is the revised version where our changes in the text are highlighted in red.
\\
\\
\noindent Respectfully,\\
the Authors.

\clearpage

%\section*{Improved TVCG submission}

\vspace{0.5cm}

{\scriptsize\begin{alltt}
-------------------- Review 17707 --------------------

Although finite-element-based solutions to the RTE are well known in many areas, this topic is underexplored in
computer graphics. The solution proposed in the paper seems practical, and the paper has the potential to encourage
future research along this direction.

\tocomment{
The main issue of this paper is the lack of rendering evaluations: the only actual rendered results are in Figure 1, and the P_5 result is only slightly better than classical diffusion approximation and arguably worse than flux-limited diffusion. Thus, the practical usefulness of the proposed technique for graphics applications remain unclear.
}
\end{alltt}}

\begin{reply}{}{}
We agree with the reviewer on that the practical usefulness of our technique is unclear. In fact, at this stage, we admit that in comparison to flux-limited diffusion, our technique does not appear to be competitive. We mention this in the conclusion of our paper. At this point, the value in our submission is, that it closes a knowledge gap in the graphics community on the PN-method, presents the real-valued PN-equations and also introduces an interesting approach for solving general coupled PDE's. Making this method competitive in comparison to other deterministic approaches is our plan going forward.
\end{reply}

{\scriptsize\begin{alltt}
-------------------- Review 17724 --------------------
This paper is huge in terms of scope, and it opens exploration in multiple ways, even with very few results.
This paper is a nice contribution to the E&I Track.

There are some open questions that might need some discussion:
\tocomment{
- The need to create a custom CAS should be really justified, as it seems like overkill. There are some CAS libraries (SymPy for Python,
GiNaC for C++) that would probably fit perfectly this approach, and that are able to generate C++ code and LaTeX. Both of them enable exploration and parsing of their "tree". In any case, for a potential future version it might make sense to opensource the specific CAS library, write a technical report about it and cite it. Right now it adds some extra noise to the paper's story (which might be summarized as "we use a CAS library to derive and generate stencil code for the P-N equations").}
\end{alltt}}

\begin{reply}{}{}
We agree with the reviewer in that other off-the-shelve computer algebra frameworks, such as SymPy, would be suitable for the tasks of expression manipulation and automated discretization. The reason why we went for our own solution is simply, that we started with a very small lightweight implementation for initial tests and stuck with it throughout the project.
\newline
\newline
The reviewer suggests, that simply pointing to an existing computer algebra framework would allow to summarize things as "we use a CAS library to derive and generate stencil code for the P-N equations". While we agree, that the fact that we created our own computer algebra representation is of no importance to the paper, we find that covering how to use any of such a representation for automated discretization is an important aspect of our work. Computer algebra frameworks, such as SimPy, are primarily known for and used to perform symbolic computation. We feel, that it is worth covering in more detail, how such a framework can be used for automated discretization. This argument is backed up by the fact, that none of the off-the-shelf packages offers a manipulation step for discretizing a given equation. We find that implementing such a manipulation step is not trivial, especially in the presence of coupling and staggered grid locations.
\newline
\newline
So in order to account for the reviewers comment, we made changes to the text in various places, which de-emphasize the use of our own implementation.
%\begin{itemize}
%	\item abstract: We intoduce a computer algebra framework $\rightarrow$ We use a computer algebra framework
%\end{itemize}
Further, we reworked section 4, which was used to introduce our own CAS-framework. We now state, that any computer algebra representation can be used and reference SymPy as an example. We also shortened this section by removing the text which was dedicated to explaining the capabilities of our representation as those are only a subset of what most of the alternatives can do. We kept a paragraph, which summarizes the concept of such frameworks and also kept figure 4, which expresses the idea of computer algebra representations visually. Arguably, this is known by most of the readers, but we find it useful to briefly summarize the concept, because we refer to it in more detail in section 6.1, when we explain the automated discretization.
\newline
\newline
Section 6.1 explains the implementation of a discretization step, independent from the computer algebra framework used. As already mentioned, such a manipulation would have to be implemented for any framework and this section gives a good summary on how to go about it. We therefore see enough value in this section to keep it as is.
\end{reply}

{\scriptsize\begin{alltt}
\tocomment{
- The use of P-N for diffusion in participating media might be an unviable solution. The paper in the discussion section points to avenues of future work in terms of speed-up (multigrid linear solver, for instance) but there should be some avenue of future work regarding convergence. Results look adequate, but would it ever converge, given enough order in spherical harmonics? In which cases the solution is closer to the ground truth? Which kinds of participating media are better represented? From my point of view that is much more important than render time.}
\end{alltt}}

\begin{reply}{}{}
\label{rpl:gg}
We very much agree with this comment. To be clear: The $P_N$-method is only not converging, when vacuum or close-to vacuum regions are present, because the matrix $A$ is singular in these cases. If the extinction coefficient is sufficiently large throughout the domain, the method does converge (but might still require a lot of computation for that).
\newline
\newline
The presence of vacuum regions and its problem with conergence was particularily important to us, as vacuum regions are common in rendering. This problem seems to be closely related to second order methods, which we mention in the conclusion section of the paper (e.g. the work of Hansen et al.~\cite{Hansen14}). Some of those second order methods are said to have the benefit of being able to deal with vacuum regions. They are an active research topic in the neutron transport community. By solving the normal form ($A^TA$), we basically solved the naive second order form of the RTE and we found that it is able to deal with vacuum regions in the sense, that it does not break down formally (e.g. due to division by zero), but at the same time it does not converge, since $A^TA$ is still singular if $A$ happens to be singular. A paragraph in the conclusion section is dedicated to this insight.
\newline
\newline
We have not been very clear about the singularity of $A$ in the presence of vacuum. We added a comment on this in section 6.2.
\newline
\newline
There are different second order forms, such as Even-Odd-Parity (EOP), Self-adjoint angular flux (SAAF) or the least-squares form introduced by Hansen et al.~\cite{Hansen14} and evaluating their properties for rendering applications would be a very interesting avenue for future work. We mention this at the end of the conclusion section. However, these research questions really do exceed the scope of our paper, which is primarily about giving the theory and a practical implementation of the $P_N$-mehod with some validation.
\end{reply}

{\scriptsize\begin{alltt}
\tocomment{
- The linear system to be solved during runtime targets the normal form A^T A because it is not simmetric. However, the used library (Eigen) provides a variety of solvers that would probably ensure accuracy even for non-simmetric matrices (LU? SVD? QR?).
In that case, why is it better to use the normal form with a naturally higher condition number?}
\end{alltt}}

\begin{reply}{}{}
If no vacuum regions are present, the mentioned alternatives would work for both, the first order form ($A\vec{x}=\vec{b}$) and the normal form of the problem. The reason why we introduced the normal form is, that it enables the use of standard iterative methods, such as conjugate gradient. These iterative methods allow balancing the computation time against accuracy by tweaking the convergence threshold. We adjusted the text in section 6.2 to outline this decision.

As mentioned in reply~\ref{rpl:gg}, we have not been explicit about the singularity of $A$ in the presence of vacuum. In these cases, iterative solvers will not converge and solvers, such as LU factorization will not be applicable. We added a remark in the text in section 6.2.
%mention inverse of the scattering/collision operator
%introduce text referencing the lspn stuff and talking about the relation to normal form of the RTE

%When we take the normal form of the linear system $A\vec{x}=Q$ (which we arrive at after discretization), we essentially get the coefficient matrix of the naive second order form of the RTE. The neutron transport literature states, that standard first order forms are not able to cope with vacuum regions, since their solution requires inversion of the scattering/collision operator. And those break down in vacuum. We saw that solvers, such as LU were not able to converge. such as the least-squares $P_N$ method introduced by 
\end{reply}

{\scriptsize\begin{alltt}
\tocomment{
Last, I would like to congratulate the authors for their work and I would like to encourage them to keep this research until a final version, that most surely I would like to read.}
\end{alltt}}

\begin{reply}{}{}
We are really thankful for your comment and encouragement. Deterministic methods are a niche topic and it is unclear, whether they will find a place in rendering long-term. However, investigating these methods and providing a clear understanding about their traits and their applicability is valuable research for which the EI\&I track seems to be a good place. We are very happy that all reviewers seem to agree.
\end{reply}

{\scriptsize\begin{alltt}
-------------------- Review 17681 --------------------

The previous paragraphs explain the reasons why I will reference the paper. I think that beyond the CG community,
this paper can have a larger influence in many work dealing with the volume Rendering.

It is very important to make sure that the supplemental materials and the source code are released at the same time
than the paper.

\tocomment{Note to the authors: the section 3.1.2 of the supplemental material is empty. Make sure to correct for the final version of the paper.}
\end{alltt}}
\begin{reply}{}{}
We completed section 3.1.2 in the supplemental material on the real-valued spherical harmonics expansion of the phase function and also fixed some minor notation issues throughout the text.
\end{reply}

%\bibliographystyle{plain}
\bibliographystyle{eg-alpha}
\bibliography{bibliography}

\end{document}

### 2018-06-08 latex/egsr17-cgf/introduction.tex
revision 2

@@
 \section{Introduction}
+\vspace{0.1in}
 
-Simulating light transport in paticipating media remains a challenging problem for image synthesis in computer graphics.
-Due to their ability to produce unbiased results and conceptional simplicity, Monte Carlo based techniques have become the standard approach.
-The main downside of these methods are their computational demands when rendering participating media, which exhibits strong scattering or anisotropy.
+Simulating light transport in participating media remains a challenging problem for image synthesis in computer graphics.
+Due to their ability to produce unbiased results and conceptual simplicity, Monte Carlo based techniques have become the standard approach.
+The main downside of these methods are their computational demands when rendering media with strong scattering or anisotropy.
 
@@
 
-The work on path guiding techniques from recent years has shown, how approximate representations of the steady-state transport in a scene can be used to accelerate Monte Carlo integration techniques, such as path tracing.
+The work on path guiding techniques from recent years has shown how approximate representations of the steady-state transport in a scene can be used to accelerate Monte Carlo integration techniques, such as path tracing.
 Instead of generating these approximate representations using Monte Carlo methods, deterministic methods may offer a viable alternative.
-Hybrid methods could combine benefits of deterministic methods while generating unbiased results.
+Hybrid methods could combine the performance benefits of deterministic methods with accurate and unbiased Monte Carlo results.
 
 Deterministic methods also lend themselves to applications where fast approximate solutions are preferable over correct, but slowly converging results.
-For this reason, we find it important to study deterministic methods and have a solid understanding about their characteristics and performance traits for typical rendering problems.
+For this reason, we suggest it is important for volume-rendering researchers to study deterministic methods and have a solid understanding of their characteristics and performance traits for typical rendering problems.
 
-The $P_N$-method is a deterministic method, which is popular in fields like medical imaging and nuclear sciences, but has not found use in computer graphics thus far.
+The $P_N$-method is a deterministic method which is popular in fields like medical imaging and nuclear sciences, but has not found use in computer graphics thus far.
 The purpose and main contribution of our paper is to gain a solid understanding of its foundations and present a method for using it in the context of rendering.

### 2018-06-08 latex/egsr17-cgf/previous_work.tex
revision 2

@@
 In strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become prohibitively expensive.
-Consider a high albedo media like milk, where tracing paths with many scattering events is necessary.
+Consider a high albedo medium like milk, where tracing paths with many scattering events is necessary.
 
@@
 It is derived by discretizing the angular variable of the radiative transfer equation into spherical harmonics (SH).
-This gives rise to a set of coupled, complex-valued partial differential equations, called $P_N$-equations.
-The subscript $N$ refers to the truncation order of the spherical harmonics.
+This gives rise to a set of coupled, complex-valued partial differential equations, called the $P_N$-equations.
+The subscript $N$ refers to the spherical harmonics truncation order.
 
@@
 This is further strengthened by our observation, that a straight forward finite difference discretization of the $P_N$-equations produces unuseable results, due to oscillation artifacts in the solution.
-In our paper, we derive the real-valued $P_N$-equations along with a staggered-grid solver, that produces artifact free solutions.
+We derive the real-valued $P_N$-equations with a staggered-grid solver, that produces artifact free solutions.
 
@@
 pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.
-The idea behind VEF is to derive a better diffusion coefficient by seeing it as a mean to spatially blend between the diffusive and pure transport regime.
+The idea behind VEF is to derive a better diffusion coefficient by seeing it as a means to spatially blend between the diffusive and pure transport regime.
 
 It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion.
-This question is also been raised in other domains~\cite{Olson00} and resolving it is one of our motivations.
+This question has also been raised in other domains~\cite{Olson00} and resolving it is one of our motivations.
 

### 2018-06-08 latex/egsr17-cgf/rendering.tex
revision 2

@@
 L\left(\vec{x},\omega\right) = L_{ss}\left(\vec{x},\omega\right) + L_{ms}\left(\vec{x},\omega\right)
+\ .
 \end{align}
@@
 Q(\vec{x}, \omega) = L_{ss}(\vec{x}, \omega) = \sigma_s\left(\vec{x}\right)\int_\Omega{ p\left(\omega'\rightarrow\omega\right) L_{u}\left(\vec{x}, \omega'\right)\ud\omega' }
+\ .
 \end{align}
@@
 The quantity $L_u$ is the uncollided light, which was emitted from the light source and attenuated by the volume without undergoing any scattering event.
-We compute it using a few light samples per voxel, which quickly converge to a useful result for Dirac Delta light sources.
+We compute it using a few light samples per voxel, which quickly converge to a useful result for Dirac delta light sources.
 
@@
 Then we sample a new direction according to the phase function.
-Instead of continuing tracing into the new direction, we evaluate the inscattering integral using $\widehat{L}_{ms}$.
+Instead of continuing tracing into the new direction, we evaluate the in-scattering integral using $\widehat{L}_{ms}$.
 The SH coefficients at $\vec{x}$ are found by trilinear interpolation from the voxel grid of SH coefficients.

### 2018-06-08 latex/egsr17-cgf/section_conclusion.tex
revision 2

@@
 
+\vspace{0.5in}
+
 The literature in other fields often states, that the $P_N$ method---when solved in normal form like we do---is able to deal with vacuum regions.
 We found this misleading.
-The $P_N$-method in normal form indeed does not break down completely in the presence of vacuum like diffusion based methods do (due to $\sigma_t^{-1}$ in the diffusion coefficient).
+The $P_N$-method in normal form indeed does not break down completely in the presence of vacuum as diffusion based methods do (due to $\sigma_t^{-1}$ in the diffusion coefficient).
 However, in the presence of vacuum, the condition number of the system matrix becomes infinite and does not converge either.
@@
 
-Much more work needs to be done in order to make our method competitive to alternative solutions.
+Much more work needs to be done in order to make the $P_N$ method competitive in performance to alternative solutions for volume rendering.
 We believe this can be made possible by employing a multigrid scheme for solving the linear system of equations.
@@
 This is possibly due to the coupling between coefficients within a voxel, which does not work well together with the smoothing step.
-We want to study this further in the future.
+We want to study this in the future.
 
-Unique to our system is, that it uses a computer algebra representation of the equation to solve as input.
+Unique to our system is that it uses a computer algebra representation of the equation to solve as input.
 Discretization in angular and spatial domain is done using manipulation passes on the representation.
-The stencil code, which is used to propagate the system of linear equations, is generated from the expression tree.
+The stencil code, which is used to populate the system of linear equations, is generated from the expression tree.
 This way, we can easily deal with coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand.

### 2018-06-08 latex/egsr17-cgf/section_discretized_rte.tex
minor revision
revision 2

@@
 +Q\left(\vec{x}, \omega\right)\nonumber
+\quad .
 \label{eq:rte}
@@
 The left hand side (LHS) is the transport term, and we refer to the terms on the right hand side (RHS) as collision, scattering, and source term, respectively.
-The symbols $\sigma_t$, $\sigma_s$, $p$, and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term.
+The symbols $\sigma_t$, $\sigma_s$, $p$, and $Q$ refer to the extinction- and scattering coefficient, phase function and emission.
 
@@
 \mathcal{T}\left(L\right) = -\mathcal{C}\left(L\right) + \mathcal{S}\left(L\right) + Q
+\quad .
 \end{align}
 
-Deterministic methods, like ours, are derived by discretizing the angular and spatial domain.
+Deterministic methods are derived by discretizing the angular and spatial domain.
 This gives rise to a linear system of equations, which can be solved using standard methods.
@@
 \right.
+\ .
 \label{eq:sh_real_basis}
@@
 \begin{align}
-\mathcal{P}^{l, m}(f) = \int_{\Omega}f(\mathbf{x}, \omega) \SHBR^{l,m}(\omega)\,\mathrm{d}\omega = f^{l,m}
+\mathcal{P}^{l, m}(f) = \int_{\Omega}f(\vec{x}, \omega) \SHBR^{l,m}(\omega)\,\mathrm{d}\omega = f^{l,m}\left(\vec{x}\right)
+\ .
 \nonumber
@@
 L\left(\vec{x}, \omega\right)
+\ .
 \nonumber
 \end{align}
+
+\vspace{0.5in}
 
@@
 +\mathcal{P}^{l,m}\left(Q\right)
+\ .
 \label{eq:pn_operator_notation}
@@
 
-Once the $P_N$-equation have been found, the spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxel grid (using central differences for differential operators).
+Once the $P_N$-equations have been found, the spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxel grid (using central differences for differential operators).
 
@@
 The RHS vector $\vec{Q}$ is produced similarly.
-The projected operators can be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$ (see figure~\ref{fig:matrix_layout}):
+The projected operators can be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$ (see Figure~\ref{fig:matrix_layout}):
 \begin{align}
 (T+C-S)\vec{u} = A\vec{u} = \vec{Q}
+\ .
 \end{align}
@@
 These are challenging and error-prone if done by hand.
-We therefore developed a computer algebra representation, which allowed us to derive the the $P_N$-equation in a semi-automatic fashion.
+We \rev{therefore used a computer algebra representation}, which allowed us to derive and discretize the $P_N$-equation in a semi-automatic fashion.

### 2018-06-08 latex/egsr17-cgf/section_results.tex
revision 2

@@
 We use a naive implementation of a CG solver, which has been modified such that we do not need to expliticly compute the normal form of the linear system to solve.
-We use the sparse matrix representation and sparse matrix vector product from the Eigen linear algebra library.
+We use the sparse matrix representation and sparse matrix vector product from the Eigen linear algebra library (eigen.tuxfamily.org).
 
@@
 \nabla\left(\frac{1}{3\sigma_t}\nabla L^{0,0}\right)  = -Q^{0,0}
+\ .
 \label{eq:cda}
@@
 
-Since our solver can work with any PDE, which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which would generate the correct stencil code automatically.
+Since our solver can work with any PDE, which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which generates the correct stencil code automatically.
 
@@
 The problem has dimensions $7\times 7$ and is discretized with resolution $71\times 71$.
-Unit size blocks are filled with purely absorbing medium with $\sigma_a=10$ in a checkerboard pattern.
+Unit size blocks are filled with purely absorbing medium $\sigma_a=10$ in a checkerboard pattern.
 All other blocks are filled with a purely scattering medium with $\sigma_s=1$.
@@
 Due to the time-dependency, their approach is to do explicit incremental steps in time.
-We run their solver for many timesteps, in order to get a result which is close to steady state.
+We run their solver for many timesteps to get a result close to steady state.
 
@@
 \vspace{-0.2in}
-\icaption{Comparison of the result for the checkerboard test using Starmaps time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$.
+\icaption{Comparison of the result for the checkerboard test using StaRMAP's time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$.
 Our results are in good agreement.}
@@
 
-\subsection{Pointsource problem}
+\subsection{Point source problem}
 
@@
 The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even.
-We further see, that the $P_1$ solution exactly matches the results from CDA, which confirmes, that the latter is only a collapsed version of the former.
+We further see that the $P_1$ solution exactly matches the results from CDA, which confirms that the latter is only a collapsed version of the former.
 The time to solve is significant with $10m$ for $P_1$ and $45m$ with $P_5$.
-With these performance characteristics, the $P_N$-method is clearly not competitive in comparison with much diffusion and FLD solver, which are much faster.
+With these performance characteristics, the $P_N$-method is clearly not competitive in comparison with CDA and FLD solver, which are much faster.
 
@@
 We see that at order $N=5$, our method can capture indirect illumination similarily well as FLD and is significantly better than CDA as expected.
-The indirectly illuminated region at the bottom appears to be closer to the path-traced result as opposed to the solution form FLD which is very flat in that region.
+The indirectly illuminated region at the bottom appears to be closer to the path-traced result as opposed to the solution from FLD which is very flat in that region.
 However, in many other areas, $P_5$ seems to still suffer a lot from the problem of energy loss near transitions from dense regions to vacuum.
-It appears that going higher order a few steps, mitigates this problem only slowly at an immense cost of compute and storage.
+It appears that going higher order a few steps mitigates this only slowly at a high cost of compute and storage.
 
 The main characteristic of the nebulae dataset is the presence of vacuum.
-We found that having vacuum regions in the dataset will cause the condition number to become infinite and the solver practically doesnt converge.
+We found that having vacuum regions in the dataset will cause the condition number to become infinite and the solver practically does not converge.
 We therefore introduced a minimum threshold for the extinction coefficient $\sigma_t$.
-Every voxel with an extinction coefficient smaller than the threshold would be set to the threshold value.
+Every voxel with an extinction coefficient smaller than the threshold is set to the threshold value.
 In Figure~\ref{fig:results_convergence} we show the effect of the minimum threshold on the convergence behavior.
-As the threshold increases, convergence improves.
+As it increases, convergence improves.
 
@@
 The convergence deteriorates as the threshold decreases.
-Once it reaches zero, the presence of pure vaccuum makes the condition number infinite.}
+Once it reaches zero, the presence of pure vacuum makes the condition number infinite.}
 \label{fig:results_convergence}

### 2018-06-08 latex/egsr17-cgf/solver.tex
revision 2

@@
 The truncation order $N$ is the key input parameter to the solver.
-With higher values, manual implementation of the solver from the equations would be ardous, error-prone and time-consuming.
+With higher values, manual implementation of the solver from the equations would be arduous, error-prone and time-consuming.
 We therefore decided to make use of the computer algebra representation and designed our solver around it.
@@
 
-\begin{figure}[b]
+\vspace{0.6in}
+
+Stencil generation entails discretizing a PDE at a hypothetical center voxel $(i, j, k)$ (assumed to mean the voxel center most of the time).
+Finite differences create weighted references to other voxels (e.g.
+$i+1, j, k$).
+After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and offsets (Figure~\ref{fig:stencile_pipeline}).
+Voxel $(i, j, k)$ will only be known during runtime, when the stencil is executed for a particular unknown (row).
+Then the offsets can be used to find the component index into the solution-vector, and weights can be evaluated for concrete world space position.
+We refer to the space with the hypothetical voxel $(i, j, k)$ at the center as stencil space.
+
+\vspace{0.1in}
+
+\begin{figure}[h]
 \centering
@@
 \end{figure}
-
-\vspace{0.6in}
-
-Stencil generation entails discretizing a PDE at a hypothetical center voxel $(i, j, k)$ (assuming to mean the voxel center most of the time).
-Finite differences create weighted references to other voxels (e.g.
-$i+1, j, k$).
-After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and offsets (Figure~\ref{fig:stencile_pipeline}).
-$(i, j, k)$ will only be known during runtime, when the stencil is executed for a particular unknown (row).
-Then the offsets can be used to find the component index into the solution-vector, and weights can be evaluated for concrete world space position.
-We refer to the space with the hypthetical voxel $(i, j, k)$ at the center as stencil space.
 
@@
 The subtree of the differential operator node is duplicated for different offsets to the discrete variable $(i, j, k)$.
-Since this offset only applies to the substree, a stack is maintained by the parser to manage scope.
+Since this offset only applies to the subtree, a stack is maintained by the parser to manage scope.
 Whenever the parser encounters the continuous variable $\vec{x}$ in the expression tree, its node in the tree is replaced by the discrete substitute, currently on top of the stack.
@@
 Factorization into canonical form is done as a manipulation pass on the mathematical expression tree.
-The result allows to implement the code generation in a straight forward fashion.
+The result allows us to implement the code generation in a straightforward fashion.
 For each term, the $ijk$-offset is retrieved from the unknown.
@@
 This means, that during the discretization step, the discrete location $(i, j, k)$ (at which the unknown is meant to be evaluated) might not coincide with the location of the unknown.
-To solve this, depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at $(i, j, k)$ from their defined locations.
+To solve this, depending on how the two are located relativeto each other, the parser returns an expression which interpolates the coefficients at $(i, j, k)$ from their defined locations.
 If those happen to coincide, the coefficient itsself is returned.
@@
 The stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver.
-The runtime executes the stencil for every voxel to propagate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.
+The runtime executes the stencil for every voxel to populate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.
 
@@
 Unfortunately, it is non-symmetric due to the transport term and not diagonal dominant, which rules out many \rev{iterative} methods for solving linear systems.
-\rev{Iterative methods are useful, as they allow balancing accuracy against performance by tweaking the convergence threshold.} We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead.
-This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number.
+\rev{Iterative methods are useful, as they allow balancing accuracy against performance by tweaking the convergence threshold.} We address this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead.
+This gives a symmetric and positive definite system matrix $A^TA$, albeit with a higher condition number.
 Investigation of other solution schemes (e.g.
@@
 Whenever the stencil accumulates coefficients into a boundary location, the framework either ignores the write operation (Dirichlet BC) or accumulates into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC).
-This is done changing the index of the written component under the hood.
+This is done by changing the index of the written component.
 

### 2018-06-25 latex/egsr17-cgf/previous_work.tex
applied camera ready fixes

@@
 
-Light transport in participating media is governed by the radiative transfer equation~(RTE), first studied in the context of astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later introduced to computer graphics by Kajiya~\cite{Kajiya86}.
+Light transport in participating media is governed by the radiative transfer equation~(RTE), first studied in the context of astrophysics by Chandrasekhar~\cite{Chandrasekhar60} and later introduced to computer graphics by Kajiya~\cite{Kajiya86}.
 Today, this equation is typically solved using Monte Carlo methods.
 In strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become prohibitively expensive.
-Consider a high albedo medium like milk, where tracing paths with many scattering events is necessary.
+For example in the case of a high albedo medium like milk where tracing paths with many a huge number of scattering events is necessary.
+
+%Consider a high albedo medium like milk, where tracing paths with many scattering events is necessary.
 
@@
 The $P_N$-method has a long history in other fields and was never applied in graphics.
-Kajia~\cite{Kajiya84} explained the theory, but did not give any details on implementation or how to solve it.
+Kajiya~\cite{Kajiya84} explained the theory, but did not give any details on implementation or how to solve it.
 In fact, as Max~\cite{Max95} pointed out, it is not clear if Kajiya succeeded at all at applying the method, as all of the results in his paper were produced with a simpler method.
-This is further strengthened by our observation, that a straight forward finite difference discretization of the $P_N$-equations produces unuseable results, due to oscillation artifacts in the solution.
-We derive the real-valued $P_N$-equations with a staggered-grid solver, that produces artifact free solutions.
+This is further strengthened by our observation that a straightforward finite difference discretization of the $P_N$-equations produces unusable results, due to oscillation artifacts in the solution.
+We derive the real-valued $P_N$-equations with a staggered-grid solver, that produces artifact-free solutions.
 
@@
 CDA suffers from severe energy loss close to regions with strong density gradients.
-The problem can be addressed by a modification known as Variable Eddington factor (VEF), which nonlinearly adjusts the diffusion coefficient to improve the solution near density gradients and low-density regions.
+The problem can be addressed by a modification known as the Variable Eddington factor (VEF) method, which nonlinearly adjusts the diffusion coefficient to improve the solution near density gradients and low-density regions.
 Flux-limited diffusion, developed in the context of astrophysics by Levermore et al.~\cite{Levermore81} and later introduced to graphics by Koerner et al.~\cite{Koerner14}, is the most prominent example.
 
-VEF is based on the observation, that the behavior of volumetric transport is closely linked to the density of the material: In the diffusive regime, thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
+VEF is based on the observation that the behavior of volumetric transport is closely linked to the density of the material: In the diffusive regime, thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
 In the absence of scattering (e.g.
@@
 
-It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion.
+It is an open and unresolved question whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion.
 This question has also been raised in other domains~\cite{Olson00} and resolving it is one of our motivations.

### 2018-06-25 latex/egsr17-cgf/section_conclusion.tex
applied camera ready fixes

@@
 We derived and presented the real-valued $P_N$-equations, along with a staggered grid solver for solving them.
-We showed, how to use the results in a rendering system and ran our solver for various standard problems for validation.
+We showed how to use the results in a rendering system and ran our solver for various standard problems for validation.
 
@@
 
-The literature in other fields often states, that the $P_N$ method---when solved in normal form like we do---is able to deal with vacuum regions.
+The literature in other fields often states that the $P_N$ method---when solved in normal form like we do---is able to deal with vacuum regions.
 We found this misleading.
@@
 The stencil code, which is used to populate the system of linear equations, is generated from the expression tree.
-This way, we can easily deal with coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand.
+This way, we can easily deal with coupled PDEs and avoid the time consuming and error prone process of writing stencil code by hand.
 
@@
 These variations have been introduced to address certain problems of the standard $P_N$-method, such as ringing artifacts, dealing with vacuum regions and general convergence.
-Our solver can be applied to any (potentially coupled) PDE and therefore can generate stencil code for all these variations by simply expressing the respective PDE's in our computer algebra representation and providing this as an input to our solver.
+Our solver can be applied to any (potentially coupled) PDE and therefore can generate stencil code for all these variations by simply expressing the respective PDEs in our computer algebra representation and providing this as an input to our solver.
 This would allow an exhaustive comparison of all these methods and we consider this as future work.

### 2018-06-25 latex/egsr17-cgf/section_discretized_rte.tex
adding gitignore
applied camera ready fixes

@@
 
-Our method derives from the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
+Our method derives from the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position in direction $\omega$ at point $\vec{x}$:
 \begin{align}
@@
 For the $P_N$-method, the angular variable is first discretized, using a truncated spherical harmonics expansion.
-This results in the $P_N$-equations, a system of coupled PDE's that still depend on a continuous spatial variable.
+This results in the $P_N$-equations, a system of coupled PDEs that still depend on a continuous spatial variable.
 
 The number of equations grows with the truncation order $N$.
-This is why the discretization is hard and error prone if done by hand.
+This is why the discretization is laborious and difficult to do without errors if done by hand.
 We therefore use a computer algebra representation to automate this process.
@@
 \includegraphics[width=\columnwidth]{figures/fig_matrix_layout.pdf}
+%\missingfigure{figures fig matrix layout}
 \vspace{-0.2in}

### 2018-06-25 latex/egsr17-cgf/section_results.tex
adding gitignore
applied camera ready fixes

@@
 \begin{subfigure}{0.49\columnwidth}
-%\missingfigure{test}
 \includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered_starmap.png}
@@
 \begin{subfigure}{0.49\columnwidth}
-%\missingfigure{test2}
 \includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered.png}
@@
 \vspace{-0.2in}
-\icaption{Comparison of the result for the checkerboard test using StaRMAP's time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$.
+\icaption{Comparison of the result for the checkerboard test using \textsf{StaRMAP}'s time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$.
 Our results are in good agreement.}
@@
 
-As can be seen in figure~\ref{fig:vs_starmap}, the results from our solver are in good agreement with the results from Seibold et al.~\cite{Seibold14} and verify the correctness of our implementation.
+As can be seen in Figure~\ref{fig:vs_starmap}, the results from our solver are in good agreement with the results from Seibold et al.~\cite{Seibold14} and verify the correctness of our implementation.
 Converging to a residual of $10e^{-10}$ takes $0.27s$ for $P_1$ and $25s$ for $P_5$.
@@
 \includegraphics[width=\columnwidth]{figures/pointsource_pn.pdf}
-%\missingfigure{test}
+%\missingfigure{figures fig pointsource pn}
 \caption{$P_N$ vs. ground truth}
@@
 \begin{subfigure}{0.45\columnwidth}
-%\missingfigure{test2}
 \includegraphics[width=\columnwidth]{figures/pointsource_p5.pdf}
+%\missingfigure{figures fig pointsource p5}
 \caption{$P_5$ vs. CDA and FLD}
@@
 Finally, we run our solver on a procedural cloud dataset to get an idea of its performance in more practical applications.
-Figure~\ref{fig:teaser}~\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function.
-We see that at order $N=5$, our method can capture indirect illumination similarily well as FLD and is significantly better than CDA as expected.
+Figure~\ref{fig:teaser}\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function.
+We see that at order $N=5$, our method can capture indirect illumination similarly well as FLD and is significantly better than CDA as expected.
 The indirectly illuminated region at the bottom appears to be closer to the path-traced result as opposed to the solution from FLD which is very flat in that region.
@@
 \centering
-%\missingfigure{test}
 \includegraphics[width=0.9\columnwidth]{figures/fig_nebulae_p1_convergence.pdf}
-%\vspace{-0.2in}
+%\missingfigure{figures fig nebulae p1 convergence}
 \vspace{-0.1in}

### 2018-06-29 latex/egsr17-cgf/introduction.tex
final version with figures of smaller filesize

@@
 
-The work on path guiding techniques from recent years has shown how approximate representations of the steady-state transport in a scene can be used to accelerate Monte Carlo integration techniques, such as path tracing.
+The work on path-guiding techniques from recent years has shown how approximate representations of the steady-state transport in a scene can be used to accelerate Monte Carlo integration techniques, such as path tracing.
 Instead of generating these approximate representations using Monte Carlo methods, deterministic methods may offer a viable alternative.
@@
 \begin{itemize}
-	\item We derive and present the time-independent real-valued $P_N$-equations. To our knowledge, we are the first to derive and write down their compact form.
+	\item We derive and present the time-independent real-valued $P_N$-equations and write them down in a very concise and compact form which we have not found anywhere else in the literature.
 	\item We introduce a staggered-grid solver, for which we generate stencil code automatically from a computer algebra representation of the $P_N$-equations.
-This allows us to deal with the increasingly complex equations, which the $P_N$-method produces for higher order.
-It further allows our solver to be used for any (potentially coupled) partial differential equation, which results in a system of linear equations after discretization.
+This allows us to deal with the increasingly complex equations which the $P_N$-method produces for higher order.
+It further allows our solver to be used for any (potentially coupled) partial differential equations, which result in a system of linear equations after discretization.
 	\item Finally, we compare the $P_N$-method for higher orders against flux-limited diffusion and ground truth Monte Carlo integration.
@@
 This representation is also a key component of our solver, which we present in Section~\ref{sec:pnsolver}.
-Section~\ref{sec:rendering} discusses application of the solution to the context of rendering.
+Section~\ref{sec:rendering} discusses application of the solution in the context of rendering.
 We compare our $P_N$-solver against flux-limited diffusion for a set of standard problems in Section~\ref{sec:results}.

### 2018-06-29 latex/egsr17-cgf/previous_work.tex
final version with figures of smaller filesize

@@
 In strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become prohibitively expensive.
-For example in the case of a high albedo medium like milk where tracing paths with many a huge number of scattering events is necessary.
+For example in the case of a high albedo medium like milk where tracing paths with a huge number of scattering events is necessary.
 
@@
 
-VEF is based on the observation that the behavior of volumetric transport is closely linked to the density of the material: In the diffusive regime, thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
+VEF is based on the observation that the behavior of volumetric transport is closely linked to the density of the material: in the diffusive regime, thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines.
 In the absence of scattering (e.g.

### 2018-06-29 latex/egsr17-cgf/section_car.tex
revision 2
adding gitignore
final version with figures of smaller filesize

@@
 \centering
-\includegraphics[width=0.9\columnwidth]{figures/fig_car.pdf}
+\includegraphics[width=0.9\columnwidth]{figures/fig_car_small.pdf}
+%\missingfigure{figures fig car}
 \vspace{-0.16in}
@@
 \rev{Using the computer algebra representation, we perform the derivation steps required to arrive at the real-valued $P_N$-equations (the derivation steps shown in the supplemental material were almost all rendered from the expression tree).
-More importantly, we use the representation to perform automated discretization and generate the stencil code used by our solver.
+More importantly, we use the representation to perform the discretization and generate the stencil code used by our solver.
 This is detailed in section~\ref{sec:solver_precomputation}}.

### 2018-06-29 latex/egsr17-cgf/section_results.tex
final version with figures of smaller filesize

@@
 
-Since our solver can work with any PDE, which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which generates the correct stencil code automatically.
+Since our solver can work with any PDE which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which generates the correct stencil code automatically.
 
@@
 \begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\columnwidth]{figures/pointsource_pn.pdf}
+\includegraphics[width=\columnwidth]{figures/pointsource_pn_small.pdf}
 %\missingfigure{figures fig pointsource pn}
@@
 \begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\columnwidth]{figures/pointsource_p5.pdf}
+\includegraphics[width=\columnwidth]{figures/pointsource_p5_small.pdf}
 %\missingfigure{figures fig pointsource p5}
@@
 The time to solve is significant with $10m$ for $P_1$ and $45m$ with $P_5$.
-With these performance characteristics, the $P_N$-method is clearly not competitive in comparison with CDA and FLD solver, which are much faster.
+With these performance characteristics, our $P_N$-solver is clearly not competitive in comparison with CDA and FLD solver, which are much faster.
 
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{figures/fig_nebulae_p1_convergence.pdf}
+\includegraphics[width=0.9\columnwidth]{figures/fig_nebulae_p1_convergence_small.pdf}
 %\missingfigure{figures fig nebulae p1 convergence}

### 2018-06-29 latex/egsr17-cgf/solver.tex
adding gitignore
applied camera ready fixes
final version with figures of smaller filesize

@@
 %\missingfigure{$P_N$-solver overview: generate stencil code $\rightarrow$ build system $\rightarrow$ solve $\rightarrow$ render}
-\includegraphics[width=\textwidth]{figures/fig_pipeline.pdf}
+\includegraphics[width=\textwidth]{figures/fig_pipeline_small.pdf}
+%\missingfigure{figures fig pipeline}
 \vspace{-0.2in}
@@
 The first is a precomputation (Section~\ref{sec:solver_precomputation}), which is excuted once for every single value of $N$.
-This steps runs a partial evaluation on the mathematical expression tree and applies the spatial discretization in a reference space we call stencil space.
+This step runs a partial evaluation on the mathematical expression tree and applies the spatial discretization in a reference space we call stencil space.
 The precomputation step automatically generates source code from the resulting expression tree.
@@
 The stencil is a pattern of indices into the solution vector, along with values.
-It expresses how the sum of the weighted solution vector, components relate to the RHS for a given unknown in the system and therefore contains most information required to fill the system matrix $A$ and RHS-vector $\vec{Q}$ row by row.
+It expresses how the sum of the weighted solution vector components relate to the RHS for a given unknown in the system and therefore contains most information required to fill the system matrix $A$ and RHS-vector $\vec{Q}$ row by row.
 Note that while the coefficients may change, the sparsity pattern of the stencil is identical for different rows.
@@
 %\missingfigure{test}
-\includegraphics[width=0.7\columnwidth]{figures/fig_stencil_pipeline.pdf}
+\includegraphics[width=0.7\columnwidth]{figures/fig_stencil_pipeline_small.pdf}
+%\missingfigure{figures fig stencil pipeline}
 %\vspace{-0.2in}
@@
 \hspace{-.2in}
-\includegraphics[width=0.6\columnwidth]{figures/fig_discretization.pdf}
+\includegraphics[width=0.6\columnwidth]{figures/fig_discretization_small.pdf}
+%\missingfigure{figures fig discretization}
 %\end{center}
@@
 \hspace{-.2in}
-\includegraphics[width=0.6\columnwidth]{figures/fig_codegen.pdf}
+\includegraphics[width=0.6\columnwidth]{figures/fig_codegen_small.pdf}
+%\missingfigure{figures fig codegen}
 %\end{center}
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{figures/fig_staggered_grids.pdf}
+\includegraphics[width=0.9\columnwidth]{figures/fig_staggered_grids_small.pdf}
+%\missingfigure{figures fig staggered grids}
 %\vspace{-0.2in}
@@
 \hspace{-.1in}
-\includegraphics[width=0.6\columnwidth]{figures/fig_staggered_interpolation.pdf}
+\includegraphics[width=0.6\columnwidth]{figures/fig_staggered_interpolation_small.pdf}
+%\missingfigure{figures fig staggered interpolation}
 \vspace{-.1in}
@@
 \end{wrapfigure}Our solver supports placement of coefficients at arbitrary staggered grid locations (see Figure~\ref{fig:staggeredgrid}).
-This means, that during the discretization step, the discrete location $(i, j, k)$ (at which the unknown is meant to be evaluated) might not coincide with the location of the unknown.
-To solve this, depending on how the two are located relativeto each other, the parser returns an expression which interpolates the coefficients at $(i, j, k)$ from their defined locations.
-If those happen to coincide, the coefficient itsself is returned.
+This means that during the discretization step, the discrete location $(i, j, k)$ (at which the unknown is meant to be evaluated) might not coincide with the location of the unknown.
+To solve this, depending on how the two are located relative to each other, the parser returns an expression which interpolates the coefficients at $(i, j, k)$ from their defined locations.
+If those happen to coincide, the coefficient itself is returned.
 This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center.

### 2018-07-15 arxiv/paper.tex
Adding arxiv version

\documentclass{egpubl}
\usepackage{egsr18-e}

% --- for  Annual CONFERENCE
% \ConferenceSubmission % uncomment for Conference submission
% \ConferencePaper      % uncomment for (final) Conference Paper
% \STAR                 % uncomment for STAR contribution
% \Tutorial             % uncomment for Tutorial contribution
% \ShortPresentation    % uncomment for (final) Short Conference Presentation
%
% --- for  CGF Journal
% \JournalSubmission    % uncomment for submission to Computer Graphics Forum
% \JournalPaper         % uncomment for final version of Journal Paper
%
% --- for  CGF Journal: special issue
 %\SpecialIssueSubmission    % uncomment for submission to Computer Graphics Forum, special issue
%\SpecialIssuePaper         % uncomment for final version of Journal Paper, special issue
%
% --- for  EG Workshop Proceedings
% \WsSubmission    % uncomment for submission to EG Workshop
\WsPaper         % uncomment for final version of EG Workshop contribution
%
 \electronicVersion % can be used both for the printed and electronic version

% !! *please* don't change anything above
% !! unless you REALLY know what you are doing
% ------------------------------------------------------------------------

% for including postscript figures
% mind: package option 'draft' will replace PS figure by a filname within a frame
\ifpdf \usepackage[pdftex]{graphicx} \pdfcompresslevel=9

\else \usepackage[dvips]{graphicx} \fi

\PrintedOrElectronic

% prepare for electronic version of your document
\usepackage{t1enc,dfadobe}

\usepackage{egweblnk}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{subcaption} % subfigure
%\usepackage{units}
% \usepackage{array}
% \usepackage{multirow}
% \usepackage{caption}
% \usepackage{dsfont}
% \usepackage{booktabs}
% \usepackage{rotating}
% \usepackage[ruled,linesnumbered,vlined,algo2e]{algorithm2e}
% \usepackage{bm}
\usepackage{overpic}
\usepackage{contour}
\usepackage{xspace}
\usepackage[normalem]{ulem} % for strikeout \sout

\usepackage{wrapfig}

\usepackage[ruled,linesnumbered,vlined,algo2e]{algorithm2e}
% For backwards compatibility to old LaTeX type font selection.
% Uncomment if your document adheres to LaTeX2e recommendations.
% \let\rm=\rmfamily    \let\sf=\sffamily    \let\tt=\ttfamily
% \let\it=\itshape     \let\sl=\slshape     \let\sc=\scshape
% \let\bf=\bfseries

\usepackage{scalerel} % very small subscripts
%\usepackage{datetime} % for timestamping
%\usepackage[yyyymmdd,hhmmss]{datetime}

\newcommand{\ud}{\,\mathrm{d}} % for integrals
\newcommand{\SHBR}{Y_{\mathbb{R}}} % for integrals
\newcommand{\SHBC}{Y_{\mathbb{C}}} % for integrals
\newcommand{\iu}{{i\mkern1mu}}
\newcommand{\icaption}[1]{\caption{#1}}
%\newcommand{\icaption}[1]{\caption{\emph{#1}}} % italic caption
\newcommand{\colvec}[3]{\ensuremath{\begin{pmatrix}#1\\#2\\#3\end{pmatrix}}}

%\newcommand{\rev}[1]{\textcolor{red}{#1}}
\newcommand{\rev}[1]{\textcolor{black}{#1}}

\newcommand{\wc}{.}

\makeatletter
\def\mathcolor#1#{\@mathcolor{#1}}
\def\@mathcolor#1#2#3{%
  \protect\leavevmode
  \begingroup\color#1{#2}#3\endgroup
}
\makeatother

\definecolor{orange}{rgb}{1,0.5,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{red}{rgb}{1,0.0,0}

%\newcommand{\realn}{\mathbb{R}}
%\newcommand{\vect}[1]{\mathbf{#1}}
%\newcommand{\ud}{\,\mathrm{d}} % for integrals
%\newcommand\norm[1]{\left\lVert#1\right\rVert}
%\newcommand{\COMMENT}[3]{\textcolor{#2}{{\small[\textsc{#1}: #3}]}}  % For work in progress using in-text comments
%\newcommand{\DK}[1]{\COMMENT{DK}{red}{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\usepackage{fullpage}
%\usepackage{lmodern}
%\usepackage[OT1]{fontenc}

%\usepackage[nottoc]{tocbibind} % Add entry for references to \tableofcontents
%\usepackage[squaren]{SIunits}

%\usepackage{xcolor}

%\usepackage{listings}% see \lstset below for package settings
%\usepackage{subfigure}
%\usepackage{SIunits}

%\usepackage{graphicx}
%\usepackage{caption}
%\captionsetup{width=0.9\textwidth}
\usepackage{titletoc}

%\usepackage{geometry}
%\usepackage{lipsum}
%\usepackage{fancyhdr}
%\pagestyle{fancy}

%\usepackage{textcomp}

%\usepackage[parfill]{parskip}
%\usepackage{mathtools}

%\definecolor{visusblue}{RGB}{19 124 190} % Color taken from VISUS slides
%\DeclarePairedDelimiterX{\norm}[1]{\lVert}{\rVert}{#1}

\newcommand{\nocontentsline}[3]{}
\newcommand{\tocless}[2]{\bgroup\let\addcontentsline=\nocontentsline#1{#2}\egroup}

% end of prologue

\usepackage{todonotes}

\captionsetup{labelfont=bf,textfont=it}

% ---------------------------------------------------------------------
% EG author guidelines plus sample file for EG publication using LaTeX2e input
% D.Fellner, v2.02, Jan 25, 2017

\title%
{
%Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media
$P_N$-Method for Multiple Scattering in Participating Media
}

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
%\author[D. Fellner \& S. Behnke]
%{\parbox{\textwidth}{\centering D.\,W. Fellner\thanks{Chairman Eurographics Publications Board}$^{1,2}$
%        and S. Behnke$^{2}$
%        S. Spencer$^2$\thanks{Chairman Siggraph Publications Board}
%        }
%        \\
%% For Computer Graphics Forum: Please use the abbreviation of your first name.
%{\parbox{\textwidth}{\centering $^1$TU Darmstadt \& Fraunhofer IGD, Germany\\
%         $^2$Graz University of Technology, Institute of Computer Graphics and Knowledge Visualization, Austria
%%        $^2$ Another Department to illustrate the use in papers from authors
%%             with different affiliations
%       }
%}
%}

\author[D. Koerner \& J. Portsmouth \& W. Jakob]
{
%Compiled on \today
%paper1027
David Koerner$^{1}$
\qquad Jamie Portsmouth$^{2}$
\qquad Wenzel Jakob$^{3}$
%\qquad Ralf Habel$^{2}$
%\qquad Wojciech Jarosz$^{1,4}$
\\
% For Computer Graphics Forum: Please use the abbreviation of your first name.
$^1$University of Stuttgart
\qquad $^2$Solid Angle
\qquad $^3$\'{E}cole polytechnique f\'{e}d\'{e}rale de Lausanne (EPFL)
%\qquad $^4$Dartmouth College
}

% ------------------------------------------------------------------------

% if the Editors-in-Chief have given you the data, you may uncomment
% the following five lines and insert it here
%
% \volume{36}   % the volume in which the issue will be published;
% \issue{1}     % the issue number of the publication
% \pStartPage{1}      % set starting page

%-------------------------------------------------------------------------
\begin{document}
% uncomment for using teaser
\teaser
{
\centering
\begin{subfigure}{0.24\linewidth}
%\missingfigure{}
\includegraphics[width=\linewidth]{nebulae_ms_cda.png}
\vspace{-0.17in}
\caption{Classical diffusion}
\label{fig:teaser_cda}
\end{subfigure}%
\hspace{0.002\linewidth}
\begin{subfigure}{0.24\linewidth}
\includegraphics[width=\linewidth]{nebulae_ms_p5.png}
\vspace{-0.17in}
\caption{$P_5$ (ours)}
\label{fig:nebulae_ours}
\label{fig:teaser_pn}
\end{subfigure}%
\hspace{0.002\linewidth}
\begin{subfigure}{0.24\linewidth}
%\missingfigure{}
\includegraphics[width=\linewidth]{nebulae_ms_fld.png}
\vspace{-0.17in}
\caption{Flux-limited diffusion}
\label{fig:teaser_fld}
\end{subfigure}%
\hspace{0.002\linewidth}
\begin{subfigure}{0.24\linewidth}
%\missingfigure{}
\includegraphics[width=\linewidth]{nebulae_ms_groundtruth.png}
\vspace{-0.17in}
\caption{Brute force path tracing}
\label{fig:teaser_gt}
\end{subfigure}%
%\includegraphics[width=\linewidth]{figures/eg_new}
\vspace{-0.1in}
\icaption
{
Flux-limited diffusion (\subref{fig:teaser_fld}) is an extension to the classical diffusion approximation (\subref{fig:teaser_cda}). It improves accuracy, but is based on ad-hoc assumptions about volumetric transport. We add the $P_N$-method (\subref{fig:teaser_pn}), which allows solution of the RTE with configurable accuracy, to the toolbox of methods in volume rendering and investigate its benefits and trade-offs against flux-limited diffusion.
}
\label{fig:teaser}
}

\maketitle

%-------------------------------------------------------------------------

\begin{abstract}
Rendering highly scattering participating media using brute force path tracing is a challenge. The diffusion approximation reduces the problem to solving a simple linear partial differential equation. Flux-limited diffusion introduces non-linearities to improve the accuracy of the solution, especially in low optical depth media, but introduces several ad-hoc assumptions. Both methods are based on a spherical harmonics expansion of the radiance field that is truncated after the first order. In this paper, we investigate the open question of whether going to higher spherical harmonic orders provides a viable improvement to these two approaches. Increasing the order introduces a set of complex coupled partial differential equations (the \emph{$P_N$-equations}), whose growing number make them difficult to work with at higher orders. We thus \rev{use} a computer algebra framework for representing and manipulating the underlying mathematical equations, and use it to derive the real-valued $P_N$-equations for arbitrary orders. We further present a staggered-grid $P_N$-solver and generate its stencil code directly from the expression tree of the $P_N$-equations. Finally, we discuss how our method compares to prior work for various standard problems.
%\begin{classification} % according to http://www.acm.org/class/1998/
%\CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Raytracing}
%\end{classification}
\end{abstract}

%-------------------------------------------------------------------------

\tocless\section{Introduction}
%\vspace{0.1in}

Simulating light transport in participating media remains a challenging problem for image synthesis in computer graphics. Due to their ability to produce unbiased results and conceptual simplicity, Monte Carlo based techniques have become the standard approach~\cite{Novak18}. The main downside of these methods are their computational demands when rendering media with strong scattering or anisotropy.

Deterministic methods have enjoyed less popularity, because they suffer from discretization artifacts, produce biased results, cannot be coupled easily with surface rendering problems and are trickier to implement. However, their appeal lies in the fact that they produce a global solution across the whole domain and have better performance for certain problems~\cite{Brunner02}.

The work on path-guiding techniques from recent years~\cite{Muller17} has shown how approximate representations of the steady-state transport in a scene can be used to accelerate Monte Carlo integration techniques, such as path tracing. Instead of generating these approximate representations using Monte Carlo methods, deterministic methods may offer a viable alternative. Hybrid methods could combine the performance benefits of deterministic methods with accurate and unbiased Monte Carlo results.
Deterministic methods also lend themselves to applications where fast approximate solutions are preferable over correct, but slowly converging results.

For these reasons, we suggest it is important for volume-rendering researchers to study deterministic methods and have a solid understanding of their characteristics and performance traits for typical rendering problems.

The $P_N$-method is a deterministic method of solving the radiative transfer equation~(RTE) which is used in other fields such as medical imaging and nuclear sciences, but has not found use in computer graphics thus far. The purpose and main contribution of our paper is to gain a solid understanding of its foundations and present a method for using it in the context of rendering. In particular, we present these theoretical and practical contributions:
\begin{itemize}
	\item We derive and present the time-independent real-valued $P_N$-equations and write them down in a very concise and compact form which we have not found anywhere else in the literature.
	\item We introduce a staggered-grid solver, for which we generate stencil code automatically from a computer algebra representation of the $P_N$-equations. This allows us to deal with the increasingly complex equations which the $P_N$-method produces for higher order. It further allows our solver to be used for any (potentially coupled) partial differential equations, which result in a system of linear equations after discretization.
	\item Finally, we compare the $P_N$-method for higher orders against flux-limited diffusion and ground truth Monte Carlo integration.
\end{itemize}

In the next section, we will discuss related work and its relation to our contribution. In Section~\ref{sec:discretized_rte} we revisit the deterministic approach to light transport simulation in participating media and outline the discretization using spherical harmonics. In Section~\ref{sec:car} we introduce our computer algebra representation, which we required to derive the real-valued $P_N$-equations, presented in Section~\ref{sec:real_valued_pn_eq}. This representation is also a key component of our solver, which we present in Section~\ref{sec:pnsolver}. Section~\ref{sec:rendering} discusses application of the solution in the context of rendering. We compare our $P_N$-solver against flux-limited diffusion for a set of standard problems in Section~\ref{sec:results}. Finally, Section~\ref{sec:conclusion} concludes with a summary and review of future work.

%-------------------------------------------------------------------------

\tocless\section{Previous work}

Light transport in participating media is governed by the RTE, first studied in the context of astrophysics by Chandrasekhar~\cite{Chandrasekhar60} and later introduced to computer graphics by Kajiya~\cite{Kajiya86}. In computer graphics today, this equation is typically solved using Monte Carlo methods~\cite{Novak18}. However in strongly scattering or highly anisotropic media these methods can become prohibitively expensive, for example in the case of a high albedo medium such as milk where tracing paths with a huge number of scattering events is necessary.

%Consider a high albedo medium like milk, where tracing paths with many scattering events is necessary.
In contrast to path-tracing, the $P_N$-method~\cite{Brunner02} gives a solution by solving a system of linear equations. It is derived by discretizing the angular variable of the radiative transfer equation into spherical harmonics (SH). This gives rise to a set of coupled, complex-valued partial differential equations, called the $P_N$-equations. The subscript $N$ refers to the spherical harmonics truncation order.

The $P_N$-method has a long history in other fields and was never applied in graphics. Kajiya~\cite{Kajiya84} explained the theory, but did not give any details on implementation or how to solve it. In fact, as Max~\cite{Max95} pointed out, it is not clear if Kajiya succeeded at all at applying the method, as all of the results in his paper were produced with a simpler method. This is further strengthened by the fact that a straightforward finite difference discretization of the $P_N$-equations produces unusable results, due to oscillation artifacts in the solution ~\cite{Seibold14}. We use a staggered-grid solver, motivated by the solver of Seibold et al.~\cite{Seibold14}, that produces artifact-free solutions (see Figure~\ref{fig:artifacts}).
\begin{figure}[t]
\centering
\begin{subfigure}{0.45\columnwidth}
%\missingfigure{test}
\includegraphics[width=\columnwidth]{checkerboard2d_p1_collocated.png}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
%\missingfigure{test2}
\includegraphics[width=\columnwidth]{checkerboard2d_p1_staggered.png}
\end{subfigure}%
%\vspace{-0.1in}
\icaption{Solving the 2D checkerboard problem using naive collocated grids produces oscillating artifacts (left). Our solver uses staggered grids and produces artifact free results (right). \vspace{-0.1in}}
\label{fig:artifacts}
\end{figure}

Related to the $P_N$-method, the classical diffusion approximation (CDA) is a deterministic method which arrives at a solution by solving a system of linear equations. It corresponds to the $P_N$-equations when $N=1$ (truncation after the first SH order), which can be collapsed to a simple diffusion equation, giving the method its name. CDA has a long history in other domains, such as astrophysics and nuclear physics~\cite{Brunner02}, and was introduced to graphics by Stam \cite{Stam95}.

CDA suffers from severe energy loss close to regions with strong density gradients~\cite{Koerner14}. The problem can be addressed by a modification known as the Variable Eddington factor (VEF) method~\cite{Brunner02}, which non-linearly adjusts the diffusion coefficient to improve the solution near density gradients and low-density regions. Flux-limited diffusion, developed in the context of astrophysics by Levermore et al.~\cite{Levermore81} and later introduced to graphics by Koerner et al.~\cite{Koerner14}, is the most prominent example.
VEF is based on modifying the closure of the moment hierarchy in a way which more accurately than CDA models the transition from the diffusive regime (in which photons effectively undergo a random walk) to the transport regime (in which photons travel along straight lines) as the medium opacity decreases. VEF methods produce better results than CDA for isotropic media, but they do not provide a satisfactory treatment of anisotropic media.

Although the $P_N$-method should provide an increasingly accurate solution to the full RTE including anisotropy as the truncation moment order is increased, at the expense of more computation time, it is still an open and unresolved question how high a truncation order is needed for a particular problem in order to produce significantly better results than first order non-linear diffusion methods (FLD and VEF). This question has also been raised in other domains~\cite{Olson00}.

%\vspace{1in}

%\begin{itemize}
  %\item Alternative deterministic methods
  %\begin{itemize}
  %  \item Heuristics \cite{Kaplanyan10} \cite{Elek14}
  %  \item Lattice Boltzmann Methods \cite{Geist04}
  %  \item Discrete Ordinates \cite{Languenou95}
  %\end{itemize}
  %However, among all the deterministic methods, the diffusion approximation has been the most popular due to its intuition and simplicity.

%\end{itemize}

\newpage

%-------------------------------------------------------------------------
\tocless\section{Discretized Radiative Transfer Equation \label{sec:discretized_rte}}

Our method derives from the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position in direction $\omega$ at point $\vec{x}$:
\begin{align}
%\label{eq:rte}
\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
=&
-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\nonumber\\
&
+\sigma_s\left(\vec{x}\right) \int_{\Omega}
{
p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega' \right)\ud\omega'
}\nonumber\\
&
+Q\left(\vec{x}, \omega\right)\nonumber
\  .
\label{eq:rte}
\end{align}

The left hand side (LHS) is the transport term, and we refer to the terms on the right hand side (RHS) as collision, scattering, and source term, respectively. The symbols $\sigma_t$, $\sigma_s$, $p$, and $Q$ refer to the extinction- and scattering coefficient, phase function and emission.

The RTE is often given in operator notation, where transport, collision, and scattering are expressed as operators $\mathcal{T}$, $\mathcal{C}$ and $\mathcal{S}$, which are applied to the radiance field $L$:
\begin{align}
\mathcal{T}\left(L\right) = -\mathcal{C}\left(L\right) + \mathcal{S}\left(L\right) + Q
\ .
\end{align}

Deterministic methods are derived by discretizing the angular and spatial domain. This gives rise to a linear system of equations, which can be solved using standard methods. For the $P_N$-method, the angular variable is first discretized, using a truncated spherical harmonics expansion. This results in the $P_N$-equations, a system of coupled PDEs that still depend on a continuous spatial variable.

The number of equations grows with the truncation order $N$. This is why the discretization is laborious and difficult to do without errors if done by hand. We therefore use a computer algebra representation to automate this process. After giving an outline of the general discretization in this section, we will present our computer algebra representation in the next section. The $P_N$-equations that result from our automated discretization are given in Section~\ref{sec:real_valued_pn_eq}.

Since the radiance field $L$ is real, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$ as follows~\cite{Jarosz09}:
\begin{align}
\SHBR^{l,m}=
\left\{
\begin{array}{lr}
\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
\SHBC^{l,m}, & \text{for } m = 0\\
\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}-\left(-1\right)^m\SHBC^{l,m}\right) & \text{for } m > 0
\end{array}
\right.
\label{eq:sh_real_basis}
\end{align}

We express the projection into the spherical harmonics basis functions with a projection operator $\mathcal{P}$:
\begin{align}
\mathcal{P}^{l, m}(f) = \int_{\Omega}f(\vec{x}, \omega) \SHBR^{l,m}(\omega)\,\mathrm{d}\omega = f^{l,m}\left(\vec{x}\right)
\ .
\nonumber
\end{align}

The $P_N$-equations are derived by first expressing all direction-dependent parameters in spherical harmonics. The radiance field $L$ in Equation~\ref{eq:rte} is therefore replaced by its SH reconstruction $\widehat{L}$, introducing an error due to truncation at order $N$:
\begin{align}
\widehat{L}\left(\vec{x}, \omega\right) =
\sum_{l=0}^{N}
{
\sum_{m=-l}^{l}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
}
}
\approx
L\left(\vec{x}, \omega\right)
\ .
\nonumber
\end{align}
%\vspace{0.5in}

After substitution, all angular parameters are expressed in terms of spherical harmonics, but they still depend on the continuous angular variable $\omega$. As a next step, we project each term of the RTE into spherical harmonics, using the projection operator $\mathcal{P}$. This produces a single equation for each $l,m$-pair. The $P_N$-equations therefore can be written as:
\begin{align}
\mathcal{P}^{l,m}\mathcal{T}\left(\widehat{L}\right)
=
-\mathcal{P}^{l,m}\mathcal{C}\left(\widehat{L}\right)
+\mathcal{P}^{l,m}\mathcal{S}\left(\widehat{L}\right)
+\mathcal{P}^{l,m}\left(Q\right)
\ .
\label{eq:pn_operator_notation}
\end{align}

Once the $P_N$-equations have been found, the spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxel grid (using central differences for differential operators).

Following this discretization, the radiance field $L$, is represented as a set of SH coefficients per voxel. Flattening these over all voxels into a single vector, gives the solution vector $\vec{u}$. The RHS vector $\vec{Q}$ is produced similarly. The projected operators can be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$ (see Figure~\ref{fig:matrix_layout}):
\begin{align}
(T+C-S)\vec{u} = A\vec{u} = \vec{Q}
\ .
\end{align}

$T$, $C$, $S$ are matrices, which result from the discretized transport, collision and scattering operators respectively.

\begin{figure}[t]
\centering
%\missingfigure{test}
\includegraphics[width=\columnwidth]{fig_matrix_layout_small.pdf}
%\missingfigure{figures fig matrix layout}
\vspace{-0.3in}
\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
\label{fig:matrix_layout}
\end{figure}

\tocless\section{Computer Algebra Representation \label{sec:car}}

So far, we have only given the $P_N$-equations in high-level operator notation (Equation~\ref{eq:pn_operator_notation}). Carrying out the full derivation creates large, unwieldy equations and requires a string of expansions and applications of identities. These are challenging to manipulate if done by hand. We \rev{therefore used a computer algebra representation}, which allowed us to derive and discretize the $P_N$-equation in a semi-automatic fashion (Figure~\ref{fig:car})).

\rev{It represents the equations using a tree of mathematical expressions, which represent numbers, symbols and other expression types, such as integrals, derivatives, sums, products and functions. Further, manipulators can be executed on these expression trees to perform substitution, constant folding, reordering of nested integrals, application of identities and more complex operations. Finally, frontends allow rendering the expression tree into different forms, such as \LaTeX~and C++ source code. While we ended up implementing our own lightweight framework, off-the-shelf packages, such as SymPy (www.sympy.org), exist and would be equally suitable for our use case.}

\begin{figure}[h]
%\vspace{1in}
\centering
\includegraphics[width=0.95\columnwidth]{fig_car_small.pdf}
%\missingfigure{figures fig car}
\icaption{\rev{A} computer algebra framework allows us to represent equations as mathematical expressions trees. It further provides a set of functions for manipulating the tree according to valid mathematical operations, such as the binomial expansion above. Frontends allow generation of source code from the expression tree.\vspace{-0.2in} \label{fig:car}}
\end{figure}

\rev{Using the computer algebra representation, we perform the derivation steps required to arrive at the real-valued $P_N$-equations (the derivation steps shown in the Appendix were almost all rendered from the expression tree). More importantly, we use the representation to perform the discretization and generate the stencil code used by our solver. This is detailed in section~\ref{sec:solver_precomputation}}.

\tocless\section{Real-valued $P_N$-Equations \label{sec:real_valued_pn_eq}}

With the help of \rev{a} computer algebra representation framework, we are able to easily derive and work with the large and unwieldy $P_N$-equations. We present here the final real-valued $P_N$-equations, for a general $N$ and in three dimensions, in a very compact form which we have not found elsewhere in the literature. The derivation of the real-valued equations is rather long and takes many pages to describe in detail. We therefore give only the final result in this section, and present the full derivation for reference in Appendix~\ref{appendix}.

Since the real-valued SH bases (Equation~\ref{eq:sh_real_basis}) have different definitions for $m<0$, $m=0$ or $m>0$,  we get different projections for $\mathcal{S}^{l,m}$, depending on the sign of $m$.
%\vspace{5in}
%\begin{align}
%&-\frac{1}{2}c^{\scaleto{l-1,m-1}{4pt}}
%\partial_y
%L^{\scaleto{l-1,-m+1}{4pt}}
%%\\
%+\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
%\partial_y
%L^{\scaleto{l+1,-m+1}{4pt}}
%%\\
%-\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m+1}{4pt}}
%\partial_y
%L^{\scaleto{l-1,-m-1}{4pt}}
%\nonumber
%\\&
%+\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{\scaleto{l+1,m+1}{4pt}}
%\partial_y
%L^{\scaleto{l+1,-m-1}{4pt}}
%%\\
%+\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}c^{\scaleto{l-1,m-1}{4pt}}
%\partial_x
%L^{\scaleto{l-1,m-1}{4pt}}
%\nonumber
%\\&
%-\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}e^{{l-1,m+1}}
%\partial_x
%L^{\scaleto{l-1,m+1}{4pt}}
%%\\
%+\frac{1}{2}f^{\scaleto{l+1,m+1}{4pt}}
%\partial_x
%L^{\scaleto{l+1,m+1}{4pt}}
%%\\
%-\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
%\partial_x
%L^{\scaleto{l+1,m-1}{4pt}}
%\nonumber
%\\&
%+a^{\scaleto{l-1,m}{4pt}}
%\partial_z
%L^{\scaleto{l-1,m}{4pt}}
%%\\
%+b^{\scaleto{l+1,m}{4pt}}
%\partial_z
%L^{\scaleto{l+1,m}{4pt}}
%%\\
%+\sigma_t L^{\scaleto{l,m}{4pt}}
%%\\
%-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
%%\\
%= Q^{\scaleto{l,m}{4pt}}
%\label{eq:rpn_m_<_z}
%\end{align}

For $m=0$ we have
\begin{align}
&
\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_x L^{\scaleto{l-1,1}{4pt}}
-\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_x L^{\scaleto{l+1,1}{4pt}}
%\\&
\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_y L^{\scaleto{l-1,-1}{4pt}}
\nonumber
\\&
-\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_y L^{\scaleto{l+1,-1}{4pt}}
%\\&
+a^{\scaleto{l-1,0}{4pt}}\partial_z L^{\scaleto{l-1,0}{4pt}}
+b^{\scaleto{l+1,0}{4pt}}\partial_z L^{\scaleto{l+1,0}{4pt}}
\nonumber
\\&
+\sigma_t L^{\scaleto{l,m}{4pt}}
-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
= Q^{\scaleto{l,m}{4pt}}\quad .
\label{eq:rpn_m_=_z}
\end{align}
%and for $m>0$:
%\begin{align}
%&
%\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_x L^{\scaleto{l-1,m+1}{4pt}}
%%\\&
%-\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_x L^{\scaleto{l+1,m+1}{4pt}}
%%\\&
%-\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m-1}{4pt}}\partial_x L^{\scaleto{l-1,m-1}{4pt}}
%\nonumber
%\\&
%+\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{l+1,-m+1}\partial_x L^{\scaleto{l+1,m-1}{4pt}}
%%\\&
%+\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_y L^{\scaleto{l-1,-m-1}{4pt}}
%\nonumber
%\\&
%-\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_y L^{\scaleto{l+1,-m-1}{4pt}}
%%\nonumber
%%\\&
%+\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}e^{\scaleto{l-1,-m+1}{4pt}}%\partial_y L^{\scaleto{l-1,-m+1}{4pt}}
%\nonumber
%\\&
%-\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}f^{\scaleto{l+1,-m+1}{4pt}}%\partial_y L^{\scaleto{l+1,-m+1}{4pt}}
%%\\&
%+a^{\scaleto{l-1,-m}{4pt}}\partial_z L^{\scaleto{l-1,m}{4pt}}
%%\nonumber
%%\\&
%+b^{\scaleto{l+1,-m}{4pt}}\partial_z L^{\scaleto{l+1,m}{4pt}}
%\nonumber
%\\&
%+\sigma_t L^{\scaleto{l,m}{4pt}}
%-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
%= Q^{\scaleto{l,m}{4pt}}
%%\nonumber
%\label{eq:rpn_m_>_z}
%\end{align}

For $m<0$ (upper sign) and $m>0$ (lower sign) we have
\begin{align}
&
\frac{1}{2}c^{\scaleto{l-1,\pm m-1}{4pt}}\partial_x L^{\scaleto{l-1,m\mp 1}{4pt}}
%\\&
-\frac{1}{2}d^{\scaleto{l+1,\pm m-1}{4pt}}\partial_x L^{\scaleto{l+1,m\mp 1}{4pt}}
%\\&
-\frac{1}{2}\beta_x^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m\pm 1}{4pt}}\partial_x L^{\scaleto{l-1,m\pm 1}{4pt}}
\nonumber
\\&
+\frac{1}{2}\beta_x^{\scaleto{m}{4pt}}f^{l+1,\pm m+1}\partial_x L^{\scaleto{l+1,m\pm 1}{4pt}}
%\\&
\mp \frac{1}{2}c^{\scaleto{l-1,\pm m-1}{4pt}}\partial_y L^{\scaleto{l-1,-m\pm 1}{4pt}}
\nonumber
\\&
\pm \frac{1}{2}d^{\scaleto{l+1,\pm m-1}{4pt}}\partial_y L^{\scaleto{l+1,-m \pm 1}{4pt}}
%\nonumber
%\\&
\mp \beta_y^{\scaleto{m}{4pt}}\frac{1}{2}e^{\scaleto{l-1,\pm m+1}{4pt}}\partial_y L^{\scaleto{l-1,-m\mp 1}{4pt}}
\nonumber
\\&
\pm \beta_y^{\scaleto{m}{4pt}}\frac{1}{2}f^{\scaleto{l+1,\pm m+1}{4pt}}\partial_y L^{\scaleto{l+1,-m\mp 1}{4pt}}
%\\&
+a^{\scaleto{l-1,\pm m}{4pt}}\partial_z L^{\scaleto{l-1,\mp m}{4pt}}
%\nonumber
%\\&
+b^{\scaleto{l+1,\pm m}{4pt}}\partial_z L^{\scaleto{l+1,\mp m}{4pt}}
\nonumber
\\&
+\sigma_t L^{\scaleto{l,m}{4pt}}
-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
= Q^{\scaleto{l,m}{4pt}}
\quad ,
%\nonumber
\label{eq:rpn_m_>_z}
\end{align}

with
\begin{align*}
\label{eq:real_sh_basis}
\beta_x^{m}=
\left\{
\begin{array}{ll}
0, & \text{for } m = -1\\
\frac{2}{\sqrt{2}}, & \text{for } m \neq 1\\
1, & \text{otherwise }
\end{array}
\right.
,\quad
\beta_y^{m}=
\left\{
\begin{array}{ll}
\frac{2}{\sqrt{2}}, & \text{for } m = -1\\
0, & \text{for } m \neq 1\\
1, & \text{otherwise }
\end{array}
\right.
\end{align*}

and
\begin{align*}
&
a^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}} \qquad
b^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
\\&
c^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \qquad
d^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
\\&
e^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \qquad
f^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
\end{align*}
\begin{align*}
\lambda_l=\sqrt{\frac{4\pi}{2l+1}}
\quad .
\end{align*}

In the next section we will present the solver that we use to solve the $P_N$-equations.
\vspace{1in}

%-------------------------------------------------------------------------
\begin{figure*}[t!]
\centering
%\missingfigure{$P_N$-solver overview: generate stencil code $\rightarrow$ build system $\rightarrow$ solve $\rightarrow$ render}
\includegraphics[width=\textwidth]{fig_pipeline_small.pdf}
%\missingfigure{figures fig pipeline}
\vspace{-0.2in}
\icaption{Overview of our $P_N$-solver. After generating the stencil source code from the expression trees representing the $P_N$-equations, the linear system $A\vec{u}=\vec{Q}$ is built using RTE parameter fields and additional user input, such as grid resolution and type of boundary conditions. The resulting system is solved for $\vec{u}$, which is then used in our rendering application.}
\label{fig:pnsolver}
\end{figure*}

\vspace{-0.75in}

\tocless\section{$P_N$-Solver \label{sec:pnsolver}}

The truncation order $N$ is the key input parameter to the solver. With higher values, manual implementation of the solver from the equations would be arduous, error-prone and time-consuming. We therefore decided to make use of the computer algebra representation and designed our solver around it.

The solver consists of two components. The first is a precomputation (Section~\ref{sec:solver_precomputation}), which is executed once for every single value of $N$. This step runs a partial evaluation on the mathematical expression tree and applies the spatial discretization in a reference space we call stencil space. The precomputation step automatically generates source code from the resulting expression tree.

The generated stencil code is compiled with the runtime component (Section~\ref{sec:solver_runtime}) of our solver. This component receives the actual problem as input, including grid resolution and RTE parameter fields. It then builds the linear system and solves it using standard methods. An overview of the solver is given in Figure~\ref{fig:pnsolver}.

\tocless\subsection{Precomputation \label{sec:solver_precomputation}}

The result of the precomputation is a stencil, which can be used during runtime to build the linear system for a given problem. The stencil is a pattern of indices into the solution vector, along with values. It expresses how the sum of the weighted solution vector components relate to the RHS for a given unknown in the system, and therefore contains most information required to fill the system matrix $A$ and RHS-vector $\vec{Q}$ row-by-row. Note that while the coefficients may change, the sparsity pattern of the stencil is identical for different rows.

\vspace{0.8in}

Stencil generation entails discretizing a PDE at a hypothetical center voxel $(i, j, k)$ (assumed to mean the voxel center most of the time). Finite differences create weighted references to other voxels (e.g. $i+1, j, k$). After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and offsets (Figure~\ref{fig:stencile_pipeline}). Voxel $(i, j, k)$ will only be known during runtime, when the stencil is executed for a particular unknown (row). Then the offsets can be used to find the component index into the solution-vector, and weights can be evaluated for concrete world space position. We refer to the space with the hypothetical voxel $(i, j, k)$ at the center as stencil space.

\vspace{0.1in}

\begin{figure}[h]
\centering
%\missingfigure{test}
\includegraphics[width=0.7\columnwidth]{fig_stencil_pipeline_small.pdf}
%\missingfigure{figures fig stencil pipeline}
%\vspace{-0.2in}
\icaption{Creating the stencil code requires several steps, usually done by hand. We express the given problem in a computer algebra representation and use it to fully automate the process.}
\label{fig:stencile_pipeline}
\end{figure}

\vspace{0.2in}

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.2in}
\includegraphics[width=0.6\columnwidth]{fig_discretization_small.pdf}
%\missingfigure{figures fig discretization}
%\end{center}
\end{wrapfigure}
The spatial discretization is done by parsing the expression tree from the root. The discrete substitute for the continuous position variable $\vec{x}$ is initialized with $ijk$. Differential operator nodes are replaced by a subtree, which expresses the finite difference approximation (including voxel-size factor $h$). The subtree of the differential operator node is duplicated for different offsets to the discrete variable $(i, j, k)$. Since this offset only applies to the subtree, a stack is maintained by the parser to manage scope. Whenever the parser encounters the continuous variable $\vec{x}$ in the expression tree, its node in the tree is replaced by the discrete substitute, currently on top of the stack. Nested differential operators yield higher order finite difference stencils as expected.

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.2in}
\includegraphics[width=0.6\columnwidth]{fig_codegen_small.pdf}
%\missingfigure{figures fig codegen}
%\end{center}
\end{wrapfigure}
Factorization into canonical form is done as a manipulation pass on the mathematical expression tree. The result allows us to implement the code generation in a straightforward fashion. For each term, the $ijk$-offset is retrieved from the unknown. The factor-expression, which is multipled with the unknown, is extracted from the tree and used to generate source code for evaluating the factor-expression during runtime (including calls for evaluating RTE-parameter fields).

\begin{figure}[h]

%\vspace{1in}
\centering
\includegraphics[width=0.8\columnwidth]{fig_staggered_grids_small.pdf}
%\missingfigure{figures fig staggered grids}
%\vspace{-0.2in}
\icaption{\emph{Staggering} distributes the coefficients of the solution vector $\vec{u}$ onto four disjoint grids, indicated by the symbols, in a way which guarantees second order accuracy~\cite{Seibold14}.\vspace{-.2in}}
\label{fig:staggeredgrid}
\end{figure}

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.1in}
\includegraphics[width=0.6\columnwidth]{fig_staggered_interpolation_small.pdf}
%\missingfigure{figures fig staggered interpolation}
\vspace{-.25in}
%\end{center}
\end{wrapfigure}Our solver supports placement of coefficients at arbitrary staggered grid locations (see Figure~\ref{fig:staggeredgrid}). This means that during the discretization step, the discrete location $(i, j, k)$ (at which the unknown is meant to be evaluated) might not coincide with the location of the unknown. To solve this, depending on how the two are located relative to each other, the parser returns an expression which interpolates the coefficients at $(i, j, k)$ from their defined locations. If those happen to coincide, the coefficient itself is returned. This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center. We use the staggering scheme introduced by Seibold et al.~\cite{Seibold14} for their solver \textsf{StaRMAP}, which they proved ensures second order accuracy, i.e. a truncation error of $O(h^2)$, and prevents the growth of non-physical oscillations in the solution (see their paper for the full details of the staggered coefficient placement).

%\subsection{Stencil Code Generation}

% After applying the discretization step to the expression tree of the $P_N$-equations, it is used to generate the stencil code. In numerical analysis, a stencil is an arrangement of voxels and weights, that relate values at different locations to each other and form the basis for propagating rows in the system matrix $A$ and RHS vector $\vec{Q}$ with values. The name comes from the fact, that the geometric structure and weights of the configuration do not change, when applied to different voxels. The same is true for the $P_N$-equations and we use this fact to generate a single function, which propagates rows in $A$ and $\vec{Q}$ for a given voxel.

%The $P_N$-equations express, how each coefficient of a voxel depends on other coefficients of the same or adjacent voxels. The unknowns in the terms give information about the coefficient index and voxel offset, and therefore identify a column offset in the matrix $A$. The factors to these coefficients may contain evaluations of RTE parameters, such as $\sigma_t$. Therefore, these factors can not be determined during stencil generation, but are rendered into code expressions, which are executed as part of the stencil function during runtime. Because the stencil code has been generated in stencil space relative to the voxel at $(0,0,0)$, we can run the same stencil code for every voxel, by simply applying an offset accordingly.

%\subsection{System Building and Solving}
\tocless\subsection{Runtime}
\label{sec:solver_runtime}

The stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver. The runtime executes the stencil for every voxel to populate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.

The number of rows is determined by the number of voxels times the number of coefficients per voxel (see Figure~\ref{fig:matrix_layout}) and can therefore become very large for high resolution and high truncation order. The matrix $A$ is square and extremely sparse, due to the local structure of the finite differences discretization. Unfortunately, it is non-symmetric due to the transport term and not diagonal dominant, which rules out many \rev{iterative} methods for solving linear systems. \rev{Iterative methods are useful, as they allow balancing accuracy against performance by tweaking the convergence threshold.} We address this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definite system matrix $A^TA$, albeit with a higher condition number. Investigation of other solution schemes (e.g. multigrid) would be an interesting avenue for future work. \rev{However, more importantly, in the presence of vacuum regions, the matrix $A$ becomes singular and the system cannot be solved at all. This requires the introduction of a minimum threshold for the extinction coefficient $\sigma_t$.}

Our solver supports both Neumann and Dirichlet boundary conditions (BC). They are handled transparently by the code which generates the stencil. Whenever the stencil accumulates coefficients into a boundary location, the framework either ignores the write operation (Dirichlet BC) or accumulates into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC). This is done by changing the index of the written component.

%To respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels. This requires careful management and bookkeeping of coefficient indices, which is done transparently by the runtime code.

%\begin{figure}[h]
%\centering
%\begin{subfigure}{0.45\columnwidth}
%%\missingfigure{test}
%\includegraphics[width=\columnwidth]{figures/fig_staggered_grids.pdf}
%\end{subfigure}%
%\hspace{0.05\columnwidth}
%\begin{subfigure}{0.45\columnwidth}
%\missingfigure{test2}
%\end{subfigure}%
%\vspace{-0.2in}
%\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
%\label{fig:staggeredgrid}
%\end{figure}

%Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.

%\subsection*{CDA vs. $P_1$}
%\begin{itemize}
  %\item CDA is a degenerated form of $P_1$. It is derived by isolating the flux-vector on one side of the vector-equation formed by the $l=1$ SH-band equations. This isolation requires division by the extinction coefficient, introducing a $\frac{1}{\sigma_t}$ factor which diverges as $\sigma_t$ approaches zero. Thresholding to some minimum is required, in order to be able to solve the system for vacuum regions.
  %\item $P_1$ does not require any thresholding of $\sigma_t$ as it does not contain $\sigma_t$ as a denominator. It therefore can deal with vacuum regions without modifications.
  %\item (needs validation) Further, in the presence of vacuum or near vacuum regions, the condition number for CDA is higher than for $P_1$, because of small extinction values in the denominator of the diffusion coefficient.
  %\item Using the normal form for CDA will further increase the condition number when vacuum regions are present and significantly decreases convergence.
%\end{itemize}

%-------------------------------------------------------------------------
\tocless\section{Rendering \label{sec:rendering}}

We use an approach similar to Koerner et al.~\cite{Koerner14}, where we separate the radiance field into single scattered light $L_{ss}$ and multiple scattered light $L_{ms}$:
\begin{align}
L\left(\vec{x},\omega\right) = L_{ss}\left(\vec{x},\omega\right) + L_{ms}\left(\vec{x},\omega\right)
\ .
\end{align}
The single scattered light is folded into the emission term $Q$:
\begin{align}
Q(\vec{x}, \omega) = L_{ss}(\vec{x}, \omega) = \sigma_s\left(\vec{x}\right)\int_\Omega{ p\left(\omega'\rightarrow\omega\right) L_{u}\left(\vec{x}, \omega'\right)\ud\omega' }
\ .
\end{align}

This means that our solver will solve for the multiple scattered light $L_{ms}$. The quantity $L_u$ is the ``uncollided'' light, which was emitted from the light source and attenuated by the volume without undergoing any scattering event.
We compute it using a few light samples per voxel, which quickly converges to a useful result for Dirac delta light sources.

Running the solver gives solution vector $\vec{u}$. We then un-stagger the solution by interpolating all coefficients to voxel centers. The additional coefficients at boundary voxels are no longer needed. This operation is represented as a matrix that produces a three-dimensional voxel grid with SH coefficients for order $N$ at the center of each voxel.

For rendering, we use a simple forward path tracing approach, where we start tracing from the camera. At the first scattering event, we use next event estimation to account for $L_{ss}$. Then we sample a new direction according to the phase function. Instead of continuing tracing into the new direction, we evaluate the in-scattering integral using $\widehat{L}_{ms}$. The SH coefficients at $\vec{x}$ are found by trilinear interpolation from the voxel grid of SH coefficients.

%-------------------------------------------------------------------------
\tocless\section{Results \label{sec:results}}

In this section, we present results for a set of standard problems in order to validate and evaluate our method. We also compare against classical diffusion (CDA) and flux-limited diffusion (FLD).

Our computer algebra framework and the precomputation has been implemented in Python. The runtime component of our solver has been implemented in C++. We use a naive implementation of a CG solver, which has been modified such that we do not need to explicitly compute the normal form of the linear system to solve. We use the sparse matrix representation and sparse matrix vector product from the Eigen linear algebra library (eigen.tuxfamily.org).

The solver for classical diffusion is based on the diffusion equation, which is derived by collapsing the $P_1$-equations:
\begin{align}
\nabla\left(\frac{1}{3\sigma_t}\nabla L^{0,0}\right)  = -Q^{0,0}
\ .
\label{eq:cda}
\end{align}

Since our solver can work with any PDE which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which generates the correct stencil code automatically.

Since FLD is based on a non-linear diffusion equation, we were not able to use our system in the same way. Our implementation closely follows the implementation in~\cite{Koerner14} (though ours runs on CPU) and we refer to their paper for more details.

\tocless\subsection{2D checkerboard}

First we ran our solver on the 2D checkerboard, a very common test case in other fields. The problem has dimensions $7\times 7$ and is discretized with resolution $71\times 71$. Unit size blocks are filled with purely absorbing medium $\sigma_a=10$ in a checkerboard pattern. All other blocks are filled with a purely scattering medium with $\sigma_s=1$.

Solving the standard checkerboard problem allows us to validate our solver against the \textsf{StaRMAP} solver from Seibold et al.~\cite{Seibold14}, which solves for the time-dependent and complex-valued $P_N$-equations on a staggered grid, but in the 2D case only. The 2D case is derived by assuming that all SH coefficients, RTE parameters and boundary conditions are z-independent. This causes all SH coefficients and moment equations for which $l+m$ is odd to vanish. Due to the time-dependency, their approach is to do explicit incremental steps in time. We run their solver for many timesteps to get a result close to steady state.

\begin{figure}[!t]
\centering
\begin{subfigure}{0.47\columnwidth}
\includegraphics[width=\columnwidth]{pointsource_pn_small.pdf}
%\missingfigure{figures fig pointsource pn}
\caption{$P_N$ vs. ground truth}
\label{fig:pointsource_pn}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.47\columnwidth}
\includegraphics[width=\columnwidth]{pointsource_p5_small.pdf}
%\missingfigure{figures fig pointsource p5}
\caption{$P_5$ vs. CDA and FLD}
\label{fig:pointsource_p5}
\end{subfigure}%
%\vspace{-0.1in}
\icaption{Lineplot through the 3D solution of our solver for the point source problem for various order $N$ (left). Solution for $P_5$ compared against classical diffusion, flux-limited diffusion and analytical solution (right). \vspace{-0.14in}}
\label{fig:pointsource}
\end{figure}

\begin{figure}[h]
\centering
\begin{subfigure}{0.49\columnwidth}
\includegraphics[width=\columnwidth]{checkerboard2d_p1_neumann_staggered_starmap.png}
\end{subfigure}%
\hspace{0.01\columnwidth}
\begin{subfigure}{0.49\columnwidth}
\includegraphics[width=\columnwidth]{checkerboard2d_p1_neumann_staggered.png}
\end{subfigure}%
\vspace{-0.1in}
\icaption{Comparison of the result (here the $L^{0,0}$ coefficient, or \emph{fluence}) for the checkerboard test using \textsf{StaRMAP}'s time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$. Our results are in good agreement.}
\label{fig:vs_starmap}
\end{figure}

As can be seen in Figure~\ref{fig:vs_starmap}, the results from our solver are in good agreement with the results from Seibold et al.~\cite{Seibold14} and verify the correctness of our implementation. Converging to a residual of $10e^{-10}$ takes $0.27s$ for $P_1$ and $25s$ for $P_5$.

\tocless\subsection{Point source problem}

We also run our solver for the point source problem, a single point light in a homogeneous medium. This not only helps to validate our implementation for the 3D case, but also provides information on the accuracy of these methods. We use the Grosjean approximation, which was introduced by D'Eon et al.\cite{dEon11} as a very accurate approximation to the ground truth solution.

\begin{figure}[!t]
\centering
\includegraphics[width=0.7\columnwidth]{fig_nebulae_p1_convergence_small.pdf}
%\missingfigure{figures fig nebulae p1 convergence}
%\vspace{-0.1in}
\icaption{Convergence behaviour of our solver with $N=1$ for the nebulae dataset and for varying minimum thresholds of the extinction coefficient $\sigma_t$. Threshold values and an estimate for the condition number of $A$ (MATLAB's \texttt{condest} function) are shown next to the plots. The convergence deteriorates as the threshold decreases. Once it reaches zero, the presence of pure vacuum makes the condition number infinite. \vspace{-0.2in}}
\label{fig:results_convergence}
\end{figure}

For our test case, we choose a FD resolution of $80\times80\times80$, an extinction coefficient $\sigma_t=8.0$ and albedo $\alpha=0.9$. We run the solver for different truncation values $N$. In Figure~\ref{fig:pointsource}~\subref{fig:pointsource_pn}, we see that the solution becomes increasingly accurate for higher truncation order. The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even. We further see in ~\ref{fig:pointsource}~\subref{fig:pointsource_p5} that the $P_1$ solution exactly matches the results from CDA, which confirms that the latter is only a collapsed version of the former. The time to solve is significant with $10m$ for $P_1$ and $45m$ with $P_5$. With these performance characteristics, our $P_N$-solver is clearly not competitive in comparison with our CDA and FLD solver, which are much faster.

\tocless\subsection{Nebulae}

Finally, we run our solver on a procedural cloud dataset to get an idea of its performance in more practical applications. Figure~\ref{fig:teaser}\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function, with path-traced ground truth in Figure~\ref{fig:teaser}\subref{fig:teaser_gt}. We see that at order $N=5$, our method can capture indirect illumination similarly well as FLD (\ref{fig:teaser}\subref{fig:teaser_fld}) and is significantly better than CDA (\ref{fig:teaser}\subref{fig:teaser_cda}) as expected. The indirectly illuminated region at the bottom appears to be closer to the path-traced result as opposed to the solution from FLD which is very flat in that region. However, in many other areas, $P_5$ seems to still suffer a lot from the problem of energy loss near transitions from dense regions to vacuum. It appears that going higher order a few steps mitigates this only slowly at a high cost of compute and storage.
The main characteristic of the nebulae dataset is the presence of vacuum. We found that having vacuum regions in the dataset will cause the condition number to become infinite and the solver practically does not converge. We therefore introduced a minimum threshold for the extinction coefficient $\sigma_t$. Every voxel with an extinction coefficient smaller than the threshold is set to the threshold value. In Figure~\ref{fig:results_convergence} we show the effect of the minimum threshold on the convergence behaviour. As it increases, convergence improves.

%\begin{table}[h]
%\begin{center}
%\begin{tabular}{ c | c c c }
% & \textbf{2D Checkerboard} & \textbf{Point Source} & \textbf{Nebulae} \\
%\hline
%$P_1$ & 0.27s & 10m (64\times 64\times 64) & cell6 \\
%$P_2$ & 2.4s & cell8 & cell9 \\
%$P_3$ & 5.8s & cell8 & cell9 \\
%$P_4$ & 12.7s & cell8 & cell9 \\
%$P_5$ & 25s & ? & ?
%\end{tabular}
%\end{center}
%\icaption{Performance measurements for different datasets and varying truncation order $N$}
%\label{table2}
%\end{table}

%\begin{itemize}
%	\item mention performance.
%	\item run for various values of $N$
%	\item convergence behaviour in vacuum. Clarify
%	\item run for anisotropic problem
%\end{itemize}

%-------------------------------------------------------------------------
\newpage
\tocless\section{Conclusion \label{sec:conclusion}}

In this paper we introduced the $P_N$-method to the toolbox of deterministic methods for rendering participating media in computer graphics. We derived and presented the real-valued $P_N$-equations, along with a staggered grid solver with numerical stencils constructed automatically from the equations via a computer algebra system. We showed how to use the results in a rendering system and ran our solver for various standard problems for validation.

We originally set out to understand how non-linear diffusion methods compare to the $P_N$-method for increasing order. Our results indicate that although the lack of higher moments makes the FLD solution overly smooth, its energy conserving nature and comparably small resource footprint make it a better approach at present for most graphics applications compared to the $P_N$-method, which becomes increasingly costly for higher values of $N$.

The literature in other fields often states that the $P_N$ method---when solved in normal form as we do---is able to deal with vacuum regions. We found this misleading. The $P_N$-method in normal form indeed does not break down completely in the presence of vacuum as diffusion based methods do (due to $\sigma_t^{-1}$ in the diffusion coefficient). However, in the presence of vacuum, the condition number of the system matrix becomes infinite and does not converge either. Therefore $P_N$ based methods also require minimum thresholding of the extinction coefficient and offer no benefit for vacuum regions.

Much more work needs to be done in order to make the $P_N$-method competitive in performance to alternative solutions for volume rendering. We believe this can be made possible by employing a multigrid scheme for solving the linear system of equations. We implemented a multigrid solver, but did not find the expected performance improvements. This is possibly due to the coupling between coefficients within a voxel, which does not work well together with the smoothing step. We want to study this in the future.

Unique to our system is that it uses a computer algebra representation of the equation to solve as input. Discretization in angular and spatial domain is done using manipulation passes on the representation. The stencil code, which is used to populate the system of linear equations, is generated from the expression tree. This way, we can easily deal with coupled PDEs and avoid the time consuming and error prone process of writing stencil code by hand.

When researching the application of the $P_N$-method in other fields, we came across a rich variety of variations, such as simplified $P_N$ ($SP_N$)~\cite{Ryan10}, filtered $P_N$ ($FP_N$)~\cite{Radice13}, diffusion-correction $P_N$ ($DP_N$)~\cite{Schaefer11} and least-squares $P_N$ ($LSP_N$)~\cite{Hansen14}. These variations have been introduced to address certain problems of the standard $P_N$-method, such as ringing artifacts, dealing with vacuum regions and general convergence. Our solver can be applied to any (potentially coupled) PDE and therefore can generate stencil code for all these variations by simply expressing the respective PDEs in our computer algebra representation and providing this as an input to our solver. This would allow an exhaustive comparison of all these methods and we consider this as future work.

Finally, since the approach of our solver is very generic, we also would like to explore its application to other simulation problems in computer graphics.

% --------------------------------------------------------------------------------------------------

\section*{Acknowledgements}

We thank the anonymous reviewers for their valuable and encouraging comments and feedback. We also thank Martin Frank and Benjamin Seibold for the very valuable discussions and answers to our questions. We thank Bernd Eberhardt for feedback and support. This project has been partially funded by the MSC-BW project.

%-------------------------------------------------------------------------

\bibliographystyle{eg-alpha}
%\bibliographystyle{eg-alpha-doi}

\tocless\bibliography{paper}

%-------------------------------------------------------------------------

\newpage

\onecolumn

\renewcommand{\familydefault}{\sfdefault}

\begin{appendices}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tocless\section{Full derivation of the $P_N$-equations \label{appendix}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{appendices}

This appendix presents the detailed derivations of the complex-valued and real-valued $P_N$-equations. This was actually performed semi-automatically using our computer algebra representation of the equations, guaranteeing correctness, and here we just report the result of each stage in the derivation.

The starting point is the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
% RTE ------------
\begin{align}
%\label{eq:rte}
\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
=&
-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\nonumber\\
&
+\sigma_s\left(\vec{x}\right) \int_{\Omega}
{
p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega' \right)\ud\omega'
} \ . \nonumber\\
&
+Q\left(\vec{x}, \omega\right)\nonumber
\end{align}
where the left hand side (LHS) is the transport term, and we refer to the terms on the right hand side (RHS) as collision, scattering, and source term, respectively. The symbols $\sigma_t$, $\sigma_s$, $p$, and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term.

The derivation of the $P_N$-equations is then done in two steps. First, the directional-dependent quantities are replaced by their SH-projected counterparts. For example the radiance field $L$ is replaced by its SH projection. This way the quantities are expressed in spherical harmonics, but still depend on direction $\omega$. In the second step, the RTE is projected into spherical harmonics, which is done by multiplying each term with the complex conjugate of the SH basis functions.

The SH basis functions are complex, which produces complex coefficients and complex $P_N$-equations. However, there are also real SH basis functions, which are defined in terms of the complex SH basis functions and which produce real coefficients and reconstructions. Since the radiance field $L$ is real, it is more convenient to work with the real SH basis functions.

In the next section, the complex $P_N$-equations are derived. In order to give the derivation a clearer structure, the two steps mentioned above are applied to each term individually in a separate subsection. The section concludes by putting all derived terms together. In Section~\ref{sec:real} the analogous derivation is followed to obtain the real $P_N$-equations which are used in the article.

\setcounter{tocdepth}{2}
\tableofcontents
\addtocontents{toc}{\protect\setcounter{tocdepth}{3}}

\section{Derivation of the complex-valued $P_N$-equations}

Deriving the $P_N$-equations consists of two main steps. First, all angular dependent quantities in the RTE are expressed in terms of spherical harmonics (SH) basis functions. After this, the RTE still depends on the angular variable. Therefore, the second step projects each term of the RTE by multiplying with the complex conjugate of the SH basis functions, followed by integration over solid angle to integrate out the angular variable. This gives an equation for each spherical harmonics coefficient.

Spherical Harmonics are a set of very popular and well known functions on the sphere. The complex-valued SH basis functions are given by
% complex valued SH ------------------
\begin{align}
\SHBC^{l,m}(\omega) = \SHBC^{l,m}(\theta, \phi)=
\begin{cases}
(-1)^m\sqrt{\frac{2l+1}{4\pi}\frac{(l-m)!}{(l+m)!}}e^{im\phi}P^{l,m}\left(\operatorname{cos}\left(\theta\right)\right), & \text{for $m\ge0$}\\
\left(-1\right)^m\overline{\SHBC^{l\left|m\right|}}(\theta, \phi), & \text{for $m<0$}
\end{cases}
\end{align}
where $P^{l,m}$ are the associated Legendre polynomials. The $\left(-1\right)^m$ factor is called the Condon-Shortley phase and is not part of the associated Legendre Polynomial (unlike some other definitions).

\subsection{Projecting Radiative Transfer Quantities}

% -----------------------------
\subsubsection{Radiance Field $L$ and Emission Field $Q$}
\label{sec:complex_proj_L}

Radiative transfer quantities, which depend on position $\vec{x}$ and angle $\omega$, are projected into spatially dependent SH coefficients for each SH basis function:
\begin{align}
L^{l,m}\left(\vec{x}\right)
&=
\int_\Omega{L\left(\vec{x}, \omega\right)\overline{\SHBC^{l,m}}\ud\omega} \ . \nonumber\\
Q^{l,m}\left(\vec{x}\right)
&=
\int_\Omega{Q\left(\vec{x}, \omega\right)\overline{\SHBC^{l,m}}\ud\omega} \ . \nonumber
\end{align}
The function is completely reconstructed by using all SH basis functions up to infinite order. The $P_N$-equations introduce a truncation error by only using SH basis functions up to order $N$ for the reconstruction $\hat{L}$ and $\hat{Q}$:
\begin{align}
\label{eq:sh_exp_L}
L\left(\vec{x}, \omega\right)
\approx
\hat{L}\left(\vec{x}, \omega\right) =
\sum_{l=0}^{N}
{
\sum_{m=-l}^{l}
{
L^{l,m}\left(\vec{x}\right)\SHBC^{l,m}\left(\omega\right)
}
}
=
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right)\SHBC^{l,m}\left(\omega\right)
} \ . \nonumber\\
Q\left(\vec{x}, \omega\right)
\approx
\hat{Q}\left(\vec{x}, \omega\right) =
\sum_{l=0}^{N}
{
\sum_{m=-l}^{l}
{
Q^{l,m}\left(\vec{x}\right)\SHBC^{l,m}\left(\omega\right)
}
}
=
\sum_{l,m}
{
Q^{l,m}\left(\vec{x}\right)\SHBC^{l,m}\left(\omega\right)
} \ .
\end{align}

% -------------------------------------
\subsubsection{Phase Function}
\label{sec:complex_proj_phase}

Throughout our article, we assume an isotropic phase function, which only depends on the angle between incident and outgoing vector $\omega_i$ and $\omega_o$ (note that in the graphics literature, these would often be called anisotropic). We will see later in section~\ref{sec:complex_scattering_term}, that this allows us to fix the outgoing vector $\omega_o$ at the pole axis $\vec{e}_3$ and compute the phase function SH coefficients by just varying the incident vector $\omega_i$.
%\operatorname{cos}\theta
\begin{align*}
p^{l,m}
=
\int_\Omega
{
p\left(\omega_i\cdot\vec{e}_3\right)
\SHBC^{l,m}\left(\omega_i\right)
\ud\omega_i
} \ .
\nonumber
\end{align*}

The expansion of the phase function can be further simplified because the phase function is rotationally symmetric around the pole axis $\vec{e}_3$. Consider the definition of the spherical harmonics basis function $\SHBC^{l,m}$:
\begin{align*}
\SHBC^{l,m}(\theta, \phi) = C^{l,m}e^{im\phi}P^{l,m}(\operatorname{cos}(\theta)) \ .
\end{align*}
Now we apply a rotation $R(\alpha)$ of $\alpha$ degrees around the pole axis. In spherical harmonics, this is expressed as:
\begin{align*}
\rho_{R(\alpha)}(\SHBC^{l,m}) = e^{-i m\alpha}\SHBC^{l,m} \ .
\end{align*}
If the phase function is rotationally symmetric around the pole axis, we have:
\begin{align*}
\rho_{R(\alpha)}(p) = p \ .
\end{align*}
and in spherical harmonics this would be:
\begin{align*}
\sum_{l,m}
{
e^{-i m\alpha}
p^{l,m}
\SHBC^{l,m} }\left(\omega_i\right)
=
\sum_{l,m}
{
p^{l,m}
\SHBC^{l,m}\left(\omega_i\right)
} \ .
\end{align*}
By equating coefficients we get:
\begin{align*}
p^{l,m} = p^{l,m}e^{-i m\alpha} \ .
\end{align*}

Since $e^{-i m\alpha}=1$ for all $\alpha$ only when $m=0$, we can conclude that $p^{l,m} = 0$ for all $m\ne0$. This means that for a function which is rotationally symmetric around the pole axis, only the $m=0$ coefficients will be valid. Therefore, our phase function reconstruction for a fixed outgoing vector ($\omega_o=\vec{e}_3$) only requires SH coefficients with $m=0$:
\begin{align}
\label{eq:sh_exp_phase}
p(\omega_i) =
\sum_l
{
p^{l0}
\SHBC^{l0}(\omega_i)
} \ .
\end{align}
%

\subsection{Projecting Terms of the RTE}

% ------------------------------------------------------------
\subsubsection{Transport Term}
\label{sec:complex_transport_term}

The transport term of the RTE is given as
\begin{align*}
(\omega\cdot\nabla)L(\vec{x}, \omega)
\end{align*}
Replacing $L$ with its expansion gives:
\begin{align*}
\left(\omega\cdot\nabla\right)
\left(
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right )
\SHBC^{l,m}\left(\omega\right)
}
\right)
\end{align*}

Next we multiply with $\overline{\SHBC^{l'm'}}(\omega)$ and integrate over solid angle:
\begin{align*}
\int_\Omega
{
\overline{Y^{l'm'}}(\omega)(\omega\cdot\nabla)
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right)
\SHBC^{l,m}\left(\omega\right)
}
}
\ud\omega
\end{align*}

We can pull the spatial derivative out of the integral to get:
\begin{align}
\nabla\cdot\int_\Omega
{
\omega\overline{\SHBC^{l'm'}}(\omega)
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right)
\SHBC^{l,m}\left(\omega\right)
}
\ud\omega
}
\label{eq:complex_transport_a}
\end{align}

We apply the following recursive relation for the spherical harmonics basis functions:
\begin{align}
\omega\;\overline{\SHBC^{l,m}} = \frac{1}{2}
\begin{pmatrix}
\ c^{l-1, m-1}\overline{\SHBC^{l-1,m-1}} - d^{l+1, m-1}\overline{\SHBC^{l+1,m-1}} - e^{l-1, m+1}\overline{\SHBC^{l-1,m+1}} + f^{l+1, m+1}\overline{\SHBC^{l+1,m+1}}\\
i\left(-c^{l-1, m-1}\overline{\SHBC^{l-1,m-1}} + d^{l+1, m-1}\overline{\SHBC^{l+1,m-1}} - e^{l-1, m+1}\overline{\SHBC^{l-1,m+1}} + f^{l+1, m+1}\overline{\SHBC^{l+1,m+1}}\right) \\
2\left(a^{l-1, m}\overline{\SHBC^{l-1,m}}+b^{l+1, m}\overline{\SHBC^{l+1,m}}\right)
\end{pmatrix}
\label{eq:recursive_relation}
\end{align}

with
\begin{align*}
a^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}} \, , \qquad
b^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}} \, , \qquad
c^{l,m}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}}\\
d^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}} \, , \qquad
e^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \, , \qquad
f^{l,m}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
\end{align*}

Note that the signs for the $x$- and $y$- component depend on the handedness of the coordinate system in which the SH basis functions are defined. Using this in Equation~\ref{eq:complex_transport_a} gives
\begin{align*}
\begin{pmatrix}
\frac{1}{2}\partial_x\\
\frac{i}{2}\partial_y\\
\partial_z
\end{pmatrix}
\cdot
\int_\Omega
&
\begin{pmatrix}
\ c^{l'-1, m'-1}\overline{\SHBC^{l'-1,m'-1}} - d^{l'+1, m'-1}\overline{\SHBC^{l'+1,m'-1}} - e^{l'-1, m'+1}\overline{\SHBC^{l'-1,m'+1}} + f^{l'+1, m'+1}\overline{\SHBC^{l'+1,m'+1}}\\
-c^{l'-1, m'-1}\overline{\SHBC^{l'-1,m'-1}} + d^{l'+1, m'-1}\overline{\SHBC^{l'+1,m'-1}} - e^{l'-1, m'+1}\overline{\SHBC^{l'-1,m'+1}} + f^{l'+1, m'+1}\overline{\SHBC^{l'+1,m'+1}} \\
a^{l'-1, m'}\overline{\SHBC^{l'-1,m'}}+b^{l'+1, m'}\overline{\SHBC^{l'+1,m'}}
\end{pmatrix}
&\sum_{l,m}{
L^{l,m}\left(\vec{x}\right )\SHBC^{l,m}\left(\omega\right)
}
\ud\omega
\end{align*}

Integrating the vector term over solid angle can be expressed as separate solid angle integrals over each component. These integrals over a sum of terms are split into separate integrals. We arrive at:
\begin{align*}
\begin{pmatrix}
\frac{1}{2}\partial_x\\
\frac{i}{2}\partial_y\\
\partial_z
\end{pmatrix}
\cdot
\begin{pmatrix}
\ c^{l'-1, m'-1}\sum_{l,m}{L^{l,m}\left(\vec{x}\right )\int_\Omega{\overline{\SHBC^{l'-1,m'-1}}\left(\omega\right)\SHBC^{l,m}\left(\omega\right)\ud\omega}} \quad - \quad ...\\
-c^{l'-1, m'-1}\sum_{l,m}{L^{l,m}\left(\vec{x}\right )\int_\Omega{\overline{\SHBC^{l'-1,m'-1}}\left(\omega\right)\SHBC^{l,m}\left(\omega\right)\ud\omega}} \quad + \quad ... \\
a^{l'-1, m'}\sum_{l,m}{L^{l,m}\left(\vec{x}\right )\int_\Omega{\overline{\SHBC^{l'-1,m'}}\left(\omega\right)\SHBC^{l,m}\left(\omega\right)\ud\omega}} \quad + \quad ...
\end{pmatrix}
\end{align*}

Applying the orthogonality property to the solid angle integrals will will select specific $l,m$ in each term:
\begin{align*}
\begin{pmatrix}
\frac{1}{2}\partial_x\\
\frac{i}{2}\partial_y\\
\partial_z
\end{pmatrix}
\cdot
\begin{pmatrix}
\ c^{l-1, m-1}L^{l-1,m-1} - d^{l+1, m-1}L^{l+1,m-1} - e^{l-1, m+1}L^{l-1,m+1} + f^{l+1, m+1}L^{l+1,m+1}\\
-c^{l-1, m-1}L^{l-1,m-1} + d^{l+1, m-1}L^{l+1,m-1} - e^{l-1, m+1}L^{l-1,m+1} + f^{l+1, m+1}L^{l+1,m+1} \\
a^{l-1, m}L^{l-1,m}+b^{l+1, m}L^{l+1,m}
\end{pmatrix}
\end{align*}

Which gives the final moment equation for the transport term:
\begin{align*}
=
&\frac{1}{2}\partial_x\left(c^{l-1, m-1}L^{l-1,m-1} - d^{l+1, m-1}L^{l+1,m-1} - e^{l-1, m+1}L^{l-1,m+1} + f^{l+1, m+1}L^{l+1,m+1}\right) + \\
&\frac{i}{2}\partial_y\left( -c^{l-1, m-1}L^{l-1,m-1} + d^{l+1, m-1}L^{l+1,m-1} - e^{l-1, m+1}L^{l-1,m+1} + f^{l+1, m+1}L^{l+1,m+1} \right) + \\
&\partial_z\left( a^{l-1, m}L^{l-1,m}+b^{l+1, m}L^{l+1,m} \right)
\\
=
&
\frac{1}{2}c^{l-1, m-1}\partial_x L^{l-1,m-1} - \frac{1}{2}d^{l+1, m-1}\partial_x L^{l+1,m-1} - \frac{1}{2}e^{l-1, m+1}\partial_x L^{l-1,m+1} + \frac{1}{2}f^{l+1, m+1}\partial_x L^{l+1,m+1}+
\\
&-\frac{i}{2}c^{l-1, m-1}\partial_y L^{l-1,m-1} + \frac{i}{2}d^{l+1, m-1}\partial_y L^{l+1,m-1} - \frac{i}{2}e^{l-1, m+1}\partial_y L^{l-1,m+1} + \frac{i}{2}f^{l+1, m+1}\partial_y L^{l+1,m+1}+
\\
&
a^{l-1, m}\partial_z L^{l-1,m}+b^{l+1, m}\partial_z L^{l+1,m} \ .
\end{align*}

% ------------------------------------------------------------
\subsubsection{Collision Term}

The collision term of the RTE is given as:
\begin{align*}
-\sigma_t\left(\vec{x}\right)L\left(\vec{x}, \omega\right)
\end{align*}
We first replace the radiance field $L$ with its spherical harmonics expansion:
\begin{align*}
-\sigma_t\left(\vec{x}\right)
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right )\SHBC^{l,m}\left(\omega\right)
}
\end{align*}
Multiplying with $\overline{\SHBC^{l'm'}}$ and integrating over solid angle gives, after pulling some factors out of the integral:
\begin{align*}
&-\sigma_t\left(\vec{x}\right)\sum_{l,m}{L^{l,m}\left(\vec{x}\right )\int_\Omega\overline{\SHBC^{l'm'}}\left(\omega\right)\SHBC^{l,m}\left(\omega\right)\ud\omega}\\
&= -\sigma_t\left(\vec{x}\right)\sum_{l,m}{L^{l,m}\left(\vec{x}\right )\delta_{ll'}\delta_{mm'}}\\
&= -\sigma_t\left(\vec{x}\right)L^{l,m}\left(\vec{x}\right ) \ .
\end{align*}

% ------------------------------------------------------------
\subsubsection{Scattering Term}
\label{sec:complex_scattering_term}

The scattering term in the RTE is given as:
\begin{align*}
\sigma_s(\vec{x})\int_{\Omega}p(\vec{x}, \omega'\cdot\omega)L(\vec{x}, \omega')\ud\omega'
\end{align*}

The phase function used in isotropic scattering medium only depends on the angle between incident and outgoing direction and therefore is rotationally symmetric around the pole-defining axis. This property allows us to define a rotation $R(\omega)$, which rotates the phase function such that the pole axis aligns with direction vector $\omega$. The rotated phase function is defined as:
\begin{align*}
\rho_{R(\omega)}(p)
\end{align*}
where $\rho$ is the rotation operator, which can be implemented by applying the inverse rotation $R(\omega)^{-1}$ to the arguments of $p$. With this rotated phase function, we now can express the integral of the scattering operator as a convolution denoted with the symbol $\circ$:
\begin{align}
\int_{\Omega}p(\vec{x}, \omega'\cdot\omega)L(\vec{x}, \omega')\ud\omega'
&=
L\circ \rho_{R(\omega)}(p) \nonumber\\
&=
\int_{\Omega'}{L(\vec{x}, \omega')\rho_{R(\omega)}(p)(\omega')\ud\omega'} \nonumber\\
&= \langle L,  \rho_{R(\omega)}(p)\rangle \ .
\label{eq:complex_scatt_conv}
\end{align}
As we evaluate the inner product integral of the convolution, the phase function rotates along with the argument $\omega$.

We now use the spherical harmonics expansions of $L$ (Equation~\ref{eq:sh_exp_L}) and $p$ (Equation~\ref{eq:sh_exp_phase}) in the definition for the inner product of our convolution (Equation~\ref{eq:complex_scatt_conv}):
\begin{align*}
\langle L,  \rho_{R(\omega)}(p)\rangle = \left < \sum_{l,m}{L^{l,m}(\vec{x}) \SHBC^{l,m}}, \rho_{R(\omega)}\left ( \sum_{l}{p^{l0}\SHBC^{l0}} \right )\right> \ .
\end{align*}
Due to linearity of the inner product operator, we can pull out the non-angular dependent parts of the expansions:
\begin{align*}
\langle L,  \rho_{R(\omega)}(p)\rangle
&=
\sum_{l,m}
{
L^{l,m}(\vec{x})
\left<
\SHBC^{l,m},
\rho_{R(\omega)}
\left(\sum_l{p^{l0} \SHBC^{l0}}\right)
\right>
} \ .
\end{align*}
and further:
\begin{align*}
\langle L,  \rho_{R(\omega)}(p)\rangle
&=
\sum_{l'}
{
\sum_{l,m}
{
p^{l'0}L^{l,m}(\vec{x})
\left<\SHBC^{l,m}, \rho_{R(\omega)}\left( \SHBC^{l'0} \right)\right>
}
} \ .
\end{align*}

The rotation $\rho_{R(\omega)}$ of a function with frequency $l$ gives a function of frequency $l$ again. In addition the spherical harmonics basis functions $\SHBC^{l,m}$ are orthogonal. We therefore have:
\begin{align*}
\left<
\SHBC^{l,m}, \rho_{R(\omega)}\left(\SHBC^{l'm'}\right)
\right> = 0       \qquad    \text{for all}\ \ l\ne l'
\end{align*}
which further simplifies our inner product integral to:
\begin{align*}
\langle L,  \rho_{R(\omega)}(p)\rangle
&=
\sum_{l,m}
{
p^{l0}L^{l,m}(\vec{x})
\left<
\SHBC^{l,m}, \rho_{R(\omega)}\left(\SHBC^{l0} \right )
\right>
} \ .
\end{align*}

What remains to be resolved is the inner product. We use the fact that the spherical harmonics basis functions $ \SHBC^{l,m}$ are eigenfunctions of the inner product integral operator in the equation above:
\begin{align*}
\left<
\SHBC^{l,m}, \rho_{R(\omega)}\left ( \SHBC^{l0} \right )\right> = \lambda_l \SHBC^{l,m}
\end{align*}
with
\begin{align*}
\lambda_l=\sqrt{\frac{4\pi}{2l+1}} \ .
\end{align*}
Replacing the inner product gives:
\begin{align*}
\langle L,  \rho_{R(\omega)}(p)\rangle
&=
\sum_{l,m}
{
\lambda_l
p^{l0}L^{l,m}(\vec{x})
\SHBC^{l,m}
} \ .
\end{align*}

This allows us to express the scattering term using SH expansions of phase function $p$ and radiance field $L$:
\begin{align*}
\sigma_s(\vec{x})\int_{\Omega}p(\vec{x}, \omega'\cdot\omega)L(\vec{x}, \omega')\ud\omega'
&=
\sigma_s(\vec{x})\langle L,  \rho_{R(\omega)}(p)\rangle
\\
&=
\sigma_s(\vec{x})
\sum_{l,m}
{
\lambda_l
p^{l0}L^{l,m}(\vec{x})
\SHBC^{l,m}
} \ .
\end{align*}

However, we haven't done a spherical harmonics expansion of the scattering term itself. It is still a scalar function which depends on direction $\omega$. We thus project the scattering term into spherical harmonics by multiplying with $\overline {\SHBC^{l'm'}}$ and integrating over solid angle $\omega$. We further pull out all factors which do not depend on $\omega$, and apply the SH orthogonality property to arrive at the scattering term of the complex-valued $P_N$-equations:
\begin{align}
&
\int_{\Omega}
{
\overline{\SHBC^{l'm'}}(\omega)
\sigma_s(\vec{x})
\sum_{l,m}
{
\lambda_l
p^{l0}L^{l,m}(\vec{x})
\SHBC^{l,m}\left(\omega\right)
}
\ud\omega
}
\nonumber\\
=&
\lambda_l
\sigma_s(\vec{x})
p^{l0}L^{l,m}(\vec{x})
\sum_{l,m}
{
\int_{\Omega}
{
\overline{\SHBC^{l'm'}}(\omega)
\SHBC^{l,m}\left(\omega\right)
\ud\omega
}
}
\nonumber\\
=&
\lambda_l
\sigma_s(\vec{x})
p^{l0}L^{l,m}(\vec{x})
\sum_{l,m}
{
\delta_{ll'}\delta_{mm'}
}
\nonumber\\
=&
\lambda_l
\sigma_s(\vec{x})
p^{l0}L^{l,m}(\vec{x}) \ .
\end{align}

% --------------------------------------------------------------
\subsubsection{Emission Term}

The emission term of the RTE is given as:
\begin{align}
Q\left(\vec{x}, \omega\right)
\end{align}
The derivation of the SH projected term is equivalent to the derivation of the projected collision term. Replacing the emission field with its SH projection and multiplying the term with the conjugate complex of $\SHBC$ results, after applying the orthogonality property, in:
\begin{align}
Q^{l,m}\left(\vec{x}, \omega\right)
\end{align}

% --------------------------------------------------------------
\subsection{Final Equation}

We arrive at the complex-valued $P_N$-equations after putting all the projected terms together:
\begin{align*}
\frac{1}{2}c^{l-1, m-1}\partial_x L^{l-1,m-1} - \frac{1}{2}d^{l+1, m-1}\partial_x L^{l+1,m-1} - \frac{1}{2}e^{l-1, m+1}\partial_x L^{l-1,m+1} + \frac{1}{2}f^{l+1, m+1}\partial_x L^{l+1,m+1}+
\\
-\frac{i}{2}c^{l-1, m-1}\partial_y L^{l-1,m-1} + \frac{i}{2}d^{l+1, m-1}\partial_y L^{l+1,m-1} - \frac{i}{2}e^{l-1, m+1}\partial_y L^{l-1,m+1} + \frac{i}{2}f^{l+1, m+1}\partial_y L^{l+1,m+1} +
\\
a^{l-1, m}\partial_z L^{l-1,m}+b^{l+1, m}\partial_z L^{l+1,m}
=
-\sigma_t\left(\vec{x}\right)L^{l,m}\left(\vec{x}\right )
+
\lambda_l
\sigma_s(\vec{x})
p^{l0}L^{l,m}(\vec{x}) + Q^{l,m}\left(\vec{x}, \omega\right) \ .
\end{align*}

\section{Derivation of the real-valued $P_N$-equations \label{sec:real}}

The real-valued $P_N$-equations are derived similar to their complex-valued counterpart, except that the real-valued SH basis functions $\SHBR$ are used instead of the complex-valued SH basis functions. The real-valued SH basis functions are defined in terms of complex-valued SH basis functions as follows:
% real valued SH ------------------
\begin{align}
\label{eq:real_sh_basis}
\SHBR^{l,m}=
\left\{
\begin{array}{lr}
\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
\SHBC^{l,m}, & \text{for } m = 0\\
\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}+\left(-1\right)^m\SHBC^{l,m}\right), & \text{for } m > 0
\end{array}
\right.
\end{align}
Note we use the subscript $\mathbb{R}$ and $\mathbb{C}$ do distinguish between real- and complex-valued SH basis functions respectively.

\subsection{Projecting Radiative Transfer Quantities}

% --------------------------------------------
\subsubsection{Radiance Field $L$ and Emission Field $Q$}
\label{sec:real_proj_L}

As with the complex-valued case, the angular dependent quantities are projected into SH coefficients. Here, those coefficients will be real-valued, since we use the real-valued SH basis.
\begin{align}
L^{l,m}\left(\vec{x}\right)
&=
\int_\Omega{L\left(\vec{x}, \omega\right)\SHBR^{l,m}\ud\omega} \ . \nonumber\\
Q^{l,m}\left(\vec{x}\right)
&=
\int_\Omega{Q\left(\vec{x}, \omega\right)\SHBR^{l,m}\ud\omega} \ . \nonumber
\end{align}

The reconstruction $\hat{L}$ and $\hat{Q}$, is found by a truncated linear combination of SH basis functions weighted by their respective coefficients:
\begin{align}
\label{eq:real_sh_exp_L}
\hat{L}\left(\vec{x}, \omega\right) =
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
} \ .
\\
\hat{Q}\left(\vec{x}, \omega\right) =
\sum_{l,m}
{
Q^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
} \ .
\end{align}
We later will have to apply identities and properties for the complex-valued SH basis functions and therefore need to expand the real-valued basis function in $\hat{L}$. The real-valued basis function is different depending on $m$ and therefore gives different expansions for the sign of $m$:
\begin{align}
\hat{L}\left(\vec{x}, \omega\right)
=&
\left\{
\begin{array}{lr}
\sum_{l,m}L^{l,m}\left(\vec{x}\right)\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
\sum_{l,m}L^{l,m}\left(\vec{x}\right)\SHBC^{l,m}, & \text{for } m = 0\\
\sum_{l,m}L^{l,m}\left(\vec{x}\right)\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}-\left(-1\right)^m\SHBC^{l,m}\right), & \text{for } m > 0
\end{array}
\right.
\nonumber\\
=&
\left\{
\begin{array}{lr}
\sum_{l}\sum_{m=-l}^{-1}L^{l,m}\left(\vec{x}\right)\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
L^{l,0}\left(\vec{x}\right)\SHBC^{l,0}, & \text{for } m = 0\\
\sum_{l}\sum_{m=1}^{l}L^{l,m}\left(\vec{x}\right)\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}-\left(-1\right)^m\SHBC^{l,m}\right), & \text{for } m > 0
\end{array}
\right.
\nonumber\\
=&
\sum_{l=0}^{N}
\left(
\sum_{m=-l}^{-1}
{
L^{{l,m}}\left (\vec{x} \right )\left(\frac{i}{\sqrt{2}}\SHBC^{l, m}(\omega )-\frac{i}{\sqrt{2}}\left({-1}\right)^{m}\SHBC^{l, -m}(\omega )\right)
}
\right.
\nonumber\\
&
+L^{l,0}\left (\vec{x} \right )\SHBC^{l, 0}(\omega )
\nonumber\\
&
+
\left.
\sum_{m=1}^{l}
{
L^{{l,m}}\left (\vec{x} \right )\left(\frac{1}{\sqrt{2}}\SHBC^{l, -m}(\omega )+\frac{1}{\sqrt{2}}\left({-1}\right)^{m}\SHBC^{l, m}(\omega )\right)
}
\right)
\nonumber\\=&
\frac{i}{\sqrt{2}}\left(\sum_{l=0}^{N}{\sum_{m=-l}^{-1}{L^{{l,m}}\left (\vec{x} \right )\SHBC^{l, m}(\omega )}}\right)
-\frac{i}{\sqrt{2}}\left(\sum_{l=0}^{N}{\sum_{m=-l}^{-1}{L^{{l,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\SHBC^{l, -m}(\omega )}}\right)
\nonumber\\&
+\sum_{l=0}^{N}{L^{{l,0}}\left (\vec{x} \right )\SHBC^{l, 0}(\omega )}
\nonumber\\&
+\frac{1}{\sqrt{2}}\left(\sum_{l=0}^{N}{\sum_{m=1}^{l}{L^{{l,m}}\left (\vec{x} \right )\SHBC^{l, -m}(\omega )}}\right)
+\frac{1}{\sqrt{2}}\left(\sum_{l=0}^{N}{\sum_{m=1}^{l}{L^{{l,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\SHBC^{l, m}(\omega )}}\right) \ .
\end{align}

% ----------------------------------------------------
\subsubsection{Phase Function}

The real-valued spherical harmonics expansion of the phase function follows the same derivation as the complex-valued expansion from section~\ref{sec:complex_proj_phase}. We first fix the outgoing direction vector $\omega_o$ to always align with the $z$-axis ($\omega_o=\vec{e}_3$). We compute the spherical harmonics projection over incident direction vector $\omega_i$, using the real-valued spherical harmonics basis functions:
\begin{align*}
p^{l,m}
=
\int_\Omega
{
p\left(\omega_i\cdot\vec{e}_3\right)
\SHBR^{l,m}\left(\omega_i\right)
\ud\omega_i
} \ .
\nonumber
\end{align*}

Phase functions which only depend on the angle between incident and outgoing vectors are rotationally symmetric around the pole axis. Like with the complex-values spherical harmonics basis functions, such a rotation $R$ of angle $\alpha$ around the pole axis is given by:
\begin{align*}
\rho_{R(\alpha)}(\SHBR^{l,m}) = e^{-i m\alpha}\SHBR^{l,m} \ .
\end{align*}
We formulate symmetry around the pole axis with the following constraint:
\begin{align*}
\sum_{l,m}
{
e^{-i m\alpha}
p^{l,m}
\SHBR^{l,m} }\left(\omega_i\right)
=
\sum_{l,m}
{
p^{l,m}
\SHBR^{l,m}\left(\omega_i\right)
} \ .
\end{align*}
By comparing coefficients we get
\begin{align*}
p^{l,m} = p^{l,m}e^{-i m\alpha} \ .
\end{align*}

From this we can infer that $p^{l,m} = 0$ for all $m\ne0$, if the phase function is rotationally symmetric around the pole axis. We therefore have the same property as we have with complex-valued expansions of functions, which are symmetric about the pole axis: only the $m=0$ coefficients are needed for reconstruction. We therefore have
\begin{align}
\label{eq:sh_exp_phase}
p(\omega_i) =
\sum_l
{
p^{l0}
\SHBR^{l0}(\omega_i)
} \ .
\end{align}

\subsection{Projecting Terms of the RTE}

% -----------------------------------------------------
\subsubsection{Transport Term}

The transport term of the RTE is given as
\begin{align}
(\omega\cdot\nabla)L(\vec{x}, \omega)
=
\omega_{x}\partial_xL\left (\vec{x} ,\omega \right )+\omega_{y}\partial_yL\left (\vec{x} ,\omega \right )+\omega_{z}\partial_zL\left (\vec{x} ,\omega \right ) \ .
\label{eq:real_transport_term}
\end{align}

To improve readability, we first project the term into SH by multiplying with the conjugate complex of the SH basis, and replace $L$ by its SH expansion afterwards. This order was reversed, when we derived the complex-valued $P_N$-equation in section~\ref{sec:complex_transport_term}.

We now multiply Equation~\ref{eq:real_transport_term} with the real-valued SH basis and integrate over solid angle. However, the SH basis is different for $m'<0$, $m'=0$ and $m'>0$, and therefore will give us different $P_N$-equations depending on $m'$. We will go through the derivation in detail for the $m'<0$ case and give the results for the other cases at the end Multiplying the expanded transport term with the SH basis for $m'<0$ and integrating over solid angle gives:
\begin{align*}
&\int{\left(\frac{-i}{\sqrt{2}}\overline{Y^{l', m'}}(\omega )-\frac{-i}{\sqrt{2}}\left({-1}\right)^{m'}\overline{Y^{l', -m'}}(\omega )\right)\left(\omega_{x}\partial_xL\left (\vec{x} ,\omega \right )+\omega_{y}\partial_yL\left (\vec{x} ,\omega \right )+\omega_{z}\partial_zL\left (\vec{x} ,\omega \right )\right)\ud\omega}
\\
=&
\int-\frac{i}{\sqrt{2}}\overline{Y^{l', m'}}(\omega )\omega_{x}\partial_xL\left (\vec{x} ,\omega \right )-\frac{i}{\sqrt{2}}\overline{Y^{l', m'}}(\omega )\omega_{y}\partial_yL\left (\vec{x} ,\omega \right )-\frac{i}{\sqrt{2}}\overline{Y^{l', m'}}(\omega )\omega_{z}\partial_zL\left (\vec{x} ,\omega \right )
\\&
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\overline{Y^{l', -m'}}(\omega )\omega_{x}\partial_xL\left (\vec{x} ,\omega \right )+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\overline{Y^{l', -m'}}(\omega )\omega_{y}\partial_yL\left (\vec{x} ,\omega \right )
\\&
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\overline{Y^{l', -m'}}(\omega )\omega_{z}\partial_zL\left (\vec{x} ,\omega \right )\ud\omega \ .
\end{align*}

After expanding the integrand and splitting the integral, we apply the recursive relation from Equation~\ref{eq:recursive_relation} to get:
\begin{align*}
&
\frac{i}{\sqrt{2}}\frac{1}{2}c^{{l'-1,m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'-1}}(\omega )\ud\omega}
-\frac{i}{\sqrt{2}}\frac{1}{2}d^{{l'+1,m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'-1}}(\omega )\ud\omega}
\\&
-\frac{i}{\sqrt{2}}\frac{1}{2}e^{{l'-1,m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'+1}}(\omega )\ud\omega}
+\frac{i}{\sqrt{2}}\frac{1}{2}f^{{l'+1,m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-\frac{i}{\sqrt{2}}\frac{i}{2}c^{{l'-1,m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'-1}}(\omega )\ud\omega}
+\frac{i}{\sqrt{2}}\frac{i}{2}d^{{l'+1,m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'-1}}(\omega )\ud\omega}
\\&
-\frac{i}{\sqrt{2}}\frac{i}{2}e^{{l'-1,m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'+1}}(\omega )\ud\omega}
+\frac{i}{\sqrt{2}}\frac{i}{2}f^{{l'+1,m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-\frac{i}{\sqrt{2}}a^{{l'-1,m'}}\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'}}(\omega )\ud\omega}
-\frac{i}{\sqrt{2}}b^{{l'+1,m'}}\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'}}(\omega )\ud\omega}
\\&
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{2}c^{{l'-1,-m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'-1}}(\omega )\ud\omega}
\\&
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{2}d^{{l'+1,-m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'-1}}(\omega )\ud\omega}
\\&
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{2}e^{{l'-1,-m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'+1}}(\omega )\ud\omega}
\\&
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{2}f^{{l'+1,-m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'+1}}(\omega )\ud\omega}
\\&
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{2}c^{{l'-1,-m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'-1}}(\omega )\ud\omega}
\\&
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{2}d^{{l'+1,-m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'-1}}(\omega )\ud\omega}
\\&
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{2}e^{{l'-1,-m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'+1}}(\omega )\ud\omega}
\\&
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{2}f^{{l'+1,-m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'+1}}(\omega )\ud\omega}
\\&
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}a^{{l'-1,-m'}}\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'}}(\omega )\ud\omega}+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}b^{{l'+1,-m'}}\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'}}(\omega )\ud\omega} \ .
\end{align*}

Before we further expand the radiance field $L$ into its SH expansion, we will simplify coefficients by using the following relations:
\begin{align}
a^{l,m} = a^{l,-m}, \qquad
b^{l,m} = b^{l,-m}, \qquad
c^{l,m} = e^{l,-m}, \qquad
d^{l,m} = f^{l,-m} \ .
\label{eq:recursion_identities}
\end{align}

This allows us to rewrite the equation above as:
\begin{align*}
-i\alpha_c\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'-1}}(\omega )\ud\omega}
%\\
+\left({-1}\right)^{m'}i&\alpha_c\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'+1}}(\omega )\ud\omega}
\\
+i\alpha_d\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'-1}}(\omega )\ud\omega}
%\\
-\left({-1}\right)^{m'} i &\alpha_d\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'+1}}(\omega )\ud\omega}
\\
-i\alpha_e \int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'+1}}(\omega )\ud\omega}
%\\
+\left({-1}\right)^{m'}i &\alpha_e \int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'-1}}(\omega )\ud\omega}
\\
+i\alpha_f \int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'+1}}(\omega )\ud\omega}
%\\
-\left({-1}\right)^{m'}i &\alpha_f \int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'-1}}(\omega )\ud\omega}
% ---------------------------------------
\\
+\alpha_c\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'-1}}(\omega )\ud\omega}
%\\
+\left({-1}\right)^{m'}&\alpha_c\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'+1}}(\omega )\ud\omega}
% ---------------------------------------
\\
-\alpha_e\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'+1}}(\omega )\ud\omega}
%\\
-\left({-1}\right)^{m'}&\alpha_e\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'-1}}(\omega )\ud\omega}
% ---------------------------------------
\\
+\alpha_f\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'+1}}(\omega )\ud\omega}
%\\
+\left({-1}\right)^{m'}&\alpha_f\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'-1}}(\omega )\ud\omega}
\\
% ---------------------------------------------
-\alpha_d\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'-1}}(\omega )\ud\omega}
%\\
% ---------------------------------------
-\left({-1}\right)^{m'}&\alpha_d\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'+1}}(\omega )\ud\omega}
\\
% ---------------------------------------
-\alpha_a\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, m'}}(\omega )\ud\omega}
%\\
+\left({-1}\right)^{m'}&\alpha_a\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y^{l'-1, -m'}}(\omega )\ud\omega}
% ---------------------------------------
\\
-\alpha_b\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, m'}}(\omega )\ud\omega}
%\\
+\left({-1}\right)^{m'}&\alpha_b\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y^{l'+1, -m'}}(\omega )\ud\omega}
\end{align*}
with
\begin{align*}
\alpha_c = \frac{i}{\sqrt{2}}\frac{1}{2}c^{{l'-1,m'-1}}
,\qquad
%\\
\alpha_e = \frac{i}{\sqrt{2}}\frac{1}{2}e^{{l'-1,m'+1}}
,\qquad
%\\
\alpha_d = \frac{i}{\sqrt{2}}\frac{1}{2}d^{{l'+1,m'-1}}
\\
\alpha_f = \frac{i}{\sqrt{2}}\frac{1}{2}f^{{l'+1,m'+1}}
,\qquad
%\\
\alpha_a = \frac{i}{\sqrt{2}}a^{{l'-1,m'}}
,\qquad
%\\
\alpha_b = \frac{i}{\sqrt{2}}b^{{l'+1,m'}} \ .
\end{align*}

In the next step, we substitute the radiance field function $L$ with its spherical harmonics expansion and arrive at the following expression after further expansions and transformations:
\begin{align}
-i\alpha_c\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, m'-1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term1}
%\\
+\left({-1}\right)^{m'}i&\alpha_c\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, -m'+1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term2}
\label{eq:real_transport_expansion_unsimplified_term1_term2}
\\
+i\alpha_d\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, m'-1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term3}
%\\
-\left({-1}\right)^{m'} i &\alpha_d\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, -m'+1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term4}
\label{eq:real_transport_expansion_unsimplified_term3_term4}
\\
-i\alpha_e \partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, m'+1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term5}
%\\
+\left({-1}\right)^{m'}i &\alpha_e \partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, -m'-1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term6}
\\
+i\alpha_f \partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, m'+1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term7}
%\\
-\left({-1}\right)^{m'}i &\alpha_f \partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, -m'-1}}(\omega )\ud\omega}
% ---------------------------------------
%\label{eq:real_transport_expansion_unsimplified_term8}
\\
+\alpha_c\partial_x\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, m'-1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term9}
%\\
+\left({-1}\right)^{m'}&\alpha_c\partial_x\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, -m'+1}}(\omega )\ud\omega}
% ---------------------------------------
%\label{eq:real_transport_expansion_unsimplified_term10}
\\
-\alpha_e\partial_x\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, m'+1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term11}
%\\
-\left({-1}\right)^{m'}&\alpha_e\partial_x\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, -m'-1}}(\omega )\ud\omega}
% ---------------------------------------
%\label{eq:real_transport_expansion_unsimplified_term12}
\\
+\alpha_f\partial_x\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, m'+1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term13}
%\\
+\left({-1}\right)^{m'}&\alpha_f\partial_x\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, -m'-1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term14}
\\
% ---------------------------------------
-\alpha_d\partial_x\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, m'-1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term15}
%\\
% ---------------------------------------
-\left({-1}\right)^{m'}&\alpha_d\partial_x\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, -m'+1}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term16}
\\
% ---------------------------------------
-\alpha_a\partial_z\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, m'}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term17}
% ---------------------------------------
%\\
+\left({-1}\right)^{m'}&\alpha_a\partial_z\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, -m'}}(\omega )\ud\omega}
% ---------------------------------------
%\label{eq:real_transport_expansion_unsimplified_term18}
\\
-\alpha_b\partial_z\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, m'}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term19}
%\\
+\left({-1}\right)^{m'}&\alpha_b\partial_z\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, -m'}}(\omega )\ud\omega}
%\label{eq:real_transport_expansion_unsimplified_term20}
\end{align}

The real-valued $P_N$-equation have an intricate structure which causes many terms to cancel out. We take the first two terms (Equation~\ref{eq:real_transport_expansion_unsimplified_term1_term2}) of the $P_N$-equations and apply the following orthogonality property of SH:
\begin{align}
\int_\Omega{\SHBR^{l_1, m_1}\overline{\SHBC^{l_2, m_2}}}\ud\omega
=
\left\{
\begin{array}{lr}
\frac{i}{\sqrt{2}}
\left(
\delta_{\scaleto{\substack{l_1=l_2\\m_1=m_2}}{9pt}}
-\left({-1}\right)^{m_1}
\delta_{\scaleto{\substack{l_1=l_2\\m_1=-m_2}}{9pt}}
\right)
, & \text{for } m_1 < 0
\\
\delta_{\scaleto{\substack{l_1=l_2\\m_1=m_2}}{9pt}}, & \text{for } m_1 = 0
\\
\frac{1}{\sqrt{2}}
\left(
\delta_{\scaleto{\substack{l_1=l_2\\m_1=-m_2}}{9pt}}
+\left({-1}\right)^{m_1}
\delta_{\scaleto{\substack{l_1=l_2\\m_1=m_2}}{9pt}}
\right)
, & \text{for } m_1 > 0
\end{array}
\right.  \ .
\label{eq:real_orthogonality_property_with_complex}
\end{align}
This way we get for the first two terms:
\begin{align*}
%-i &\alpha_c\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)
%\int{Y_{\mathbb{R}}^{l,m}\overline{Y^{l'-1, m'-1}}(\omega )\ud\omega}
&-i\alpha_c\partial_y
\sum_{l=0}^{N}
\sum_{m=-l}^{-1}L^{l,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
i\delta_{\scaleto{\substack{l=l'-1\\m=m'-1}}{9pt}}
%\\
+i\alpha_c\partial_y
\sum_{l=0}^{N}
\sum_{m=-l}^{-1}L^{l,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
i
\left({-1}\right)^{m}
\delta_{\scaleto{\substack{l=l'-1\\m=-m'+1}}{9pt}}
\\&
-i\alpha_c\partial_y
\sum_{l=0}^{N}
L^{l,0}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{\substack{l=l'-1\\0=m'-1}}{9pt}}
%\\
-i\alpha_c\partial_y
\sum_{l=0}^{N}
\sum_{m=1}^{l}L^{l,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{\substack{l=l'-1\\m=-m'+1}}{9pt}}
\\&
-i\alpha_c\partial_y
\sum_{l=0}^{N}
\sum_{m=1}^{l}L^{l,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{\substack{l=l'-1\\m=m'-1}}{9pt}}
%\\
%\int{Y_{\mathbb{R}}^{l,m}\overline{Y^{l'-1, -m'+1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}i\alpha_c\partial_y
\sum_{l=0}^{N}
\sum_{m=-l}^{-1}L^{l,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
i
\delta_{\scaleto{\substack{l=l'-1\\m=-m'+1}}{9pt}}
\\&
-\left({-1}\right)^{m'}i\alpha_c\partial_y
\sum_{l=0}^{N}
\sum_{m=-l}^{-1}L^{l,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
i
\left({-1}\right)^{m}
\delta_{\scaleto{\substack{l=l'-1\\m=m'-1}}{9pt}}
%\\
+
\left({-1}\right)^{m'}i\alpha_c\partial_y
\sum_{l=0}^{N}
L^{l,0}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{\substack{l=l'-1\\0=-m'+1}}{9pt}}
\\&
+
\left({-1}\right)^{m'}i\alpha_c\partial_y
\sum_{l=0}^{N}
\sum_{m=1}^{l}L^{l,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{\substack{l=l'-1\\m=m'-1}}{9pt}}
%\\
+
\left({-1}\right)^{m'}i\alpha_c\partial_y
\sum_{l=0}^{N}
\sum_{m=1}^{l}L^{l,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{\substack{l=l'-1\\m=-m'+1}}{9pt}} \ .
\end{align*}

We apply the delta function for the sums which run over the variable $l$:
\begin{align}
\sum_{l=0}^{N}\sum_{m=a}^{b}L^{l,m}\delta_{\scaleto{\substack{l=x\\m=y}}{9pt}}=\sum_{m=a}^{b}L^{x,m}\delta_{\scaleto{m=y}{3pt}} \ .
\end{align}
We get for the first two terms of the transport term of the $P_N$-equation (Equation~\ref{eq:real_transport_expansion_unsimplified_term1_term2}):
\begin{align*}
&
%-i &\alpha_c\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)
%\int{Y_{\mathbb{R}}^{l,m}\overline{Y^{l'-1, m'-1}}(\omega )\ud\omega}
\mathcolor{red}{-i}
%&
\mathcolor{red}{\alpha_c\partial_y
\sum_{m=-l'+1}^{-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
i\delta_{\scaleto{m=m'-1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{blue}{
+i
}
%&
\mathcolor{blue}{
\alpha_c\partial_y
\sum_{m=-l'+1}^{-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
i
\left({-1}\right)^{m}
\delta_{\scaleto{m=-m'+1}{4pt}}
}
\\&
% -------------------------------------------------
\mathcolor{blue}
{
-i
}
%&
\mathcolor{blue}
{
\alpha_c\partial_y
L^{l'-1,0}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{m'=1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{black}
{
-i
}
%&
\mathcolor{black}
{
\alpha_c\partial_y
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{m=-m'+1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{blue}
{
-i
}
%&
\mathcolor{blue}
{
\alpha_c\partial_y
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{m=m'-1}{4pt}}
}
\\&
% -------------------------------------------------
\mathcolor{blue}
{
+\left({-1}\right)^{m'}i
}
%&
\mathcolor{blue}{
\alpha_c\partial_y
\sum_{m=-l'+1}^{-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
i
\delta_{\scaleto{m=-m'+1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{red}
{
-\left({-1}\right)^{m'}i
}
%&
\mathcolor{red}
{
\alpha_c\partial_y
\sum_{m=-l'+1}^{-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
i
\left({-1}\right)^{m}
\delta_{\scaleto{m=m'-1}{4pt}}
}
% -------------------------------------------------
\\&
\mathcolor{blue}
{
+
\left({-1}\right)^{m'}i
}
%&
\mathcolor{blue}
{
\alpha_c\partial_y
L^{l'-1,0}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{m'=1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{blue}
{
+
\left({-1}\right)^{m'}i
}
%&
\mathcolor{blue}
{
\alpha_c\partial_y
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{m=m'-1}{4pt}}
}
\\&
% -------------------------------------------------
\mathcolor{black}
{
+
\left({-1}\right)^{m'}i
}
%&
\mathcolor{black}
{
\alpha_c\partial_y
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{m=-m'+1}{4pt}}
}
\end{align*}

The variables $l'$ and $m'$ specify a particular equation within the given set of $P_N$-equations. We remember that $m'$ originated from multiplying the transport term with the real-valued SH basis function $\SHBR$ for the projection. The real-valued basis function is different for the sign of $m'$ and we derived the transport term of the $P_N$-equations under the assumption of $m'<0$ (different equations have to be derived for $m'=0$ and $m'>0$). We are able to greatly simplify the terms above when considering the parity of $m'$ and that $m'<0$.

The blue terms in the equation above all vanish, since the sums run over all negative (or positive) $m$, up to $-1$ (or l), while the Kronecker deltas in the blue terms only become non-zero for values $m>0$ (or $m<0$). This is because we derived these terms by multiplying with the real-valued SH basis function for $m'<0$.

Consider the seventh and 10th term from the equation above. Due to $\delta_{m=m'-1}$ or $\delta_{m=-m'+1}$, an even $m$ is selected if $m'$ is odd and vice versa. Therefore, we have $(-1)^m(-1)^{m'}=-1$. This causes term one and seven (red) to vanish and term four and ten (black) to collapse into one term.

Therefore, the first two terms in the expansion (Equation~\ref{eq:real_transport_expansion_unsimplified_term1_term2}), simplify to:
\begin{align*}
&-i\alpha_c\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, m'-1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}i\alpha_c\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'-1, -m'+1}}(\omega )\ud\omega}
\\
&=-\frac{2}{\sqrt{2}}i
\alpha_c\partial_y
L^{l'-1,-m'+1}\left (\vec{x}\right)
\\
&=-\frac{2}{\sqrt{2}}i
\frac{i}{\sqrt{2}}\frac{1}{2}c^{{l'-1,m'-1}}
\partial_y
L^{l'-1,-m'+1}\left (\vec{x}\right)
\\
&=
\frac{1}{2}c^{{l'-1,m'-1}}
\partial_y
L^{l'-1,-m'+1}\left (\vec{x}\right) \ .
\end{align*}

The terms in Equation~\ref{eq:real_transport_expansion_unsimplified_term3_term4} are derived in the same way with the difference, that the signs are reversed and that we have $l'+1$ instead of $l'-1$. However, this does not affect the simplification:
\begin{align*}
&i\alpha_d\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, m'-1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}i\alpha_c\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)\int{Y_{\mathbb{R}}^{l,m}\overline{\SHBC^{l'+1, -m'+1}}(\omega )\ud\omega}
\\
&=-\frac{2}{\sqrt{2}}i
\alpha_d\partial_y
L^{l'-1,-m'+1}\left (\vec{x}\right)
\\
&=-\frac{2}{\sqrt{2}}i
\frac{i}{\sqrt{2}}\frac{1}{2}d^{{l'+1,m'-1}}
\partial_y
L^{l'+1,-m'+1}\left (\vec{x}\right)
\\
&=
\frac{1}{2}d^{{l'+1,m'-1}}
\partial_y
L^{l'+1,-m'+1}\left (\vec{x}\right) \ .
\end{align*}
Carrying out the same simplifications for the remaining terms, results in the following real-valued $P_N$-equations for $m'<0$:
\begin{align*}
&-\frac{1}{2}c^{{l'-1,m'-1}}
\partial_y
L^{l'-1,-m'+1}
%\\
+\frac{1}{2}d^{{l'+1,m'-1}}
\partial_y
L^{l'+1,-m'+1}
%\\
-\frac{1}{2}\beta^{m'}e^{{l'-1,m'+1}}
\partial_y
L^{l'-1,-m'-1}
\\&
+\frac{1}{2}\beta^{m'}f^{{l'+1,m'+1}}
\partial_y
L^{l'+1,-m'-1}
%\\
+\frac{1}{2}c^{{l'-1,m'-1}}
\partial_x
L^{l'-1,m'-1}
\\&
-\frac{1}{2}\delta_{\scaleto{m'\neq -1}{4pt}}e^{{l'-1,m'+1}}
\partial_x
L^{l'-1,m'+1}
%\\
+\frac{1}{2}\delta_{\scaleto{m'\neq -1}{4pt}}f^{{l'+1,m'+1}}
\partial_x
L^{l'+1,m'+1}
%\\
-\frac{1}{2}d^{{l'+1,m'-1}}
\partial_x
L^{l'+1,m'-1}
\\&
+a^{{l'-1,m'}}
\partial_z
L^{l'-1,m'}
%\\
+b^{{l'+1,m'}}
\partial_z
L^{l'+1,m'} \ .
\end{align*}
with
\begin{align}
\label{eq:real_sh_basis}
\beta^{x}=
\left\{
\begin{array}{lr}
\frac{2}{\sqrt{2}}, & \text{for } \vert x\vert = 1\\
1, & \text{for } \vert x\vert \neq 1
\end{array}
\right. \ .
\end{align}

We now carry out the same derivation for the assumption of $m'>0$. We multiply Equation~\ref{sec:complex_transport_term} with the definition of the real-valued SH basis for $m'>0$ and get:
\begin{align*}
\int{\left(\frac{1}{\sqrt{2}}\overline{Y_{\mathbb{C}}^{l', -m'}}(\omega )+\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\overline{Y_{\mathbb{C}}^{l', m'}}(\omega )\right)\left(\omega_{x}\partial_xL\left (\vec{x} ,\omega \right )+\omega_{y}\partial_yL\left (\vec{x} ,\omega \right )+\omega_{z}\partial_zL\left (\vec{x} ,\omega \right )\right)\ud\omega}
\end{align*}

We expand the integrand and split the integral. Then we apply the recursive relation from Equation~\ref{eq:recursive_relation} and get:
\begin{align*}
&
\frac{1}{\sqrt{2}}\frac{1}{2}c^{{l'-1,-m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'-1}}(\omega )\ud\omega}
-\frac{1}{\sqrt{2}}\frac{1}{2}d^{{l'+1,-m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'-1}}(\omega )\ud\omega}
\\&
-\frac{1}{\sqrt{2}}\frac{1}{2}e^{{l'-1,-m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'+1}}(\omega )\ud\omega}
+\frac{1}{\sqrt{2}}\frac{1}{2}f^{{l'+1,-m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'+1}}(\omega )\ud\omega}
\\&
-\frac{1}{\sqrt{2}}\frac{i}{2}c^{{l'-1,-m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'-1}}(\omega )\ud\omega}
+\frac{1}{\sqrt{2}}\frac{i}{2}d^{{l'+1,-m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'-1}}(\omega )\ud\omega}
\\&
-\frac{1}{\sqrt{2}}\frac{i}{2}e^{{l'-1,-m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'+1}}(\omega )\ud\omega}
+\frac{1}{\sqrt{2}}\frac{i}{2}f^{{l'+1,-m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'+1}}(\omega )\ud\omega}
\\&
+\frac{1}{\sqrt{2}}a^{{l'-1,-m'}}\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'}}(\omega )\ud\omega}
+\frac{1}{\sqrt{2}}b^{{l'+1,-m'}}\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'}}(\omega )\ud\omega}
\\&
+\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{2}c^{{l'-1,m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
\\&
-\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{2}d^{{l'+1,m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
\\&
-\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{2}e^{{l'-1,m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
+\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{2}f^{{l'+1,m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{2}c^{{l'-1,m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
\\&
+\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{2}d^{{l'+1,m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
\\&
-\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{2}e^{{l'-1,m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
+\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{2}f^{{l'+1,m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
+\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}a^{{l'-1,m'}}\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'}}(\omega )\ud\omega}
+\frac{1}{\sqrt{2}}\left({-1}\right)^{m'}b^{{l'+1,m'}}\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'}}(\omega )\ud\omega}
\end{align*}

We simplify these using the identities from Equation~\ref{eq:recursion_identities}:
\begin{align*}
&
\alpha_c\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'-1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}\alpha_c\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
-\alpha_d\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'-1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}\alpha_d\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-\alpha_e\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'+1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}\alpha_e\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
\\&
+\alpha_f\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'+1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}\alpha_f\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
\\&
-i \alpha_c\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'-1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}i \alpha_c\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
+i \alpha_d\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'-1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}i \alpha_d\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-i \alpha_e\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'+1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}i \alpha_e\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
\\&
+i \alpha_f\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'+1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}i \alpha_f\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
\\&
+\alpha_a\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, -m'}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}\alpha_a\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'}}(\omega )\ud\omega}
\\&
+\alpha_b\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, -m'}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}\alpha_b\int{\partial_zL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'}}(\omega )\ud\omega}
\end{align*}
with
\begin{align*}
\alpha_c = \frac{1}{\sqrt{2}}\frac{1}{2}c^{{l'-1,-m'-1}}
,\qquad
%\\
\alpha_e = \frac{1}{\sqrt{2}}\frac{1}{2}e^{{l'-1,-m'+1}}
,\qquad
%\\
\alpha_d = \frac{1}{\sqrt{2}}\frac{1}{2}d^{{l'+1,-m'-1}}
\\
\alpha_f = \frac{1}{\sqrt{2}}\frac{1}{2}f^{{l'+1,-m'+1}}
,\qquad
%\\
\alpha_a = \frac{1}{\sqrt{2}}a^{{l'-1,-m'}}
,\qquad
%\\
\alpha_b = \frac{1}{\sqrt{2}}b^{{l'+1,-m'}}  \ .
\end{align*}

We substitute the radiance field function L with its spherical harmonics expansion and arrive
at the following expression after further expansions and transformations:
\begin{align*}
&
\alpha_c\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, -m'-1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}\alpha_c\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
-\alpha_d\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, -m'-1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}\alpha_d\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-\alpha_e\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, -m'+1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}\alpha_e\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
\\&
+\alpha_f\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, -m'+1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}\alpha_f\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
\\&
-i \alpha_c\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, -m'-1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}i \alpha_c\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
+i \alpha_d\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, -m'-1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}i \alpha_d\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-i \alpha_e\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, -m'+1}}(\omega )\ud\omega}
-\left({-1}\right)^{m'}i \alpha_e\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
\\&
+i \alpha_f\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, -m'+1}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}i \alpha_f\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
\\&
+\alpha_a\partial_z\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, -m'}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}\alpha_a\partial_z\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'}}(\omega )\ud\omega}
\\&
+\alpha_b\partial_z\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, -m'}}(\omega )\ud\omega}
+\left({-1}\right)^{m'}\alpha_b\partial_z\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'}}(\omega )\ud\omega}
\end{align*}
Again we apply the identity given in Equation~\ref{eq:real_orthogonality_property_with_complex}. For the first two terms we for example get:
\begin{align*}
&
%-i &\alpha_c\partial_y\sum_{l,m}L^{l,m}\left (\vec{x}\right)
%\int{Y_{\mathbb{R}}^{l,m}\overline{Y^{l'-1, m'-1}}(\omega )\ud\omega}
\mathcolor{red}{}
%&
\mathcolor{red}{\alpha_c\partial_x
\sum_{m=-l'+1}^{-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{i}{\sqrt{2}}
\delta_{\scaleto{m=-m'-1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{blue}{
-
}
%&
\mathcolor{blue}{
\alpha_c\partial_x
\sum_{m=-l'+1}^{-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{i}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{m=m'+1}{4pt}}
}
\\&
% -------------------------------------------------
\mathcolor{blue}
{
+
}
%&
\mathcolor{blue}
{
\alpha_c\partial_x
L^{l'-1,0}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{-m'=1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{black}
{
+
}
%&
\mathcolor{black}
{
\alpha_c\partial_x
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{m=m'+1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{blue}
{
+
}
%&
\mathcolor{blue}
{
\alpha_c\partial_x
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{m=-m'-1}{4pt}}
}
\\&
% -------------------------------------------------
\mathcolor{blue}
{
-\left({-1}\right)^{-m'}
}
%&
\mathcolor{blue}{
\alpha_c\partial_x
\sum_{m=-l'+1}^{-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{i}{\sqrt{2}}
\delta_{\scaleto{m=m'+1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{red}
{
+\left({-1}\right)^{-m'}
}
%&
\mathcolor{red}
{
\alpha_c\partial_x
\sum_{m=-l'+1}^{-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{i}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{m=-m'-1}{4pt}}
}
% -------------------------------------------------
\\&
\mathcolor{blue}
{
-
\left({-1}\right)^{-m'}
}
%&
\mathcolor{blue}
{
\alpha_c\partial_x
L^{l'-1,0}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{-m'=1}{4pt}}
}
%\\
% -------------------------------------------------
\mathcolor{blue}
{
-
\left({-1}\right)^{-m'}
}
%&
\mathcolor{blue}
{
\alpha_c\partial_x
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{m=-m'-1}{4pt}}
}
\\&
% -------------------------------------------------
\mathcolor{black}
{
-
\left({-1}\right)^{-m'}
}
%&
\mathcolor{black}
{
\alpha_c\partial_x
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{m=m'+1}{4pt}}
}
\end{align*}
As with the $m'<0$ case, the blue and red terms cancel each other out, leaving only the black terms. The first two terms of the real-valued $P_N$-equations for the transport term therefore are:
\begin{align*}
&
\alpha_c\partial_x
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\delta_{\scaleto{m=m'+1}{4pt}}
-
\left({-1}\right)^{-m'}
\alpha_c\partial_x
\sum_{m=1}^{l'-1}L^{l'-1,m}\left (\vec{x}\right)
\frac{1}{\sqrt{2}}
\left({-1}\right)^{m}
\delta_{\scaleto{m=m'+1}{4pt}}
\\&
=
\frac{2}{\sqrt{2}}
\alpha_c\partial_x
L^{l'-1,m'+1}\left (\vec{x}\right)
=
\frac{2}{\sqrt{2}}
\left(\frac{1}{2\sqrt{2}}c^{l'-1, -m'-1}\right)\partial_x
L^{l'-1,m'+1}\left (\vec{x}\right)
\\&
=
\frac{1}{2}c^{l'-1,-m'-1}L^{l'-1,m'+1}\left (\vec{x}\right) \ .
\end{align*}

Following this through for the remaining terms gives us the real-valued $P_N$-equations for $m'>0$:
\begin{align}
&
\frac{1}{2}c^{l'-1,-m'-1}\partial_x L^{l'-1,m'+1}\left (\vec{x}\right)
%\\&
-\frac{1}{2}d^{l'+1,-m'-1}\partial_x L^{l'+1,m'+1}\left (\vec{x}\right)
%\\&
-\frac{1}{2}\beta^{m'}e^{l'-1,m'-1}\partial_x L^{l'-1,m'-1}\left (\vec{x}\right)
\nonumber
\\&
\frac{1}{2}\beta^{m'}f^{l'+1,-m'+1}\partial_x L^{l'+1,m'-1}\left (\vec{x}\right)
%\\&
\frac{1}{2}c^{l'-1,-m'-1}\partial_y L^{l'-1,-m'-1}\left (\vec{x}\right)
%\\&
-\frac{1}{2}d^{l'+1,-m'-1}\partial_y L^{l'+1,-m'-1}\left (\vec{x}\right)
\nonumber
\\&
\delta_{\scaleto{m'\neq 1}{4pt}}\frac{1}{2}e^{l'-1,-m'+1}\partial_y L^{l'-1,-m'+1}\left (\vec{x}\right)
%\\&
-\delta_{\scaleto{m'\neq 1}{4pt}}\frac{1}{2}f^{l'+1,-m'+1}\partial_y L^{l'+1,-m'+1}\left (\vec{x}\right)
%\\&
a^{l'-1,-m'}\partial_z L^{l'-1,m'}\left (\vec{x}\right)
\nonumber
\\&
b^{l'+1,-m'}\partial_z L^{l'+1,m'}\left (\vec{x}\right)
\nonumber
\end{align}

Finally the $m'=0$ case needs to be derived. The derivation starts very similar to the complex-valued $P_N$-equations as in this case, the real-valued SH basis function is identical to the complex-valued SH basis function. We multiply Equation~\ref{sec:complex_transport_term} with the definition of the real-valued SH basis for $m'=0$ and get:
\begin{align*}
\int{\overline{Y_{\mathbb{C}}^{l', m'}}(\omega )\left(\omega_{x}\partial_xL\left (\vec{x} ,\omega \right )+\omega_{y}\partial_yL\left (\vec{x} ,\omega \right )+\omega_{z}\partial_zL\left (\vec{x} ,\omega \right )\right)\ud\omega}
\end{align*}

Expanding the integrand and applying the recursion relation (Equation~\ref{eq:recursion_identities}) produces the following set of terms:
\begin{align*}
&
\frac{1}{2}c^{{l'-1,m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
-\frac{1}{2}e^{{l'-1,m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
-\frac{1}{2}d^{{l'+1,m'-1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
+\frac{1}{2}f^{{l'+1,m'+1}}\int{\partial_xL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-\frac{i}{2}c^{{l'-1,m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
-\frac{i}{2}e^{{l'-1,m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
+\frac{i}{2}d^{{l'+1,m'-1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
+\frac{i}{2}f^{{l'+1,m'+1}}\int{\partial_yL\left (\vec{x} ,\omega \right )\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
+a^{{l'-1,m'}}\int{\overline{Y_{\mathbb{C}}^{l'-1, m'}}(\omega )\partial_zL\left (\vec{x} ,\omega \right )\ud\omega}
+b^{{l'+1,m'}}\int{\overline{Y_{\mathbb{C}}^{l'+1, m'}}(\omega )\partial_zL\left (\vec{x} ,\omega \right )\ud\omega}
\end{align*}

Again we will replace the radiance field $L$ with its real-valued SH projection and get:
\begin{align*}
&
\frac{1}{2}c^{{l'-1,m'-1}}\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
-\frac{1}{2}e^{{l'-1,m'+1}}\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
-\frac{1}{2}d^{{l'+1,m'-1}}\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
+\frac{1}{2}f^{{l'+1,m'+1}}\partial_x\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
-\frac{i}{2}c^{{l'-1,m'-1}}\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'-1}}(\omega )\ud\omega}
-\frac{i}{2}e^{{l'-1,m'+1}}\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'-1, m'+1}}(\omega )\ud\omega}
\\&
+\frac{i}{2}d^{{l'+1,m'-1}}\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'-1}}(\omega )\ud\omega}
+\frac{i}{2}f^{{l'+1,m'+1}}\partial_y\sum_{l,m}L^{l,m}\int{\SHBR^{l,m}\overline{Y_{\mathbb{C}}^{l'+1, m'+1}}(\omega )\ud\omega}
\\&
+a^{{l'-1,m'}}\partial_z\sum_{l,m}L^{l,m}\int{\overline{Y_{\mathbb{C}}^{l'-1, m'}}(\omega )\SHBR^{l,m}\ud\omega}
+b^{{l'+1,m'}}\partial_z\sum_{l,m}L^{l,m}\int{\overline{Y_{\mathbb{C}}^{l'+1, m'}}(\omega )\SHBR^{l,m}\ud\omega}
\end{align*}

These terms also have an intricate structure where many terms cancel out and simplify. This is seen once we apply the SH orthogonality property (Equation~\ref{eq:real_orthogonality_property_with_complex}) and further consider that $m'=0$. We show this for the first two terms, which expand to:
\begin{align*}
&
\mathcolor{red}
{
\frac{1}{2}c^{{l'-1,-1}}\frac{i}{\sqrt{2}}\partial_x\left(\sum_{m=-l'+1}^{-1}{L^{{l'-1,m}}\left (\vec{x} \right )\delta_{-1,m}}\right)
}
\mathcolor{blue}
{
-\frac{1}{2}c^{{l'-1,-1}}\frac{i}{\sqrt{2}}\partial_x\left(\sum_{m=-l'+1}^{-1}{L^{{l'-1,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{-1,-m}}\right)
}
\\&
\mathcolor{blue}
{
\frac{1}{2}c^{{l'-1,-1}}\partial_x\left(L^{{l'-1,0}}\left (\vec{x} \right )\delta_{-1,0}\right)
}
\frac{1}{2}c^{{l'-1,-1}}\frac{1}{\sqrt{2}}\partial_x\left(\sum_{m=1}^{l'-1}{L^{{l'-1,m}}\left (\vec{x} \right )\delta_{-1,-m}}\right)
\\&
\mathcolor{blue}
{
\frac{1}{2}c^{{l'-1,-1}}\frac{1}{\sqrt{2}}\partial_x\left(\sum_{m=1}^{l'-1}{L^{{l'-1,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{-1,m}}\right)
}
\mathcolor{blue}
{
-\frac{1}{2}e^{{l'-1,1}}\frac{i}{\sqrt{2}}\partial_x\left(\sum_{m=-l'+1}^{-1}{L^{{l'-1,m}}\left (\vec{x} \right )\delta_{1,m}}\right)
}
\\&
\mathcolor{red}
{
\frac{1}{2}e^{{l'-1,1}}\frac{i}{\sqrt{2}}\partial_x\left(\sum_{m=-l'+1}^{-1}{L^{{l'-1,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{1,-m}}\right)
}
\mathcolor{blue}
{
-\frac{1}{2}e^{{l'-1,1}}\partial_x\left(L^{{l'-1,0}}\left (\vec{x} \right )\delta_{1,0}\right)
}
\\&
\mathcolor{blue}
{
-\frac{1}{2}e^{{l'-1,1}}\frac{1}{\sqrt{2}}\partial_x\left(\sum_{m=1}^{l'-1}{L^{{l'-1,m}}\left (\vec{x} \right )\delta_{1,-m}}\right)
}
-\frac{1}{2}e^{{l'-1,1}}\frac{1}{\sqrt{2}}\partial_x\left(\sum_{m=1}^{l'-1}{L^{{l'-1,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{1,m}}\right)
\end{align*}

Again the blue terms vanish since the delta functions will never be non-zero under the sums. The red terms cancel each other out since $c^{l,-1}=e^{l,1}$ and $-1^m=-1$ for $m=-1$. The terms in black simplify to:
\begin{align*}
&
\frac{1}{2}c^{{l'-1,-1}}\frac{1}{\sqrt{2}}\partial_x\left(\sum_{m=1}^{l'-1}{L^{{l'-1,m}}\left (\vec{x} \right )\delta_{-1,-m}}\right)
-\frac{1}{2}e^{{l'-1,1}}\frac{1}{\sqrt{2}}\partial_x\left(\sum_{m=1}^{l'-1}{L^{{l'-1,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{1,m}}\right)
\\&
=
\frac{1}{2}c^{{l'-1,-1}}\frac{1}{\sqrt{2}}\partial_xL^{{l'-1,1}}\left (\vec{x} \right )
-\frac{1}{2}c^{{l'-1,-1}}\frac{1}{\sqrt{2}}\partial_xL^{{l'-1,1}}\left (\vec{x} \right )\left({-1}\right)^{1}
\\&
=
\frac{1}{\sqrt{2}}c^{{l'-1,-1}}\partial_xL^{{l'-1,1}}\left (\vec{x} \right ) \ .
\end{align*}
Similar simplifications apply to the remaining terms of the SH expansion of the transport term for $m=0$, resulting in the final expression:
\begin{align*}
&
\frac{1}{\sqrt{2}}c^{{l'-1,-1}}\partial_x L^{{l'-1,1}}\left (\vec{x} \right )
-\frac{1}{\sqrt{2}}d^{{l'+1,-1}}\partial_x L^{{l'+1,1}}\left (\vec{x} \right )
\\&
\frac{1}{\sqrt{2}}c^{{l'-1,-1}}\partial_y L^{{l'-1,-1}}\left (\vec{x} \right )
-\frac{1}{\sqrt{2}}d^{{l'+1,-1}}\partial_y L^{{l'+1,-1}}\left (\vec{x} \right )
\\&
a^{{l'-1,0}}\partial_z L^{{l'-1,0}}\left (\vec{x} \right)
+b^{{l'+1,0}}\partial_z L^{{l'+1,0}}\left (\vec{x} \right)
\end{align*}

\subsubsection{Collision Term}

The collision term of the RTE is given as:
\begin{align*}
-\sigma_t\left(\vec{x}\right)L\left(\vec{x}, \omega\right)
\end{align*}
We first replace the radiance field $L$ with its real-valued SH expansion:
\begin{align*}
-\sigma_t\left(\vec{x}\right)
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right )\SHBR^{l,m}\left(\omega\right)
}
\end{align*}

In order to project the term into SH, we have to multiply with the real-valued SH basis function and integrate over solid angle. Since the basis function is different depending on $m'<0$, $m=0$ or $m>0$, we have to derive separate $P_N$-equations for each case.

We first derive the SH projection of the collision term for the case $m'<0$. Multiplying with the SH basis and integrating over solid angle gives, after some further transformations and application of the SH orthogonality property:
\begin{align*}
&
\mathcolor{black}
{
-\frac{i}{\sqrt{2}}\sigma_t\left (\vec{x} \right )\frac{i}{\sqrt{2}}\sum_{m=-l'}^{-1}{L^{{l',m}}\left (\vec{x} \right )\delta_{m',m}}
}
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\sigma_t\left (\vec{x} \right )\frac{i}{\sqrt{2}}\sum_{m=-l'}^{-1}{L^{{l',m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{m',-m}}
}
\\&
\mathcolor{blue}
{
-\frac{i}{\sqrt{2}}\sigma_t\left (\vec{x} \right )L^{{l',0}}\left (\vec{x} \right )\delta_{m',0}
}
\mathcolor{red}
{
-\frac{i}{\sqrt{2}}\sigma_t\left (\vec{x} \right )\frac{1}{\sqrt{2}}\sum_{m=1}^{l'}{L^{{l',m}}\left (\vec{x} \right )\delta_{m',-m}}
}
\\&
\mathcolor{blue}
{
-\frac{i}{\sqrt{2}}\sigma_t\left (\vec{x} \right )\frac{1}{\sqrt{2}}\sum_{m=1}^{l'}{L^{{l',m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{m',m}}
}
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_t\left (\vec{x} \right )\frac{i}{\sqrt{2}}\sum_{m=-l'}^{-1}{L^{{l',m}}\left (\vec{x} \right )\delta_{-m',m}}
}
\\&
\mathcolor{black}
{
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_t\left (\vec{x} \right )\frac{i}{\sqrt{2}}\sum_{m=-l'}^{-1}{L^{{l',m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{-m',-m}}
}
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_t\left (\vec{x} \right )L^{{l',0}}\left (\vec{x} \right )\delta_{-m',0}
}
\\&
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_t\left (\vec{x} \right )\frac{1}{\sqrt{2}}\sum_{m=1}^{l'}{L^{{l',m}}\left (\vec{x} \right )\delta_{-m',-m}}
}
\mathcolor{red}
{
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_t\left (\vec{x} \right )\frac{1}{\sqrt{2}}\sum_{m=1}^{l'}{L^{{l',m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{-m',m}}
}
\end{align*}

As for the transport term derivation, the blue terms vanish due to the delta function being always zero under the sum. The red terms cancel each other out. The remaining term (black) determines the SH projection of the collision term for $m'<0$:
\begin{align}
\sigma_t L^{l',m'}
\end{align}
The derivation for the SH projection of the collision term for $m>0$ follows the same structure and likewise results in:
\begin{align}
\sigma_t L^{l',m'}
\end{align}
The real-values SH projection of the collision term for $m=0$ also is:
\begin{align}
\sigma_t L^{l',m'}
\end{align}

\subsubsection{Scattering Term}

The scattering term is given as a convolution of the radiance field $L$ with the phase function $p$ using a rotation $R_\omega$:
\begin{align*}
&
\sigma_s(\vec{x})\int_{\Omega'}p(\vec{x}, \omega'\cdot\omega)L(\vec{x}, \omega')\ud\omega'
\\
&= \sigma_s(\vec{x})(L\circ \rho_{R(\omega)}(p))(\omega)
\end{align*}
where the convolution can be also expressed as a inner product integral:
\begin{align*}
(L\circ \rho_{R(\omega)}(p)) &= \int_{\Omega'}{L(\vec{x}, \omega')\rho_{R(\omega)}(p)(\omega')\ud\omega'} \\
&= \langle L,  \rho_{R(\omega)}(p)\rangle \ .
\end{align*}

We substitute $L$ with its real-valued SH-expansion (Equation~\ref{eq:real_sh_exp_L}) in the inner product integral and perform some further factorizations to get:
%\begin{align*}
%L\left(\vec{x}, \omega\right)=
%&
%\frac{i}{\sqrt{2}}\left(\sum_{l=0}^{N}{\sum_{m=-l}^{-1}{L^{{l,m}}\left (\vec{x} \right )\SHBC^{l, m}(\omega )}}\right)-\frac{i}{\sqrt{2}}\left(\sum_{l=0}^{N}{\sum_{m=-l}^{-1}{L^{{l,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\SHBC^{l, -m}(\omega )}}\right)
%\\&
%+\sum_{l=0}^{N}{L^{{l,0}}\left (\vec{x} \right )\SHBC^{l, 0}(\omega )}+\frac{1}{\sqrt{2}}\left(\sum_{l=0}^{N}{\sum_{m=1}^{l}{L^{{l,m}}\left (\vec{x} \right )\SHBC^{l, -m}(\omega )}}\right)
%\\&
%+\frac{1}{\sqrt{2}}\left(\sum_{l=0}^{N}{\sum_{m=1}^{l}{L^{{l,m}}\left (\vec{x} \right )\left({-1}\right)^{m}\SHBC^{l, m}(\omega )}}\right)
%\end{align*}
\begin{align*}
&
\frac{i}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{l'}^{N}{
\sum_{m=-l}^{-1}{
L^{{l,m}}\left (\vec{x} \right )
f^{l'0}
\left<
\SHBC^{l, m}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l'0}
\right)
\right>
}
}
}
\\
-
&
\frac{i}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{l'}^{N}{
\sum_{m=-l}^{-1}{
\left({-1}\right)^{m}
L^{{l,m}}\left (\vec{x} \right )
f^{l'0}
\left<
\SHBC^{l, -m}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l'0}
\right)
\right>
}
}
}
\\
+
&
\sum_{l=0}^{N}{
\sum_{l'}^{N}{
L^{{l,0}}\left (\vec{x} \right )
f^{l'0}
\left<
\SHBC^{l, 0}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l'0}
\right)
\right>
}
}
\\
+
&
\frac{1}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{l'}^{N}{
\sum_{m=1}^{l}{
L^{{l,m}}\left (\vec{x} \right )
f^{l'0}
\left<
\SHBC^{l, -m}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l'0}
\right)
\right>
}
}
}
\\
+
&
\frac{1}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{l'}^{N}{
\sum_{m=1}^{l}{
\left({-1}\right)^{m}
L^{{l,m}}\left (\vec{x} \right )
f^{l'0}
\left<
\SHBC^{l, m}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l'0}
\right)
\right>
}
}
}
\end{align*}

The spherical harmonics basis functions $\SHBC^{lm}$ are orthogonal. We therefore have $\left < Y^{lm}, \rho_{R(\omega)}\left(Y^{l'm'}\right) \right > = 0$, for all $l\ne l'$, which further simplifies our scattering operator to
\begin{align}
&
\frac{i}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{m=-l}^{-1}{
L^{{l,m}}\left (\vec{x} \right )
f^{l0}
\left<
\SHBC^{l, m}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l0}
\right)
\right>
}
}
\\
-
&
\frac{i}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{m=-l}^{-1}{
\left({-1}\right)^{m}
L^{{l,m}}\left (\vec{x} \right )
f^{l0}
\left<
\SHBC^{l, -m}(\omega )
, \rho_{R\left(\omega\right)}
\left(
Y^{l0}
\right)
\right>
}
}
\\
+
&
\sum_{l=0}^{N}{
L^{{l,0}}\left (\vec{x} \right )
f^{l0}
\left<
\SHBC^{l, 0}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l0}
\right)
\right>
}
\\
+
&
\frac{1}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{m=1}^{l}{
L^{{l,m}}\left (\vec{x} \right )
f^{l0}
\left<
\SHBC^{l, -m}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l0}
\right)
\right>
}
}
\\
+
&
\frac{1}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{m=1}^{l}{
\left({-1}\right)^{m}
L^{{l,m}}\left (\vec{x} \right )
f^{l0}
\left<
\SHBC^{l, m}(\omega )
, \rho_{R\left(\omega\right)}
\left(
\SHBC^{l0}
\right)
\right>
}
}
\end{align}

What remains to be resolved are the inner products. We use the fact that the spherical harmonics basis functions $ \SHBC^{lm}$ are eigenfunctions of the inner product integral operator in the equation above, i.e.
\begin{align}
\left < \SHBC^{lm}, \rho_{R(\omega)}\left ( \SHBC^{l0} \right )\right > = \lambda_l \SHBC^{lm}
\end{align}
which results in:
\begin{align}
&
\frac{i}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{m=-l}^{-1}{
L^{{l,m}}\left (\vec{x} \right )
f^{l0}
\lambda_l
\SHBC^{l, m}(\omega )
}
}
\\
-
&
\frac{i}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{m=-l}^{-1}{
\left({-1}\right)^{m}
L^{{l,m}}\left (\vec{x} \right )
f^{l0}
\lambda_l
\SHBC^{l, -m}(\omega )
}
}
\\
+
&
\sum_{l=0}^{N}{
L^{{l,0}}\left (\vec{x} \right )
f^{l0}
\lambda_l
\SHBC^{l, 0}(\omega )
}
\\
+
&
\frac{1}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{m=1}^{l}{
L^{{l,m}}\left (\vec{x} \right )
f^{l0}
\lambda_l
\SHBC^{l, -m}(\omega )
}
}
\\
+
&
\frac{1}{\sqrt{2}}
\sum_{l=0}^{N}{
\sum_{m=1}^{l}{
\left({-1}\right)^{m}
L^{{l,m}}\left (\vec{x} \right )
f^{l0}
\lambda_l
\SHBC^{l, m}(\omega )
}
} \ .
\end{align}

The next step is to project the scattering term into real-valued SH. Again we will have to use different terms for $m<0$, $m=0$ and $m>0$ due to the definition of the real-valued SH basis functions. Multiplying with the real-valued SH basis function for $m<0$ and after applying further transformations, we get:
\begin{align*}
&
\frac{i}{\sqrt{2}}\sigma_s\left (\vec{x} \right )\frac{i}{\sqrt{2}}p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\sum_{m=-l'}^{-1}{L^{{l',m}}\left (\vec{x} \right )\delta_{m',m}}
\mathcolor{blue}
{
-\frac{i}{\sqrt{2}}\sigma_s\left (\vec{x} \right )\frac{i}{\sqrt{2}}p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\sum_{m=-l'}^{-1}{\left({-1}\right)^{m}L^{{l',m}}\left (\vec{x} \right )\delta_{m',-m}}
}
\\&
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\sigma_s\left (\vec{x} \right )L^{{l',0}}\left (\vec{x} \right )p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\delta_{m',0}
}
\mathcolor{red}
{
+\frac{i}{\sqrt{2}}\sigma_s\left (\vec{x} \right )\frac{1}{\sqrt{2}}p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\sum_{m=1}^{l'}{L^{{l',m}}\left (\vec{x} \right )\delta_{m',-m}}
}
\\&
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\sigma_s\left (\vec{x} \right )\frac{1}{\sqrt{2}}p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\sum_{m=1}^{l'}{\left({-1}\right)^{m}L^{{l',m}}\left (\vec{x} \right )\delta_{m',m}}
}
\mathcolor{blue}
{
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_s\left (\vec{x} \right )\frac{i}{\sqrt{2}}p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\sum_{m=-l'}^{-1}{L^{{l',m}}\left (\vec{x} \right )\delta_{-m',m}}
}
\\&
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_s\left (\vec{x} \right )\frac{i}{\sqrt{2}}p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\sum_{m=-l'}^{-1}{\left({-1}\right)^{m}L^{{l',m}}\left (\vec{x} \right )\delta_{-m',-m}}
\mathcolor{blue}
{
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_s\left (\vec{x} \right )L^{{l',0}}\left (\vec{x} \right )p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\delta_{-m',0}
}
\\&
\mathcolor{blue}
{
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_s\left (\vec{x} \right )\frac{1}{\sqrt{2}}p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\sum_{m=1}^{l'}{L^{{l',m}}\left (\vec{x} \right )\delta_{-m',-m}}
}
\mathcolor{red}
{
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\sigma_s\left (\vec{x} \right )\frac{1}{\sqrt{2}}p^{{l',0}}\left (\vec{x} \right )\lambda_{{l'}}\sum_{m=1}^{l'}{\left({-1}\right)^{m}L^{{l',m}}\left (\vec{x} \right )\delta_{-m',m}}
}
\end{align*}
Again, the blue terms vanish, because the delta functions will always be zero under the sum. The red terms cancel each other out. The black terms reduce to:
\begin{align*}
-\sigma_s\left(\vec{x}\right)\lambda_{l'}p^{l',0}\left(\vec{x}\right)L^{l',m'}\left(\vec{x}\right)
\end{align*}
The same happens for the derivation for $m>0$ and $m=0$ resulting in the same term.

\subsubsection{Emission Term}

The derivation of the real-valued SH projection of the emission term is exactly the same as for the collision and scattering term.
After replacing the emission term $Q$ with its real-valued SH expansion, we multiply by the real-valued SH basis function for $m<0$ and integrate over solid angle. After some transformations we arrive at the following expression:
\begin{align*}
&
-\frac{i}{\sqrt{2}}\frac{i}{\sqrt{2}}\sum_{m=-l'}^{-1}{Q^{{l',m}}\left (\vec{x} \right )\delta_{m',m}}
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\frac{i}{\sqrt{2}}\sum_{m=-l'}^{-1}{Q^{{l',m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{m',-m}}
}
\\&
\mathcolor{blue}
{
-\frac{i}{\sqrt{2}}\delta_{m',0}Q^{{l',0}}\left (\vec{x} \right )
}
\mathcolor{red}
{
-\frac{i}{\sqrt{2}}\frac{1}{\sqrt{2}}\sum_{m=1}^{l'}{Q^{{l',m}}\left (\vec{x} \right )\delta_{m',-m}}
}
\\&
\mathcolor{blue}
{
-\frac{i}{\sqrt{2}}\frac{1}{\sqrt{2}}\sum_{m=1}^{l'}{Q^{{l',m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{m',m}}
}
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{\sqrt{2}}\sum_{m=-l'}^{-1}{Q^{{l',m}}\left (\vec{x} \right )\delta_{-m',m}}
}
\\&
-\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{i}{\sqrt{2}}\sum_{m=-l'}^{-1}{Q^{{l',m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{-m',-m}}
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\delta_{-m',0}Q^{{l',0}}\left (\vec{x} \right )
}
\mathcolor{blue}
{
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{\sqrt{2}}\sum_{m=1}^{l'}{Q^{{l',m}}\left (\vec{x} \right )\delta_{-m',-m}}
}
\\&
\mathcolor{red}
{
+\frac{i}{\sqrt{2}}\left({-1}\right)^{m'}\frac{1}{\sqrt{2}}\sum_{m=1}^{l'}{Q^{{l',m}}\left (\vec{x} \right )\left({-1}\right)^{m}\delta_{-m',m}}
}
\end{align*}

Again, the blue terms vanish and the red terms cancel each other out. The black terms collapses to:
\begin{align*}
Q^{l',m'}
\end{align*}

for $m<0$, $m=0$ and $m>0$.

\subsection{Final equation}

Putting all projected terms from previous subsections together, we get for $m=0$:
\begin{align}
&
\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_x L^{\scaleto{l-1,1}{4pt}}
-\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_x L^{\scaleto{l+1,1}{4pt}}
\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_y L^{\scaleto{l-1,-1}{4pt}}
\nonumber
\\&
-\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_y L^{\scaleto{l+1,-1}{4pt}}
%\\&
a^{\scaleto{l-1,0}{4pt}}\partial_z L^{\scaleto{l-1,0}{4pt}}
+b^{\scaleto{l+1,0}{4pt}}\partial_z L^{\scaleto{l+1,0}{4pt}}
\nonumber
\\&
+\sigma_t L^{\scaleto{l,m}{4pt}}
-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
= Q^{\scaleto{l,m}{4pt}} \ .
\end{align}
For $m<0$:
\begin{align}
&-\frac{1}{2}c^{\scaleto{l-1,m-1}{4pt}}
\partial_y
L^{\scaleto{l-1,-m+1}{4pt}}
%\\
+\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
\partial_y
L^{\scaleto{l+1,-m+1}{4pt}}
%\\
-\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m+1}{4pt}}
\partial_y
L^{\scaleto{l-1,-m-1}{4pt}}
\nonumber
\\&
+\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{\scaleto{l+1,m+1}{4pt}}
\partial_y
L^{\scaleto{l+1,-m-1}{4pt}}
%\\
+\frac{1}{2}c^{\scaleto{l-1,m-1}{4pt}}
\partial_x
L^{\scaleto{l-1,m-1}{4pt}}
\nonumber
\\&
-\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}e^{{l-1,m+1}}
\partial_x
L^{\scaleto{l-1,m+1}{4pt}}
%\\
+\frac{1}{2}\delta_{\scaleto{m\neq -1}{4pt}}f^{\scaleto{l+1,m+1}{4pt}}
\partial_x
L^{\scaleto{l+1,m+1}{4pt}}
%\\
-\frac{1}{2}d^{\scaleto{l+1,m-1}{4pt}}
\partial_x
L^{\scaleto{l+1,m-1}{4pt}}
\nonumber
\\&
+a^{\scaleto{l-1,m}{4pt}}
\partial_z
L^{\scaleto{l-1,m}{4pt}}
%\\
+b^{\scaleto{l+1,m}{4pt}}
\partial_z
L^{\scaleto{l+1,m}{4pt}}
%\\
+\sigma_t L^{\scaleto{l,m}{4pt}}
%\\
-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
%\\
= Q^{\scaleto{l,m}{4pt}} \ .
\end{align}
And for $m>0$:
\begin{align}
&
\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_x
%\\&
-\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_x L^{\scaleto{l+1,m+1}{4pt}}
%\\&
-\frac{1}{2}\beta^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m-1}{4pt}}\partial_x L^{\scaleto{l-1,m-1}{4pt}}
\nonumber
\\&
+\frac{1}{2}\beta^{\scaleto{m}{4pt}}f^{l+1,-m+1}\partial_x L^{\scaleto{l+1,m-1}{4pt}}
%\\&
+\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_y L^{\scaleto{l-1,-m-1}{4pt}}
\nonumber
\\&
-\frac{1}{2}d^{\scaleto{l+1,-m-1}{4pt}}\partial_y L^{\scaleto{l+1,-m-1}{4pt}}
%\nonumber
%\\&
+\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}e^{\scaleto{l-1,-m+1}{4pt}}\partial_y L^{\scaleto{l-1,-m+1}{4pt}}
\nonumber
\\&
-\delta_{\scaleto{m\neq 1}{4pt}}\frac{1}{2}f^{\scaleto{l+1,-m+1}{4pt}}\partial_y L^{\scaleto{l+1,-m+1}{4pt}}
%\\&
+a^{\scaleto{l-1,-m}{4pt}}\partial_z L^{\scaleto{l-1,m}{4pt}}
%\nonumber
%\\&
+b^{\scaleto{l+1,-m}{4pt}}\partial_z L^{\scaleto{l+1,m}{4pt}}
\nonumber
\\&
+\sigma_t L^{\scaleto{l,m}{4pt}}
-\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
= Q^{\scaleto{l,m}{4pt}} \ .
\end{align}
Where we defined:
\begin{align*}
\beta^{x}=
\left\{
\begin{array}{ll}
\frac{2}{\sqrt{2}}, & \text{for } \vert x\vert = 1\\
1, & \text{for } \vert x\vert \neq 1
\end{array}
\right.
,\quad
\delta_{x\neq y}=
\left\{
\begin{array}{ll}
1, & \text{for } x \neq y\\
0, & \text{for } x = y
\end{array}
\right. \ .
\end{align*}
and
\begin{align*}
&
a^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}} \, , \qquad
b^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
\\&
c^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \, , \qquad
d^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
\\&
e^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}} \, , \qquad
f^{\scaleto{l,m}{4pt}}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
\end{align*}
\begin{align*}
\lambda_l=\sqrt{\frac{4\pi}{2l+1}} \ .
\end{align*}

These equations can be written in a more compact form by using $\pm$ and $\mp$ to write the equations for $m<0$ and $m>0$ as one. This gives the expressions of Section~\ref{sec:real_valued_pn_eq}.

%\begin{align*}
%&-\frac{1}{2}c^{\scaleto{l'-1,m'-1}{4pt}}
%\partial_y
%L^{\scaleto{l'-1,-m'+1}{4pt}}
%%\\
%+\frac{1}{2}d^{\scaleto{l'+1,m'-1}{4pt}}
%\partial_y
%L^{\scaleto{l'+1,-m'+1}{4pt}}
%%\\
%-\frac{1}{2}\beta^{\scaleto{m'}{4pt}}e^{\scaleto{l'-1,m'+1}{4pt}}
%\partial_y
%L^{\scaleto{l'-1,-m'-1}{4pt}}
%\\&
%+\frac{1}{2}\beta^{\scaleto{m'}{4pt}}f^{\scaleto{l'+1,m'+1}{4pt}}
%\partial_y
%L^{\scaleto{l'+1,-m'-1}{4pt}}
%%\\
%+\frac{1}{2}\delta_{\scaleto{m'\neq -1}{4pt}}c^{\scaleto{l'-1,m'-1}{4pt}}
%\partial_x
%L^{\scaleto{l'-1,m'-1}{4pt}}
%\\&
%-\frac{1}{2}\delta_{\scaleto{m'\neq -1}{4pt}}e^{{l'-1,m'+1}}
%\partial_x
%L^{\scaleto{l'-1,m'+1}{4pt}}
%%\\
%+\frac{1}{2}f^{\scaleto{l'+1,m'+1}{4pt}}
%\partial_x
%L^{\scaleto{l'+1,m'+1}{4pt}}
%%\\
%-\frac{1}{2}d^{\scaleto{l'+1,m'-1}{4pt}}
%\partial_x
%L^{\scaleto{l'+1,m'-1}{4pt}}
%\\&
%+a^{\scaleto{l'-1,m'}{4pt}}
%\partial_z
%L^{\scaleto{l'-1,m'}{4pt}}
%%\\
%+b^{\scaleto{l'+1,m'}{4pt}}
%\partial_z
%L^{\scaleto{l'+1,m'}{4pt}}
%\end{align*}
%\vspace{1cm}

\end{document}

### 2018-07-15 arxiv/paper.tex
First round of fixes for v2 of arxiv paper

@@
 %\label{eq:rte}
-\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
+\left(\omega\cdot\nabla\right)L\left(\vec{x}, \omega \right)
 =&
@@
 These are challenging to manipulate if done by hand.
-We \rev{therefore used a computer algebra representation}, which allowed us to derive and discretize the $P_N$-equation in a semi-automatic fashion (Figure~\ref{fig:car})).
+We \rev{therefore used a computer algebra representation}, which allowed us to derive and discretize the $P_N$-equation in a semi-automatic fashion (Figure~\ref{fig:car}).
 
@@
 Since the real-valued SH bases (Equation~\ref{eq:sh_real_basis}) have different definitions for $m<0$, $m=0$ or $m>0$,  we get different projections for $\mathcal{S}^{l,m}$, depending on the sign of $m$.
+Note that in cases where $L^{l,m}$ would appear with $l<0$ or $|m|>l$, this term is implicitly set to zero.
 %\vspace{5in}
@@
 For $m=0$ we have
-\begin{align}
-&
-\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_x L^{\scaleto{l-1,1}{4pt}}
--\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_x L^{\scaleto{l+1,1}{4pt}}
-%\\&
-\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_y L^{\scaleto{l-1,-1}{4pt}}
-\nonumber
-\\&
--\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_y L^{\scaleto{l+1,-1}{4pt}}
-%\\&
-+a^{\scaleto{l-1,0}{4pt}}\partial_z L^{\scaleto{l-1,0}{4pt}}
-+b^{\scaleto{l+1,0}{4pt}}\partial_z L^{\scaleto{l+1,0}{4pt}}
-\nonumber
-\\&
-+\sigma_t L^{\scaleto{l,m}{4pt}}
--\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
+\begin{eqnarray}
+&&\frac{1}{\sqrt{2}} \left( 
+         c^{\scaleto{l-1,-1}{4pt}} \partial_x L^{\scaleto{l-1,1}{4pt}}
+        -d^{\scaleto{l+1,-1}{4pt}} \partial_x L^{\scaleto{l+1,1}{4pt}}  \right)
+\nonumber \\
+&& \frac{1}{\sqrt{2}} \left(
+         c^{\scaleto{l-1,-1}{4pt}} \partial_y L^{\scaleto{l-1,-1}{4pt}}
+        -d^{\scaleto{l+1,-1}{4pt}} \partial_y L^{\scaleto{l+1,-1}{4pt}}  \right)
+\nonumber \\
+&&+ \,a^{\scaleto{l-1,0}{4pt}} \partial_z L^{\scaleto{l-1,0}{4pt}}
+  + \,b^{\scaleto{l+1,0}{4pt}} \partial_z L^{\scaleto{l+1,0}{4pt}} 
+  +\,\sigma_t L^{\scaleto{l,m}{4pt}}
+  -\,\sigma_s \lambda_{\scaleto{l}{4pt}} p^{\scaleto{l,0}{4pt}} L^{\scaleto{l,m}{4pt}}
  = Q^{\scaleto{l,m}{4pt}}\quad .
 \label{eq:rpn_m_=_z}
-\end{align}
+\end{eqnarray}
 %and for $m>0$:
@@
 \begin{align}
-&
-\frac{1}{2}c^{\scaleto{l-1,\pm m-1}{4pt}}\partial_x L^{\scaleto{l-1,m\mp 1}{4pt}}
-%\\&
--\frac{1}{2}d^{\scaleto{l+1,\pm m-1}{4pt}}\partial_x L^{\scaleto{l+1,m\mp 1}{4pt}}
-%\\&
--\frac{1}{2}\beta_x^{\scaleto{m}{4pt}}e^{\scaleto{l-1,m\pm 1}{4pt}}\partial_x L^{\scaleto{l-1,m\pm 1}{4pt}}
-\nonumber
-\\&
-+\frac{1}{2}\beta_x^{\scaleto{m}{4pt}}f^{l+1,\pm m+1}\partial_x L^{\scaleto{l+1,m\pm 1}{4pt}}
-%\\&
-\mp \frac{1}{2}c^{\scaleto{l-1,\pm m-1}{4pt}}\partial_y L^{\scaleto{l-1,-m\pm 1}{4pt}}
-\nonumber
-\\&
-\pm \frac{1}{2}d^{\scaleto{l+1,\pm m-1}{4pt}}\partial_y L^{\scaleto{l+1,-m \pm 1}{4pt}}
-%\nonumber
-%\\&
-\mp \beta_y^{\scaleto{m}{4pt}}\frac{1}{2}e^{\scaleto{l-1,\pm m+1}{4pt}}\partial_y L^{\scaleto{l-1,-m\mp 1}{4pt}}
-\nonumber
-\\&
-\pm \beta_y^{\scaleto{m}{4pt}}\frac{1}{2}f^{\scaleto{l+1,\pm m+1}{4pt}}\partial_y L^{\scaleto{l+1,-m\mp 1}{4pt}}
-%\\&
+  \frac{1}{2} c^{\scaleto{l-1,\pm m-1}{4pt}} \partial_x L^{\scaleto{l-1,m\mp 1}{4pt}}
+&-\frac{1}{2} d^{\scaleto{l+1,\pm m-1}{4pt}} \partial_x L^{\scaleto{l+1,m\mp 1}{4pt}}
+\nonumber \\
+ -\frac{1}{2} \beta_x^{\scaleto{m}{4pt}} e^{\scaleto{l-1, m\pm 1}{4pt}} \partial_x L^{\scaleto{l-1,m\pm 1}{4pt}}
+&+\frac{1}{2} \beta_x^{\scaleto{m}{4pt}} f^{\scaleto{l+1,\pm m+1}{4pt}} \partial_x L^{\scaleto{l+1,m\pm 1}{4pt}}
+\nonumber \\
+ \mp \frac{1}{2} c^{\scaleto{l-1,\pm m-1}{4pt}} \partial_y L^{\scaleto{l-1,-m\pm 1}{4pt}}
+&\pm \frac{1}{2} d^{\scaleto{l+1,\pm m-1}{4pt}} \partial_y L^{\scaleto{l+1,-m \pm 1}{4pt}}
+\nonumber \\
+ \mp \frac{1}{2} \beta_y^{\scaleto{m}{4pt}} e^{\scaleto{l-1,\pm m+1}{4pt}} \partial_y L^{\scaleto{l-1,-m\mp 1}{4pt}}
+&\pm \frac{1}{2} \beta_y^{\scaleto{m}{4pt}} f^{\scaleto{l+1,\pm m+1}{4pt}} \partial_y L^{\scaleto{l+1,-m\mp 1}{4pt}}
+\nonumber \\
 +a^{\scaleto{l-1,\pm m}{4pt}}\partial_z L^{\scaleto{l-1,\mp m}{4pt}}
-%\nonumber
-%\\&
 +b^{\scaleto{l+1,\pm m}{4pt}}\partial_z L^{\scaleto{l+1,\mp m}{4pt}}
-\nonumber
-\\&
-+\sigma_t L^{\scaleto{l,m}{4pt}}
+&+\sigma_t L^{\scaleto{l,m}{4pt}}
 -\sigma_s\lambda_{\scaleto{l}{4pt}}p^{\scaleto{l,0}{4pt}}L^{\scaleto{l,m}{4pt}}
-= Q^{\scaleto{l,m}{4pt}}
-\quad ,
-%\nonumber
+= Q^{\scaleto{l,m}{4pt}}  \;
 \label{eq:rpn_m_>_z}
 \end{align}
-
 with
@@
 %\vspace{-0.2in}
-\icaption{\emph{Staggering} distributes the coefficients of the solution vector $\vec{u}$ onto four disjoint grids, indicated by the symbols, in a way which guarantees second order accuracy~\cite{Seibold14}.\vspace{-.2in}}
+\icaption{\emph{Staggering} distributes the coefficients of the solution vector $\vec{u}$ onto disjoint grids, indicated by the symbols, in a way which guarantees second order accuracy~\cite{Seibold14}.\vspace{-.2in}}
 \label{fig:staggeredgrid}
@@
 
-\begin{wrapfigure}{r}{0.6\columnwidth}
+\begin{wrapfigure}{r}{0.5\columnwidth}
 %\begin{center}
 \hspace{-.1in}
-\includegraphics[width=0.6\columnwidth]{fig_staggered_interpolation_small.pdf}
+\includegraphics[width=0.5\columnwidth]{fig_staggered_interpolation_small.pdf}
 %\missingfigure{figures fig staggered interpolation}
@@
 We use the staggering scheme introduced by Seibold et al.~\cite{Seibold14} for their solver \textsf{StaRMAP}, which they proved ensures second order accuracy, i.e.
-a truncation error of $O(h^2)$, and prevents the growth of non-physical oscillations in the solution (see their paper for the full details of the staggered coefficient placement).
+a truncation error of $O(h^2)$, and prevents the growth of non-physical oscillations in the solution.
+This involves placing the SH coefficients onto disjoint grids --- 4 in 2D as in Figure~\ref{fig:staggeredgrid}, and 8 in 3D (see their paper for the full details of the staggered coefficient placement).
+\vspace{-.15in}
 
@@
 %\subsection{System Building and Solving}
-\tocless\subsection{Runtime}
-\label{sec:solver_runtime}
+\tocless\subsection{Runtime \label{sec:solver_runtime}}
 
@@
 %-------------------------------------------------------------------------
+
 \newpage
@@
 \end{align}
-Note we use the subscript $\mathbb{R}$ and $\mathbb{C}$ do distinguish between real- and complex-valued SH basis functions respectively.
+Note we use the subscript $\mathbb{R}$ and $\mathbb{C}$ to distinguish between real- and complex-valued SH basis functions respectively.
 
@@
 -\frac{1}{\sqrt{2}}d^{\scaleto{l+1,-1}{4pt}}\partial_x L^{\scaleto{l+1,1}{4pt}}
-\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_y L^{\scaleto{l-1,-1}{4pt}}
++\frac{1}{\sqrt{2}}c^{\scaleto{l-1,-1}{4pt}}\partial_y L^{\scaleto{l-1,-1}{4pt}}
 \nonumber
@@
 &
-\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_x
+\frac{1}{2}c^{\scaleto{l-1,-m-1}{4pt}}\partial_x L^{\scaleto{l-1,-m-1}{4pt}}
 %\\&

### 2018-07-15 arxiv/paper.tex
Minor typesetting changes.

@@
 Since the real-valued SH bases (Equation~\ref{eq:sh_real_basis}) have different definitions for $m<0$, $m=0$ or $m>0$,  we get different projections for $\mathcal{S}^{l,m}$, depending on the sign of $m$.
-Note that in cases where $L^{l,m}$ would appear with $l<0$ or $|m|>l$, this term is implicitly set to zero.
+(Note that in cases where $L^{l,m}$ would appear with $l<0$ or $|m|>l$, this term is implicitly set to zero.)
 %\vspace{5in}
@@
 \nonumber \\
-&& \frac{1}{\sqrt{2}} \left(
+&+& \frac{1}{\sqrt{2}} \left(
          c^{\scaleto{l-1,-1}{4pt}} \partial_y L^{\scaleto{l-1,-1}{4pt}}
@@
 \nonumber \\
-&&+ \,a^{\scaleto{l-1,0}{4pt}} \partial_z L^{\scaleto{l-1,0}{4pt}}
+&+& \,a^{\scaleto{l-1,0}{4pt}} \partial_z L^{\scaleto{l-1,0}{4pt}}
   + \,b^{\scaleto{l+1,0}{4pt}} \partial_z L^{\scaleto{l+1,0}{4pt}} 
@@
   -\,\sigma_s \lambda_{\scaleto{l}{4pt}} p^{\scaleto{l,0}{4pt}} L^{\scaleto{l,m}{4pt}}
- = Q^{\scaleto{l,m}{4pt}}\quad .
+ = Q^{\scaleto{l,m}{4pt}} \ . \quad
 \label{eq:rpn_m_=_z}
